### 2024-10-07 kw/1-intro.tex
Add materials

% !TEX root = main.tex

\chapter{Introduction}
\label{ch:intro}
\noindent
Quantum computing is attracting significant attention from academia and
industry, offering distinct advantages over classical computing by enabling the
development of substantially faster algorithms.
%
A prime example is \citet{shor1994algorithms}'s algorithm, which efficiently
factorizes integers in polynomial time.

When developing quantum programs, \emph{quantum circuit simulators} are
essential.
%
Quantum programs, written in high-level languages, are typically compiled to
quantum circuits described by specific \emph{quantum circuit description
	languages}~\cite{wang2022qdiff, paltenghi2023morphq}.
%
These circuits are executed either on real quantum computers or simulators
running on classical devices.
%
Owing to the high cost of quantum computing resources, developers predominantly
rely on simulators.

Unfortunately, the correctness of existing quantum circuit simulators has not
been explored, primarily due to the lack of formal semantics for quantum
circuit description languages.
%
Since formal semantics are crucial for verifying language implementations,
researchers have defined formal semantics for numerous real-world languages,
such as
%
JavaScript~\cite{maffeis2008operational,guha2010essence,politz2012tested,bodin2014trusted,park2015kjs,park2021jiset},
%
C~\cite{blazy2009mechanized,leroy2009formal,ellison2012executable,krebbers2014formal,memarian2016depths},
%
Java~\cite{farzan2004formal,klein2006machine,bogdanas2015kjava},
%
Standard ML~\cite{kumar2014cakeml,lee2007towards},
%
WebAssembly~\cite{watt2018mechanising,watt2021two},
%
Rust~\cite{jung2017rustbelt},
%
Python~\cite{politz2013python},
%
PHP~\cite{filaretti2014executable},
%
Solidity~\cite{jiao2020semantic},
%
and even the POSIX shell~\cite{greenberg2019executable},
%
often using proof assistants or semantic frameworks like
K~\cite{rosu2010overview}.
%
Particularly, \emph{executable} formal semantics has been beneficial for
testing, serving as a testing oracle to aid in bug
detection~\cite{watt2023wasmref,schumi2021spectest,park2021jest} and test suite
generation~\cite{park2023feature}.
%
However, the semantics of quantum circuit description languages has not been
formalized yet.
%
Notably, OpenQASM~\cite{cross2017open}, a representative language supported by
various major quantum programming platforms~\cite{qiskit, cirq, q-sharp},
defines its syntax using BNF but explains its semantics in prose, lacking
formal rigor.
%
This gap hinders the systematic verification and testing of quantum circuit
simulators.

In this work, we present the first executable formal semantics for OpenQASM
2.0.
%
We chose version 2.0 over the newer OpenQASM 3.0 due to its established
stability and extensive software support.
%
Our formalization employs Coq, from which we extract OCaml code to achieve
executability.
%
The unique characteristics of quantum circuits pose distinct challenges in
developing formal semantics, compared to traditional programming languages.
%
Specifically, we address two challenges:
%
(1) taming \emph{nondeterminism} and (2) ensuring \emph{physical consistency}.

The first challenge stems from the nondeterminism in quantum circuit execution,
a fundamental aspect of quantum mechanics.
%
Quantum circuit simulators mimic this nondeterministic behavior through
\emph{sampling-based} executions.
%
While outcomes may vary between executions, they collectively conform to a
specific probability distribution over multiple runs.
%
Since executable formal semantics should have deterministic results, we adopt
the notion of \emph{exact inference}, i.e., calculating the probability
distribution of possible outcomes, from probabilistic programming.
%
For \emph{static quantum circuits}, featuring end-of-circuit measurements,
exact inference is relatively straightforward, as probability distributions can
be computed by multiplying matrices corresponding to circuits' gates.
%
However, applications like quantum teleportation and iterative phase
estimation~\cite{dobsicek2007arbitrary} involve \emph{dynamic quantum
	circuits}~\cite{dynamic-circuit}.
%
These circuits modify the control flow based on mid-circuit measurements, and
their effects cannot be captured through matrix multiplication.

To enable exact inference for dynamic circuits, we adopt physicist
\citet{everett1957relative}'s \emph{many-worlds interpretation} (MWI), which
posits that each nondeterministic event generates multiple parallel
\emph{worlds} with varying outcomes.
%
Our semantics maintains a list of worlds, starting with one world and splitting
it as execution progresses.
%
A potential concern with MWI is that large circuits may create an excessive
number of worlds, leading to unreasonably long execution times and hindering
the use as a testing oracle.
%
To address this, we introduce two optimizations:
%
\emph{world merging} and \emph{efficient reset}, both aimed at reducing the
number of worlds.

The second challenge relates to the physical laws governing quantum circuits.
%
Unlike traditional programming languages for high-level computations, quantum
circuit description languages specify behavior of circuits running on quantum
hardware, bound to physical principles.
%
Physicists use \emph{density matrices} to model quantum states, and these
matrices and their transformations must fulfill specific constraints, stated
through five postulates~\cite{scherer2019mathematics}.
%
Formal semantics must align with this physical model because deviating from
these constraints would yield probability distributions unrealizable in actual
quantum devices, failing as a testing oracle for simulators.
%
Therefore, verifying the \emph{physical consistency}, i.e., adherence to these
postulates, of our semantics is critical.

To prove the physical consistency of our semantics, we introduce OpenQASMCore,
a core language of OpenQASM.
%
The simplified nature of OpenQASMCore facilitates a streamlined proof of
physical consistency.
%
We define OpenQASMCore's big-step operational semantics and prove its physical
consistency in Coq.
%
We then indirectly define OpenQASM's semantics through desugaring from OpenQASM
to OpenQASMCore.
%
As OpenQASM programs are essentially OpenQASMCore programs with syntactic
sugar, OpenQASM's physical consistency naturally follows.

To demonstrate the effectiveness of our formal executable semantics as a
testing oracle, we tested real-world quantum programming platforms,
Qiskit~\cite{qiskit} and staq~\cite{staq}, specifically a simulator and its
optimization passes in Qiskit and optimizers in staq, and identified five
previously unknown bugs with distinct root causes.
%
For testing, we collected three OpenQASM benchmark suites~\cite{cross2017open,
	li2023qasmbench, tomesh2022supermarq} and generated random circuits, some of
them designed to target Qiskit's optimization passes.
%
Of the five bugs, one was in Qiskit's transpiler, three in Qiskit's
optimization passes, and one in staq's optimizer.
%
All the bugs were confirmed by the developers and two have been already fixed.

Overall, our contributions are as follows:
%
\begin{itemize}
	%
	\item We introduce OpenQASMCore, a core language of OpenQASM
	      (\cref{ch:openqasmcore}).
	      %
	\item We define the executable formal semantics of OpenQASMCore, propose its
	      optimizations, and prove its physical consistency using Coq
	      (\cref{ch:consistency}).
	      %
	\item We indirectly define the semantics of OpenQASM by introducing desugaring from
	      OpenQASM to OpenQASMCore (\cref{ch:openqasm}).
	      %
	\item We show the effectiveness of our optimizations in reducing execution times and
	      validate the utility of our executable semantics as a testing oracle by
	      identifying five previously unknown bugs in real-world quantum programming
	      platforms (\cref{ch:evaluation}).
	      %
\end{itemize}
%
We then discuss related work (\cref{ch:related}) and conclude
(\cref{ch:conclusion}).

### 2024-10-07 kw/2-openqasmcore.tex
Add materials

% !TEX root = paper.tex

\chapter{Core Language: OpenQASMCore}
\label{ch:openqasmcore}
\noindent
In this chapter, we define the syntax of OpenQASMCore
(\cref{ch:openqasmcore:syntax}) and provide examples of quantum circuits
written in OpenQASMCore (\cref{ch:openqasmcore:example}).
%
These examples serve to offer a concise introduction to quantum computing and
aid in the intuitive understanding of OpenQASMCore's semantics.

\section{Syntax}
\label{ch:openqasmcore:syntax}

\begin{figure}[t]
	\[
		\begin{array}{@{}l}
			\text{Bit value} \quad b \in \{0,1\} \qquad
			\text{Real number} \quad r \in \mathbb{R} \qquad
			\text{Classical bit} \quad c \in \mathbb{N} \qquad
			\text{Qubit} \quad q \in \mathbb{N} \\
			\text{Statement} \quad s ::= \inop
			\ |\ \iseq{s}{s}
			\ |\ \iif{c}{b}{s}
			\ |\ \irotate{r}{r}{r}{q}
			\ |\ \icnot{q}{q}
			\ |\ \imeasure{q}{c}
			\ |\ \ireset{q}
		\end{array}
	\]
	\caption{Syntax of OpenQASMCore}
	\label{fig:syntax-openqasmcore}
\end{figure}

\noindent
\cref{fig:syntax-openqasmcore} defines the syntax of OpenQASMCore.
%
The metavariable $b$ ranges over (classical) bit values, each of which is
either $0$ or $1$.
%
The metavariable $r$ ranges over real numbers.
%
The metavariable $c$ ranges over natural numbers when used to label classical
bits.
%
Similarly, the metavariable $q$ ranges over natural numbers when used to label
qubits (quantum bits).
%
The numbers of classical bits and qubits used by each program (i.e., circuit)
are finite.
%
In a program with $n$ classical bits and $m$ qubits, classical bits are
numbered from $0$ to $n-1$, and qubits are numbered from $0$ to $m-1$.
%
Since classical bits and qubits are used in different syntactic contexts, it is
always apparent whether a specific natural number corresponds to a classical
bit or a qubit.
%
Finally, the metavariable $s$ ranges over statements, and a program consists of
a single statement.

The language provides seven kinds of statements.
%
The first three kinds are commonly used in classical computing.
%
$\inop$ is a no-op, and $\iseq{s_1}{s_2}$ is a sequencing statement that
sequentially executes $s_1$ and $s_2$.
%
$\iif{c}{b}{s}$ is a conditional statement that executes $s$ if the value of
classical bit $c$ equals $b$.
%
The next four kinds are specific to quantum computing.
%
$\irotate{r_1}{r_2}{r_3}{q}$ is a rotation gate that alters the state of qubit
$q$ by rotating it by $r_1$, $r_2$, and $r_3$ radians around certain axes.
%
$\icnot{q_1}{q_2}$ is a controlled NOT (CNOT) gate that flips qubit $q_2$ if
qubit $q_1$ equals $1$.
%
$\imeasure{q}{c}$ is a measurement statement that measures qubit $q$, yielding a
non-deterministic result, which is then assigned to classical bit $c$.
%
$\ireset{q}$ is a reset statement that prepares qubit $q$ in $\ket{0}$.
%
Further details on these four quantum-specific statements are provided in
\cref{ch:openqasmcore:example}.

\section{Examples}
\label{ch:openqasmcore:example}

\noindent
We now provide example quantum circuits while introducing basic concepts in
quantum programming.
%
These concepts are standard in the physics
literature~\cite{hayashi2014introduction}.
%
First, we explain the execution of a single-qubit circuit, which is simpler
than a multi-qubit circuit.
%
We present the same single-qubit circuit twice: once using state vectors
(\cref{ch:openqasmcore:example:vector}) and once using density matrices
(\cref{ch:openqasmcore:example:matrix}).
%
State vectors offer an intuitive explanation of quantum circuits, while density
matrices are closely related the postulates to which our formal semantics
adheres.
%
Next, we show a multi-qubit circuit to illustrate how the concepts from a
single-qubit circuit can be generalized (\cref{ch:openqasmcore:example:multi}).

\subsection{Single-qubit circuit (with state vector)}
\label{ch:openqasmcore:example:vector}

\noindent
A \emph{qubit} is the base unit of quantum computing.
%
Unlike a classical bit that can take only values 0 or 1, a qubit exists in a
\emph{superposition} state, represented as
$\ket{\psi}=\alpha\ket{0}+\beta\ket{1}$, where $\alpha$ and $\beta$ are complex
numbers, and $\ket{0}$ and $\ket{1}$ represent states 0 and 1, respectively.
%
Measurement of a qubit yields either 0 or 1, with probabilities $|\alpha|^2$
and $|\beta|^2$, respectively.
%
The sum of these probabilities must be equal to 1, satisfying
$|\alpha|^2+|\beta|^2=1$.
%
We refer to $\ket{\psi}$ as the \emph{state vector} of the qubit since it
denotes a column vector as follows:
%
\[
	%
	\ket{0}=\mtrx{1\\0}
	%
	\qquad\qquad
	%
	\ket{1}=\mtrx{0\\1}
	%
	\qquad\qquad
	%
	\ket{\psi}=\alpha\ket{0}+\beta\ket{1}=\mtrx{\alpha\\\beta}
	%
\]

\cref{fig:example:code} presents an example quantum circuit implemented in
OpenQASMCore, while \cref{fig:example:circuit} provides a visual representation
of the circuit.
%
It is a dynamic circuit because it incorporates a measurement of a qubit in the
middle of execution (line 2), changing the control flow based on the
measurement outcome (line 3).
%
The circuit employs one qubit, $q_0$, and two classical bits, $c_0$ and $c_1$.
%
By convention, the initial state of the qubit is $\ket{0}$, and the initial
value of each classical bit is $0$.
%
We now explain the circuit's execution line by line.

\paragraph{Line 1}

We apply the $\ugate{\frac{\pi}{2}}{0}{\pi}$ gate to qubit $q_0$.
%
This gate is known as the \emph{Hadamard gate} and denoted as $H$ in the
diagram.
%
Its purpose is to create a superposition.
%
The effect of any rotation gate ($\mathsf{U}$ gate) is represented by matrix
multiplication with the state vector.
%
The matrix representing the Hadamard gate is
$\frac{1}{\sqrt{2}}\smtrx{1&1\\1&-1}$.
%
(A general formula for computing the matrix representing an arbitrary rotation
gate is discussed in \cref{ch:consistency:semantics}.)
%
Consequently, the state vector changes as follows after the gate:
%
\[
	%
	\frac{1}{\sqrt{2}}\mtrx{1&1\\1&-1}\ket{0}
	%
	=\frac{1}{\sqrt{2}}\mtrx{1&1\\1&-1}\mtrx{1\\0}
	%
	=\frac{1}{\sqrt{2}}\mtrx{1\\1}
	%
	=\frac{1}{\sqrt{2}}\ket{0}+\frac{1}{\sqrt{2}}\ket{1}
	%
\]

\begin{figure}[t]
	\centering
	\begin{subfigure}{0.3\textwidth}
		$
			\begin{array}{@{}r@{~}|l@{}}
				\iseq
				{{\scriptsize\texttt{1}} & \irotate{\frac{\pi}{2}}{0}{\pi}{0}}
				{                                                                          \\ \iseq
				{{\scriptsize\texttt{2}} & \imeasure{0}{0}}
				{                                                                          \\ \iseq
				{{\scriptsize\texttt{3}} & \iif{0}{0}{\irotate{\frac{\pi}{2}}{0}{\pi}{0}}}
				{                                                                          \\
				{\scriptsize\texttt{4}}  & \imeasure{0}{1}
				} } }
			\end{array}
		$
		\caption{OpenQASMCore}
		\label{fig:example:code}
	\end{subfigure}
	\hspace{2em}
	\begin{subfigure}{0.4\textwidth}
		\centering
		\begin{quantikz}[wire types={q,c,c},row sep={0.5cm,between origins}]
			\lstick{$q_0$} & \gate{H} & \meter{} \wire[d][1]{c} & \gate{H} & \meter{} \wire[d][2]{c}&\\
			\lstick{$c_0$} & & & \octrl[vertical wire=c]{-1} & &\\
			\lstick{$c_1$} & & & & &\\
		\end{quantikz}
		\caption{Diagram}
		\label{fig:example:circuit}
	\end{subfigure}
	\caption{An example single-qubit dynamic quantum circuit}
	\label{fig:example}
\end{figure}

\paragraph{Line 2}

We measure qubit $q_0$ and assign the outcome to classical bit $c_0$.
%
According to the many-worlds interpretation, this measurement splits the world
into two branches: one where $0$ is obtained and another where $1$ is obtained.
%
In the first branch, $0$ is assigned to $c_0$, while the superposition of $q_0$
collapses to $\ket{0}$.
%
Similarly, in the second branch, $1$ is assigned to $c_0$, while the
superposition of $q_0$ collapses to $\ket{1}$.
%
Each branch is reachable from the previous world with a probability of $1/2$.
%
We represent a world with a tuple consisting of a state vector, classical bits,
and the corresponding probability of reaching that branch.
%
Therefore, we get $(\smtrx{1\\0},[0,0],\frac{1}{2})$ and
$(\smtrx{0\\1},[1,0],\frac{1}{2})$.

\paragraph{Line 3}

If the value of $c_0$ equals $0$, we apply the Hadamard gate to qubit $q_0$
once again.
%
In the first world, $(\smtrx{1\\0},[0,0],\frac{1}{2})$, $c_0$ holds $0$, and
the gate is applied to $q_0$, resulting in
$(\frac{1}{\sqrt{2}}\smtrx{1\\1},[0,0],\frac{1}{2})$.
%
In the second world, $(\smtrx{0\\1},[1,0],\frac{1}{2})$, $c_0$ contains $1$, so
no changes occur.
%
Consequently, we obtain two worlds:
$(\frac{1}{\sqrt{2}}\smtrx{1\\1},[0,0],\frac{1}{2})$ and
$(\smtrx{0\\1},[1,0],\frac{1}{2})$.

\paragraph{Line 4}

We measure qubit $q_0$ and assign the outcome to classical bit $c_1$.
%
In the first world, the measurement yields $0$ or $1$ with probabilities of
$1/2$.
%
Consequently, the world splits into two: $(\smtrx{1\\0},[0,0],\frac{1}{4})$ and
$(\smtrx{0\\1},[0,1],\frac{1}{4})$.
%
In the second world, the measurement always yields $1$.
%
Therefore, the world does not split and only updates the value of $c_1$,
resulting in $(\smtrx{0\\1},[1,1],\frac{1}{2})$.
%
After the measurement, we have three worlds:
$(\smtrx{1\\0},[0,0],\frac{1}{4})$, $(\smtrx{0\\1},[0,1],\frac{1}{4})$, and
$(\smtrx{0\\1},[1,1],\frac{1}{2})$.
%
From this, we can conclude that the circuit outputs, i.e., the values in the
classical bits, are either $00$, $01$, or $11$, with probabilities of $1/4$,
$1/4$, and $1/2$, respectively.

\subsection{Single-qubit circuit (with density matrix)}
\label{ch:openqasmcore:example:matrix}

\noindent
The \emph{density matrix}, which is the linear combination of outer products of
a state vector with itself, is used to represent the state of a qubit as a
matrix instead of a vector. For example, consider a density matrix with a
single outer product:
%
\[
	%
	\rho
	%
	=\mtrx{\alpha\\\beta}\mtrx{\alpha\\\beta}^*
	%
	=\mtrx{\alpha\\\beta}\mtrx{\alpha^*&\beta^*}
	%
	=\mtrx{\alpha\alpha^*&\alpha\beta^*\\\beta\alpha^*&\beta\beta^*}
	%
\]
%
Note that $\alpha^*$ is the complex conjugate of a complex number $\alpha$, and
$M^*$ is the conjugate transpose of a matrix $M$.
%
The diagonal elements of a density matrix correspond to the probabilities of
each measurement outcome because $\alpha\alpha^*$ equals $|\alpha|^2$ for every
complex number $\alpha$.
%
We now explain the execution of \cref{fig:example}'s circuit using density
matrices.
%
The initial density matrix is $\smtrx{1\\0}\smtrx{1&0}=\smtrx{1&0\\0&0}$.

\paragraph{Line 1}

We apply the Hadamard gate to $q_0$.
%
The effect of a rotation gate is represented by matrix multiplication with the
density matrix.
%
Specifically, the updated density matrix is given by $M\rho M^*$, where $M$ is
the matrix representing the gate and $\rho$ is the current density matrix.
%
Since the Hadamard gate is represented as
$\frac{1}{\sqrt{2}}\smtrx{1&1\\1&-1}$, the density matrix is updated as
follows:
%
\[
	%
	\frac{1}{\sqrt{2}}\mtrx{1&1\\1&-1}\cdot\mtrx{1&0\\0&0}\cdot\frac{1}{\sqrt{2}}\mtrx{1&1\\1&-1}
	%
	=\frac{1}{2}\mtrx{1&1\\1&1}
	%
\]
%
This result matches the result obtained using the state vector:
%
\[
	\frac{1}{2}\mtrx{1&1\\1&1}=\frac{1}{\sqrt{2}}\mtrx{1\\1}\cdot\frac{1}{\sqrt{2}}\mtrx{1&1}.
\]

\paragraph{Line 2}

We measure $q_0$ and assign the outcome to $c_0$.
%
For each possible measurement outcome, there exists a corresponding matrix,
which is used to compute the probability of the outcome and the density matrix
after the measurement.
%
The matrices for outcomes $0$ and $1$ are $\smtrx{1&0\\0&0}$ and
$\smtrx{0&0\\0&1}$, respectively.
%
The probability of the outcome is $P=\trace{\rho M}$, and the density matrix
after the measurement is $\frac{1}{P}M\rho M$, where $M$ is the matrix
corresponding to an outcome and $\rho$ is the current density matrix.
%
Note that $\trace{M}$ is the \emph{trace} of $M$, i.e., the sum of $M$'s
diagonal elements.
%
We compute the probability of $0$ being the outcome and the updated density
matrix as follows:
%
\[
	%
	\trace{\frac{1}{2}\mtrx{1&1\\1&1}\mtrx{1&0\\0&0}}
	%
	=\trace{\frac{1}{2}\mtrx{1&0\\1&0}}
	%
	=\frac{1}{2}
	%
	\qquad
	%
	\frac{1}{\sfrac{1}{2}}\cdot\mtrx{1&0\\0&0}\cdot\frac{1}{2}\mtrx{1&1\\1&1}\cdot\mtrx{1&0\\0&0}
	%
	=\mtrx{1&0\\0&0}
	%
\]
%
Similarly, we compute the probability of $1$ being the outcome and the updated
density matrix:
%
\[
	%
	\trace{\frac{1}{2}\mtrx{1&1\\1&1}\mtrx{0&0\\0&1}}
	%
	=\trace{\frac{1}{2}\mtrx{0&1\\0&1}}
	%
	=\frac{1}{2}
	%
	\qquad
	%
	\frac{1}{\sfrac{1}{2}}\cdot\mtrx{0&0\\0&1}\cdot\frac{1}{2}\mtrx{1&1\\1&1}\cdot\mtrx{0&0\\0&1}
	%
	=\mtrx{0&0\\0&1}
	%
\]
%
As a result, we obtain two worlds: $(\smtrx{1&0\\0&0},[0,0],\frac{1}{2})$ and
$(\smtrx{0&0\\0&1},[1,0],\frac{1}{2})$.
%
This result is also consistent with the result obtained using the state vector:
%
$\smtrx{1&0\\0&0}=\smtrx{1\\0}\smtrx{1&0}$ and
$\smtrx{0&0\\0&1}=\smtrx{0\\1}\smtrx{0&1}$.

\paragraph{Lines 3--4}

Similar to the process using state vectors, we end up with three worlds in the
end: $(\smtrx{1&0\\0&0},[0,0],\frac{1}{4})$,
$(\smtrx{0&0\\0&1},[0,1],\frac{1}{4})$, and
$(\smtrx{0&0\\0&1},[1,1],\frac{1}{2})$.

\subsection{Multi-qubit circuit}
\label{ch:openqasmcore:example:multi}

\noindent
Both state vectors and density matrices easily extend to multi-qubit systems.
%
The state of an $n$-qubit system is represented by a state vector of $2^n$
entries or a density matrix of dimension $2^n \times 2^n$.
%
For example, in a $2$-qubit system, a measurement can produce the outcomes
$00$, $01$, $10$, or $11$.
%
A system that produces these outcomes with corresponding probabilities
$|\alpha|^2$, $|\beta|^2$, $|\gamma|^2$, and $|\delta|^2$ is represented by the
following state vector and density matrix:
%
\[
	%
	\ket{\psi}
	%
	=\alpha\ket{00}+\beta\ket{01}+\gamma\ket{10}+\delta\ket{11}
	%
	=\alpha\mtrx{1\\0\\0\\0}+\beta\mtrx{0\\1\\0\\0}+\gamma\mtrx{0\\0\\1\\0}+\delta\mtrx{0\\0\\0\\1}
	%
	=\mtrx{\alpha\\\beta\\\gamma\\\delta}
	%
\]
\[
	%
	\rho
	%
	=\mtrx{\alpha\\\beta\\\gamma\\\delta}\mtrx{\alpha\\\beta\\\gamma\\\delta}^*
	%
	=\mtrx{\alpha\\\beta\\\gamma\\\delta}\mtrx{\alpha^*&\beta^*&\gamma^*&\delta^*}
	%
	=\mtrx{
		%
		\alpha\alpha^*&\alpha\beta^*&\alpha\gamma^*&\alpha\delta^*\\
		%
		\beta\alpha^*&\beta\beta^*&\beta\gamma^*&\beta\delta^*\\
		%
		\gamma\alpha^*&\gamma\beta^*&\gamma\gamma^*&\gamma\delta^*\\
		%
		\delta\alpha^*&\delta\beta^*&\delta\gamma^*&\delta\delta^*
		%
	}
	%
\]

\begin{figure}[t]
	\centering
	\begin{subfigure}{0.3\textwidth}
		$
			\begin{array}{@{}r@{~}|l@{}}
				\iseq
				{{\scriptsize\texttt{1}}                          & \irotate{\frac{\pi}{2}}{0}{\pi}{0}}
				{                                                                                       \\
				\iseq
				{{\scriptsize\texttt{2}}                          & \icnot{0}{1}}
				{                                                                                       \\
				\iseq
				{{\scriptsize\texttt{3}}                          & \imeasure{0}{0}}
				{ \\{\scriptsize\texttt{4}} & \imeasure{1}{1}}
				}
				}
			\end{array}
		$
		\caption{OpenQASMCore}
		\label{fig:example-multi:code}
	\end{subfigure}
	\hspace{2em}
	\begin{subfigure}{0.4\textwidth}
		\centering
		\begin{quantikz}[wire types={q,q,c,c},row sep={0.4cm,between origins}]
			\lstick{$q_0$} & \gate{H} & \ctrl{1} & \meter{} \wire[d][2]{c} & & \\
			\lstick{$q_1$} & & \targ{} & & \meter{} \wire[d][2]{c} & \\
			\lstick{$c_0$} & & & & & \\
			\lstick{$c_1$} & & & & &
		\end{quantikz}
		\caption{Diagram}
		\label{fig:example-multi:circuit}
	\end{subfigure}
	\caption{An example multi-qubit quantum circuit}
	\label{fig:example-multi}
\end{figure}

\cref{fig:example-multi} illustrates an example multi-qubit quantum circuit.
%
The circuit utilizes two qubits, $q_0$ and $q_1$, as well as two classical
bits, $c_0$ and $c_1$.
%
Each qubit is initially set to $\ket{0}$, resulting in the following initial
state vector and density matrix of the whole system:
%
\[
	%
	\ket{00}=
	%
	\mtrx{1\\0\\0\\0}
	%
	\qquad\qquad\qquad\qquad
	%
	\mtrx{1\\0\\0\\0}\mtrx{1\\0\\0\\0}^*
	%
	=\mtrx{1&0&0&0\\0&0&0&0\\0&0&0&0\\0&0&0&0}
	%
\]
%
We now explain the circuit's execution line by line.

\paragraph{Line 1}

We apply the Hadamard gate to $q_0$.
%
Since the density matrix is of dimension $4\times4$, the matrix representing
the gate also needs to be of dimension $4\times4$.
%
This matrix can be obtained by taking the tensor product of the original
$2\times2$ matrix representing the Hadamard gate and an identity matrix.
%
The matrix representing the application of the Hadamard gate to $q_0$ is as
follows:
%
\[
	%
	\frac{1}{\sqrt{2}}\mtrx{1&1\\1&-1}\otimes\mtrx{1&0\\0&1}
	%
	=\frac{1}{\sqrt{2}}\mtrx{1&0&1&0\\0&1&0&1\\1&0&-1&0\\0&1&0&-1}
	%
\]
%
A general formula for computing a $2^n\times2^n$ matrix from a $2\times2$
matrix is discussed in \cref{ch:consistency:semantics}.
%
Using the above matrix, we compute the density matrix after the gate as
follows:
%
\[
	%
	\frac{1}{\sqrt{2}}\mtrx{1&0&1&0\\0&1&0&1\\1&0&-1&0\\0&1&0&-1}
	%
	\cdot
	%
	\mtrx{1&0&0&0\\0&0&0&0\\0&0&0&0\\0&0&0&0}
	%
	\cdot
	%
	\frac{1}{\sqrt{2}}\mtrx{1&0&1&0\\0&1&0&1\\1&0&-1&0\\0&1&0&-1}
	%
	= \frac{1}{2}\mtrx{1&0&1&0\\0&0&0&0\\1&0&1&0\\0&0&0&0}
	%
\]
%
The diagonal elements imply that the measurement outcome of the system is
either $00$ or $10$ with probabilities $1/2$.
%
In other words, the measurement outcome of $q_0$ has a probability of $1/2$ of
being $0$ or $1$, while the measurement outcome of $q_1$ is always $0$.
%
This aligns with the intuition that the Hadamard gate applied to $q_0$ should
create a superposition in $q_0$ only, leaving $q_1$ unaffected.

\paragraph{Line 2}

We apply the CNOT gate to $q_0$ and $q_1$, with $q_0$ serving as the
\emph{control} and $q_1$ as the \emph{target}.
%
Intuitively, it flips the target if the control is $1$.
%
The purpose of this gate is to \emph{entangle} the qubits, establishing a
quantum mechanical dependency between them.
%
The CNOT gate operates on two distinct qubits and is represented by a
$4\times4$ matrix: $\Smtrx{1&0&0&0\\0&1&0&0\\0&0&0&1\\0&0&1&0}$.
%
We generalize this matrix for a system with more than two qubits in
\cref{ch:consistency:semantics}.
%
We compute the density matrix after the gate as follows:
%
\[
	%
	\mtrx{1&0&0&0\\0&1&0&0\\0&0&0&1\\0&0&1&0}
	%
	\cdot
	%
	\frac{1}{2}\mtrx{1&0&1&0\\0&0&0&0\\1&0&1&0\\0&0&0&0}
	%
	\cdot
	%
	\mtrx{1&0&0&0\\0&1&0&0\\0&0&0&1\\0&0&1&0}
	%
	=\frac{1}{2}\mtrx{1&0&0&1\\0&0&0&0\\0&0&0&0\\1&0&0&1}
	%
\]
%
The diagonal elements indicate that the measurement outcome of the system can
be either $00$ or $11$ with probabilities of $1/2$.
%
This implies that measuring one qubit allows determining the measurement
outcome of the other qubit, demonstrating the entanglement between $q_0$ and
$q_1$.

\paragraph{Line 3}

We measure $q_0$ and assign the outcome to $c_0$.
%
The matrices corresponding to each measurement outcome are generalized using
tensor product, similar to the matrices representing rotation gates.
%
The matrices corresponding to outcomes $0$ and $1$ from the measurement of
$q_0$ are as follows:
%
\[
	%
	\mtrx{1&0\\0&0}\otimes\mtrx{1&0\\0&1}
	%
	=\mtrx{1&0&0&0\\0&1&0&0\\0&0&0&0\\0&0&0&0}
	%
	\qquad\qquad
	%
	\mtrx{0&0\\0&1}\otimes\mtrx{1&0\\0&1}
	%
	=\mtrx{0&0&0&0\\0&0&0&0\\0&0&1&0\\0&0&0&1}
	%
\]
%
We compute the probability of $0$ being the outcome and the updated density
matrix as follows:
%
\[
	%
	\btrace{\frac{1}{2}\mtrx{1&0&0&1\\0&0&0&0\\0&0&0&0\\1&0&0&1}\mtrx{1&0&0&0\\0&1&0&0\\0&0&0&0\\0&0&0&0}}
	%
	=\btrace{\frac{1}{2}\mtrx{1&0&0&0\\0&0&0&0\\0&0&0&0\\1&0&0&0}}
	%
	=\frac{1}{2}
	%
\]
\[
	%
	\frac{1}{\sfrac{1}{2}}
	%
	\cdot\mtrx{1&0&0&0\\0&1&0&0\\0&0&0&0\\0&0&0&0}
	%
	\cdot\frac{1}{2}\mtrx{1&0&0&1\\0&0&0&0\\0&0&0&0\\1&0&0&1}
	%
	\cdot\mtrx{1&0&0&0\\0&1&0&0\\0&0&0&0\\0&0&0&0}
	%
	=\mtrx{1&0&0&0\\0&0&0&0\\0&0&0&0\\0&0&0&0}
	%
\]
%
We can compute the probability of $1$ being the outcome and the updated density
matrix in a similar fashion.
%
Consequently, we obtain two worlds:
$(\Smtrx{1&0&0&0\\0&0&0&0\\0&0&0&0\\0&0&0&0},[0,0],\frac{1}{2})$ and
$(\Smtrx{0&0&0&0\\0&0&0&0\\0&0&0&0\\0&0&0&1},[1,0],\frac{1}{2})$.

\paragraph{Line 4}

We measure $q_1$ and assign the outcome to $c_1$.
%
The matrices corresponding to the outcomes $0$ and $1$ from the measurement of
$q_1$ are as follows:
%
\[
	%
	\mtrx{1&0\\0&1}\otimes\mtrx{1&0\\0&0}
	%
	=\mtrx{1&0&0&0\\0&0&0&0\\0&0&1&0\\0&0&0&0}
	%
	\qquad\qquad
	%
	\mtrx{1&0\\0&1}\otimes\mtrx{0&0\\0&1}
	%
	=\mtrx{0&0&0&0\\0&1&0&0\\0&0&0&0\\0&0&0&1}
	%
\]
%
Using these matrices, we can compute that the probability of the outcome being
$0$ is $1$ in the first world, and the probability of the outcome being $1$ is
$1$ in the second world.
%
Therefore, the execution terminates with two worlds:
%
$(\Smtrx{1&0&0&0\\0&0&0&0\\0&0&0&0\\0&0&0&0},[0,0],\frac{1}{2})$ and
$(\Smtrx{0&0&0&0\\0&0&0&0\\0&0&0&0\\0&0&0&1},[1,1],\frac{1}{2})$.

### 2024-10-07 kw/3-consistency.tex
Add materials

% !TEX root = main.tex
\chapter{Executable Formal Semantics and Its Physical Consistency}
\label{ch:consistency}

\noindent
In this section, we define the executable formal semantics of OpenQASMCore
(\cref{ch:consistency:semantics}) and introduce two optimization techniques
(\cref{ch:consistency:optimization}).
%
We then prove the physical consistency of the semantics
(\cref{ch:consistency:consistency}).
%
The semantics and the proof are mechanized in Coq and publicly
available~\cite{artifact}.

\section{Semantics}
\label{ch:consistency:semantics}

\begin{figure}[t]
	\[
		\begin{array}{@{}rr@{~}c@{~}l@{\qquad\qquad}r@{~}c@{~}l@{}}
			\text{Density matrix}  & \rho   & \in                                                    & \bigcup_{n\in\mathbb{N}}\mathcal{M}_{2^n,2^n}(\mathbb{C}) &
			\sunitary{\rho}{M}     & =      & M\rho M^*                                                                                                            \\
			\text{Classical state} & \sigma & \in                                                    & \bigcup_{n\in\mathbb{N}}\mathbb{Z}_n\rightarrow \{0,1\}   &
			\sprob{\rho}{M}        & =      & \trace{\rho M}                                                                                                       \\
			\text{Probability}     & P      & \in                                                    & \mathbb{R}_{[0,1]}                                        &
			\smeasure{\rho}{M}     & =      & \frac{1}{\sprob{\rho}{M}}M\rho M                                                                                     \\
			\text{World}           & w      & ::=                                                    & (\rho,\sigma,P)                                           &
			\qubits{\rho}          & =      & n\ \text{if}\ \rho\in\mathcal{M}_{2^n,2^n}(\mathbb{C})
		\end{array}
	\]
	\caption{Semantic elements and functions}
	\label{fig:semantic-elements}
\end{figure}

\noindent
\cref{fig:semantic-elements} shows the definitions related to the semantics of
OpenQASMCore.
%
The metavariable $\rho$ ranges over density matrices, which are complex
matrices of dimensions $2^n\times2^n$, where $n$ is a natural number.
%
The metavariable $\sigma$ ranges over classical states, which are mappings from
classical bits to their values.
%
The metavariable $P$ ranges over probabilities of a world, which are real
numbers inclusively between $0$ and $1$.
%
The metavariable $w$ ranges over worlds, which are triples consisting of a
density matrix, a classical state, and a probability.
%
The functions $\embox{gate}$, $\embox{prob}$, and $\embox{msr}$ compute the
updated density matrix after a gate, the probability of a measurement outcome,
and the updated density matrix after a measurement, respectively.
%
Each function takes a density matrix and a matrix representing a gate/outcome
as inputs, and the result is defined only when the two matrices are of the same
dimension.
%
In addition, $\smeasure{\rho}{M}$ is also undefined when $\sprob{\rho}{M}$ is
$0$.
%
The function $\embox{qbts}$ returns the number of qubits in the system
represented by a given density matrix.

\begin{figure}[t]
	\[
		\small
		\begin{array}{@{}c@{}}
			\fbox{$\eval{\rep{w}}{s}{\rep{w}}$}
			\qquad\qquad
			\inferrule
			{ \rep{\eval{w}{s}{\rep{w'}}} }
			{ \eval{\rep{w}}{s}{\rep{\rep{w'}}} }
			\quad[\textsc{Eval-Many}]
			\qquad\qquad
			\eval{w}{\inop}{w}
			\quad[\textsc{Eval-Nop}]
			\\[1.5em]
			\inferrule
			{ \eval{w}{s_1}{\rep{w'}}                                                                 \\ \eval{\rep{w'}}{s_2}{\rep{w''}} }
			{ \eval{w}{\iseq{s_1}{s_2}}{\rep{w''}} }
			\quad[\textsc{Eval-Seq}]
			\qquad
			\inferrule
			{
			w=(\rho,\sigma,P)                                                                         \\
				\sigma(c)\not=b
			}
			{ \eval{w}{\iif{c}{b}{s}}{w} }
			\quad[\textsc{Eval-If-False}]
			\\[1.5em]
			\inferrule
			{
			w=(\rho,\sigma,P)                                                                         \\
			\sigma(c)=b                                                                               \\
				\eval{w}{s}{\rep{w'}}
			}
			{ \eval{w}{\iif{c}{b}{s}}{\rep{w'}} }
			\quad[\textsc{Eval-If-True}]
			\\[1.5em]
			\inferrule
			{ \qubits{\rho}=n }
			{
				\eval{(\rho,\sigma,P)}{\irotate{r_1}{r_2}{r_3}{q}}{(\sunitary{\rho}{\mrotate{n,q,r_1,r_2,r_3}},\sigma,P)}
			}
			\quad[\textsc{Eval-Rotate}]
			\\[1.5em]
			\inferrule
			{ \qubits{\rho}=n }
			{
				\eval{(\rho,\sigma,P)}{\icnot{q_1}{q_2}}{(\sunitary{\rho}{\mcnot{n,q_1,q_2}},\sigma,P)}
			}
			\quad[\textsc{Eval-Cnot}]
			\\[1.5em]
			\inferrule
			{
			\qubits{\rho}=n                                                                           \\
			\sprob{\rho}{\mproj{n,q,0}} \neq 0                                                        \\
			\sprob{\rho}{\mproj{n,q,1}} \neq 0                                                        \\\\
			w_0=(\smeasure{\rho}{\mproj{n,q,0}},\sigma[c\mapsto0],P\times\sprob{\rho}{\mproj{n,q,0}}) \\
				w_1=(\smeasure{\rho}{\mproj{n,q,1}},\sigma[c\mapsto1],P\times\sprob{\rho}{\mproj{n,q,1}})
			}
			{ \eval{(\rho,\sigma,P)}{\imeasure{q}{c}}{w_0,w_1} }
			\quad[\textsc{Eval-Measure}]
			\\[1.5em]
			\inferrule
			{
			\qubits{\rho}=n                                                                           \\
			\sprob{\rho}{\mproj{n,q,1}}=0                                                             \\\\
				w_0=(\smeasure{\rho}{\mproj{n,q,0}},\sigma[c\mapsto0],P\times\sprob{\rho}{\mproj{n,q,0}})
			}
			{ \eval{(\rho,\sigma,P)}{\imeasure{q}{c}}{w_0} }
			\quad[\textsc{Eval-Measure-0}]
			\\[1.5em]
			\inferrule
			{
			\qubits{\rho}=n                                                                           \\
			\sprob{\rho}{\mproj{n,q,0}}=0                                                             \\\\
				w_1=(\smeasure{\rho}{\mproj{n,q,1}},\sigma[c\mapsto1],P\times\sprob{\rho}{\mproj{n,q,1}})
			}
			{ \eval{(\rho,\sigma,P)}{\imeasure{q}{c}}{w_1} }
			\quad[\textsc{Eval-Measure-1}]
			\\[1.5em]
			\inferrule
			{
			w=(\rho,\sigma,P)                                                                         \\
			c_{\embox{\tiny reset}} \notin \dom{\sigma}                                               \\\\
				\eval {w} {
					\iseq{
						\imeasure{q}{c_{\embox{\tiny reset}}}
					}{
						\iif{c_{\embox{\tiny reset}}}{1}{\irotate{\pi}{0}{\pi}{q}}
					}} {\rep{w'}}
			} { \eval {w} {\ireset{q}} {\rep{w'}}}
			\quad[\textsc{Eval-Reset}]
		\end{array}
	\]
	\caption{Semantics of OpenQASMCore}
	\label{fig:semantics}
\end{figure}

\cref{fig:semantics} defines the big-step operational semantics of OpenQASMCore
in the form of \fbox{$\eval{\rep{w}}{s}{\rep{w}}$}.
%
Following the notation used by \citet{vytiniotis2013halo} and
\citet{coughlin2014fissile}, an overline above material represents zero or more
repetitions of the material, separated by commas, with a subscript added to
each metavariable in the material.
%
Thus, $\eval{\rep{w}}{s}{\rep{w'}}$ denotes
$\eval{w_1,w_2,\ldots,w_n}{s}{w_1',w_2',\ldots,w_m'}$, indicating that the
worlds $w_1$, $w_2$, \ldots, $w_n$ transform into $w_1'$, $w_2'$, \ldots,
$w_m'$ when $s$ is executed.

Most rules handle the case where a single world is provided as input, and they
are generalized to accommodate multiple worlds through rule
	[\textsc{Eval-Many}].
%
This rule states that when multiple worlds are given, the statement runs in
each world, and the output is determined by collecting the outputs from each
individual run.
%
Note that when overlines are nested, the outermost one expands first, while the
copies of each inner overline expand independently and may have different
lengths.

The semantics of classical statements are conventional, as defined by rules
	[\textsc{Eval-Nop}], [\textsc{Eval-Seq}], [\textsc{Eval-If-False}], and
	[\textsc{Eval-If-True}].
%
$\inop$ does not alter the world state.
%
$\iseq{s_1}{s_2}$ executes $s_1$ first in the given world and subsequently
executes $s_2$ in every world generated by $s_1$.
%
$\iif{c}{b}{s}$ executes $s$ if the value of $c$ equals $b$.

Rule [\textsc{Eval-Rotate}] defines the semantics of a rotation gate, which
updates only the density matrix.
%
The definition of the matrix $\mrotate{n,q,r_1,r_2,r_3}$ can be found in
\cref{fig:gen22}, which illustrates the generalization of $2\times2$ matrices
representing gates/measurements operating on a single qubit.
%
The function $\embox{gen}_{n,q}$ returns a $2^n\times2^n$ matrix that affects
the $q$-th qubit (indexed from zero) in an $n$-qubit system by generalizing a
given $2\times2$ matrix.
%
Note that $\identity{m}$ represents the identity matrix of dimension
$2^m\times2^m$.
%
The effect of the $\ugate{r_1}{r_2}{r_3}$ gate on a single qubit is denoted as
$\mrotate{r_1,r_2,r_3}$ and is generalized by $\embox{gen}_{n,q}$ to form
$\mrotate{n,q,r_1,r_2,r_3}$.

Rule [\textsc{Eval-Cnot}] defines the semantics of the CNOT gate, which also
updates only the density matrix.
%
\cref{fig:cnot} defines the matrix $\mcnot{n,q_1,q_2}$, which flips the
$q_2$-th qubit using the $q_1$-th qubit as control in an $n$-qubit system, by
generalizing the $4\times4$ matrix $\mcnot{}$.
%
Unlike rotation matrices, $\mcnot{}$ is not simply generalized by tensor
products with identity matrices due to its ability to operate on two qubits,
which may be non-adjacent.
%
To address this, we introduce the matrix $\mswap{n,q_1,q_2}$, which swaps the
$q_1$-th and $q_2$-th qubits in an $n$-qubit system, and employ this matrix to
define $\mcnot{n,q_1,q_2}$.
%
Since $\mswap{n,q_1,q_2}$ is meaningful only when $q_1$ differs from $q_2$, it
is an identity matrix when $q_1$ equals $q_2$.
%
For $q_1\not=q_2$, $\mswap{n,q_1,q_2}$ is recursively defined, beginning with
the base case $\mswap{}$, which swaps the qubits in a $2$-qubit system.
%
We then define $\mswap{n,0,n-1}$, which swaps the $0$-th and $(n-1)$-th qubits
by swapping the $0$-th and $1$-st, then the $1$-st and $(n-1)$-th, and finally
$0$-th and $1$-st.
%
Subsequently, we define $\mswap{n,q_1,q_2}$ using tensor products.
%
Finally, $\mcnot{n,q_1,q_2}$ is defined by swapping the $0$-th and $q_1$-th
qubits, then the $1$-st and $q_2$-th qubits, flipping the $1$-st qubit using
the $0$-th qubit as control, and finally swapping back the $0$-th and $q_1$-th
qubits as well as the $1$-st and $q_2$-th qubits.

Rules [\textsc{Eval-Measure}], [\textsc{Eval-Measure-0}], and
	[\textsc{Eval-Measure-1}] define the semantics of a measurement statement.
%
Rule [\textsc{Eval-Measure}] handles the case where both $0$ and $1$ are
possible outcomes by splitting the world into two, while rules
	[\textsc{Eval-Measure-0}] and [\textsc{Eval-Measure-1}] handle the cases where
only one of them is a possible outcome.
%
The matrices $\mproj{n,q,0}$ and $\mproj{n,q,1}$ are defined by generalizing
$\mproj{0}$ and $\mproj{1}$ using tensor products, as shown in
\cref{fig:gen22}.

Rule [\textsc{Eval-Reset}] defines the semantics of a reset statement, which
collapses a qubit's superposition and sets its state to $\ket{0}$.
%
To achieve this, we introduce a fresh classical bit that is not used by the
circuit and assign the measurement result to the bit.
%
If the measurement outcome is $0$, the qubit is already in the state $\ket{0}$.
%
If the outcome is $1$, the qubit is in the state $\ket{1}$, and we flip it by
applying the $\ugate{\pi}{0}{\pi}$ gate.
%
Due to the measurement, the world splits into two unless the qubit is already
in the states $\ket{0}$ or $\ket{1}$ before the measurement.

When executing a statement as a single program, the input is a single world
where every qubit is $\ket{0}$ and every classical bit is $0$. The probability
of the world is $1$.

\begin{figure}[t]
	\[
		\begin{array}{@{}r@{~}c@{~}l@{\quad}r@{~}c@{~}l@{}}
			\multicolumn{6}{c}{
				\generalize{n,q}{M}=\identity{q}\otimes{M}\otimes\identity{n-q-1}
			}                                                                          \\[1em]
			\mrotate{r_1,r_2,r_3}        & = &
			\mtrx{
			e^{-i(r_2+r_3)/2}\cos(r_1/2) &
			-e^{-i(r_2-r_3)/2}\sin(r_1/2)                                              \\
			e^{i(r_2-r_3)/2}\sin(r_1/2)  &
			e^{i(r_2+r_3)/2}\cos(r_1/2)
			}
			                             &
			\mrotate{n,q,r_1,r_2,r_3}    & = & \generalize{n,q}{\mrotate{r_1,r_2,r_3}}
			\\[1em]
			\multicolumn{6}{c}{
			\mproj{0}=\smtrx{1           & 0                                           \\0                  & 0} \qquad
				  \mproj{n,q,0}=\generalize{n,q}{\mproj{0}} \qquad
			\mproj{1}=\smtrx{0           & 0                                           \\0 & 1} \qquad
				  \mproj{n,q,1}=\generalize{n,q}{\mproj{1}}
			}
		\end{array}
	\]
	\caption{Generalization of $2\times2$ matrices}
	\label{fig:gen22}
\end{figure}

\begin{figure}[t]
	\[
		\begin{array}{@{}r@{~}c@{~}l@{\quad}r@{~}c@{~}l@{}}
			\mswap{}          & = & \Smtrx{1                                                                                    & 0 & 0 & 0 \\0&0&1&0\\0&1&0&0\\0                 & 0 & 0                         & 1}
			                  &
			\mswap{n,q_1,q_2} & = & \begin{cases}
				                        I_n\qquad\qquad\qquad\qquad\qquad\text{if}\ q_1=q_2                      \\
				                        \mswap{}\ \ \qquad\qquad\qquad\qquad\text{if}\ n=2\land q_1=0\land q_2=1 \\
				                        (\mswap{}\otimes\identity{n-2})
				                        (\identity{1}\otimes\mswap{n-1,0,n-2})
				                        (\mswap{}\otimes\identity{n-2})                                          \\
				                        \quad\qquad\qquad\qquad\qquad
				                        \text{if}\ n>2\land q_1=0\land q_2=n-1                                   \\
				                        \identity{q_1}\otimes\mswap{q_2-q_1+1,0,q_2-q_1}\otimes\identity{n-q_2-1}
				                        \ \qquad\quad\text{otherwise}
			                        \end{cases}
			\\[2em]
			\mcnot{}          & = & \Smtrx{1                                                                                    & 0 & 0 & 0 \\0&1&0&0\\0&0&0&1\\0                 & 0 & 1                                                                                           & 0}
			                  &
			\mcnot{n,q_1,q_2} & = & \mswap{n,0,q_1}\mswap{n,1,q_2}(\mcnot{}\otimes\identity{n-2})\mswap{n,1,q_2}\mswap{n,0,q_1}
		\end{array}
	\]
	{\footnotesize$(0\le q_1<n,0\le q_2<n)$}
	\caption{Generalization of CNOT}
	\label{fig:cnot}
\end{figure}

\section{Optimizations}
\label{ch:consistency:optimization}

\noindent
We propose optimization techniques for our executable formal semantics.
%
Since we split worlds at each measurement following MWI, the number of worlds
exponentially increases as the number of measurements grows.
%
This prevents the execution of large quantum circuits in a reasonable time.
%
To address this issue, we introduce optimizations that reduce the number of
worlds.

The key idea behind these optimizations is the notion of a \emph{mixed state},
which represents different quantum states and their probabilities using a
single density matrix.
%
For example, consider the density matrix $\rho=\tfrac{1}{2}\smtrx{1&1\\1&1}$
for a superposed state of
$\tfrac{1}{\sqrt{2}}\ket{0}+\tfrac{1}{\sqrt{2}}\ket{1}$.
%
When measuring the qubit, it yields either $0$ or $1$, each with a probability
of $\frac{1}{2}$.
%
The corresponding density matrices are $\rho_0=\smtrx{1&0\\0&0}$ and
$\rho_1=\smtrx{0&0\\0&1}$.
%
Instead of maintaining two matrices with their respective probabilities, a
single matrix can represent both possibilities through a weighted sum:
%
$\rho_{\embox{\tiny mixed}} = \tfrac{1}{2}\rho_0+\tfrac{1}{2}\rho_1 =
	\tfrac{1}{2}\smtrx{1&0\\0&1}$.

Unfortunately, we cannot always use mixed states to avoid splitting worlds
because we need to consider not only qubits and probabilities but also
classical bits.
%
Without using mixed states, we have two post-measurement worlds,
$(\rho_0,[0],\frac{1}{2})$ and $(\rho_1,[1],\frac{1}{2})$.
%
We cannot simply replace them with a single world containing
$\rho_{\embox{\tiny mixed}}$ because we cannot retain the classical bit
information.
%
Neither $(\rho_{\embox{\tiny mixed}},[0],1)$ nor $(\rho_{\embox{\tiny
			mixed}},[1],1)$ is correct.
%
To have only a single world, we would need to expand the density matrix to
incorporate the classical bit as if it were a qubit, necessitating a $4\times
	4$ matrix for a 2-qubit system, with sixteen entries.
%
However, this is less efficient than simply splitting the world into two,
requiring two $2\times 2$ matrices, which have eight entries in total.
%
Therefore, we should utilize mixed states only when we do not need to add
classical bit information to the density matrix.

As a solution, we identify two cases where mixed states can be used without
concerning classical bits and optimize each case by replacing two worlds with a
single world containing a mixed state.
%
This leads to two optimizations: \emph{world merging} and \emph{efficient
	reset}.
%
\cref{fig:optimization} defines the operational semantics for optimized
executions.
%
We now describe each optimization in detail.

\paragraph{World Merging}

Rule [\textsc{Eval-Merge}] defines the world-merging optimization, which can be
applied to the evaluation of any statements.
%
During execution, when two distinct worlds with identical classical states but
different quantum states emerge, we merge their quantum states into a single
density matrix, creating a mixed state while preserving the identical classical
state.
%
This rule guarantees that the number of worlds during the execution of a
circuit with $n$ different classical bits is upper-bounded by $2^n$ because the
system can result in a maximum of $2^n$ different classical states.

\paragraph{Efficient Reset}

Rule [\textsc{Eval-Reset-Opt}] enhances the efficiency of evaluating reset
statements by replacing Rule [\textsc{Eval-Reset}].
%
In the case of Rule [\textsc{Eval-Reset}], a reset statement assigns a
measurement outcome from a qubit to a classical bit.
%
This classical bit is solely utilized for conditionally flipping the qubit and
is never used thereafter.
%
Therefore, we can simply remove the classical bit from the world and employ a
mixed state instead of splitting the world into two.
%
Adopting this idea, Rule [\textsc{Eval-Reset-Opt}] updates the density matrix
but does not split the world.

\begin{figure}
	\[
		\small
		\begin{array}{@{}c@{}}
			\inferrule
			{
			\eval {\rep{w}} {s} {w'_0,\ldots,w'_i,\ldots,w'_j,\ldots,w'_n}      \\\\
			w'_i = (\rho_i, \sigma, P_i)                                        \\
			w'_j = (\rho_j, \sigma, P_j)                                        \\
			\rho_{ij} = \tfrac{P_i}{P_i+P_j}\rho_i + \tfrac{P_j}{P_i+P_j}\rho_j \\
			} {
			\eval {\rep{w}} {s} {w'_0,\ldots,w'_{i-1},w'_{i+1},\ldots,w'_{j-1},w'_{j+1},\ldots,w'_n,(\rho_{ij}, \sigma, P_i + P_j)}			}
			\quad[\textsc{Eval-Merge}]
			\\[3em]
			\inferrule
			{
			\qubits{\rho}=n                                                     \\
			\rho_0 = \smeasure{\rho}{\mproj{n,q,0}}                             \\
			p_0 = \sprob{\rho}{\mproj{n,q,0}}                                   \\\\
			\rho_1 = \smeasure{\rho}{\mproj{n,q,1}}                             \\
			p_1 = \sprob{\rho}{\mproj{n,q,1}}                                   \\\\
				\rho' = p_0\rho_0 + p_1\sunitary{\rho_1}{U_{\pi,0,\pi}}
			} {
				\eval {(\rho,\sigma,P)} {\ireset{q}} {(\rho',\sigma,P)}
			}
			\quad[\textsc{Eval-Reset-Opt}]
		\end{array}
	\]
	\caption{Optimized semantics}
	\label{fig:optimization}
\end{figure}

\section{Physical Consistency}
\label{ch:consistency:consistency}

\noindent
We introduce the postulates of quantum computing and prove that the optimized
version of the proposed formal semantics adheres to the postulates.
%
\citet{scherer2019mathematics} summarizes the laws of quantum physics within
five postulates.
%
The exact quote of the postulates can be found in our companion
report~\cite{supp}.
%
To ensure clarity and consistency with the terminologies defined in this paper,
we rephrase the original postulates by presenting three postulates:
\cref{pos:gate} (corresponding to postulate 4), \cref{pos:measurement}
(corresponding to postulates 2 and 3), and \cref{pos:density} (corresponding to
postulate 5).
%
We ignore postulate 1 due to its irrelevance to the functionality of
OpenQASMCore.

The first postulate describes the operation of gates:
%
\begin{postulate}[Gate]
	%
	\label{pos:gate}
	%
	The change of a density matrix from $\rho$ to $\rho'$ by a gate is described by
	$\rho'=M\rho M^*$ where $M$ is a unitary matrix, i.e., $M^*M=I$.
	%
\end{postulate}
%
\noindent
%
Since OpenQASMCore has two kinds of gate statements, we prove that each
satisfies the postulate.
%
\begin{theorem}[Rotation Gate]
	%
	\label{thm:rotation}
	%
	Suppose that the following holds:
	%
	$\eval{(\rho,\sigma,P)}{\irotate{r_1}{r_2}{r_3}{q}}{(\rho',\sigma',P')}$.
	%
	Then, there exists a unitary matrix $M$ such that $\rho'=M\rho M^*$.
	%
\end{theorem}
%
\begin{proof}
	%
	By rule [\textsc{Eval-Rotate}], $\rho'=M\rho M^*$ where
	$M=\mrotate{\qubits{\rho},q,r_1,r_2,r_3}$. Therefore, it is enough to show that
	$M$ is unitary. We can easily show that a $2\times2$ matrix
	$\mrotate{r_1,r_2,r_3}$ and the identity matrix of any dimension are unitary.
	Since the tensor product of unitary matrices is also unitary, $M$ is unitary.
	%
\end{proof}
%
\begin{theorem}[Cnot Gate]
	%
	\label{thm:cnot}
	%
	Suppose that the following holds:
	%
	$\eval{(\rho,\sigma,P)}{\icnot{q_1}{q_2}}{(\rho',\sigma',P')}$.
	%
	Then, there exists a unitary matrix $M$ such that $\rho'=M\rho M^*$.
	%
\end{theorem}
%
\begin{proof}
	%
	By rule [\textsc{Eval-Cnot}], $\rho'=M\rho M^*$ where
	$M=\mcnot{\qubits{\rho},q_1,q_2}$, it is enough to show that $M$ is unitary. By
	induction on $n$, we can show that $\mswap{n,0,n-1}$ is unitary for any $n$.
	Since the tensor product of unitary matrices is also unitary,
	$\mswap{n,q_1,q_2}$ is unitary. It is easy to show that the $4\times4$ matrix
	$\mcnot{}$ is unitary. Since the product of unitary matrices is also unitary,
	$M$ is unitary.
	%
\end{proof}

The second postulate describes the results of measurements:
%
\begin{postulate}[Measurement]
	%
	\label{pos:measurement}
	%
	For a quantum system described by a density matrix $\rho$, a physical
	observable to be measured is represented by a self-adjoint matrix $A$, whose
	dimension is the same as $\rho$.
	%
	Each measurement outcome corresponds to $\lambda$, an eigenvalue of $A$.
	%
	The probability of the outcome corresponding to $\lambda$ is $\trace{\rho M}$
	where $M$ is the projection onto the eigenspace of $\lambda$.
	%
	The change of a density matrix from $\rho$ to $\rho'$ by the measurement is
	described by $\rho'=\frac{1}{\trace{\rho M}}M\rho M$.
	%
\end{postulate}
%
\noindent
%
OpenQASMCore has only one kind of measurement statement, and we show that the
semantics of the statement follows the postulate:
%
\begin{theorem}[Measurement]
	%
	\label{thm:measure}
	%
	Suppose that the following holds:
	%
	$\eval{(\rho,\sigma,P)}{\imeasure{q}{c}}{\rep{w'}}$.
	%
	Then, $\exists A.\forall w'\in\{\rep{w'}\}.\exists\lambda$ such that the
	followings hold:
	%
	(1) $A$ is self-adjoint;
	%
	(2) $\lambda$ is an eigenvalue of $A$;
	%
	(3) $M$ is the projection operator onto the eigenspace of $\lambda$;
	%
	(4) $w'=(\rho',\sigma',P')$;
	%
	(5) $\rho'=\frac{1}{\trace{\rho M}}M\rho M$;
	%
	(6) $\frac{P'}{P}=\trace{\rho M}$.
	%
\end{theorem}
%
\begin{proof}
	%
	By rules [\textsc{Eval-Measure}], [\textsc{Eval-Measure-0}], and
		[\textsc{Eval-Measure-1}], $w'\in\{\rep{w'}\}$ is either
	$(\smeasure{\rho}{\mproj{n,q,0}},\sigma[c\mapsto0],P\times\sprob{\rho}{\mproj{n,q,0}})$
	or
	$(\smeasure{\rho}{\mproj{n,q,1}},\sigma[c\mapsto1],P\times\sprob{\rho}{\mproj{n,q,1}})$
	where $n=\qubits{\rho}$.
	%
	Let $A=\identity{q}\otimes\smtrx{1&0\\0&-1}\otimes\identity{n-q-1}$.
	%
	We can easily show that $1$ and $-1$ are eigenvalues of $A$, and
	$\mproj{n,q,0}$ and $\mproj{n,q,1}$ are their corresponding projections.
	%
\end{proof}

The last postulate describes the properties of density matrices:
%
\begin{postulate}[Density Matrix]
	%
	\label{pos:density}
	%
	A density matrix $\rho$ satisfies the following:
	%
	(1) $\rho$ is self-adjoint: $\rho^*=\rho$;
	%
	(2) $\rho$ is positive semi-definite: $\rho\ge0$, i.e., $z^*\rho z\ge 0$ for every
	column vector $z$;
	%
	(3) $\trace{\rho}=1$.
	%
\end{postulate}
%
\noindent
%
We show that for every OpenQASMCore program, if its execution starts with a
valid density matrix, then the execution ends with valid density matrices.
%
\begin{theorem}[Density Matrix]
	%
	Suppose that the following holds:
	%
	$\eval{(\rho,\sigma,P)}{s}{\rep{w'}}$.
	%
	If $\rho$ is self-adjoint and positive semi-definite and $\trace{\rho}=1$, then
	for every $(\rho',\sigma',P')\in\{\rep{w'}\}$, $\rho'$ is self-adjoint and
	positive semi-definite and $\trace{\rho'}=1$.
	%
\end{theorem}
%
\begin{proof} We prove it by structural induction on $s$.
	%
	\begin{itemize}
		%
		\item $s=\inop$.
		      %
		      By rule [\textsc{Eval-Nop}], $\rho'=\rho$.
		      %
		\item $s=\iseq{s_1}{s_2}$.
		      %
		      We prove it using rule [\textsc{Eval-Seq}] and the induction hypothesis.
		      %
		\item $s=\iif{c}{b}{s'}$.
		      %
		      If $\sigma(c)\not=b$, $\rho'=\rho$ by rule [\textsc{Eval-If-False}].
		      %
		      If $\sigma(c)=b$, we prove it using rule [\textsc{Eval-If-True}] and the
		      induction hypothesis.
		      %
		\item $s=\irotate{r_1}{r_2}{r_3}{q}$ or $s=\icnot{q_1}{q_2}$.
		      %
		      By \cref{thm:rotation,thm:cnot}, $\rho'=M\rho M^*$ where $M$ is unitary.
		      %
		      Basic linear algebra allows us to prove it using that $M$ is unitary.
		      %
		\item $s=\imeasure{q}{c}$.
		      %
		      By \cref{thm:measure}, $\rho'=\frac{1}{\trace{\rho M}}M\rho M$ where $M$ is a
		      projection.
		      %
		      Basic linear algebra allows us to prove it using that $M$ is a projection.
		      %
		\item $s=\ireset{q}$.
		      %
		      Rule [\textsc{Eval-Reset-Opt}] yields $\rho'=p_0\rho_0 +
			      p_1\sunitary{\rho_1}{X}$ where $p_0$ and $p_1$ correspond to measurement
		      outcomes of $q$ being 0 and 1, respectively, and $X$ representing a unitary
		      operator.
		      %
		      By proving that probabilities are non-negative and their sum is unity, and that
		      positive semi-definiteness and self-adjointness are preserved under convex
		      combinations, $\rho'$ retains the required properties.
	\end{itemize}
	%
	Finally, the application of [\textsc{Eval-Merge}] preserves the three
	properties of density matrices.
	%
	If $\rho' = \frac{P_0}{P_0+P_1}\rho_1 + \frac{P_1}{P_0+P_1}\rho_2$, which is a
	convex combination of $\rho_1$ and $\rho_2$, all properties are preserved,
	analogous to the $s=\ireset{q}$ case.
	%
\end{proof}

\noindent
%
The initial state in OpenQASMCore is a state where every qubit is $\ket{0}$,
which is straightforward to prove that it is a valid density matrix.
%
Therefore, we ensure that the execution of every OpenQASMCore program ends with
valid density matrices.

### 2024-10-07 kw/4-openqasm.tex
Add materials

% !TEX root = main.tex

\chapter{Desugaring from OpenQASM to OpenQASMCore}
\label{ch:openqasm}

\noindent
In this section, we present the syntax of OpenQASM 2.0
(\cref{ch:openqasm:syntax}) and define desugaring from OpenQASM to
OpenQASMCore (\cref{ch:openqasm:desugaring}).
%
This enables us to indirectly define the executable formal semantics of
OpenQASM and naturally achieve the physical consistency of OpenQASM.

\section{Syntax of OpenQASM}
\label{ch:openqasm:syntax}

\begin{figure}[t]
	\[
		\begin{array}{@{}rrrll@{}}
			\text{Identifier}        & x                            & \multicolumn{1}{c}{\in} & \embox{Id}                                                                \\
			\text{Length/Index}      & n                            & \multicolumn{1}{c}{\in} & \mathbb{N}                                                                \\
			\text{Program}           & p                            & ::=                     & \rep{d};\rep{S}                                                           \\
			\text{Declaration}       & d                            & ::=                     & \dcreg{x_c}{n}                             & \text{Classical register}    \\
			                         &                              & |                       & \dqreg{x_q}{n}                             & \text{Quantum register}      \\
			                         &                              & |                       & \dgate{x_g}{\rep{x_r}}{\rep{x_q}}{\rep{g}} & \text{Custom gate}           \\
			\text{Gate operation}    & g                            & ::=                     & \irotate{e}{e}{e}{a_q}                     & \text{Rotation}              \\
			                         &                              & |                       & \icnot{a_q}{a_q}                           & \text{Cnot}                  \\
			                         &                              & |                       & \icustom{x_g}{\rep{e}}{\rep{a_q}}          & \text{Custom gate}           \\
			\text{Expression}        & e                            & ::=                     & r                                          & \text{Real number}           \\
			                         &                              & |                       & x_r                                        & \text{Real number parameter} \\
			                         &                              & |                       & e+e\quad|\quad\cdots                       & \text{Math operations}       \\
			\text{Argument}          & a                            & ::=                     & x[n]
			                         & \text{Quantum/classical bit}                                                                                                       \\
			                         &                              & |                       & x                                          & \text{Register}              \\
			\text{Statement}         & S                            & ::=                     & Q                                          & \text{Quantum operation}     \\
			                         &                              & |                       & \iif{x_c}{n}{Q}                            & \text{Conditional}           \\
			\text{Quantum operation} & Q                            & ::=                     & g                                          & \text{Gate operation}        \\
			                         &                              & |                       & \imeasure{a_q}{a_c}                        & \text{Measurement}           \\
			                         &                              & |                       & \ireset{a_q}                               & \text{Reset}                 \\
		\end{array}
	\]
	\caption{Syntax of OpenQASM}
	\label{fig:openqasm-syntax}
\end{figure}

\noindent
The syntax of OpenQASM is outlined in \cref{fig:openqasm-syntax}.
%
The metavariable $x$ ranges over identifiers, which serve as names for
classical registers, quantum registers, custom gates, and real-number
parameters of custom gates.
%
Although these various kinds of names are not syntactically differentiated, we
denote them with $x_c$, $x_q$, $x_g$, and $x_r$ in the syntax to provide
clarity regarding the role of an identifier in each context.
%
The metavariable $n$ ranges over natural numbers, which are used as the lengths
of registers and indices to access specific bits within registers.
%
The metavariable $p$ ranges over programs, which consist of declarations and
statements.

A declaration $d$ declares either a classical register, a quantum register, or
a custom gate.
%
$\dcreg{x_c}{n}$ declares a classical register named $x_c$ of length $n$.
%
Similarly, $\dqreg{x_q}{n}$ declares a quantum register named $x_q$ of length
$n$.
%
$\dgate{x_g}{\rep{x_r}}{\rep{x_q}}{\rep{g}}$ defines a custom gate named $x_g$
that takes zero or more real numbers $\rep{x_r}$ and one or more qubits
$\rep{x_q}$.
%
The behavior of a custom gate is specified by a sequence of gate operations
$\rep{g}$, which apply gates to qubits.
%
The scope of a custom gate begins after its definition.
%
Therefore, each custom gate can only refer to built-in or already-defined gates
and cannot be recursive.

A gate operation is either a rotation gate $\irotate{e_1}{e_2}{e_3}{a_q}$, a
CNOT gate $\icnot{a_q}{a_q'}$, or a custom gate
$\icustom{x_g}{\rep{e}}{\rep{a_q}}$.
%
A rotation gate requires three expressions, namely $e_1$, $e_2$, and $e_3$,
which evaluate to real numbers.
%
These expressions can either be a real number $r$, a real-number parameter
$x_r$ (only used in the body of a custom gate), or mathematical operations
between expressions (e.g., $e+e$).
%
Additionally, a rotation gate takes a quantum argument $a_q$.
%
Note that an argument $a$ is either a single bit $x[n]$ or a register $x$;
%
we use $a_q$ for a quantum argument $x_q[n]$ or $x_q$, and $a_c$ for a
classical argument $x_c[n]$ or $x_c$.
%
When a rotation gate operates on a register, it independently applies the
rotation of the same degree to every bit in the register.
%
CNOT and custom gates are similar to rotation gates but can have multiple
quantum arguments.
%
If multiple registers are provided to a gate, their lengths must be equal, and
the gate is independently applied to every index of the registers.
%
For example, $\icnot{x}{y}$ means $\icnot{x[i]}{y[i]}$ for all $i$ where $x$
and $y$ have the same length, and $\icnot{x[0]}{y}$ means $\icnot{x[0]}{y[i]}$
for all $i$.

A statement $S$ is either a quantum operation $Q$ or a conditional quantum
operation $\iif{x_c}{n}{Q}$.
%
Unlike the conditional statement in OpenQASMCore, which checks the value of a
single bit, the conditional statement in OpenQASM compares the value of a
classical register, interpreted as a binary number using the bit at index zero
as the low order bit, with a natural number.

A quantum operation $Q$ is either a gate operation $g$, a measurement
$\imeasure{a_q}{a_c}$, or a reset $\ireset{a_q}$.
%
A measurement operation can measure either a single qubit or a quantum
register.
%
In the latter case, the outcome should be assigned to a classical register of
the same length.
%
A reset operation assigns a new state $\ket{0}$ to a single qubit or each qubit
of a quantum register.

In fact, for the sake of presentation, the syntax defined in
\cref{fig:openqasm-syntax} differs from the syntax of real OpenQASM 2.0 in a
few aspects:
%
\begin{itemize}
	%
	\item We omit \emph{include} statements, which are used to include code from another
	      file in the current code.
	      %
	      They can be replaced with proper code through preprocessing.
	      %
	\item We omit \emph{opaque gate} declarations, which declare gates that can be
	      physically implemented but cannot be described in OpenQASM.
	      %
	      It is impossible to determine the behavior of opaque gates, and they do not
	      accord with our goal of defining executable formal semantics.
	      %
	\item We omit \emph{barrier} statements, which disable optimization across
	      themselves.
	      %
	      They do not affect the semantics and can be simply replaced by $\inop$ during
	      desugaring.
	      %
	\item We require all declarations to precede statements, whereas the real OpenQASM
	      syntax allows the interleaving of declarations and statements.
	      %
	      They can be easily reordered to follow our restriction through preprocessing.
	      %
\end{itemize}

\section{Desugaring}
\label{ch:openqasm:desugaring}

\noindent
The desugaring process from OpenQASM to OpenQASMCore involves five steps:
%
unrolling operations to prevent registers from being used as arguments
(\cref{ch:openqasm:desugaring:unrolling});
%
inlining custom gates to remove their definitions and usages
(\cref{ch:openqasm:desugaring:inlining});
%
decomposing each conditional into nested conditionals to handle comparisons
with natural numbers (\cref{ch:openqasm:desugaring:conditional});
%
representing qubits or classical bits with natural numbers instead of
identifiers to remove register declarations
(\cref{ch:openqasm:desugaring:register});
%
evaluating expressions to real numbers in rotation gate operations
(\cref{ch:openqasm:desugaring:expr}).

\subsection{Unrolling operations}
\label{ch:openqasm:desugaring:unrolling}

\noindent
We unroll each quantum operation that takes a quantum register as an argument
into a sequence of operations, each operating on a single qubit.
%
This process resembles loop unrolling, which transforms a loop with a finite
number of iterations into a sequence of individual statements.
%
For instance, $\icnot{x}{y}$ is desugared as
$\icnot{x[0]}{y[0]};\icnot{x[1]}{y[1]}$, assuming both $x$ and $y$ have lengths
of 2.
%
It is a precondition in OpenQASM that the involved registers must be of equal
length; otherwise, the desugaring fails.
%
This step addresses the disparity between OpenQASM and OpenQASMCore regarding
the acceptance of registers as arguments;
%
while every operation in OpenQASM allows registers, a statement in OpenQASMCore
does not.

\subsection{Inlining custom gates}
\label{ch:openqasm:desugaring:inlining}

\noindent
We inline custom gates to eliminate their definitions and usages because
OpenQASMCore lacks the notion of a custom gate.
%
The inlining process is straightforward as none of the custom gates are
recursive.
%
If the inlining process encounters a reference to a custom gate that is not
defined, it will result in failure.

\subsection{Decomposing conditionals}
\label{ch:openqasm:desugaring:conditional}

\noindent
We desugar conditional statements, which allow comparison with a natural
number, whereas OpenQASMCore only allows comparison with a bit.
%
We decompose each conditional statement into nested conditionals that compare
each bit of the register with the corresponding digit in the binary
representation of the natural number.
%
For instance, $\iif{x}{6}{\cdots}$ is desugared as
$\iif{x[2]}{1}{\iif{x[1]}{1}{\iif{x[0]}{0}{\cdots}}}$, where the length of $x$
is $3$.

\subsection{Representing qubits and classical bits with natural numbers}
\label{ch:openqasm:desugaring:register}

\noindent
We represent qubits and classical bits using natural numbers instead of
identifiers because OpenQASMCore prohibits the declaration of registers and
requires each qubit or classical bit to be represented by a natural number.
%
Conceptually, we merge all quantum registers into a single register and
substitute each qubit access with the corresponding index within the merged
register;
%
this process is repeated for classical registers.

\subsection{Evaluating expressions}
\label{ch:openqasm:desugaring:expr}

\noindent
We substitute every expression, which is used as an argument for a rotation
gate, with the actual real number denoted by the expression by evaluating it.
%
Note that rotation gates in OpenQASMCore accept only real-number constants as
arguments, whereas OpenQASM allows arbitrary expressions.
%
This substitution is possible because all real-number parameters have been
replaced by arguments through inlining, and, at this stage, every expression
contains no variables but solely consists of constants and operations involving
them.

### 2024-10-07 kw/5-evaluation.tex
Add materials

% !TEX root = paper.tex
\chapter{Evaluation} \label{ch:evaluation}

\noindent
We evaluate our approach with the following two research questions:
%
\begin{itemize}
	%
	\item RQ1. Effectiveness of optimizations: How significantly do the optimizations
	      reduce execution times? (\cref{ch:evaluation:rq1})
	      %
	\item RQ2. Effectiveness as a testing oracle: How many real-world bugs can be
	      detected using the semantics as a testing oracle? (\cref{ch:evaluation:rq2})
	      %
\end{itemize}
%
To facilitate the execution of arbitrary OpenQASM programs using the semantics,
we extracted OCaml code from the Coq mechanization of OpenQASMCore's semantics.
Additionally, we implemented a desugarer in OCaml to transform OpenQASM into
OpenQASMCore.
%
All experiments were conducted on a MacBook Pro, equipped with an M1 Max CPU
and 32GB of memory.
%
We used Qiskit Terra v0.25.3 throughout the evaluation.

\begin{figure}[t]
	\centering
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[width=\textwidth]{imgs/opt-bench.pdf}
		\caption{Across different optimization levels}
		\label{fig:result-opt-performance}
	\end{subfigure}
	\hspace{1em}
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[width=\textwidth]{imgs/sim-bench.pdf}
		\caption{Semantics vs. Qiskit's simulator}
		\label{fig:result-performance}
		\vspace{0.9em}
	\end{subfigure}
	\caption{Execution times of OpenQASM circuits (log scale)}
\end{figure}

\section{RQ1: Effectiveness of Optimizations}
\label{ch:evaluation:rq1}

\noindent
We evaluate the effectiveness of the optimizations in reducing execution times
by comparing three settings: Level 0 (no optimizations), Level 1 (with world
merging), and Level 2 (both optimizations).
%
These optimizations particularly decrease the execution times of dynamic
circuits, which perform mid-circuit measurements, by reducing the creation of
new worlds during measurements.
%
Due to the scarcity of dynamic circuits in available OpenQASM benchmark suites,
we generated 100 random dynamic circuits and used them for evaluation.
%
Each circuit comprises five qubits and average 62.8 random operations, where
each operation is either a reset, a measurement, or a gate chosen from Qiskit's
built-in gate set, consisting of 52 different gates.
%
Additionally, each operation has a 50\% chance of being conditional.
%
By properly setting the probability of selecting each operation, each circuit
has 5.7 measurements and 4.9 resets on average.

Our experimental results indicate a significant reduction in execution times
due to these optimizations.
%
\cref{fig:result-opt-performance} shows a box plot of the execution times across
different settings, with green dots representing average times, annotated in
black text.
%
Two optimizations collectively achieve an average reduction of 85\%.
%
Specifically, world merging achieves an average reduction of 71\%, while
efficient reset leads to a 49\% reduction.
%
World merging is more effective because it can be applied to any statements,
which may create two worlds with the same classical state.
%
This technique would be especially effective in circuits with measurements
considerably more than classical bits, as it limits the number of worlds
regardless of the number of measurements.
%
In contrast, unlike the typical reset, efficient reset does not exponentially
increase the number of worlds with each reset.
%
This benefit exponentially grows with the number of reset statements, making
efficient reset particularly effective for circuits with frequent resets.

To better understand the performance characteristics of the semantics, we
further compare its execution times with those of a real-world simulator.
%
We utilized \C{aer\_simulator} from Qiskit, with parallel execution disabled.
%
For this experiment, we employed 141 circuits from three benchmark suites, in
addition to the 100 randomly generated dynamic circuits.
%
The first suite is OpenQASM 2.0 specification~\cite{openqasm2}, where we
selected six example circuits by excluding one with ten qubits.
%
The second suite is QASMBench~\cite{li2023qasmbench}, where we selected 35
circuits from 81 by excluding those with nine or more qubits and one with eight
qubits.
%
The third suite, Supermarq~\cite{tomesh2022supermarq}, offers eight benchmark
generators, from which we generated 100 circuits.
%
Details on the simulator names and circuit sizes are available in the companion
report~\cite{supp}.

To compare performance, we executed the semantics only once for each circuit,
while we executed the simulator multiple times for each circuit.
%
This approach stems from the fact that our semantics computes the exact
probability distribution, eliminating the need for repeated runs.
%
In contrast, the simulator yields a nondeterministic outcome in each execution,
requiring multiple runs to approximate the probability distribution.

\cref{fig:result-performance} presents the geometric mean of execution times for
the circuits using each method: our semantics with optimization Level 2 or
\C{aer\_simulator}.
%
We employ the geometric mean due to significant variances in execution times
across circuits, with the slowest being 100,000 times longer than the fastest.
%
The left bar in the figure indicates the execution time for the semantics,
while the right bars represent the simulator's execution times, varying by the
number of runs.
%
The results demonstrate that a single semantics run is roughly equivalent in
time to $2^{19}$ simulator runs.
%
However, this does not imply that the semantics' performance is exceedingly
poor, as its capabilities fundamentally differ from those of the simulators,
which cannot compute the exact probability distribution regardless of the
number of runs.

\section{RQ2: Effectiveness as a Tesing Oracle}
\label{ch:evaluation:rq2}

\noindent
To evaluate the effectiveness of our semantics as a testing oracle, we tested a
simulator, \C{aer\_simulator}, provided by Qiskit using the semantics.
%
Simulating an OpenQASM circuit in Qiskit involves three steps:
%
(1) transpilation, which converts OpenQASM code into a circuit in Qiskit's
internal representation,
%
(2) optional optimization passes that transform the circuit, and
%
(3) simulation, which executes the circuit using a simulator.
%
By saying ``we tested a simulator,'' we mean testing all three steps, with the
simulator employed in the third step.
%
We tested steps 2 and 3 separately for effectiveness, utilizing step 1 in both
cases, as steps 2 and 3 rely on step 1.

To test optimization passes with an OpenQASM circuit, we first transpile (step
1) and optimize (step 2) the circuit, with all optimization passes enabled.
%
We then convert the optimized circuit back to OpenQASM code and execute both
the original and the optimized OpenQASM circuits using the semantics.
%
Finally, we compare the resulting probability distributions from both
executions, and any discrepancy indicates a bug in the optimization passes.

To generate circuits that trigger as many optimizations as possible, we
examined the optimization passes and developed a circuit generator tailored to
this task.
%
Below are all the optimization passes in Qiskit, except for
\C{CrosstalkAdaptiveSchedule}, along with the operations we insert into
circuits to activate them.
%
The \C{CrosstalkAdaptiveSchedule} pass is excluded because it optimizes
circuits based on a user-provided hardware description.
%
\begin{itemize}[leftmargin=*]
	%
	\item \C{collect\_cliffords}/\C{collect\_linear\_functions}:
	      %
	      Each of them replaces a block of Clifford/linear gates with a single gate.
	      %
	      We insert a block of Clifford/linear gates.
	      %
	\item \C{commutative\_cancellation}:
	      %
	      It removes a pair of adjacent gates that are self-inverse.
	      %
	      We insert a sequence of self-inverse gates.
	      %
	\item \C{commutative\_inverse\_cancellation}:
	      %
	      It removes a pair of possibly non-adjacent inverse gates by exploiting
	      commutation relations.
	      %
	      We insert a sequence consisting of an arbitrary gate at the beginning, its
	      inverse at the end, and gates commuting with them in between.
	      %
	\item \C{consolidate\_blocks}:
	      %
	      It replaces a sequence of consecutive gates with a single gate.
	      %
	      We insert a sequence of gates.
	      %
	\item \C{cx\_cancellation}:
	      %
	      It removes a pair of adjacent CNOT gates.
	      %
	      We insert a sequence of CNOT gates.
	      %
	\item \C{echo\_rzx\_weyl\_decomposition}:
	      %
	      It rewrites a two-qubit gate using the Weyl decomposition.
	      %
	      We insert a block of two-qubit gates.
	      %
	\item \C{hoare\_opt}:
	      %
	      It applies Hoare logic to selective gates.
	      %
	      We insert a block of gates handled by Hoare logic.
	      %
	\item \C{inverse\_cancellation}:
	      %
	      It removes a pair of adjacent inverse gates.
	      %
	      We insert an arbitrary gate and its inverse. \item
	      \C{optimize\_1q\_commutation}:
	      %
	      It optimizes the circuit by exploiting a 1-qubit gate commuting with a 2-qubit
	      gate.
	      %
	      We insert a 2-qubit gate and a 1-qubit gate commuting with it.
	      %
	\item \C{optimize\_cliffords}:
	      %
	      It replaces a sequence of consecutive Clifford gates over the same qubit with a
	      single gate.
	      %
	      We insert a sequence of Clifford gates over a single qubit.
	      %
	\item \C{optimize\_swap\_before\_measure}:
	      %
	      It removes a SWAP gate followed by a measurement.
	      %
	      We insert a SWAP gate followed by a measurement.
	      %
	\item \C{remove\_diagonal\_gates\_before\_measure}:
	      %
	      It removes a diagonal gate followed by a measurement.
	      %
	      We insert a diagonal gate followed by a measurement.
	      %
	\item \C{remove\_reset\_in\_zero\_state}:
	      %
	      It removes a reset on a qubit in $\ket{0}$ state.
	      %
	      We insert a reset on a qubit in $\ket{0}$ state.
	      %
	\item \C{reset\_after\_measure\_simplification}:
	      %
	      It removes a reset preceded by a measurement.
	      %
	      We insert a reset preceded by a measurement.
	      %
\end{itemize}

To test simulation with an OpenQASM circuit, we transpile (step 1) and then
simulate (step 3) the circuit, with all optimization passes disabled.
%
As the simulator is sampling-based and produces nondeterministic results, we
must compare the outcomes from multiple simulator runs to the probability
distribution computed by the semantics.
%
For this comparison, we employ the chi-square goodness-of-fit
test~\cite{pearson1900criterion}.
%
This test evaluates whether observed outcomes are likely sampled from a given
probability distribution.
%
The p-value computed by the test indicates the likelihood of the observed
outcomes occurring if the null hypothesis (the observations are sampled from
the distribution) is true.
%
A low p-value suggests that the outcomes are improbable under the null
hypothesis, strongly indicating that the observations do not conform to the
expected probability distribution.
%
Consistent with standard statistical hypothesis testing practices, we interpret
a p-value below 0.01 (=1\%) as evidence of divergence between the semantics and
the simulator.

\begin{table}[t]
	\centering
	\caption{Real-world bugs found by our approach}
	\label{table:bugs}
	{\footnotesize
		\begin{tabular}{@{}lllllll@{}}
			\toprule
			ID    & Location      & Platform & Status    & Novelty   & Description                                   \\ \midrule
			Bug 1 & transpilation & Qiskit   & fixed     & new       & losing classical conditionals                 \\
			Bug 2 & optimization  & Qiskit   & fixed     & duplicate & wrong commutation analysis                    \\
			Bug 3 & optimization  & Qiskit   & confirmed & new       & incorrectly optimizing SWAP and measurement   \\
			Bug 4 & optimization  & Qiskit   & confirmed & new       & Hoare optimizer ignoring classical conditions \\
			Bug 5 & optimization  & Qiskit   & confirmed & new       & Hoare optimizer misusing gate cancellation    \\
			Bug 6 & optimization  & staq     & fixed     & new       & losing conditionals                           \\ \bottomrule
		\end{tabular}
	}
\end{table}

Our results demonstrate the effectiveness of the semantics as a testing oracle.
%
We generated 40,000 static circuits and 10,000 dynamic circuits using the
optimization-triggering circuit generator to test both optimization and
simulation.
%
When testing simulation, we ran the simulator 1,024 times for each circuit.
%
From testing, we discovered five bugs in Qiskit, detailed as Bugs 1 to 5 in
\cref{table:bugs}.
%
To the best of our knowledge, this is the first systematic testing to identify
bugs pertained to physical inconsistencies within real-world quantum
programming platforms.

These bugs were identified through optimization testing, with Bug 1 also
detected during simulation testing as it is due to incorrect transpilation.
%
Initially, out of 50,000 circuits, 10,000 (all the dynamic circuits) failed in
optimization testing, and 9,037 yielded p-values below 0.01 in simulation
testing.
%
A manual examination of a few circuits revealed Bug 1.
%
As Bug 1 led to numerous failures, hindering the discovery of other bugs, we
repeated the optimization and simulation testing after fixing Bug 1.
%
This resulted in 2,716 failures in optimization testing and a few circuits with
p-values below 0.01 in simulation testing.
%
Further manual investigation of the failed circuits in optimization testing
identified Bugs 2 to 5:
%
30 failures caused by Bug 2, 99 by Bug 3, 2,586 probably caused by Bug 4, and 1
by Bug 5.
%
While we cannot manually analyze all 2,586 failures, they have conditionals and
persist even when only the \C{hoare\_opt} pass is enabled, leading us to
suspect that all are attributable to Bug 4.
%
Conversely, no additional bugs were identified from circuits with low p-values,
which is consistent with the expectation that approximately 0.01\% of the
circuits should exhibit p-values below 0.01, by definition of a p-value.

Of the identified bugs, four, excluding Bug 2, were previously unknown.
%
Bug 2, although not fixed in the latest release of Qiskit used for our
experiments, had been identified and fixed in the development version.
%
We reported the four new bugs, which were all confirmed by the Qiskit
developers, and Bug 1 was fixed subsequent to our report.

We also conducted experiments with staq, an open-source library for quantum
circuits~\cite{staq}.
%
Since staq provides optimizers but not simulators, we limited our tests to
optimization using the same 50,000 generated circuits.
%
In staq v3.3, we found one previously unknown bug, referred to as Bug 6 in
\cref{table:bugs}, which was subsequently confirmed and fixed by the staq
developers.
%
The circuits used for the experiment are specifically designed for testing the
optimization passes in Qiskit.
%
Therefore, designing a circuit generator tailored for staq's optimization
passes would reveal additional bugs.

\begin{figure}[t]
	\centering
	\includegraphics[width=0.64\textwidth]{imgs/1-2-sample-test.pdf}
	\caption{Number of detected bugs per number of simulator runs (x-axis is log scale)}
	\label{fig:onesample}
\end{figure}

We now compare testing simulation using the semantics to existing approaches to
test quantum circuit simulators, \textsc{QDiff}~\cite{wang2022qdiff} and
MorphQ~\cite{paltenghi2023morphq}.
%
\textsc{QDiff} employs a differential testing technique, executing the same
quantum circuit on multiple simulators and checking for output discrepancies.
%
MorphQ, in contrast, utilizes a metamorphic testing technique that generates
multiple semantically equivalent quantum circuits, executing them on a
simulator to compare their outputs.

At first glance, our method may appear similar to the existing ones, as both
methods compare independently-obtained results that should be identical if no
bugs exist.
%
However, our approach directly compares simulator outcomes with the probability
distribution computed by the semantics, known as the \emph{one-sample test} in
statistics.
%
In contrast, \textsc{QDiff} and MorphQ compare two sets of simulator outcomes
to determine if they are sampled from the same probability distribution,
referred to as the \emph{two-sample test}.
%
This distinction arises from the absence of a testing oracle computing the
probability distribution in their approaches.

Our case study demonstrates that our method can detect bugs with significantly
fewer simulator runs than existing approaches, due to the difference in
statistical confidence between one-sample and two-sample tests.
%
We utilized the 9,037 circuits affected by Bug 1 for this study.
%
For varying values of $N$, ranging from 1 to 1,024, we executed each circuit on
the simulator $N$ times.
%
The outcomes were compared with the probability distribution computed by the
semantics using the one-sample test.
%
Additionally, we ran the circuit on the corrected version of the simulator $N$
times, comparing the two sets of outcomes using the two-sample test.
%
In both tests, a p-value below 0.01 indicates successful bug detection.
%
\cref{fig:onesample} presents the results, illustrating that the one-sample test
is more effective than the two-sample test.
%
The x-axis represents $N$, while the y-axis shows the number of circuits in
which Bug 1 is detected.
%
As a larger sample size increases statistical confidence and lowers the
p-value, larger $N$ leads to the discovery of bugs across more circuits.
%
However, regardless of $N$, the one-sample test finds bugs in more circuits
than the two-sample test.

When testing optimization, our approach is even more advantageous than existing
methods.
%
Our method can directly compare distributions without any sampling, while
existing approaches must still rely on the less efficient two-sample test.

### 2024-10-07 kw/6-related.tex
Add materials

% !TEX root = main.tex

\chapter{Related Work}
\label{ch:related}

\paragraph{Formal Semantics for Programming Languages}

Researchers have defined formal semantics for various real-world programming
languages.
%
\Cref{ch:intro} already provides the list of studies of formal
semantics for each language.
%
Therefore, this section emphasizes their methodologies and compares them with
the approach proposed in this work.

While numerous studies have directly defined the formal semantics of a complete
real-world language, this approach necessitates significant manual effort.
%
To circumvent this issue, some studies have chosen to define formal semantics
by defining a small core language and then desugaring the surface language to
the core
language~\cite{guha2010essence,politz2012tested,memarian2016depths,politz2013python}.
%
This approach aligns with our work, where we introduce OpenQASMCore and perform
desugaring from OpenQASM to OpenQASMCore.

Another way to reduce the burden of defining formal semantics is to utilize
semantic frameworks, such as K~\cite{rosu2010overview}, PLT
Redex~\cite{felleisen2009semantics}, and Lem~\cite{mulligan2014lem}.
%
These semantic frameworks are also beneficial because they allow the semantics
to be executable without any additional work.
%
K is a rewrite-based semantic framework, in which the semantics of many
languages have been
formalized~\cite{park2015kjs,ellison2012executable,bogdanas2015kjava,filaretti2014executable,jiao2020semantic}.
%
PLT Redex is a domain-specific language for specifying operational semantics
and has been used to describe the semantics of various
languages~\cite{guha2010essence,politz2013python}.

Researchers have defined formal semantics using proof assistants as well.
%
Coq~\cite{barras1997coq} has been widely employed to mechanize formal
semantics~\cite{bodin2014trusted,blazy2009mechanized,jung2017rustbelt,watt2021two}.
%
Isabelle/HOL~\cite{nipkow2002isabelle} is another popular proof assistant used
for defining formal semantics~\cite{klein2006machine,watt2018mechanising}.
%
Semantics defined with proof assistants are often executable, such as shown by
\citet{bodin2014trusted}, who extract OCaml code from Coq.
%
In this work, we also mechanize the semantics in Coq and extract OCaml code.

One notable approach to obtaining formal semantics with low manual effort is to
automatically extract mechanized semantics from a structured natural language
specification, as proposed and adopted in JavaScript by \citet{park2021jiset}.
%
However, this approach is not applicable to OpenQASM due to its lack of
sufficient structure in the specification.

When formal semantics is executable, it is a standard approach to validate its
conformance using existing test suites.
%
Most traditional programming languages provide official conformance test
suites, such as Test262~\cite{test262} for JavaScript and the Java
Compatibility Kit~\cite{jck} for Java.
%
However, quantum computing is still in its early stages, and OpenQASM 2.0 does
not have an official test suite.
%
Consequently, we assess the conformance of the semantics using example circuits
from the OpenQASM 2.0 specification~\cite{openqasm2} and benchmark circuits
from QASMBench~\cite{li2023qasmbench}.
%
Several studies have demonstrated the practical utility of executable semantics
in identifying bugs in language
implementations~\cite{watt2023wasmref,schumi2021spectest,park2021jest} and
automatically generating test suites of high coverage~\cite{park2023feature}.

\paragraph{Quantum Programming Languages}

Researchers have proposed high-level quantum programming languages, which are
complementary to quantum circuit description languages, such as OpenQASM, the
main focus of this work.
%
While high-level languages allow programmers to express complex quantum
algorithms using abstractions, circuit description languages are tightly
coupled with real quantum hardware and often serve as the compilation targets
for high-level languages.

In the early days of quantum computing, quantum programming languages were
defined by incorporating quantum computations into classical languages.
%
For instance, QCL~\cite{omer2003structured} and qGCL~\cite{sanders2000quantum}
introduce imperative features alongside quantum computations, while quantum
lambda calculus~\cite{maymin1997extending},
QML~\cite{altenkirch2005functional}, and Quipper~\cite{green2013quipper} enable
quantum computations within functional languages.

Recently, researchers have proposed quantum programming languages to address
specific difficulties in quantum programming.
%
Here, we discuss a few notable examples.
%
Silq~\cite{bichsel2020silq} enables programmers to safely and concisely perform
\emph{uncomputation}, which means ignoring temporary qubits during computation
without any side effects.
%
Twist~\cite{yuan2022twist} aids in reasoning about \emph{purity}, i.e., the
absence of entanglement, through its novel type system.
%
Tower~\cite{yuan2022tower} allows the use of random-access memory for
implementing pointer-based data structures.
%
Qunity~\cite{voichick2023qunity} generalizes language constructs in classical
programming, thereby establishing a unified approach to quantum and classical
computing.
%
\citet{Feng2021Hoare} propose a Hoare logic for quantum programs by defining
classical-quantum states, which are capable of describing states of dynamic
circuits.
%
\textsc{voqc}~\cite{Hietala2021voqc} is the first verified optimizer for quantum
circuits, utilizing a language called \textsc{sqir}, compatible with OpenQASM.
%
Embedded in Coq, \textsc{sqir} facilitates the verification of \textsc{voqc}'s
correctness.

A notable exception is the work of \citet{paykin2017qwire}, which introduces
\textsc{Qwire}, a quantum circuit description language that can be manipulated
using a classical language.
%
Like our semantics, \textsc{Qwire}'s semantics is defined in terms of density
matrices.
%
Its type system guarantees the preservation of the trace of the density matrix,
ensuring that any well-typed circuit maintains a trace of 1 during execution.
%
This aligns with the last condition specified by \cref{pos:density}.
%
However, \textsc{Qwire} does not enforce the other properties required by the
density matrix postulate, whereas in this work, we prove the physical
consistency of the proposed semantics.

\paragraph{Testing Quantum Programming Platforms}

As mentioned in \Cref{ch:evaluation:rq2}, \textsc{QDiff}~\cite{wang2022qdiff}
and MorphQ~\cite{paltenghi2023morphq} are differential testing and metamorphic
testing techniques for quantum programming platforms, respectively.
%
Although these techniques are effective in identifying bugs in quantum
programming platforms, all the identified bugs are due to software crashes, but
not due to outcomes not following the correct probability distributions.
%
These approaches do not utilize a testing oracle to compute the correct
probability distributions, thereby limiting their bug detection capabilities.

### 2024-10-07 kw/7-conclusion.tex
Add materials

% !TEX root = main.tex
\chapter{Conclusion}
\label{ch:conclusion}

\noindent
In this work, we propose the first executable formal semantics for OpenQASM
2.0, a representative quantum circuit description language.
%
To address the nondeterministic nature of quantum circuits, our semantics
computes the probability distribution of possible outcomes.
%
We adopt MWI from the physics literature to compute the probability
distribution even for dynamic quantum circuits and optimize the semantics using
mixed states.
%
Additionally, we prove the physical consistency of the semantics, which implies
adherence to the postulates of quantum mechanics.
%
To simplify the proof, we define OpenQASMCore as a core language of OpenQASM
and desugar OpenQASM to OpenQASMCore.
%
Our evaluation demonstrates the effectiveness of the optimizations in improving
performance and the utility of the semantics as a testing oracle, finding five
previously unknown bugs in two real-world quantum computing platforms.

### 2024-10-07 kw/main.tex
Add materials

% -*- TeX:UTF-8 -*-
%%
%% KAIST 학위논문양식 LaTeX용 (ver 0.5) 예시
%%
%% @version 0.4
%% @author  채승병 Chae,Seungbyung (mailto:chess@kaist.ac.kr)
%% @date    2004. 11. 12.
%%
%% @requirement
%% teTeX, fpTeX, teTeX 등의 LaTeX2e 배포판
%% + 은광희 님의 HLaTeX 0.991 이상 버젼 또는 홍석호 님의 HPACK 1.0
%% : 설치에 대한 자세한 정보는 http://www.ktug.or.kr을 참조바랍니다.
%%
%% @note
%% 기존에 널리 쓰여오던 차재춘 님의 학위논문양식 클래스 파일의 형식을
%% 따르지 않고 전면적으로 다시 작성하였습니다. 논문 정보 입력부분에서
%% 과거 양식과 다른 부분이 많으니 아래 예시에 맞춰 바꿔주십시오.
%%
%%
%% @acknowledgement
%% 본 예시 논문은 물리학과 박사과정 김용현 님의 호의로 제공되었습니다.
%%
%% -------------------------------------------------------------------
%% @information
%% 이 예제 파일은 hangul-ucs를 사용합니다. UTF-8 입력 인코딩으로
%% 작성되었습니다. hlatex의 hfont는 이용하지 않습니다. --2006/02/11
%% 본 템플릿은 전산학부 김민혁 교수에의해서 버그 수정되었습니다. -- 2016/11/25
%% 본 템플릿은 전산학부 김민혁 교수에의해서 추가 버그 수정되었습니다. -- 2023/06/15

% @class kaist.cls
% @options [default: doctor, korean, final]
% - doctor: 박사과정 | master : 석사과정
% - korean: 한글논문 | english: 영문논문
% - final : 최종판   | draft  : 시험판
% - pdfdoc : 선택하지 않으면 북마크와 colorlink를 만들지 않습니다.

\documentclass[master,english,final]{kaist-ucs} % 석사과정
% \documentclass[doctor,english,final]{kaist-ucs} % 박사과정

% If you want make pdf document (include bookmark, colorlink)
%\documentclass[doctor,english,final,pdfdoc]{kaist-ucs}

% kaist.cls 에서는 기본으로 dhucs, ifpdf, graphicx 패키지가 로드됩니다.
% 추가로 필요한 패키지가 있다면 주석을 풀고 적어넣으십시오,
%\usepackage{...}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage[english]{babel}
\usepackage[numbers]{natbib}
% \usepackage{indentfirst}
\usepackage{booktabs}

\usepackage{mathpartir}
\usepackage[capitalise,nameinlink]{cleveref}
\usepackage{tikz}
\usepackage{subcaption}
\usepackage{mdframed}
\usepackage{pifont}
\usepackage{xfrac}
\usepackage{enumitem}

\usetikzlibrary{quantikz2}

\crefformat{chapter}{#2\S{}#1#3}
\Crefformat{chapter}{#2Chapter~#1#3}
\crefname{postulate}{Postulate}{Postulates}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{corollary}{Corollary}[theorem]
\newtheorem{lemma}[theorem]{Lemma}

\newcommand{\SUCC}{\ding{51}}
\newcommand{\FAIL}{\ding{55}}
\newcommand{\C}[1]{{\tt\small#1}}
\newcommand{\inred}{\color{red}}
\newcommand{\todo}{{\inred TODO}}
\newcommand{\embox}[1]{\mbox{\emph{#1}}}
\newcommand{\rep}[1]{\overline{#1}}
\newcommand{\ugate}[3]{\mathsf{U}(#1,#2,#3)}
\newcommand{\cxgate}{\mathsf{CX}}
\newcommand{\inop}{\mathsf{nop}}
\newcommand{\irotate}[4]{\ugate{#1}{#2}{#3}\ #4}
\newcommand{\icnot}[2]{\cxgate\ #1\ #2}
\newcommand{\imeasure}[2]{#2:=\mathsf{Measure}\ #1}
\newcommand{\iseq}[2]{#1;#2}
\newcommand{\iif}[3]{\mathsf{if}\ #1\rightsquigarrow#2\ \mathsf{then}\ #3}
\newcommand{\eval}[3]{#1\vdash#2\Rightarrow#3}
\newcommand{\sunitary}[2]{\embox{gate}(#1,#2)}
\newcommand{\smeasure}[2]{\embox{msr}(#1,#2)}
\newcommand{\sprob}[2]{\embox{prob}(#1,#2)}
\newcommand{\trace}[1]{\textrm{tr}(#1)}
\newcommand{\btrace}[1]{\textrm{tr}\big(#1\big)}
\newcommand{\qubits}[1]{\embox{qbts}(#1)}
\newcommand{\generalize}[2]{\embox{gen}_{#1}(#2)}
\newcommand{\mtrx}[1]{\begin{pmatrix}#1\end{pmatrix}}
\newcommand{\smtrx}[1]{\big(\begin{smallmatrix}#1\end{smallmatrix}\big)}
\newcommand{\Smtrx}[1]{\Big(\begin{smallmatrix}#1\end{smallmatrix}\Big)}
\newcommand{\identity}[1]{\embox{I}_{#1}}
\newcommand{\mrotate}[1]{\embox{U}_{#1}}
\newcommand{\mcnot}[1]{\embox{CX}_{#1}}
\newcommand{\mswap}[1]{\embox{Swap}_{#1}}
\newcommand{\mproj}[1]{\embox{Proj}_{#1}}
\newcommand{\dcreg}[2]{\mathsf{Creg}\ #1[#2]}
\newcommand{\dqreg}[2]{\mathsf{Qreg}\ #1[#2]}
\newcommand{\dgate}[4]{\mathsf{Gate}\ #1(#2)\ #3=#4}
\newcommand{\icustom}[3]{#1(#2)\ #3}
\newcommand{\ireset}[1]{\mathsf{Reset}\ #1}
\newcommand{\dom}[1]{\textrm{dom}(#1)}

% @command title 논문 제목(title of thesis)
% @options [default: (none)]
% - korean: 한글제목(korean title) | english: 영문제목(english title)
\title[korean] {양자 회로 시뮬레이터의 테스트 오라클로서의 물리적으로 일관된 실행 가능한 형식 의미론}
\title[english]{Physically-Consistent Executable Formal Semantics as a Testing Oracle for Quantum Circuit Simulators}

% @note 표지에 출력되는 제목을 강제로 줄바꿈하려면 \linebreak 을 삽입.
%       \\ 나 \newline 등을 사용하면 안됩니다. (아래는 예시)
%
%\title[korean]{탄소 나노튜브의 물리적 특성에 대한\linebreak 이론 연구}
%\title[english]{Theoretical study on physical properties of\linebreak
%                carbon nanotubes}
%
% If you want to begin a new line in cover, use \linebreak .
% See examples above.
%

% @command author 저자 이름
% @param   family_name, given_name 성, 이름을 구분해서 입력
% @options [default: (none)]
% - korean: 한글이름 | chinese: 한문이름 | english: 영문이름
% 한문 이름이 없다면 빈 칸으로 두셔도 됩니다.
%
%
% If you are a foreigner , write your name in korean or your korean name.
% If you can't write native character, you can make the chinese blank empty
% Write as follow
% \author[korean]{family name in korean}{given name in korean}
% \author[chinese]{family name in your native language}{given name in your native language}
% \author[english]{family name in english}{given name in english}
%
\author[korean] {이}{강 욱}
\author[korean2] {이}{강욱}    %이름을 붙여 써 주시기 바랍니다.
\author[chinese]{李}{康 旭}
\author[english]{Lee}{Kanguk}

% @command advisor 지도교수 이름 (복수가능)
% @usage   \advisor[options]{...한글이름...}{...영문이름...}{signed|nosign}
% @options [default: major]
% - major: 주 지도교수  | coopr: 공동 지도교수
\advisor[major]{류 석 영}{Sukyoung Ryu}{signed}
\advisor[major2]{류석영}{Sukyoung Ryu}{signed}    %한글 성과 한글 이름을 모두 붙여 써 주시기 바랍니다.

% [주의] 전산학부의 경우, 전공이름(Computer Science)을 적어주시기 바랍니다. 조직명(School of Computing) 적지 말아주세요!
\advisorinfo{Professor of Computer Science} %제출승인서에 들어가는 교수님 정보, advisor's information

%\advisor[coopr]{홍 길 동}{Gil-Dong Hong}{nosign}
%\advisor[coopr2]{홍길동}{Gil-Dong Hong}{nosign}    %한글 성과 한글 이름을 모두 붙여 써 주시기 바랍니다.
%
% 지도교수 한글이름은 입력하지 않아도 됩니다.
% You may not input advisor's korean name
% like this \advisor[major]{}{Chang, Kee Joo}{signed}
%

% @command department {학과이름}{학위종류} - 아래 규칙에 따라 코드를 입력
% @command department {department code}{degree field}
%
% department code
% 2. 석박사학위논문 작성 및 제출요령 4쪽 ~ 5쪽 참고
% 또는 kaist-ucs.cls 의 % @command department 참고

% science: 이학 | engineering: 공학 | business : 경영학
% 박사논문의 경우는 학위종류를 입력하지 않아도 됩니다.
% If you write Ph.D. dissertation, you cannot input degree field.
% The third parameter : a | b | c
% a: 소속된 학과만 쓰는 옵션 (학과에만 소속되어 있는 경우에는 무조건 a를 선택해야 함)
% b: 학과 아래의, 프로그램이나 학제전공에 소속되어 있을 경우에 학과와 프로그램을 함께 쓰는 옵션
% c: 학과 아래의, 프로그램이나 학제전공에 소속되어 있을 경우에 학과를 쓰지 않고 프로그램이나 학제전공의 이름만 쓰는 옵션
%
% a: it represents only the name of department. (if you aren't in the program under the department, must choose a)
% b: it represents the names of department and the program that is under the department (consider this when you are in the program not only department)
% c: it represents only the name of program that is under the department (consider this when you are in the program not only department)
\department{CS}{engineering}{a}

% @command referee 심사위원 (석사과정 3인, 박사과정 5인)
\referee[1]{류 석 영}
\referee[2]{유 신}
\referee[3]{강 지 훈}
% \referee[4]{가 동 호}
% \referee[5]{박 태 현}
% \referee[5] {Barack Obama}
% Of course english name is available

% @command approvaldate 지도교수논문승인일
% @param   year,month,day 연,월,일 순으로 입력
\approvaldate{2023}{12}{12}

% @command refereedate 심사위원논문심사일
% @param   year,month,day 연,월,일 순으로 입력
\refereedate{2023}{12}{12}

% @command gradyear 졸업년도
\gradyear{2024}

% 본문 시작
\begin{document}

% \theoremstyle{acmdefinition}
\newtheorem{postulate}[theorem]{Postulate}

% 앞표지, 속표지, 학위논문 제출승인서, 학위논문 심사완료 검인서는
% 클래스 옵션을 final로 지정해주면 자동으로 생성되며,
% 반대로 옵션을 draft로 지정해주면 생성되지 않습니다.

% 논문 서지, 초록, 핵심 낱말, 영문 초록, 영어 핵심 낱말 (Information of thesis, abstract in korean, keywords in korean, abstract in english, keywords in english)
\thesisinfo
%% Letters of abstract in korean must be less than 500 and words of abstract in english must be less than 300.
%% Number of keywords must be less than 6.
%% Don't write english letters in the abstract in korean.
\begin{summary}
	%
	현재 양자 프로그램 개발에서는 개발자들이 양자 회로 시뮬레이터에 많이 의존하고 있다.
	%
	따라서 시뮬레이터의 정확성을 보장하는 것은 매우 중요하다.
	%
	그러나 양자 회로를 기술하는 언어에 대한 형식적 의미론이 없기 때문에 시뮬레이터의 정확성에 대한 추론이 어렵다.
	%
	본 논문에서는 대표적인 양자 회로 기술 언어인 OpenQASM 2.0에 대한 실행 가능한 형식 의미론을 제시한다.
	%
	양자 회로의 독특한 특성으로 인해, 형식 의미론을 정의할 때 크게 두 가지 어려움이 있다.
	%
	첫째, 시뮬레이터의 비결정성에도 불구하고 실행 가능한 형식 의미론은 결정론적인 결과를 내야 한다.
	%
	이를 위해 물리학계에서 제시된 다세계 해석을 이용하여 결과의 확률 분포를 계산하는 형식 의미론을 제시한다.
	%
	둘째, 실행 가능한 형식 의미론은 물리적 일관성을 보장해야 한다.
	%
	이를 위해 OpenQASM의 핵심 언어인 OpenQASMCore를 제시하고, OpenQASM에서 OpenQASMCore로의 변환(디슈가링)을 정의한다.
	%
	형식 의미론과 물리적 일관성에 대한 증명은 Coq에서 정형 검증되었다.
	%
	본 논문에서는 실행 가능한 형식 의미론의 평가로서 두 가지 최적화 기법의 효과적인 성능 향상과, 실제 양자 프로그래밍 플랫폼에서 5개의
	새로운 버그를 찾아내는 등의 실행 가능한 형식 의미론의 테스트 오라클로서의 유용성을 보여준다.
	%
\end{summary}

\begin{Korkeyword}
	양자 회로, 양자 프로그래밍 언어, 형식 의미론, 테스트 오라클, OpenQASM, OpenQASMCore, 다세계 해석, 양자 시뮬레이터,
	물리적 일관성
\end{Korkeyword}

\begin{abstract}
	In the current practice of quantum program development, programmers heavily
	rely on quantum circuit simulators.
	%
	Ensuring the correctness of these simulators is therefore crucial.
	%
	However, the lack of formal semantics for quantum circuit description
	languages, which are used to describe quantum circuits, hinders reasoning about
	simulator correctness.
	%
	In this work, we present the first executable formal semantics, which can
	function as a testing oracle, for OpenQASM 2.0, a representative quantum
	circuit description language.
	%
	The unique characteristics of quantum circuits present challenges in defining
	formal semantics.
	%
	First, executable semantics must yield deterministic results despite the
	nondeterminism of simulators.
	%
	To achieve this, our semantics computes the probability distribution of
	outcomes using the \emph{many-worlds interpretation} (MWI) from the physics
	literature.
	%
	Additionally, we propose two optimizations for the semantics.
	%
	Second, ensuring \emph{physical consistency} is essential, meaning the
	semantics must adhere to physical laws.
	%
	To facilitate this, we introduce OpenQASMCore, a core language for OpenQASM,
	simplifying the proof process, and define desugaring from OpenQASM to
	OpenQASMCore.
	%
	The semantics and proofs are mechanized in Coq.
	%
	Our evaluation shows the optimizations' effectiveness in improving performance
	and the utility of our semantics as a testing oracle by identifying five
	previously unknown bugs in two real-world quantum programming platforms.
	%
\end{abstract}

\begin{Engkeyword}
	Quantum circuit, quantum programming language, formal semantics, testing
	oracle, OpenQASM, OpenQASMCore, many-worlds interpretation, Quantum simulator, and
	physical consistency.
\end{Engkeyword}

\addtocounter{pagemarker}{1}                 % 백색별지분을 고려
\newpage

% 목차 (Table of Contents) 생성
\tableofcontents

% 표목차 (List of Tables) 생성
\listoftables

% 그림목차 (List of Figures) 생성
\listoffigures

% 위의 세 종류의 목차는 한꺼번에 다음 명령으로 생성할 수도 있습니다.
%\makecontents

%% 이하의 본문은 LaTeX 표준 클래스 report 양식에 준하여 작성하시면 됩니다.
%% 하지만 part는 사용하지 못하도록 제거하였으므로, chapter가 문서 내의
%% 최상위 분류 단위가 됩니다.
%% You cannot use 'part'

\include{1-intro}

\include{2-openqasmcore}

\include{3-consistency}

\include{4-openqasm}

\include{5-evaluation}

\include{6-related}

\include{7-conclusion}

\chapter[Method]{Method}
\label{ch:ourmethod}
\section{Quantum Communication}
To enhance the performance gain of original cognitive radio networks (CRNs),
leveraging cooperative diversity has attracted a lot of attention
\cite{SOCA1}...

\section{Examples}

\section{Applications of Quantum Communication}
\label{ch:application}
\noindent
Despite its potential to improve the throughput, spatial domain diversity was not fully considered in the studies of original CCRNs. Utilizing the spatial domain for the communications, the concept of MIMO has been adopted in many cases to increase the wireless capacity \cite{EF1,FD2}...

Let us consider an MIMO-CCRN with one PL and one SL. Each PU has one legacy
antenna and each SU has two STAR antennas. The duration of one communication
time frame is divided into two phases \cite{RVP2,ML2} ...
%%
%% 표 삽입 예시
%% Example. how to insert table
%%
\begin{table}[t]
	\caption[Enter the caption title here]{Energy stability $E$ (eV) per molecule of all meta-stable
		isomer states of C$_{60}$ opening process for forming the (5,5) cap.
		In the SW-I and SW-II, both ferromagnetic (Ferro) and paramagnetic (Para)
		spin configurations are obtained, whereas only non-magnetic configuration
		is obtained in the BF, SW-III, and CAP(5,5).
		$M$ is total magnetization $n_{\rm up}$-$n_{\rm down}$ in unit of $\mu_B$, where
		$n_{\rm up(down)}$ is the number of up (down) spins.
	}
	\label{mag-tab1}
	\begin{center}
		\begin{tabular} {ccccccccccc}
			\hline\hline
			 &               & BF & \multicolumn{2}{c}{SW-I} &       & \multicolumn{2}{c}{SW-II} & SW-III & CAP    &               \\
			\cline{4-5} \cline{7-8}
			 &               &    & Para                     & Ferro &                           & Para   & Ferro  &      &      & \\
			\hline
			 & $E$ (eV)      & 0  & 7.796                    & 7.832 &                           & 10.418 & 10.408 & 11.5 & 13.2 & \\
			 & $M$ ($\mu_B$) & 0  & 0                        & 1.94  &                           & 0      & 2.06   & 0    & 0    & \\
			\hline\hline
		\end{tabular}
	\end{center}
\end{table}

%%
%% 그림 삽입 예시
%% Example. how to insert graph
%%
%% Note. 가급적 \includegraphics 명령을 사용하십시오.
%% Recommen : Use \includegraphics to insert graph.
%%
\begin{figure}[t]
	\centerline{\includegraphics[width=12.5cm]{sample-fig1}}
	\caption[Enter the caption title here]{ Ball-and-stick models of meta-stable isomers in
		cage opening process from a C$_{60}$ buckminsterfullerene
		to a (5,5) capsule. We name them BF and CAP(5,5).
		Depending on the number of the Stone-Wales (SW) transformation,
		we call the intermediate isomers with SW-I, SW-II, and SW-III.
		Highlighted atoms are undercoordinated except BF.
	} \label{mag-fig1}
\end{figure}

\section{Discussion and Future Work}
\noindent
The achievable rates can be calculated by finding the statistics of the five signals transmitted that maximize the mutual information between $s_{t,XY}$ and $y_{t,XY}$ for $(X,Y)=(T,C), (C,N)$, and $(N,C)$ when $t=1$, and $(X,Y)=(C,R), (C,N)$, and $(N,C)$ when $t=2$ \cite{SOCA2,EF2}...

\chapter{Conclusion}
\noindent
We have proposed a novel FD MIMO-CCRN framework providing a reasonable performance improvement compared with the conventional MIMO-CCRN framework...
%%
%% 참고문헌 시작
%% bibliography
%% It can be changed but should include sufficient information.
% \bibliographystyle{abbrvnat}
\bibliographystyle{plainnat}
\bibliography{references}

%%
%% 감사의 글 시작
%% Acknowledgement
%%
% @command acknowledgement 감사의글
% @options [1 | 2 | 3 |4 ]
% - 1 : 본문과 감사의 글이 둘 다 한글일 때  | 2 : 본문은 한글인데 감사의 글이 영어일 때 | 3 :  본문과 감사의 글이 둘 다 영어일 때  | 4 : 본문은 영어인데 감사의 글이 % 한글일 때
%% It is optional.

\acknowledgment[4]
언제나 저를 바른 길로 이끌어 주시는 송익호 교수님께 큰 고마움을 느낍니다.
끝으로 오늘의 제가 있을 수 있도록 사랑으로 키워 주신 가족들에게 감사드립니다.
저의 이 작은 결실이 그분들께 조금이나마 보답이 되기를 바랍니다.

%%
%% 약력 시작
%% Curriculum Vitae
%%
% @command curriculumvitae 이력서
% @options [1 | 2 | 3 |4 ]
% - 1 : 본문과 약력이 둘 다 한글일 때  | 2 : 본문은 한글인데 약력이 영어일 때 | 3 :  본문과 약력이 둘 다 영어일 때  | 4 : 본문은 영어인데 약력이 한글일 때
%% It is optional and you can change form of this in the class file if you want.
\curriculumvitae[4]

% @environment personaldata 개인정보
% @command     name         이름
%              dateofbirth  생년월일
%              birthplace   출생지
%              domicile     본적지
%              address      주소지
%              email        E-mail 주소
% - 위 6개의 기본 필드 중에 이력서에 적고 싶은 정보를 입력
% input data only you want
\begin{personaldata}
	\name       {이 강 욱}
	\dateofbirth{1999}{08}{01}
	\birthplace {...}
	\address    {...}
\end{personaldata}

% @environment education 학력
% @options [default: (none)] - 수학기간을 입력
\begin{education}
	\item[2015. 3.\ --\ 2017. 2.] 고등학교 (2년 수료)
	\item[2017. 3.\ --\ 2022. 8.] 한국과학기술원 물리학과 (학사)
	\item[2022. 9.\ --\ 2024. 2.] 한국과학기술원 전산학부 (석사)
\end{education}

% @environment career 경력
% @options [default: (none)] - 해당기간을 입력
\begin{career}
	\item[2021. 3.\ --\ 2021. 8.] 쓰레기통 청소
\end{career}

% @environment activity 학회활동
% @options [default: (none)] - 활동내용을 입력
%%    \begin{activity}
%%        \item J. Choi, \textbf{Yong-Hyun Kim}, K.J. Chang, and D. Tomanek,
%%             \textit{Occurrence of itinerant ferromagnetism in C/BN superlattice
%%             nanotubes}, 5th Asian Workshop on First-Principles Electronic
%%             Structure Calculations, Seoul (Korea), October., 2002.
%%    \end{activity}

% @environment publication 연구업적
% @options [default: (none)] - 출판내용을 입력
\begin{publication}
	%\item \textbf{Yong-Hyun Kim}, J. Choi, K.J. Chang, and D. Tomanek,
	%     \textit{Magnetic instability in partly opened C$_{60}$ isomers},
	%     in preparation.
	\item H.-K. Min, Y. Hou, {\bf S. Park}, and I. Song, ``A computationally efficient
	scheme for feature extraction with kernel discriminant analysis," \textit{Patt.
		Recogn.}, vol.~50, no.~2, pp.~45-55, Feb. 2016 (to be published).
\end{publication}

\label{paperlastpagelabel}     % <-- 추가 부분: 마지막 페이지 위치 지정
%% 본문 끝
\end{document}

### 2024-10-07 template/sample_in_Korean.tex
Add materials

--- 2024-10-07 kw/main.tex
@@
 %%
-%% KAIST 학위논문양식 LaTeX용 (ver 0.5) 예시
+%% KAIST 학위논문양식 LaTeX용 (ver 0.4) 예시
 %%
@@
 
-\documentclass[master,english,final]{kaist-ucs} % 석사과정
-% \documentclass[doctor,english,final]{kaist-ucs} % 박사과정
+%\documentclass[master,korean,final]{kaist-ucs} % 석사과정
+\documentclass[doctor,korean,final]{kaist-ucs} % 박사과정
 
@@
 %\usepackage{...}
-\usepackage{amsmath}
-\usepackage{amssymb}
-\usepackage{amsthm}
-\usepackage[english]{babel}
-\usepackage[numbers]{natbib}
-% \usepackage{indentfirst}
-\usepackage{booktabs}
-
-\usepackage{mathpartir}
-\usepackage[capitalise,nameinlink]{cleveref}
-\usepackage{tikz}
-\usepackage{subcaption}
-\usepackage{mdframed}
-\usepackage{pifont}
-\usepackage{xfrac}
-\usepackage{enumitem}
-
-\usetikzlibrary{quantikz2}
-
-\crefformat{chapter}{#2\S{}#1#3}
-\Crefformat{chapter}{#2Chapter~#1#3}
-\crefname{postulate}{Postulate}{Postulates}
-
-\newtheorem{theorem}{Theorem}[section]
-\newtheorem{corollary}{Corollary}[theorem]
-\newtheorem{lemma}[theorem]{Lemma}
-
-\newcommand{\SUCC}{\ding{51}}
-\newcommand{\FAIL}{\ding{55}}
-\newcommand{\C}[1]{{\tt\small#1}}
-\newcommand{\inred}{\color{red}}
-\newcommand{\todo}{{\inred TODO}}
-\newcommand{\embox}[1]{\mbox{\emph{#1}}}
-\newcommand{\rep}[1]{\overline{#1}}
-\newcommand{\ugate}[3]{\mathsf{U}(#1,#2,#3)}
-\newcommand{\cxgate}{\mathsf{CX}}
-\newcommand{\inop}{\mathsf{nop}}
-\newcommand{\irotate}[4]{\ugate{#1}{#2}{#3}\ #4}
-\newcommand{\icnot}[2]{\cxgate\ #1\ #2}
-\newcommand{\imeasure}[2]{#2:=\mathsf{Measure}\ #1}
-\newcommand{\iseq}[2]{#1;#2}
-\newcommand{\iif}[3]{\mathsf{if}\ #1\rightsquigarrow#2\ \mathsf{then}\ #3}
-\newcommand{\eval}[3]{#1\vdash#2\Rightarrow#3}
-\newcommand{\sunitary}[2]{\embox{gate}(#1,#2)}
-\newcommand{\smeasure}[2]{\embox{msr}(#1,#2)}
-\newcommand{\sprob}[2]{\embox{prob}(#1,#2)}
-\newcommand{\trace}[1]{\textrm{tr}(#1)}
-\newcommand{\btrace}[1]{\textrm{tr}\big(#1\big)}
-\newcommand{\qubits}[1]{\embox{qbts}(#1)}
-\newcommand{\generalize}[2]{\embox{gen}_{#1}(#2)}
-\newcommand{\mtrx}[1]{\begin{pmatrix}#1\end{pmatrix}}
-\newcommand{\smtrx}[1]{\big(\begin{smallmatrix}#1\end{smallmatrix}\big)}
-\newcommand{\Smtrx}[1]{\Big(\begin{smallmatrix}#1\end{smallmatrix}\Big)}
-\newcommand{\identity}[1]{\embox{I}_{#1}}
-\newcommand{\mrotate}[1]{\embox{U}_{#1}}
-\newcommand{\mcnot}[1]{\embox{CX}_{#1}}
-\newcommand{\mswap}[1]{\embox{Swap}_{#1}}
-\newcommand{\mproj}[1]{\embox{Proj}_{#1}}
-\newcommand{\dcreg}[2]{\mathsf{Creg}\ #1[#2]}
-\newcommand{\dqreg}[2]{\mathsf{Qreg}\ #1[#2]}
-\newcommand{\dgate}[4]{\mathsf{Gate}\ #1(#2)\ #3=#4}
-\newcommand{\icustom}[3]{#1(#2)\ #3}
-\newcommand{\ireset}[1]{\mathsf{Reset}\ #1}
-\newcommand{\dom}[1]{\textrm{dom}(#1)}
 
@@
 % - korean: 한글제목(korean title) | english: 영문제목(english title)
-\title[korean] {양자 회로 시뮬레이터의 테스트 오라클로서의 물리적으로 일관된 실행 가능한 형식 의미론}
-\title[english]{Physically-Consistent Executable Formal Semantics as a Testing Oracle for Quantum Circuit Simulators}
+\title[korean] {탄소 나노튜브의 물리적 특성에 대한 이론 연구}
+\title[english]{Theoretical Study on Physical Properties of
+                Carbon Nanotubes}
 
@@
 %
-\author[korean] {이}{강 욱}
-\author[korean2] {이}{강욱}    %이름을 붙여 써 주시기 바랍니다.
-\author[chinese]{李}{康 旭}
-\author[english]{Lee}{Kanguk}
+\author[korean] {안}{진 현}
+\author[korean2] {안}{진현}    %이름을 붙여 써 주시기 바랍니다.
+\author[chinese]{安}{眞 玄}
+\author[english]{Ahn}{Jin-Hyun}
 
@@
 % - major: 주 지도교수  | coopr: 공동 지도교수
-\advisor[major]{류 석 영}{Sukyoung Ryu}{signed}
-\advisor[major2]{류석영}{Sukyoung Ryu}{signed}    %한글 성과 한글 이름을 모두 붙여 써 주시기 바랍니다.
+\advisor[major]{홍 길 동}{Gildong Hong}{signed}
+\advisor[major2]{홍길동}{Gildong Hong}{signed}    %한글 성과 한글 이름을 모두 붙여 써 주시기 바랍니다.
 
@@
 % c: it represents only the name of program that is under the department (consider this when you are in the program not only department)
+
 \department{CS}{engineering}{a}
 
+% @command studentid 학번(ID)
+\studentid{20100000}
+
 % @command referee 심사위원 (석사과정 3인, 박사과정 5인)
-\referee[1]{류 석 영}
-\referee[2]{유 신}
-\referee[3]{강 지 훈}
-% \referee[4]{가 동 호}
-% \referee[5]{박 태 현}
+\referee[1]{홍 길 동}
+\referee[2]{안 진 현}
+\referee[3]{정 태 성}
+\referee[4]{가 동 호}
+\referee[5]{박 태 현}
 % \referee[5] {Barack Obama}
@@
 % @param   year,month,day 연,월,일 순으로 입력
-\approvaldate{2023}{12}{12}
+\approvaldate{2020}{11}{30} 
+\fontsize{18pt}{18pt}\selectfont
 
@@
 % @param   year,month,day 연,월,일 순으로 입력
-\refereedate{2023}{12}{12}
+\refereedate{2020}{11}{30}
 
 % @command gradyear 졸업년도
-\gradyear{2024}
+\gradyear{2020}
 
@@
 
-% \theoremstyle{acmdefinition}
-\newtheorem{postulate}[theorem]{Postulate}
-
     % 앞표지, 속표지, 학위논문 제출승인서, 학위논문 심사완료 검인서는
@@
     % 반대로 옵션을 draft로 지정해주면 생성되지 않습니다.
+    % 학위논문 제출 승인서에서 자신의 전공과 교수님의 정보를 바꾸기 위해서는 첨부되어있는 제목이 kaist-ucs인 class 문서에 들어가서 ####################로 표시  
+    % 한 부분을 바꾸시길 바랍니다.
 
     % 논문 서지, 초록, 핵심 낱말, 영문 초록, 영어 핵심 낱말 (Information of thesis, abstract in korean, keywords in korean, abstract in english, keywords in english)
+    %% 한글 초록은 500자를, 영문 초록은 300 낱말을 넘지 않아야 함
+    %% 핵심 낱말은 5 개 이내로 넣음
+    %% 한글 초록에 영문 글자를 쓰지 않도록 한다.
    \thesisinfo
-%% Letters of abstract in korean must be less than 500 and words of abstract in english must be less than 300.
-%% Number of keywords must be less than 6.
-%% Don't write english letters in the abstract in korean.
+
     \begin{summary}      
-	%
-	현재 양자 프로그램 개발에서는 개발자들이 양자 회로 시뮬레이터에 많이 의존하고 있다.
-	%
-	따라서 시뮬레이터의 정확성을 보장하는 것은 매우 중요하다.
-	%
-	그러나 양자 회로를 기술하는 언어에 대한 형식적 의미론이 없기 때문에 시뮬레이터의 정확성에 대한 추론이 어렵다.
-	%
-	본 논문에서는 대표적인 양자 회로 기술 언어인 OpenQASM 2.0에 대한 실행 가능한 형식 의미론을 제시한다.
-	%
-	양자 회로의 독특한 특성으로 인해, 형식 의미론을 정의할 때 크게 두 가지 어려움이 있다.
-	%
-	첫째, 시뮬레이터의 비결정성에도 불구하고 실행 가능한 형식 의미론은 결정론적인 결과를 내야 한다.
-	%
-	이를 위해 물리학계에서 제시된 다세계 해석을 이용하여 결과의 확률 분포를 계산하는 형식 의미론을 제시한다.
-	%
-	둘째, 실행 가능한 형식 의미론은 물리적 일관성을 보장해야 한다.
-	%
-	이를 위해 OpenQASM의 핵심 언어인 OpenQASMCore를 제시하고, OpenQASM에서 OpenQASMCore로의 변환(디슈가링)을 정의한다.
-	%
-	형식 의미론과 물리적 일관성에 대한 증명은 Coq에서 정형 검증되었다.
-	%
-	본 논문에서는 실행 가능한 형식 의미론의 평가로서 두 가지 최적화 기법의 효과적인 성능 향상과, 실제 양자 프로그래밍 플랫폼에서 5개의
-	새로운 버그를 찾아내는 등의 실행 가능한 형식 의미론의 테스트 오라클로서의 유용성을 보여준다.
-	%
+    지난 10여 년간 탄소 나노튜브는 자체의 독특한 전기적, 기계적 성질로
+    인하여 다가오는 나노기술 분야의 이상적인 기초물질중의 하나로 떠오르고
+    있다. 흑연을 감는 세세한 방법에 따라 전기적 특성이 금속성에서 1eV의
+    띠간격을 가지는 반도체 특성까지 다양한 분포로 존재한다.
+    본 학위논문에서는 탄소 나노튜브의 여러 물리적 성질에 대해 고찰하는데,
+    기본적으로 제일원리 밀도함수 이론과 밀접결합근사 모형을 사용하여 전기적
+    특성과 그 제어 방법, 자기적 특성, 그리고 수송특성 등을 다루고자 한다.
     \end{summary}
@@
     \begin{Korkeyword}
-	양자 회로, 양자 프로그래밍 언어, 형식 의미론, 테스트 오라클, OpenQASM, OpenQASMCore, 다세계 해석, 양자 시뮬레이터,
-	물리적 일관성
+     가, 나, 다
     \end{Korkeyword}
@@
     \begin{abstract}
-	In the current practice of quantum program development, programmers heavily
-	rely on quantum circuit simulators.
-	%
-	Ensuring the correctness of these simulators is therefore crucial.
-	%
-	However, the lack of formal semantics for quantum circuit description
-	languages, which are used to describe quantum circuits, hinders reasoning about
-	simulator correctness.
-	%
-	In this work, we present the first executable formal semantics, which can
-	function as a testing oracle, for OpenQASM 2.0, a representative quantum
-	circuit description language.
-	%
-	The unique characteristics of quantum circuits present challenges in defining
-	formal semantics.
-	%
-	First, executable semantics must yield deterministic results despite the
-	nondeterminism of simulators.
-	%
-	To achieve this, our semantics computes the probability distribution of
-	outcomes using the \emph{many-worlds interpretation} (MWI) from the physics
-	literature.
-	%
-	Additionally, we propose two optimizations for the semantics.
-	%
-	Second, ensuring \emph{physical consistency} is essential, meaning the
-	semantics must adhere to physical laws.
-	%
-	To facilitate this, we introduce OpenQASMCore, a core language for OpenQASM,
-	simplifying the proof process, and define desugaring from OpenQASM to
-	OpenQASMCore.
-	%
-	The semantics and proofs are mechanized in Coq.
-	%
-	Our evaluation shows the optimizations' effectiveness in improving performance
-	and the utility of our semantics as a testing oracle by identifying five
-	previously unknown bugs in two real-world quantum programming platforms.
-	%
+        For the last decade, carbon nanotubes have been emerging as one
+        of ideal materials for the building block of the forthcoming
+        nanotechnology, due to their unique electrical and mechanical
+        properties. Depending on detailed wrapping-up methods, their
+        electronic properties show a wide spectrum from metals to
+        large-gap semiconductors with band gaps of 1eV.
+        In this thesis, we study various physical properties of carbon
+        nanotubes, including electrical properties and their controlling
+        methods, magnetic properties, and transport characteristics,
+        based on the first-principles density-functional theory and
+        the tight-binding model.
     \end{abstract}
@@
      \begin{Engkeyword}
-	Quantum circuit, quantum programming language, formal semantics, testing
-	oracle, OpenQASM, OpenQASMCore, many-worlds interpretation, Quantum simulator, and
-	physical consistency.
+     a, b, c
     \end{Engkeyword}
@@
     %\makecontents
-
+%% 한글로 쓴 논문에는 본문에 영문 글자를 쓰지 않는다. 다만, 꼭 필요할 때에는 ‘한글 낱말 (영문 낱말)’ 꼴로 적는다.
 %% 이하의 본문은 LaTeX 표준 클래스 report 양식에 준하여 작성하시면 됩니다.
@@
 
-\include{1-intro}
+\chapter{머릿말}
+본문을 한글로 작성할 때 머릿말로 시작을 하시는 게 좋습니다. \cite{FD1}
+인용은 다음과 같이 합니다 \cite{RVP1}-\cite{ML2}.
+인용은 뒤에 인용을 쓰는 칸이 있습니다. 참고하여서 인용하시길 바랍니다 \cite{SOCA2,EF2}.
+한글 논문에는 영어를 쓰지 마시기 바랍니다.
 
-\include{2-openqasmcore}
+\chapter{본문 작성}
 
-\include{3-consistency}
+\section{작성}
 
-\include{4-openqasm}
+장과 절 그리고 부절로 본문을 작성하실 수 있습니다.
 
-\include{5-evaluation}
+\subsection{자동}
 
-\include{6-related}
+이것들은 자동으로 차례에 들어가게 됩니다.
 
-\include{7-conclusion}
+\section{한글 논문}
 
-\chapter[Method]{Method}
-\label{ch:ourmethod}
-\section{Quantum Communication}
-To enhance the performance gain of original cognitive radio networks (CRNs),
-leveraging cooperative diversity has attracted a lot of attention
-\cite{SOCA1}...
+한글 논문에는 영어를 쓰지 마시기 바랍니다.
 
-\section{Examples}
+\chapter{그림, 표}
 
-\section{Applications of Quantum Communication}
-\label{ch:application}
-\noindent
-Despite its potential to improve the throughput, spatial domain diversity was not fully considered in the studies of original CCRNs.
-Utilizing the spatial domain for the communications, the concept of MIMO has been adopted in many cases to increase the wireless capacity \cite{EF1,FD2}...
+\section{그림과 표를 본문에서 이야기하기}
 
-Let us consider an MIMO-CCRN with one PL and one SL. Each PU has one legacy
-antenna and each SU has two STAR antennas. The duration of one communication
-time frame is divided into two phases \cite{RVP2,ML2} ...
+본문에서 그림과 표에 관해 이야기를 할 때도 인용에서처럼 하시면 됩니다.
+
 %%
@@
 \begin{table}[t]
-	\caption[Enter the caption title here]{Energy stability $E$ (eV) per molecule of all meta-stable
-		isomer states of C$_{60}$ opening process for forming the (5,5) cap.
-		In the SW-I and SW-II, both ferromagnetic (Ferro) and paramagnetic (Para)
-		spin configurations are obtained, whereas only non-magnetic configuration
-		is obtained in the BF, SW-III, and CAP(5,5).
-		$M$ is total magnetization $n_{\rm up}$-$n_{\rm down}$ in unit of $\mu_B$, where
-		$n_{\rm up(down)}$ is the number of up (down) spins.
+\caption[캡션제목 넣으십시오]{표 제목을 넣으십시오.
 }
@@
 \hline\hline
-			 &               & BF & \multicolumn{2}{c}{SW-I} &       & \multicolumn{2}{c}{SW-II} & SW-III & CAP    &               \\
+& & BF &\multicolumn{2}{c}{SW-I}&&\multicolumn{2}{c}{SW-II}&SW-III&CAP&\\
 \cline{4-5} \cline{7-8}
-			 &               &    & Para                     & Ferro &                           & Para   & Ferro  &      &      & \\
+&               &   &  Para & Ferro &&   Para &  Ferro &      &      &\\
 \hline
-			 & $E$ (eV)      & 0  & 7.796                    & 7.832 &                           & 10.418 & 10.408 & 11.5 & 13.2 & \\
-			 & $M$ ($\mu_B$) & 0  & 0                        & 1.94  &                           & 0      & 2.06   & 0    & 0    & \\
+& $E$ (eV)      & 0 & 7.796 & 7.832 && 10.418 & 10.408 & 11.5 & 13.2 &\\
+& $M$ ($\mu_B$) & 0 &     0 &  1.94 &&      0 &   2.06 &    0 &    0 &\\
 \hline\hline
@@
     \centerline{\includegraphics[width=12.5cm]{sample-fig1}}
-	\caption[Enter the caption title here]{ Ball-and-stick models of meta-stable isomers in
-		cage opening process from a C$_{60}$ buckminsterfullerene
-		to a (5,5) capsule. We name them BF and CAP(5,5).
-		Depending on the number of the Stone-Wales (SW) transformation,
-		we call the intermediate isomers with SW-I, SW-II, and SW-III.
-		Highlighted atoms are undercoordinated except BF.
+    \caption[캡션제목 넣으십시오]{그림 제목을 넣으십시오.
     } \label{mag-fig1}
@@
 
-\section{Discussion and Future Work}
-\noindent
-The achievable rates can be calculated by finding the statistics of the five signals transmitted that maximize the mutual information between $s_{t,XY}$ and $y_{t,XY}$ for $(X,Y)=(T,C), (C,N)$, and $(N,C)$ when $t=1$, and $(X,Y)=(C,R), (C,N)$, and $(N,C)$ when $t=2$ \cite{SOCA2,EF2}...
+\chapter{맺음말}
 
-\chapter{Conclusion}
-\noindent
-We have proposed a novel FD MIMO-CCRN framework providing a reasonable performance improvement compared with the conventional MIMO-CCRN framework...
+마지막은 맺음말로 하는 것을 권합니다.
+
 %%
@@
 %% bibliography
-%% It can be changed but should include sufficient information.
-% \bibliographystyle{abbrvnat}
-\bibliographystyle{plainnat}
-\bibliography{references}
+%% 위는 보기이므로 학과 또는 논문의 특성에 맞게 조정 가능함. 다만, 참고문헌마다 충분한 정보가 들어 있어야 한다.
+\begin{thebibliography}{00}
+
+\bibitem{FD1} 박상우, \underline{동시 송수신 안테나를 두 개 쓰는 협력 인지 무선통신망에 알맞은 전 이중 통신}, 한국과학기술원 석사 학위 논문, 2016.
+
+\bibitem{RVP1} 송익호, 박철훈, 김광순, 박소령, \underline{확률변수와 확률과정}, 자유아카데미, 2014.
+
+\bibitem{ML1} 송익호, 안태훈, 민황기, \underline{인지 무선에서의 광대역 주파수 검출 방법 및 장치}, 특허등록번호 10-1494966, 2015년 2월 12일.
+
+\bibitem{SOCA1} 호우위시, 이원주, 이승원, 안태훈, 이선영, 민황기, 송익호, “선형 판별 분석에서 부류안 분산 행렬의 영 공간 재공식화,” \underline{한국통신학회 2012년도 추계종합학술발표회}, 대한민국 고려대학교, 242-243쪽, 2012년 11월.
+
+\bibitem{EF1} 민황기, 안태훈, 이승원, 이성로, 송익호, “비간섭 전력 부하 감시용 고차 적률 특징을 갖는 전력 신호 인식,” \underline{한국통신학회논문지}, 제39C권, 제7호, 608-614쪽, 2014년 7월.
+
+\bibitem{FD2} S.
+Park, \textit{Full-Duplex Communication for Cooperative Cognitive Radio Networks with Two Simultaneous Transmit and Receive Antennas}, Master Thesis, Korea Adv.
+Inst.
+Science, Techn., Daejeon, Republic of Korea, 2016.
+
+\bibitem{RVP2}  I. Song, J. Bae, and S. Y. Kim, \textit{Advanced Theory of Signal Detection: Weak Signal Detection in Generalized Observations}, Springer-Verlag, 2002.
+
+\bibitem{ML2} I. Song, T. An, and J. Oh, \textit{Near ML decoding method based on metric-first search and branch length threshold,} registration no. US 8018828 B2, Sep. 13, 2011, USA.
+
+\bibitem{SOCA2} H.-K.
+Min, T.
+An, S.
+Lee, and I.
+Song, “Non-intrusive appliance load monitoring with feature extraction from higher order moments,” in \textit{Proc.
+6th IEEE Int.
+Conf.
+Service Oriented Computing, Appl.,} Kauai, HI, USA, pp.
+348-350, Dec.
+2013.
+
+\bibitem{EF2} I. Song and S. Lee, “Explicit formulae for product moments of multivariate Gaussian random variables,” \textit{Statistics, Probability Lett.,} vol. 100, pp. 27-34, May 2015.
+
+\end{thebibliography}
 
 %%
-%% 감사의 글 시작
+%% 사사 시작
 %% Acknowledgement
-%%
+%% 사사 작성은 선택사항임
 % @command acknowledgement 감사의글
@@
 % - 1 : 본문과 감사의 글이 둘 다 한글일 때  | 2 : 본문은 한글인데 감사의 글이 영어일 때 | 3 :  본문과 감사의 글이 둘 다 영어일 때  | 4 : 본문은 영어인데 감사의 글이 % 한글일 때
-%% It is optional.
 
-\acknowledgment[4]
+\acknowledgment[1]
 언제나 저를 바른 길로 이끌어 주시는 송익호 교수님께 큰 고마움을 느낍니다.
@@
 %% Curriculum Vitae
-%%
+%% 약력 작성은 선택사항임. 약력 내용도 알맞게 바꿀 수 있음
 % @command curriculumvitae 이력서
@@
 % - 1 : 본문과 약력이 둘 다 한글일 때  | 2 : 본문은 한글인데 약력이 영어일 때 | 3 :  본문과 약력이 둘 다 영어일 때  | 4 : 본문은 영어인데 약력이 한글일 때
-%% It is optional and you can change form of this in the class file if you want.
-\curriculumvitae[4]
+
+\curriculumvitae[1]
 
@@
     \begin{personaldata}
-	\name       {이 강 욱}
-	\dateofbirth{1999}{08}{01}
+        \name       {안 진 현}
+        \dateofbirth{199x}{x}{xx}
         \birthplace {...}
-	\address    {...}
+        \address    { ...}
      \end{personaldata}
@@
     \begin{education}
-	\item[2015. 3.\ --\ 2017. 2.] 고등학교 (2년 수료)
-	\item[2017. 3.\ --\ 2022. 8.] 한국과학기술원 물리학과 (학사)
-	\item[2022. 9.\ --\ 2024. 2.] 한국과학기술원 전산학부 (석사)
+        \item[2007. 3.\ --\ 2009. 2.] 고등학교 (2년 수료)
+        \item[2009. 2.\ --\ 2013. 8.] 한국과학기술원 수리과학과 (학사)
+        \item[2013. 9.\ --\ 2016. 2.] 한국과학기술원 수리과학과 (석사)
     \end{education}
@@
     \begin{career}
-	\item[2021. 3.\ --\ 2021. 8.] 쓰레기통 청소
+        \item[2013. 9.\ --\ 2016. 2.] 한국과학기술원 수리과학과 일반조교
     \end{career}
@@
 %%    \end{activity}
+%% 학회활동을 쓰고싶으시면, 이 문서와 클래스 문서의 학회활동 부분을 사용하십시오.
 
@@
     \begin{publication}
-	%\item \textbf{Yong-Hyun Kim}, J. Choi, K.J. Chang, and D. Tomanek,
-	%     \textit{Magnetic instability in partly opened C$_{60}$ isomers},
-	%     in preparation.
-	\item H.-K. Min, Y. Hou, {\bf S. Park}, and I. Song, ``A computationally efficient
-	scheme for feature extraction with kernel discriminant analysis," \textit{Patt.
-		Recogn.}, vol.~50, no.~2, pp.~45-55, Feb. 2016 (to be published).
+        \item J.
+Ahn, \textit{Analysis of Tail Probability of Interference at a Node in 2-dimensional Homogeneous Poisson Point Process}, Master Thesis, Korea Adv.
+Inst.
+Science, Techn., Daejeon, Republic of Korea, 2016.
     \end{publication}

### 2024-10-07 paper/1-intro.tex
Write main.tex

--- 2024-10-07 kw/1-intro.tex
@@
 \noindent
-Quantum computing is attracting significant attention from academia and
-industry, offering distinct advantages over classical computing by enabling the
-development of substantially faster algorithms.
-%
-A prime example is \citet{shor1994algorithms}'s algorithm, which efficiently
-factorizes integers in polynomial time.
-
-When developing quantum programs, \emph{quantum circuit simulators} are
-essential.
-%
-Quantum programs, written in high-level languages, are typically compiled to
-quantum circuits described by specific \emph{quantum circuit description
-	languages}~\cite{wang2022qdiff, paltenghi2023morphq}.
-%
-These circuits are executed either on real quantum computers or simulators
-running on classical devices.
-%
-Owing to the high cost of quantum computing resources, developers predominantly
-rely on simulators.
-
-Unfortunately, the correctness of existing quantum circuit simulators has not
-been explored, primarily due to the lack of formal semantics for quantum
-circuit description languages.
-%
-Since formal semantics are crucial for verifying language implementations,
-researchers have defined formal semantics for numerous real-world languages,
-such as
-%
-JavaScript~\cite{maffeis2008operational,guha2010essence,politz2012tested,bodin2014trusted,park2015kjs,park2021jiset},
-%
-C~\cite{blazy2009mechanized,leroy2009formal,ellison2012executable,krebbers2014formal,memarian2016depths},
-%
-Java~\cite{farzan2004formal,klein2006machine,bogdanas2015kjava},
-%
-Standard ML~\cite{kumar2014cakeml,lee2007towards},
-%
-WebAssembly~\cite{watt2018mechanising,watt2021two},
-%
-Rust~\cite{jung2017rustbelt},
-%
-Python~\cite{politz2013python},
-%
-PHP~\cite{filaretti2014executable},
-%
-Solidity~\cite{jiao2020semantic},
-%
-and even the POSIX shell~\cite{greenberg2019executable},
-%
-often using proof assistants or semantic frameworks like
-K~\cite{rosu2010overview}.
-%
-Particularly, \emph{executable} formal semantics has been beneficial for
-testing, serving as a testing oracle to aid in bug
-detection~\cite{watt2023wasmref,schumi2021spectest,park2021jest} and test suite
-generation~\cite{park2023feature}.
-%
-However, the semantics of quantum circuit description languages has not been
-formalized yet.
-%
-Notably, OpenQASM~\cite{cross2017open}, a representative language supported by
-various major quantum programming platforms~\cite{qiskit, cirq, q-sharp},
-defines its syntax using BNF but explains its semantics in prose, lacking
-formal rigor.
-%
-This gap hinders the systematic verification and testing of quantum circuit
-simulators.
-
-In this work, we present the first executable formal semantics for OpenQASM
-2.0.
-%
-We chose version 2.0 over the newer OpenQASM 3.0 due to its established
-stability and extensive software support.
-%
-Our formalization employs Coq, from which we extract OCaml code to achieve
-executability.
-%
-The unique characteristics of quantum circuits pose distinct challenges in
-developing formal semantics, compared to traditional programming languages.
-%
-Specifically, we address two challenges:
-%
-(1) taming \emph{nondeterminism} and (2) ensuring \emph{physical consistency}.
-
-The first challenge stems from the nondeterminism in quantum circuit execution,
-a fundamental aspect of quantum mechanics.
-%
-Quantum circuit simulators mimic this nondeterministic behavior through
-\emph{sampling-based} executions.
-%
-While outcomes may vary between executions, they collectively conform to a
-specific probability distribution over multiple runs.
-%
-Since executable formal semantics should have deterministic results, we adopt
-the notion of \emph{exact inference}, i.e., calculating the probability
-distribution of possible outcomes, from probabilistic programming.
-%
-For \emph{static quantum circuits}, featuring end-of-circuit measurements,
-exact inference is relatively straightforward, as probability distributions can
-be computed by multiplying matrices corresponding to circuits' gates.
-%
-However, applications like quantum teleportation and iterative phase
-estimation~\cite{dobsicek2007arbitrary} involve \emph{dynamic quantum
-	circuits}~\cite{dynamic-circuit}.
-%
-These circuits modify the control flow based on mid-circuit measurements, and
-their effects cannot be captured through matrix multiplication.
-
-To enable exact inference for dynamic circuits, we adopt physicist
-\citet{everett1957relative}'s \emph{many-worlds interpretation} (MWI), which
-posits that each nondeterministic event generates multiple parallel
-\emph{worlds} with varying outcomes.
-%
-Our semantics maintains a list of worlds, starting with one world and splitting
-it as execution progresses.
-%
-A potential concern with MWI is that large circuits may create an excessive
-number of worlds, leading to unreasonably long execution times and hindering
-the use as a testing oracle.
-%
-To address this, we introduce two optimizations:
-%
-\emph{world merging} and \emph{efficient reset}, both aimed at reducing the
-number of worlds.
-
-The second challenge relates to the physical laws governing quantum circuits.
-%
-Unlike traditional programming languages for high-level computations, quantum
-circuit description languages specify behavior of circuits running on quantum
-hardware, bound to physical principles.
-%
-Physicists use \emph{density matrices} to model quantum states, and these
-matrices and their transformations must fulfill specific constraints, stated
-through five postulates~\cite{scherer2019mathematics}.
-%
-Formal semantics must align with this physical model because deviating from
-these constraints would yield probability distributions unrealizable in actual
-quantum devices, failing as a testing oracle for simulators.
-%
-Therefore, verifying the \emph{physical consistency}, i.e., adherence to these
-postulates, of our semantics is critical.
-
-To prove the physical consistency of our semantics, we introduce OpenQASMCore,
-a core language of OpenQASM.
-%
-The simplified nature of OpenQASMCore facilitates a streamlined proof of
-physical consistency.
-%
-We define OpenQASMCore's big-step operational semantics and prove its physical
-consistency in Coq.
-%
-We then indirectly define OpenQASM's semantics through desugaring from OpenQASM
-to OpenQASMCore.
-%
-As OpenQASM programs are essentially OpenQASMCore programs with syntactic
-sugar, OpenQASM's physical consistency naturally follows.
-
-To demonstrate the effectiveness of our formal executable semantics as a
-testing oracle, we tested real-world quantum programming platforms,
-Qiskit~\cite{qiskit} and staq~\cite{staq}, specifically a simulator and its
-optimization passes in Qiskit and optimizers in staq, and identified five
-previously unknown bugs with distinct root causes.
-%
-For testing, we collected three OpenQASM benchmark suites~\cite{cross2017open,
-	li2023qasmbench, tomesh2022supermarq} and generated random circuits, some of
-them designed to target Qiskit's optimization passes.
-%
-Of the five bugs, one was in Qiskit's transpiler, three in Qiskit's
-optimization passes, and one in staq's optimizer.
-%
-All the bugs were confirmed by the developers and two have been already fixed.
-
-Overall, our contributions are as follows:
-%
-\begin{itemize}
-	%
-	\item We introduce OpenQASMCore, a core language of OpenQASM
-	      (\cref{ch:openqasmcore}).
-	      %
-	\item We define the executable formal semantics of OpenQASMCore, propose its
-	      optimizations, and prove its physical consistency using Coq
-	      (\cref{ch:consistency}).
-	      %
-	\item We indirectly define the semantics of OpenQASM by introducing desugaring from
-	      OpenQASM to OpenQASMCore (\cref{ch:openqasm}).
-	      %
-	\item We show the effectiveness of our optimizations in reducing execution times and
-	      validate the utility of our executable semantics as a testing oracle by
-	      identifying five previously unknown bugs in real-world quantum programming
-	      platforms (\cref{ch:evaluation}).
-	      %
-\end{itemize}
-%
-We then discuss related work (\cref{ch:related}) and conclude
-(\cref{ch:conclusion}).

### 2024-10-07 paper/2-background.tex
Write main.tex

% !TEX root = main.tex

\chapter{Background}
\label{ch:background}
\noindent

background

### 2024-10-24 paper/1-intro.tex
Write the intro in bullet points

@@
 \noindent
+
+\begin{itemize}
+\item Wasm
+  \begin{itemize}
+  \item is low-level bytecode language, platform-indep, ...
+  \item is compile target for Web applications
+  \item is used in wide area such as IoT, ...
+  \end{itemize}
+\item Wasm Spec
+  \begin{itemize}
+  \item descibes syntax and semantics of Wasm
+  \item should be rigorous because ... (brief explanation for two format)
+  \item has two format to describe semantics (formal\&prose notation)
+  \item formal notation is written in math expression: concise, clear, amenable to proofs, ...
+  \item prose notation is step by step instruction written English: vague (in some part), amenable to developing engines
+  \end{itemize}
+\item SpecTec
+  \begin{itemize}
+  \item is proposed to alleviate spec writing process
+  \item is accepted by Wasm committee to be an official Wasm specification tool
+  \item define AL to describe Wasm semantics in prose notation
+  \item AL is designed to be similar to prose notation (describe with fig)
+  \item implement AL interpreter to make the spec executable
+  \item executable spec passes official wasm test $\leftarrow$ correctness
+  \end{itemize}
+\item Problem
+  \begin{itemize}
+  \item no semantics in AL
+  \item prose notation is still vague unless read AL interpreter code
+  \item ???
+  \end{itemize}
+\item Contribution
+  \begin{itemize}
+  \item formalize AL syntax$\&$semantics
+  \item make prose notation clear to understand
+  \item ???
+  \end{itemize}
+\end{itemize}

### 2024-10-26 paper/3-motivation.tex
Add motivation
Compare official prose & SpecTec prose

--- 2024-10-07 paper/2-background.tex
@@
 
-\chapter{Background}
-\label{ch:background}
+\chapter{Motivation}
+\label{ch:motivation}
 \noindent
 
-background
+\newcommand{\officialp}{official prose}
+\newcommand{\spectecp}{SpecTec prose}
+
+\begin{figure}[h!]
+    \centerline{\includegraphics[width=15cm]{fig/loop}}
+    \caption[Enter the caption title here]{\texttt{loop} instruction} \label{loop-fig}
+    \centerline{\includegraphics[width=15cm]{fig/br}}
+    \caption[Enter the caption title here]{\texttt{br} instruction} \label{br-fig}
+\end{figure}
+TODO: remove redundant part in the fig
+
+It might be easy to understand how the \officialp{} explains the control flow
+of WebAssembly, if we assume that a WebAssembly code is loaded on a memory and
+a pc points to the instruction to execute.
+To describe control flow, it uses a structure named \textit{block} which
+consitutes of a instruction sequence.
+When executing the instruction sequence inside the block, it can jump to the
+starting point of the block or the end of the block.
+For example, \cref{loop-fig} is the \officialp{} \texttt{loop} instruction.
+It says that the continuation is the start of the loop and the information is
+stored in a label, which is pushed in the stack when
+entering the block.
+\cref{br-fig} is a \texttt{br} instruction in the \officialp{}.
+According to it, if the \texttt{br} instruction in the block is executed to
+branch out the block, the label is popped in the stack and jumps to the
+continuation of the label, which is the start of the loop instruction.
+
+TODO: detailed explanation of \spectecp{} w/ fig
+
+However, rather than using the notion like pc, the \spectecp{} assumes that the
+\red{machine} takes WebAssembly instructions one by one, and changes the
+internal state accordingly.
+This is because the \spectecp{} is generated automatically from the \red{dsl}
+which doesn't model the pc.
+As a result, rather than jump to the start of the loop, \spectecp{} stores the
+\texttt{loop} instruction in the label, and executes it again if \texttt{br}
+instruction is executed.
+
+don't remember end!
+
+% official spec prose notation control structure
+% seems to be pc-based
+% AL from DSL ~ formal notation: no pc
+% AL: instruction sequence input continously
+% device exit semantics for AL
+% Executable spec passes all wasm testcases
+% weird, maybe not happened in realworld, but valid wasm code
+% wrong!

### 2024-10-26 paper/3-motivation.tex
Add code example

@@
 
+It might be easy to understand how the \officialp{} explains the control flow
+of WebAssembly, if we assume that a WebAssembly code is loaded on a memory and
+a pc points to the instruction to execute.
+To describe control flow, it uses a structure named \textit{block} which
+consitutes of a instruction sequence.
+When executing the instruction sequence inside the block, it can jump to the
+starting point of the block or the end of the block.
+
+\textbf{Example 1}
+\begin{verbatim}
+  // infinite loop
+  (loop (result i32) (i32.const 42) (br 0) (unreachable) end)
+\end{verbatim}
+
+Example 1 is a WebAssembly code example that has a \texttt{loop} whose result
+type is \texttt{i32} and has a block of three instructions: \texttt{i32.const},
+\texttt{br}, \texttt{unreachable}.
+\cref{loop-fig} is the \officialp{} of \texttt{loop} instruction.
+It says that the continuation is the start of the loop, the information is
+stored in a label, and it \textit{enters} the block with the label.
+The term enter means that it pushes the label in the stack and makes pc points
+to the first instruction in the block.
+\texttt{i32.const} instruction is executed first, which just pushes the
+value in the stack.
+Then, \texttt{br} instruction is executed.
+\cref{br-fig} is the \officialp{} of \texttt{br} instruction.
+It pops the label from the stack, and makes pc points to the continuation of
+the label, which is the start of the loop instruction.
+As a result, the code example above is a infinite loop pushing a \texttt{i32}
+value \texttt{42} infinitely.
+
 \begin{figure}[h!]
@@
 
-It might be easy to understand how the \officialp{} explains the control flow
-of WebAssembly, if we assume that a WebAssembly code is loaded on a memory and
-a pc points to the instruction to execute.
-To describe control flow, it uses a structure named \textit{block} which
-consitutes of a instruction sequence.
-When executing the instruction sequence inside the block, it can jump to the
-starting point of the block or the end of the block.
-For example, \cref{loop-fig} is the \officialp{} \texttt{loop} instruction.
-It says that the continuation is the start of the loop and the information is
-stored in a label, which is pushed in the stack when
-entering the block.
-\cref{br-fig} is a \texttt{br} instruction in the \officialp{}.
-According to it, if the \texttt{br} instruction in the block is executed to
-branch out the block, the label is popped in the stack and jumps to the
-continuation of the label, which is the start of the loop instruction.
-
 TODO: detailed explanation of \spectecp{} w/ fig
 
-However, rather than using the notion like pc, the \spectecp{} assumes that the
-\red{machine} takes WebAssembly instructions one by one, and changes the
-internal state accordingly.
+However, the \spectecp{} assumes that the \red{machine} takes WebAssembly
+instructions one by one, and changes the internal state accordingly without pc.
 This is because the \spectecp{} is generated automatically from the \red{dsl}
-which doesn't model the pc.
+which uses rewrite rule to describe WebAssembly semantics.
 As a result, rather than jump to the start of the loop, \spectecp{} stores the
-\texttt{loop} instruction in the label, and executes it again if \texttt{br}
-instruction is executed.
+\texttt{loop} instruction itself in the label, and executes it again if the
+\texttt{br} instruction is executed.
+Not only that, it doesn't have block structure.
 

### 2024-10-26 paper/3-motivation.tex
Minor change

@@
 consitutes of a instruction sequence.
-When executing the instruction sequence inside the block, it can jump to the
-starting point of the block or the end of the block.
+When executing the instructions in the block, pc can be changed to the starting
+point of the block or the end of the block.
 
@@
 stored in a label, and it \textit{enters} the block with the label.
-The term enter means that it pushes the label in the stack and makes pc points
-to the first instruction in the block.
+The term \textit{enter} means that it pushes the label in the stack and makes
+pc points to the first instruction in the block.
 \texttt{i32.const} instruction is executed first, which just pushes the
-value in the stack.
+\texttt{i32} value \texttt{42} in the stack.
 Then, \texttt{br} instruction is executed.
@@
 the label, which is the start of the loop instruction.
-As a result, the code example above is a infinite loop pushing a \texttt{i32}
-value \texttt{42} infinitely.
+As a result, the code example above is a infinite loop pushing \texttt{42}
+forever.
 

### 2024-10-26 paper/3-motivation.tex
Expand to general control flow

@@
 \texttt{br}, \texttt{unreachable}.
-\cref{loop-fig} is the \officialp{} of \texttt{loop} instruction.
+\cref{loop-fig} is the \officialp{} of the \texttt{loop} instruction.
 It says that the continuation is the start of the loop, the information is
@@
 pc points to the first instruction in the block.
-\texttt{i32.const} instruction is executed first, which just pushes the
+The \texttt{i32.const} instruction is executed first, which just pushes the
 \texttt{i32} value \texttt{42} in the stack.
-Then, \texttt{br} instruction is executed.
-\cref{br-fig} is the \officialp{} of \texttt{br} instruction.
+Then, The \texttt{br} instruction is executed.
+\cref{br-fig} is the \officialp{} of the \texttt{br} instruction.
 It pops the label from the stack, and makes pc points to the continuation of
@@
 TODO: remove redundant part in the fig
+
+Similar to the control flow using the block and the label, there are many other
+control instructions and structures such as a \texttt{call} instruction that
+calls a function with a frame and a \texttt{return} instruction that returns from
+the function with the frame.
 

### 2024-10-28 paper/3-motivation.tex
Add exit label of official prose

@@
 
+\red{TODO: remove redundant part in the fig}
+
+% control flow structure in official prose
 It might be easy to understand how the \officialp{} explains the control flow
@@
 
+% an example of Wasm control flow
 \textbf{Example 1}
@@
   // infinite loop
-  (loop (result i32) (i32.const 42) (br 0) (unreachable) end)
+  (loop (result i32) (i32.const 42) (br 0) (unreachable) end) (unreachable)
 \end{verbatim}
 
-Example 1 is a WebAssembly code example that has a \texttt{loop} whose result
-type is \texttt{i32} and has a block of three instructions: \texttt{i32.const},
-\texttt{br}, \texttt{unreachable}.
-\cref{loop-fig} is the \officialp{} of the \texttt{loop} instruction.
+Example 1 is a WebAssembly code example that has \texttt{loop} and
+\texttt{unreachable}
+Here, result type of the \texttt{loop} is \texttt{i32} and the \texttt{loop}
+has a block of three instructions: \texttt{i32.const}, \texttt{br}, and
+\texttt{unreachable}.
+\cref{fig:loop} is the \officialp{} of the \texttt{loop} instruction.
 It says that the continuation is the start of the loop, the information is
@@
 Then, The \texttt{br} instruction is executed.
-\cref{br-fig} is the \officialp{} of the \texttt{br} instruction.
+\cref{fig:br} is the \officialp{} of the \texttt{br} instruction.
 It pops the label from the stack, and makes pc points to the continuation of
@@
 As a result, the code example above is a infinite loop pushing \texttt{42}
-forever.
+forever without executing \texttt{unreachable} instruction.
 
@@
     \centerline{\includegraphics[width=15cm]{fig/loop}}
-    \caption[Enter the caption title here]{\texttt{loop} instruction} \label{loop-fig}
+    \caption[Enter the caption title here]{\texttt{loop} instruction} \label{fig:loop}
     \centerline{\includegraphics[width=15cm]{fig/br}}
-    \caption[Enter the caption title here]{\texttt{br} instruction} \label{br-fig}
+    \caption[Enter the caption title here]{\texttt{br} instruction} \label{fig:br}
 \end{figure}
-TODO: remove redundant part in the fig
+
+% exit label in official prose
+There is also a special form of a behavior related to the block: an
+\textit{exit label}.
+\cref{fig:exit-label} is the \officialp of the exit label.
+It is special because this behavior is not performed without an explicit
+WebAssembly instruction.
+Rather, the behavior is performed when \textbf{the end of a block is reached}
+without control instructions or runtime error.
+The behavior is that the label is popped from the stack, and the pc changes to
+the point after the end of the block.
+
+\begin{figure}[h!]
+    \centerline{\includegraphics[width=15cm]{fig/exit}}
+    \caption[Enter the caption title here]{exit label} \label{fig:exit-label}
+\end{figure}
+
+% an example of exit label
+\textbf{Example 2}
+\begin{verbatim}
+  // exit label
+  (loop (result i32) (i32.const 42) end) (f64.const 3.14)
+\end{verbatim}
+
+Example 2 is \texttt{loop} with a block that has only \texttt{i32.const}.
+A label is pushed in the stack when a \texttt{loop} instruction is executed.
+After \texttt{i32.const} is executed, the end of the block is reached.
+As a result, exit label occurs so that the label is popped from the stack,
+pc changes to the point after the end of block: \texttt{f64.const}.
+Therefore, the code example above results in the value 42 and 3.14 pushed to
+the stack.
 

### 2024-10-28 paper/3-motivation.tex
Add control flow of SpecTec prose

@@
 
-% an example of Wasm control flow
+% an example of Wasm control flow in official prose
 \textbf{Example 1}
@@
 
-% exit label in official prose
+% exiting label in official prose
 There is also a special form of a behavior related to the block: an
-\textit{exit label}.
-\cref{fig:exit-label} is the \officialp of the exit label.
+\textit{exiting label}.
+\cref{fig:exiting-label} is the \officialp of the exiting label.
 It is special because this behavior is not performed without an explicit
@@
 \begin{figure}[h!]
-    \centerline{\includegraphics[width=15cm]{fig/exit}}
-    \caption[Enter the caption title here]{exit label} \label{fig:exit-label}
+    \centerline{\includegraphics[width=15cm]{fig/exiting}}
+    \caption[Enter the caption title here]{exiting label} \label{fig:exiting-label}
 \end{figure}
 
-% an example of exit label
+% an example of exiting label in official prose
 \textbf{Example 2}
 \begin{verbatim}
-  // exit label
+  // exiting label
   (loop (result i32) (i32.const 42) end) (f64.const 3.14)
@@
 After \texttt{i32.const} is executed, the end of the block is reached.
-As a result, exit label occurs so that the label is popped from the stack,
+As a result, exiting label occurs so that the label is popped from the stack,
 pc changes to the point after the end of block: \texttt{f64.const}.
@@
 
+% general control flow structure of official prose
 Similar to the control flow using the block and the label, there are many other
@@
 the function with the frame.
+Not only that, there are the special behaviors such as a \textit{returning from
+a function} that corresponds to the exiting label.
 
-TODO: detailed explanation of \spectecp{} w/ fig
+% control flow structure in SpecTec prose
+However, the \spectecp{} assumes different a model to explain the control flow
+of WebAssembly.
+Rather than using the notion of pc, \spectecp{} assumes that WebAssembly
+instructions are just given one by one.
+This is because the \spectecp{} is generated automatically from the
+\red{SpecTec dsl}.
+\red{SpecTec dsl} uses rewrite rule to describe WebAssembly semantics, and it
+can be seen as consuming a WebAssembly instruction and values in the stack,
+changing the stack and the store, and inputing new WebAssembly instructions.
 
-However, the \spectecp{} assumes that the \red{machine} takes WebAssembly
-instructions one by one, and changes the internal state accordingly without pc.
-This is because the \spectecp{} is generated automatically from the \red{dsl}
-which uses rewrite rule to describe WebAssembly semantics.
-As a result, rather than jump to the start of the loop, \spectecp{} stores the
-\texttt{loop} instruction itself in the label, and executes it again if the
-\texttt{br} instruction is executed.
-Not only that, it doesn't have block structure.
+% an example of Wasm control flow in official prose
+\begin{figure}[h!]
+    \centerline{\includegraphics[width=15cm]{fig/spectec-loop}}
+    \caption[Enter the caption title here]{SpecTec \texttt{loop}} \label{fig:spectec-loop}
+\end{figure}
+\begin{figure}[h!]
+    \centerline{\includegraphics[width=15cm]{fig/spectec-br}}
+    \caption[Enter the caption title here]{SpecTec \texttt{br}} \label{fig:spectec-br}
+\end{figure}
 
-don't remember end!
+Consider example 1 above.
+\texttt{loop} and \texttt{unreachable} is the input instruction for the model.
+\cref{fig:spectec-loop} is the \spectecp{} of the \texttt{loop} instruction.
+Rather than storing the point to jump as a continuation in the label, it stores
+the \texttt{loop} instruction itself and \textit{enters} the block.
+Here, \textit{enter} means that it pushes the label in the stack and inputs the
+instructions as inputs to execute.
+As a result, after the \texttt{loop} is executed, \texttt{i32.const},
+\texttt{br}, \texttt{unreachable}, and \texttt{unreachable} are remained as
+inputs to the model.
+When \texttt{i32.const} is excuted, it pushes 42 as \officialp{} does.
+However, \texttt{br} behaves a bit differently.
+\cref{fig:spectec-br} is the \spectecp{} of the \texttt{br} instruction.
+It pops the label from the stack, removes the input instructions until the end
+of the block, considers the loop instruction in the label as a new input
+instruction, and executes it again.
+Consequently, remaining inputs are \texttt{loop}, and \texttt{unreachable}.
+Therefore, it explains the behavior in different point of view, but actually
+describes the same thing.
+
+% exiting label in SpecTec prose
+However, this viewpoint is challenged by the exiting label.
+As the block structure doesn't remain intact, it is hard to describe
+\textbf{the end of the block}.
+Furthermore, the exiting label is not performed by a specific WebAssembly
+instruction, which makes it hard to model the behavior.
+To handle this problem, an administrative instruction \texttt{end} is
+introduced in \spectecp.
+When entering a block, assume that there is an \texttt{end}
 
@@
 % AL: instruction sequence input continously
-% device exit semantics for AL
+% device exiting semantics for AL
 % Executable spec passes all wasm testcases

### 2024-10-28 paper/3-motivation.tex
Update motivation

@@
 As a result, the code example above is a infinite loop pushing \texttt{42}
-forever without executing \texttt{unreachable} instruction.
+forever without executing \texttt{unreachable} instructions.
 
@@
 \textit{exiting label}.
-\cref{fig:exiting-label} is the \officialp of the exiting label.
-It is special because this behavior is not performed without an explicit
+\cref{fig:exiting-label} is the \officialp{} of the \textit{exiting label}.
+It is special because this behavior is performed without an explicit
 WebAssembly instruction.
@@
 
-Example 2 is \texttt{loop} with a block that has only \texttt{i32.const}.
-A label is pushed in the stack when a \texttt{loop} instruction is executed.
+Example 2 is a WebAssembly code example that has \texttt{loop} whose block is
+only \texttt{i32.const}, and \texttt{f32.const}
+When the \texttt{loop} instruction is executed, a label is pushed in the stack,
+and it enters the block.
 After \texttt{i32.const} is executed, the end of the block is reached.
@@
 pc changes to the point after the end of block: \texttt{f64.const}.
-Therefore, the code example above results in the value 42 and 3.14 pushed to
-the stack.
+As a result, after executing the code, the value 42 and 3.14 pushed to the
+stack.
 
@@
 control instructions and structures such as a \texttt{call} instruction that
-calls a function with a frame and a \texttt{return} instruction that returns from
-the function with the frame.
+calls a function with pushing a frame to stack and a \texttt{return}
+instruction that returns from the function with popping the frame from the
+stack.
 Not only that, there are the special behaviors such as a \textit{returning from
@@
 of WebAssembly.
-Rather than using the notion of pc, \spectecp{} assumes that WebAssembly
+Rather than using the notion of a pc, \spectecp{} assumes that WebAssembly
 instructions are just given one by one.
@@
 Consider example 1 above.
+
 \texttt{loop} and \texttt{unreachable} is the input instruction for the model.

### 2024-10-28 paper/3-motivation.tex
Explain with rewrite rule

@@
 % control flow structure in SpecTec prose
-However, the \spectecp{} assumes different a model to explain the control flow
+However, the \spectecp{} assumes a different model to explain the control flow
 of WebAssembly.
@@
 This is because the \spectecp{} is generated automatically from the
-\red{SpecTec dsl}.
-\red{SpecTec dsl} uses rewrite rule to describe WebAssembly semantics, and it
-can be seen as consuming a WebAssembly instruction and values in the stack,
-changing the stack and the store, and inputing new WebAssembly instructions.
+\red{SpecTec DSL}.
+\red{SpecTec DSL} uses rewrite rule to describe WebAssembly semantics, and it
+can be seen as consuming a WebAssembly instruction, pushing or poping a control
+structures, and inputing new WebAssembly instructions.
+It can be expressed in the following form:
+$c^* \vdash i_0, i_1, ..., i_m \leadsto c'^* \vdash i_1, ..., i_m$
+
+% challenges in SpecTec prose
+However, this viewpoint is challenged by the exiting label.
+As the block structure doesn't remain intact in the model, it is hard to
+describe \textbf{the end of the block}.
+Furthermore, the exiting label is not performed by a specific WebAssembly
+instruction, which makes it hard to model the behavior.
+To handle this problem, an administrative instruction \texttt{end} is
+introduced in \spectecp.
+When entering a block, an \texttt{end} is added to express the end of the block.
 
@@
 
+\begin{align}
+  &c^*
+  \vdash
+  loop(i32, ~ [ ~ i32.const ~ 42, ~ br ~ 0, ~ unreachable ~ ]), ~ unreachable
+\leadsto \\
+  &label(loop(i32, ~ [ ~ i32.const ~ 42; ~ br ~ 0; ~ unreachable ~ ])), c^*
+  \vdash
+  i32.const ~ 42, ~ br ~ 0, ~ unreachable, ~ end, ~ unreachable
+\leadsto \\
+  &label(loop(i32, ~ [ ~ i32.const ~ 42; ~ br ~ 0; ~ unreachable ~ ])), c^*
+  \vdash
+  ~ br ~ 0, ~ unreachable, ~ end, ~ unreachable
+\leadsto \\
+  &c^*
+  \vdash
+  loop(i32, ~ [ ~ i32.const ~ 42; ~ br ~ 0; ~ unreachable ~ ]), unreachable \\
+\end{align}
+
 Consider example 1 above.
-
-\texttt{loop} and \texttt{unreachable} is the input instruction for the model.
+If we assume some control structure $c^*$ is given, we can express the code
+like (3.1).
 \cref{fig:spectec-loop} is the \spectecp{} of the \texttt{loop} instruction.
@@
 the \texttt{loop} instruction itself and \textit{enters} the block.
-Here, \textit{enter} means that it pushes the label in the stack and inputs the
-instructions as inputs to execute.
-As a result, after the \texttt{loop} is executed, \texttt{i32.const},
-\texttt{br}, \texttt{unreachable}, and \texttt{unreachable} are remained as
-inputs to the model.
-When \texttt{i32.const} is excuted, it pushes 42 as \officialp{} does.
+Here, \textit{enter} means that it pushes the label in the stack and executes
+the instructions in the block, which means that \texttt{i32.const},
+\texttt{br}, \texttt{unreachable}, \texttt{end} and \texttt{unreachable} become
+the new inputs (3.2).
+When \texttt{i32.const} is excuted, it pushes 42 as \officialp{} does (3.3).
+The value is omitted for brevity.
 However, \texttt{br} behaves a bit differently.
@@
 of the block, considers the loop instruction in the label as a new input
-instruction, and executes it again.
-Consequently, remaining inputs are \texttt{loop}, and \texttt{unreachable}.
-Therefore, it explains the behavior in different point of view, but actually
-describes the same thing.
-
-% exiting label in SpecTec prose
-However, this viewpoint is challenged by the exiting label.
-As the block structure doesn't remain intact, it is hard to describe
-\textbf{the end of the block}.
-Furthermore, the exiting label is not performed by a specific WebAssembly
-instruction, which makes it hard to model the behavior.
-To handle this problem, an administrative instruction \texttt{end} is
-introduced in \spectecp.
-When entering a block, assume that there is an \texttt{end}
+instruction.
+Consequently, remaining inputs are \texttt{loop}, and \texttt{unreachable}
+again.
+Therefore, it explains the same behavior in different point of view.
 

### 2024-10-28 paper/3-motivation.tex
Minor changes

@@
 \red{SpecTec DSL} uses rewrite rule to describe WebAssembly semantics, and it
-can be seen as consuming a WebAssembly instruction, pushing or poping a control
-structures, and inputing new WebAssembly instructions.
+is interpreted in \spectecp{} as consuming a WebAssembly instruction, pushing
+or poping a control structures, and inputing new WebAssembly instructions.
 It can be expressed in the following form:
-$c^* \vdash i_0, i_1, ..., i_m \leadsto c'^* \vdash i_1, ..., i_m$
+$c^* \vdash i_0, i_1, ..., i_m \leadsto c'^* \vdash i'_0, ...i'_{m'}, i_1, ..., i_m$
 
@@
   loop(i32, ~ [ ~ i32.const ~ 42, ~ br ~ 0, ~ unreachable ~ ]), ~ unreachable
+  \label{eq:loop-1}
 \leadsto \\
@@
   i32.const ~ 42, ~ br ~ 0, ~ unreachable, ~ end, ~ unreachable
+  \label{eq:loop-2}
 \leadsto \\
@@
   ~ br ~ 0, ~ unreachable, ~ end, ~ unreachable
+  \label{eq:loop-3}
 \leadsto \\
@@
   loop(i32, ~ [ ~ i32.const ~ 42; ~ br ~ 0; ~ unreachable ~ ]), unreachable \\
+  \label{eq:loop-4}
 \end{align}
@@
 If we assume some control structure $c^*$ is given, we can express the code
-like (3.1).
+like \cref{eq:loop-1}.
 \cref{fig:spectec-loop} is the \spectecp{} of the \texttt{loop} instruction.
@@
 \texttt{br}, \texttt{unreachable}, \texttt{end} and \texttt{unreachable} become
-the new inputs (3.2).
-When \texttt{i32.const} is excuted, it pushes 42 as \officialp{} does (3.3).
+the new inputs like \cref{eq:loop-2}.
+When \texttt{i32.const} is excuted, it pushes 42 as \officialp{} does
+like \cref{eq:loop-3}.
 The value is omitted for brevity.
@@
 Consequently, remaining inputs are \texttt{loop}, and \texttt{unreachable}
-again.
-Therefore, it explains the same behavior in different point of view.
+again like \cref{eq:loop-4}.
+Therefore, it explains the same behavior in a different point of view.
 

### 2024-10-29 paper/3-motivation.tex
Add subsection
Add official invoke&returning Add SpecTec invoking&returning

@@
 \red{TODO: remove redundant part in the fig}
+
+\section{Control flow in official prose}
 
@@
 
-% general control flow structure of official prose
-Similar to the control flow using the block and the label, there are many other
-control instructions and structures such as a \texttt{call} instruction that
-calls a function with pushing a frame to stack and a \texttt{return}
-instruction that returns from the function with popping the frame from the
-stack.
-Not only that, there are the special behaviors such as a \textit{returning from
-a function} that corresponds to the exiting label.
+% control flow structure with function call in official prose
+Similar to the control flow using the block and the label, there is a function
+call with a frame.
+\cref{fig:invoke} is the \officialp{} of the function invocation which takes
+place by a \texttt{call} instruction.
+A frame is pushed to the stack and enters the block of the function body with a
+label.
+There is also a special behavior named returning from a function.
+\cref{fig:returning} is the \officialp{} of the returning from a function.
+When end of function is reached without without control instructions or runtime
+error, The frame is popped from the stack, and the pc points to the next
+instruction of the caller instruction.
+
+\begin{figure}[h!]
+    \centerline{\includegraphics[width=15cm]{fig/invoke}}
+    \caption[Enter the caption title here]{function invocation} \label{fig:invoke}
+    \centerline{\includegraphics[width=15cm]{fig/returning}}
+    \caption[Enter the caption title here]{returning from a function} \label{fig:returning}
+\end{figure}
+
+\section{Control flow in SpecTec prose}
 
@@
 However, this viewpoint is challenged by the exiting label.
-As the block structure doesn't remain intact in the model, it is hard to
+As the block structure doesn't remain intact in this model, it is hard to
 describe \textbf{the end of the block}.
@@
 introduced in \spectecp.
-When entering a block, an \texttt{end} is added to express the end of the block.
+When entering a block, an \texttt{end} is appended to the end of the block.
 
@@
   loop(i32, ~ [ ~ i32.const ~ 42, ~ br ~ 0, ~ unreachable ~ ]), ~ unreachable
-  \label{eq:loop-1}
-\leadsto \\
-  &label(loop(i32, ~ [ ~ i32.const ~ 42; ~ br ~ 0; ~ unreachable ~ ])), c^*
+  \label{eq:loop-1} \\
+&\leadsto
+  label(loop(i32, ~ [ ~ i32.const ~ 42; ~ br ~ 0; ~ unreachable ~ ])), c^*
   \vdash
   i32.const ~ 42, ~ br ~ 0, ~ unreachable, ~ end, ~ unreachable
-  \label{eq:loop-2}
-\leadsto \\
-  &label(loop(i32, ~ [ ~ i32.const ~ 42; ~ br ~ 0; ~ unreachable ~ ])), c^*
+  \label{eq:loop-2} \\
+  &\leadsto
+  label(loop(i32, ~ [ ~ i32.const ~ 42; ~ br ~ 0; ~ unreachable ~ ])), c^*
   \vdash
   ~ br ~ 0, ~ unreachable, ~ end, ~ unreachable
-  \label{eq:loop-3}
-\leadsto \\
-  &c^*
+  \label{eq:loop-3} \\
+&\leadsto
+  c^*
   \vdash
-  loop(i32, ~ [ ~ i32.const ~ 42; ~ br ~ 0; ~ unreachable ~ ]), unreachable \\
+  loop(i32, ~ [ ~ i32.const ~ 42; ~ br ~ 0; ~ unreachable ~ ]), unreachable
   \label{eq:loop-4}
@@
 Consider example 1 above.
-If we assume some control structure $c^*$ is given, we can express the code
+If we assume some control structure $c^*$ is given, the code can be expressed
 like \cref{eq:loop-1}.
@@
 \texttt{br}, \texttt{unreachable}, \texttt{end} and \texttt{unreachable} become
-the new inputs like \cref{eq:loop-2}.
+the new inputs in \cref{eq:loop-2}.
 When \texttt{i32.const} is excuted, it pushes 42 as \officialp{} does
-like \cref{eq:loop-3}.
+in \cref{eq:loop-3}.
 The value is omitted for brevity.
@@
 Consequently, remaining inputs are \texttt{loop}, and \texttt{unreachable}
-again like \cref{eq:loop-4}.
+again in \cref{eq:loop-4}.
 Therefore, it explains the same behavior in a different point of view.
+
+% an example of function call in official prose
+\textbf{Example 3}
+\begin{verbatim}
+  // function definition as follow
+  (func $foo (result i32) (i32.const 42))
+
+  // function call
+  (call $foo) (f32.const 3.14)
+\end{verbatim}
+
+\begin{align}
+  &c^* \vdash call(\$foo), ~ f32.const ~ 3.14 \label{eq:call-1} \\
+&\leadsto
+  Label(i32, \epsilon), ~ Frame(), ~ c^* \vdash i32.const ~ 42, ~ end, ~ f32.const ~ 3.14 \label{eq:call-2} \\
+&\leadsto
+  Label(i32, \epsilon), ~ Frame(), ~ c^* \vdash end, ~ f32.const ~ 3.14 \label{eq:call-3} \\
+&\leadsto
+  c^* \vdash f32.const ~ 3.14 \label{eq:call-4} \\
+&\leadsto
+  c^* \label{eq:call-5}
+\end{align}
+
+Example 3 contains a function definition whose body is \texttt{i32.const}, a
+\texttt{call} instruction followed by a \texttt{f32.const} instruction, which can be expressed in \cref{eq:call-1}.
+By invoking a function, it enters the block in \cref{eq:call-2}.
+After executing the body in \cref{eq:call-3}, an \texttt{end} is executed.
+In this point, this \texttt{end} represent both the end of the block and the
+function.
+As a consequence, this \text{end} perform both exiting label and returning from
+a function in \cref{eq:call-4}, and the instruction after the \texttt{call} is
+executed in \cref{eq:call-5}
+It results in the values 42 and 3.14 pushed to the stack.
 

### 2024-10-29 paper/3-motivation.tex
Add problem

@@
 
-\red{TODO: remove redundant part in the fig}
+\red{TODO: remove redundant part in the fig} \\
+\red{TODO: explanation of context: in the background? or here?} \\
 
@@
 It can be expressed in the following form:
-$c^* \vdash i_0, i_1, ..., i_m \leadsto c'^* \vdash i'_0, ...i'_{m'}, i_1, ..., i_m$
+$c^* \vdash i_0, i_1, ..., i_m \leadsto c'^* \vdash i'_0, ...i'_{m'}, i_1, ..., i_m$.
+To simply model the control flow of WebAssembly, other things not closely
+related to the control flow are omitted here.
 
@@
   \vdash
-  loop(i32, ~ [ ~ i32.const ~ 42, ~ br ~ 0, ~ unreachable ~ ]), ~ unreachable
+  loop([ ~ i32.const ~ 42, ~ br ~ 0, ~ unreachable ~ ]), ~ unreachable
   \label{eq:loop-1} \\
 &\leadsto
-  label(loop(i32, ~ [ ~ i32.const ~ 42; ~ br ~ 0; ~ unreachable ~ ])), c^*
+  label(loop([ ~ i32.const ~ 42; ~ br ~ 0; ~ unreachable ~ ])), c^*
   \vdash
@@
   &\leadsto
-  label(loop(i32, ~ [ ~ i32.const ~ 42; ~ br ~ 0; ~ unreachable ~ ])), c^*
+  label(loop([ ~ i32.const ~ 42; ~ br ~ 0; ~ unreachable ~ ])), c^*
   \vdash
@@
   \vdash
-  loop(i32, ~ [ ~ i32.const ~ 42; ~ br ~ 0; ~ unreachable ~ ]), unreachable
+  loop([ ~ i32.const ~ 42; ~ br ~ 0; ~ unreachable ~ ]), unreachable
   \label{eq:loop-4}
@@
 
+\red{TODO: Also use example 3 to explain function call in official prose}
+
 % an example of function call in official prose
@@
 &\leadsto
-  Label(i32, \epsilon), ~ Frame(), ~ c^* \vdash i32.const ~ 42, ~ end, ~ f32.const ~ 3.14 \label{eq:call-2} \\
+  Label(\epsilon), ~ Frame(), ~ c^* \vdash i32.const ~ 42, ~ end, ~ f32.const ~ 3.14 \label{eq:call-2} \\
 &\leadsto
-  Label(i32, \epsilon), ~ Frame(), ~ c^* \vdash end, ~ f32.const ~ 3.14 \label{eq:call-3} \\
+  Label(\epsilon), ~ Frame(), ~ c^* \vdash end, ~ f32.const ~ 3.14 \label{eq:call-3} \\
 &\leadsto
@@
 
+In short, the behavior of an \texttt{end} instruction 1) performs exiting
+label, 2) looks up the top of the context, and 3) performs returing from a
+function, if it is a frame.
+With this \texttt{end} instruction, it seems that \spectecp{} can model the
+WebAssembly control flow correctly.
+However, the \red{interpreter} implemented with this model doesn't work well.
+
 % official spec prose notation control structure

### 2024-10-29 paper/3-motivation.tex
Update problem

@@
 \red{TODO: remove redundant part in the fig} \\
-\red{TODO: explanation of context: in the background? or here?} \\
+\red{TODO: explanation of context(frame/label): in the background? or here?} \\
+\red{TODO: the notion of meta-level interpreter in background} \\
 
@@
 $c^* \vdash i_0, i_1, ..., i_m \leadsto c'^* \vdash i'_0, ...i'_{m'}, i_1, ..., i_m$.
+It means that given a sequence of contexts $c^*$, executing $i_0$ results in a
+new sequence of contexts $c'^*$ and instructions $i'_0, ...i'_{m'}$.
 To simply model the control flow of WebAssembly, other things not closely
-related to the control flow are omitted here.
+related to the control flow are omitted in this notation.
 
@@
 Example 3 contains a function definition whose body is \texttt{i32.const}, a
-\texttt{call} instruction followed by a \texttt{f32.const} instruction, which can be expressed in \cref{eq:call-1}.
+\texttt{call} instruction followed by a \texttt{f32.const} instruction, which
+can be expressed in \cref{eq:call-1}.
 By invoking a function, it enters the block in \cref{eq:call-2}.
@@
 
+% summary & test
 In short, the behavior of an \texttt{end} instruction 1) performs exiting
@@
 function, if it is a frame.
-With this \texttt{end} instruction, it seems that \spectecp{} can model the
-WebAssembly control flow correctly.
-However, the \red{interpreter} implemented with this model doesn't work well.
+The \spectecp{} can model the WebAssembly control flow using this \texttt{end}
+instruction.
+To show the correctness of the \spectecp{}, we run the WebAssembly tests
+including official WebAssembly test suite by executing the \spectecp{} with
+\red{definitional interpreter}.
+
+% Problem
+However, tests related to the control instruction are failed.
+At that time, there was no explicit model to describe the control flow in
+\spectecp{}, it was hard to define the problem, making it complicated to fix
+the problem.
 

### 2024-10-29 paper/3-motivation.tex
Add todo
Change function name

@@
 
+\red{TODO: explanation of the terms \officialp{} and \spectecp{}} \\
 \red{TODO: remove redundant part in the fig} \\
@@
 \red{TODO: the notion of meta-level interpreter in background} \\
+\red{TODO: try finding better terminology rather than "model"} \\
 
@@
 \begin{verbatim}
-  // function definition as follow
-  (func $foo (result i32) (i32.const 42))
+  // function definition of $push-42
+  (func $push-42 (result i32) (i32.const 42))
 
   // function call
-  (call $foo) (f32.const 3.14)
+  (call $push-42) (f32.const 3.14)
 \end{verbatim}

### 2024-10-29 paper/3-motivation.tex
Minor fix

@@
 \begin{verbatim}
-  // function definition of $push-42
-  (func $push-42 (result i32) (i32.const 42))
+  // function definition of $push42
+  (func $push42 (result i32) (i32.const 42))
 
   // function call
-  (call $push-42) (f32.const 3.14)
+  (call $push42) (f32.const 3.14)
 \end{verbatim}
@@
 \begin{align}
-  &c^* \vdash call(\$foo), ~ f32.const ~ 3.14 \label{eq:call-1} \\
+  &c^* \vdash call(\$push42), ~ f32.const ~ 3.14 \label{eq:call-1} \\
 &\leadsto

### 2024-10-29 paper/3-motivation.tex
Fix explanation related to enter

@@
 only \texttt{i32.const}, and \texttt{f32.const}
-When the \texttt{loop} instruction is executed, a label is pushed in the stack,
-and it enters the block.
+When the \texttt{loop} instruction is executed, it enters the block with a label.
 After \texttt{i32.const} is executed, the end of the block is reached.
@@
 Rather than storing the point to jump as a continuation in the label, it stores
-the \texttt{loop} instruction itself and \textit{enters} the block.
+the \texttt{loop} instruction itself in the label, and \textit{enters} the
+block.
 Here, \textit{enter} means that it pushes the label in the stack and executes
@@
 can be expressed in \cref{eq:call-1}.
-By invoking a function, it enters the block in \cref{eq:call-2}.
+By invoking a function, it pushes a frame and enters the block with a label in
+\cref{eq:call-2}.
 After executing the body in \cref{eq:call-3}, an \texttt{end} is executed.

### 2024-10-29 paper/3-motivation.tex
Minor fix

@@
 &\leadsto
-  Label(\epsilon), ~ Frame(), ~ c^* \vdash i32.const ~ 42, ~ end, ~ f32.const ~ 3.14 \label{eq:call-2} \\
+  Label(\epsilon), ~ Frame, ~ c^* \vdash i32.const ~ 42, ~ end, ~ f32.const ~ 3.14 \label{eq:call-2} \\
 &\leadsto
-  Label(\epsilon), ~ Frame(), ~ c^* \vdash end, ~ f32.const ~ 3.14 \label{eq:call-3} \\
+  Label(\epsilon), ~ Frame, ~ c^* \vdash end, ~ f32.const ~ 3.14 \label{eq:call-3} \\
 &\leadsto
@@
 &\leadsto
-  c^* \label{eq:call-5}
+  c^* \vdash \epsilon \label{eq:call-5}
 \end{align}

### 2024-10-29 paper/3-motivation.tex
Elaborate end instruction

@@
 It pops the label from the stack, removes the input instructions until the end
-of the block, considers the loop instruction in the label as a new input
-instruction.
+of the block including the \texttt{end}, considers the loop instruction in the
+label as a new input instruction.
 Consequently, remaining inputs are \texttt{loop}, and \texttt{unreachable}
@@
 the problem.
-
-% official spec prose notation control structure
-% seems to be pc-based
-% AL from DSL ~ formal notation: no pc
-% AL: instruction sequence input continously
-% device exiting semantics for AL
-% Executable spec passes all wasm testcases
-% weird, maybe not happened in realworld, but valid wasm code
-% wrong!

### 2024-10-29 paper/5-eval.tex
Write main.tex
WIP: Formalize AL semantics
Add a buggy example & explanation Step by step reduction of the example

--- 2024-10-07 paper/2-background.tex
@@
 
-\chapter{Background}
-\label{ch:background}
+\chapter{Evaluation}
+\label{ch:eval}
 \noindent
 
-background
+% an example of buggy test case
+\textbf{Example 5.1}
+\begin{verbatim}
+  // function definition of $br-returning
+  (func  $brAndReturning (br 0) (unreachable))
+
+  // function call
+  (call $brAndReturning)
+\end{verbatim}
+
+Example 5.1 is the minimul example that produces a wrong result in \spectecp{}.
+It contains a function that has only a \texttt{br} instruction.
+According to the \officialp{}, when the function \texttt{\$brAndReturning} is
+called, a frame is pushed to the stack, and the block of the function
+body is entered with a label.
+When the \texttt{br} is executed, it pops the label from the stack.
+As the end of the function is reached after the execution, returning from a
+function is performed.
+As a result, the frame is popped, so the call actually does nothing.
+
+% explaination of the bug in SpecTec prose
+\begin{align}
+  &c^* \vdash call(\$brAndReturning) \label{eq:bug-1} \\
+&\leadsto
+  Label(\epsilon), ~ Frame, ~ c^* \vdash (br ~ 0) ~ (unreachable), ~ end \label{eq:bug-2} \\
+&\leadsto
+  Frame, ~ c^* \vdash \epsilon \label{eq:bug-3}
+\end{align}
+
+In \spectecp{} the call instruction in \cref{eq:bug-1} pushes a frame and
+enters the function body \cref{eq:bug-2}.
+When the \texttt{br} is executed, it pops the label and removes the input
+instructions until the end in \cref{eq:bug-3}.
+As a result, the frame is remained after the executing the call.
+
+% simplified version of AL algorithm of call, br, and returning from a function
+\begin{verbatim}
+Call fname:
+  PushCtx Frame
+  Enter get_function_body(fname) with Label(\epsilon)
+
+Br 0:
+  PopCtx Label(instr*)
+  ExecuteSeq instr*
+
+Returning from a function:
+  PopCtx Frame
+\end{verbatim}
+
+\red{TODO: Change to array} \\
+\red{TODO: Change naming for $c$}
+\begin{align}
+  &c^* \vdash call(\$brAndReturning); ~ Enter ~ (n, c)
+  \label{eq:resolve-1} \\
+  &\leadsto c^* \vdash \epsilon; ~ Al ~ ([ ~ \pushctxi, \enteri ~ ], 0, ~ Enter ~ (n, c))
+  \label{eq:resolve-2} \\
+  &\leadsto Frame, ~ c^* \vdash \epsilon; ~ Al ~ ([ ~ \enteri ~ ], 1, ~ Enter ~ (n, c))
+  \label{eq:resolve-3} \\
+  &\leadsto Label(\epsilon), ~ Frame, ~ c^* \vdash (br ~ 0), ~ (unreachable), ~ (end ~ 2); ~
+    Enter ~ (2, Al ~ (\epsilon, 0, ~ Enter ~ (n, c)))
+  \label{eq:resolve-4} \\
+  &\leadsto Label(\epsilon), ~ Frame, ~ c^* \vdash (unreachable), ~ (end ~ 2); \\
+  &\quad\quad Al ~ ([ ~ \red{\symbol{PopCtxI}}, ~ \executeseqi ~ ], ~ 0, ~ Enter ~ (2, Al ~ (\epsilon, 0, ~ Enter ~ (n, c))))
+  \label{eq:resolve-5} \\
+  &\leadsto Frame, ~ c^* \vdash (end ~ 1); ~
+    Al ~ ([ ~ \executeseqi ~ ], ~ 0, ~ Enter ~ (1, Al ~ (\epsilon, 0, ~ Enter ~ (n, c))))
+  \label{eq:resolve-6} \\
+  &\leadsto Frame, ~ c^* \vdash (end ~ 1); ~
+    Al ~ (\epsilon, ~ 0, ~ Enter ~ (1, Al ~ (\epsilon, 0, ~ Enter ~ (n, c))))
+  \label{eq:resolve-7} \\
+  &\leadsto Frame, ~ c^* \vdash (end ~ 1); ~ Enter ~ (1, Al ~ (\epsilon, 0, ~ Enter ~ (n, c)))
+  \label{eq:resolve-8} \\
+  &\leadsto Frame, ~ c^* \vdash \epsilon; ~
+    Al ~ ([ ~ \red{\symbol{PopCtx}} ~ ], ~ 0, ~ Enter ~ (0, Al ~ (\epsilon, 0, ~ Enter ~ (n, c))))
+  \label{eq:resolve-9} \\
+  &\leadsto c^* \vdash \epsilon; ~
+    Al ~ (\epsilon, ~ 0, ~ Enter ~ (0, Al ~ (\epsilon, 0, ~ Enter ~ (n, c))))
+  \label{eq:resolve-10} \\
+  &\leadsto c^* \vdash \epsilon; ~ Enter ~ (0, Al ~ (\epsilon, 0, ~ Enter ~ (n, c)))
+  \label{eq:resolve-11} \\
+  &\leadsto c^* \vdash \epsilon; ~ Al ~ (\epsilon, 0, ~ Enter ~ (n, c))
+  \label{eq:resolve-12} \\
+  &\leadsto c^* \vdash \epsilon; ~ Enter ~ (n, c)
+  \label{eq:resolve-13}
+\end{align}

### 2024-10-29 paper/5-eval.tex
Add space

@@
 \begin{align}
-  &c^* \vdash call(\$brAndReturning); ~ Enter ~ (n, c)
+  &c^* \vdash call(\$brAndReturning); ~ \wasm ~ (n, ~ c)
   \label{eq:resolve-1} \\
-  &\leadsto c^* \vdash \epsilon; ~ Al ~ ([ ~ \pushctxi, \enteri ~ ], 0, ~ Enter ~ (n, c))
+  &\leadsto c^* \vdash \epsilon; ~ \al ~ ([ ~ \pushctxi, ~ \enteri ~ ], ~ 0, ~ \wasm ~ (n, ~ c))
   \label{eq:resolve-2} \\
-  &\leadsto Frame, ~ c^* \vdash \epsilon; ~ Al ~ ([ ~ \enteri ~ ], 1, ~ Enter ~ (n, c))
+  &\leadsto Frame, ~ c^* \vdash \epsilon; ~ \al ~ ([ ~ \enteri ~ ], ~ 1, ~ \wasm ~ (n, ~ c))
   \label{eq:resolve-3} \\
   &\leadsto Label(\epsilon), ~ Frame, ~ c^* \vdash (br ~ 0), ~ (unreachable), ~ (end ~ 2); ~
-    Enter ~ (2, Al ~ (\epsilon, 0, ~ Enter ~ (n, c)))
+    \wasm ~ (2, ~ \al ~ (\epsilon, ~ 0, ~ \wasm ~ (n, ~ c)))
   \label{eq:resolve-4} \\
   &\leadsto Label(\epsilon), ~ Frame, ~ c^* \vdash (unreachable), ~ (end ~ 2); \\
-  &\quad\quad Al ~ ([ ~ \red{\symbol{PopCtxI}}, ~ \executeseqi ~ ], ~ 0, ~ Enter ~ (2, Al ~ (\epsilon, 0, ~ Enter ~ (n, c))))
+  &\quad\quad \al ~ ([ ~ \red{\symbol{PopCtxI}}, ~ \executeseqi ~ ], ~ 0, ~ \wasm ~ (2, ~ \al ~ (\epsilon, ~ 0, ~ \wasm ~ (n, ~ c))))
   \label{eq:resolve-5} \\
   &\leadsto Frame, ~ c^* \vdash (end ~ 1); ~
-    Al ~ ([ ~ \executeseqi ~ ], ~ 0, ~ Enter ~ (1, Al ~ (\epsilon, 0, ~ Enter ~ (n, c))))
+    \al ~ ([ ~ \executeseqi ~ ], ~ 0, ~ \wasm ~ (1, ~ \al ~ (\epsilon, ~ 0, ~ \wasm ~ (n, ~ c))))
   \label{eq:resolve-6} \\
   &\leadsto Frame, ~ c^* \vdash (end ~ 1); ~
-    Al ~ (\epsilon, ~ 0, ~ Enter ~ (1, Al ~ (\epsilon, 0, ~ Enter ~ (n, c))))
+    \al ~ (\epsilon, ~ 0, ~ \wasm ~ (1, ~ \al ~ (\epsilon, ~ 0, ~ \wasm ~ (n, ~ c))))
   \label{eq:resolve-7} \\
-  &\leadsto Frame, ~ c^* \vdash (end ~ 1); ~ Enter ~ (1, Al ~ (\epsilon, 0, ~ Enter ~ (n, c)))
+  &\leadsto Frame, ~ c^* \vdash (end ~ 1); ~ \wasm ~ (1, ~ \al ~ (\epsilon, ~ 0, ~ \wasm ~ (n, ~ c)))
   \label{eq:resolve-8} \\
   &\leadsto Frame, ~ c^* \vdash \epsilon; ~
-    Al ~ ([ ~ \red{\symbol{PopCtx}} ~ ], ~ 0, ~ Enter ~ (0, Al ~ (\epsilon, 0, ~ Enter ~ (n, c))))
+    \al ~ ([ ~ \red{\symbol{PopCtx}} ~ ], ~ 0, ~ \wasm ~ (0, ~ \al ~ (\epsilon, ~ 0, ~ \wasm ~ (n, ~ c))))
   \label{eq:resolve-9} \\
   &\leadsto c^* \vdash \epsilon; ~
-    Al ~ (\epsilon, ~ 0, ~ Enter ~ (0, Al ~ (\epsilon, 0, ~ Enter ~ (n, c))))
+    \al ~ (\epsilon, ~ 0, ~ \wasm ~ (0, ~ \al ~ (\epsilon, ~ 0, ~ \wasm ~ (n, ~ c))))
   \label{eq:resolve-10} \\
-  &\leadsto c^* \vdash \epsilon; ~ Enter ~ (0, Al ~ (\epsilon, 0, ~ Enter ~ (n, c)))
+  &\leadsto c^* \vdash \epsilon; ~ \wasm ~ (0, ~ \al ~ (\epsilon, ~ 0, ~ \wasm ~ (n, ~ c)))
   \label{eq:resolve-11} \\
-  &\leadsto c^* \vdash \epsilon; ~ Al ~ (\epsilon, 0, ~ Enter ~ (n, c))
+  &\leadsto c^* \vdash \epsilon; ~ \al ~ (\epsilon, ~ 0, ~ \wasm ~ (n, ~ c))
   \label{eq:resolve-12} \\
-  &\leadsto c^* \vdash \epsilon; ~ Enter ~ (n, c)
+  &\leadsto c^* \vdash \epsilon; ~ \wasm ~ (n, ~ c)
   \label{eq:resolve-13}

### 2024-10-29 paper/5-eval.tex
Minor fix
Change notation

@@
 
-Example 5.1 is the minimul example that produces a wrong result in \spectecp{}.
-It contains a function that has only a \texttt{br} instruction.
+Example 5.1 contains a function that has only a \texttt{br} instruction, and a
+\texttt{call} insstruction.
 According to the \officialp{}, when the function \texttt{\$brAndReturning} is
@@
 As a result, the frame is remained after the executing the call.
+The problem is that the \texttt{end} express both the end of the block and the
+function,
 
@@
 
+The code above is simplified version of \spectecp{}.
+According to the semantics defined in
+
 \red{TODO: Change to array} \\
@@
   \label{eq:resolve-1} \\
-  &\leadsto c^* \vdash \epsilon; ~ \al ~ ([ ~ \pushctxi, ~ \enteri ~ ], ~ 0, ~ \wasm ~ (n, ~ c))
+  &\leadsto c^* \vdash \epsilon; ~ \al ~ ([ ~ \pushctxi, ~ \enteri ~ ], ~ \epsilon, ~ \wasm ~ (n, ~ c))
   \label{eq:resolve-2} \\
-  &\leadsto Frame, ~ c^* \vdash \epsilon; ~ \al ~ ([ ~ \enteri ~ ], ~ 1, ~ \wasm ~ (n, ~ c))
+  &\leadsto Frame, ~ c^* \vdash \epsilon; ~ \al ~ ([ ~ \enteri ~ ], ~ [ ~ Frame ~ ], ~ \wasm ~ (n, ~ c))
   \label{eq:resolve-3} \\
-  &\leadsto Label(\epsilon), ~ Frame, ~ c^* \vdash (br ~ 0), ~ (unreachable), ~ (end ~ 2); ~
-    \wasm ~ (2, ~ \al ~ (\epsilon, ~ 0, ~ \wasm ~ (n, ~ c)))
+  &\leadsto Label(\epsilon), ~ Frame, ~ c^* \vdash (br ~ \epsilon), ~ (unreachable), ~ (exiting), ~ (returning); \\
+  &\quad\quad \wasm ~ ([ ~ Label, ~ Frame ~ ], ~ \al ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (n, ~ c)))
   \label{eq:resolve-4} \\
-  &\leadsto Label(\epsilon), ~ Frame, ~ c^* \vdash (unreachable), ~ (end ~ 2); \\
-  &\quad\quad \al ~ ([ ~ \red{\symbol{PopCtxI}}, ~ \executeseqi ~ ], ~ 0, ~ \wasm ~ (2, ~ \al ~ (\epsilon, ~ 0, ~ \wasm ~ (n, ~ c))))
+  &\leadsto Label(\epsilon), ~ Frame, ~ c^* \vdash (unreachable), ~ (exiting), ~ (returning); \\
+  &\quad\quad \al ~ ([ ~ \red{\symbol{PopCtxI}}, ~ \executeseqi ~ ], ~ \epsilon, ~ \wasm ~ ([ ~ Label, ~ Frame ~ ], ~ \al ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (n, ~ c))))
   \label{eq:resolve-5} \\
-  &\leadsto Frame, ~ c^* \vdash (end ~ 1); ~
-    \al ~ ([ ~ \executeseqi ~ ], ~ 0, ~ \wasm ~ (1, ~ \al ~ (\epsilon, ~ 0, ~ \wasm ~ (n, ~ c))))
+  &\leadsto Frame, ~ c^* \vdash (returning); ~
+    \al ~ ([ ~ \executeseqi ~ ], ~ \epsilon, ~ \wasm ~ ([ ~ Frame ~ ], ~ \al ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (n, ~ c))))
   \label{eq:resolve-6} \\
-  &\leadsto Frame, ~ c^* \vdash (end ~ 1); ~
-    \al ~ (\epsilon, ~ 0, ~ \wasm ~ (1, ~ \al ~ (\epsilon, ~ 0, ~ \wasm ~ (n, ~ c))))
+  &\leadsto Frame, ~ c^* \vdash (returning); ~
+    \al ~ (\epsilon, ~ \epsilon, ~ \wasm ~ ([ ~ Frame ~ ], ~ \al ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (n, ~ c))))
   \label{eq:resolve-7} \\
-  &\leadsto Frame, ~ c^* \vdash (end ~ 1); ~ \wasm ~ (1, ~ \al ~ (\epsilon, ~ 0, ~ \wasm ~ (n, ~ c)))
+  &\leadsto Frame, ~ c^* \vdash (returning); ~ \wasm ~ ([ ~ Frame ~ ], ~ \al ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (n, ~ c)))
   \label{eq:resolve-8} \\
   &\leadsto Frame, ~ c^* \vdash \epsilon; ~
-    \al ~ ([ ~ \red{\symbol{PopCtx}} ~ ], ~ 0, ~ \wasm ~ (0, ~ \al ~ (\epsilon, ~ 0, ~ \wasm ~ (n, ~ c))))
+    \al ~ ([ ~ \red{\symbol{PopCtxI}} ~ ], ~ \epsilon, ~ \wasm ~ (\epsilon, ~ \al ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (n, ~ c))))
   \label{eq:resolve-9} \\
   &\leadsto c^* \vdash \epsilon; ~
-    \al ~ (\epsilon, ~ 0, ~ \wasm ~ (0, ~ \al ~ (\epsilon, ~ 0, ~ \wasm ~ (n, ~ c))))
+    \al ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (\epsilon, ~ \al ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (n, ~ c))))
   \label{eq:resolve-10} \\
-  &\leadsto c^* \vdash \epsilon; ~ \wasm ~ (0, ~ \al ~ (\epsilon, ~ 0, ~ \wasm ~ (n, ~ c)))
+  &\leadsto c^* \vdash \epsilon; ~ \wasm ~ (\epsilon, ~ \al ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (n, ~ c)))
   \label{eq:resolve-11} \\
-  &\leadsto c^* \vdash \epsilon; ~ \al ~ (\epsilon, ~ 0, ~ \wasm ~ (n, ~ c))
+  &\leadsto c^* \vdash \epsilon; ~ \al ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (n, ~ c))
   \label{eq:resolve-12} \\

### 2024-10-30 paper/5-eval.tex
Minor fix

@@
 As a result, the frame is remained after the executing the call.
-The problem is that the \texttt{end} express both the end of the block and the
-function,
+The problem is that the \texttt{br} should remove the end of the block but
+should not remove the end of the function, but \texttt{end} express the end of
+the block and the function at the same time.
+
+% Fix
+To fix the problem, semantics of \enteri is changed.
+The sequence of contexts that is pushed in the AL algorithm is stored,
+and when a block is entered, exiting label and returning from a function is
+appended according to the sequence.
 
@@
 
-The code above is simplified version of \spectecp{}.
-According to the semantics defined in
+The code above is simplified version of \spectecp{} of \texttt{call},
+\texttt{br}, and \texttt{returning from a function}.
+By using the semantics defined in \cref{ch:semantics}, the example 5.1 results in the same result with the \officialp{}; $call(\$brAndReturning)$ does nothing.
 

### 2024-10-30 paper/2-background.tex
WIP: Formalize AL semantics
WIP: Add Wasm background

@@
 
-background
+\section{WebAssembly}
+
+% breif overview
+WebAssembly is a low-level bytecode language that is safe, fast, portable, and
+compact~\cite{wasm}.
+It is widely used as a compile target for web applications.
+Furthermore, many areas other than Web such as edge computing~\cite{wasm-edge},
+IoT~\cite{wasm-iot}, and block chains~\cite{wasm-block} deploy the advantages
+of WebAssembly.
+
+% risk of implementation divergence
+There are dozens of WebAssembly engines; all the browsers have there own
+implementations of WebAssembly with multiple tiers, and there are many engines
+that target for specific areas including embedded systems and edge computing.
+As a consequence, implementations of WebAssembly could diverge.
+
+% rigorous standardization -> formal notation & prose notation
+To handle this problem, WebAssembly has been standardized very rigourously by
+the W3C~\cite{wasm-w3c}.
+One of the requirements to standardize a feature is to describe the semantics
+of WebAssembly in two forms: formal notation and prose notation.
+Formal notation uses mathematical rules to compactly describe the semantics,
+and it is used for proofs such as type soundness.
+On the other hand, prose notation uses psudocodes to explain the semantics
+step by step.
+People such as WebAssembly engine developers who are not familiar with
+mathematical rules utilize it.
+
+% structure: module, function, ...
+
+% execution: instantiation, invocation, ...
+
+% control structure: label frame block ...
+
+\section{SpecTec}

### 2024-10-30 paper/2-background.tex
Minor fix

@@
 It is widely used as a compile target for web applications.
-Furthermore, many areas other than Web such as edge computing~\cite{wasm-edge},
+Not only that, other areas such as edge computing~\cite{wasm-edge},
 IoT~\cite{wasm-iot}, and block chains~\cite{wasm-block} deploy the advantages
@@
 that target for specific areas including embedded systems and edge computing.
-As a consequence, implementations of WebAssembly could diverge.
+However, WebAssembly should be portable across these implementations, it risks
+the implementations divergence.
 
 % rigorous standardization -> formal notation & prose notation
-To handle this problem, WebAssembly has been standardized very rigourously by
+To mitigate this problem, WebAssembly has been standardized very rigourously by
 the W3C~\cite{wasm-w3c}.

### 2024-10-31 paper/2-background.tex
Add Wasm structure & execution

@@
 
-% rigorous standardization -> formal notation & prose notation
+% rigorous standardization -> spec is rigorous: formal notation & prose notation
 To mitigate this problem, WebAssembly has been standardized very rigourously by
 the W3C~\cite{wasm-w3c}.
-One of the requirements to standardize a feature is to describe the semantics
-of WebAssembly in two forms: formal notation and prose notation.
+Especially, WebAssembly specification is written very rigorously.
+It describes the semantics of WebAssembly in two forms: formal notation and
+prose notation.
 Formal notation uses mathematical rules to compactly describe the semantics,
 and it is used for proofs such as type soundness.
-On the other hand, prose notation uses psudocodes to explain the semantics
-step by step.
+On the other hand, prose notation uses psudocodes to explain the semantics step
+by step.
 People such as WebAssembly engine developers who are not familiar with
@@
 
-% structure: module, function, ...
+% structure
+A WebAssembly program is sequence of \textit{modules}.
+A module is the unit of deployment, loading, and compilation, which consists of
+definitions such as functions, globals, and memories.
+A module is instantiated to validate the module and allocate each definition to
+the data structure named \textit{store}.
+After the instantiation, functions can be invoked for the execution.
 
-% execution: instantiation, invocation, ...
+\begin{figure}[t]
+    \centerline{\includegraphics[width=15cm]{fig/testop}}
+    \caption[Enter the caption title here]{\texttt{testop} instruction} \label{fig:testop}
+\end{figure}
 
-% control structure: label frame block ...
+% execution
+WebAssembly execution is based on a stack machine.
+An instruction consumes values in the stack as operands, produces the
+resulting values, and pushes them to the stack.
+\cref{fig:testop} is the WebAssembly specification for the \texttt{testop}
+instruction.
+Prose notation is written in the upper part, and formal notation is written in
+the grey box.
+In the prose notation, it pops a value from the stack, performs a test
+operation, and pushes the result to the stack.
+It is written compactly in the form of reduction rule in the formal notation
+which consists of LHS, RHS, and premises: $(t.const ~ c_1) ~ t.testop$,
+$(i32.const ~ c)$, and $c = testop_t(c_1)$.
+The LHS means that when the value at the top of the stack $(t.const ~ c_1)$
+and an instruction to execute is $t.testop$.
+The RHS means that the value at the top of the stack is changed to $(i32.const
+~ c)$, and this happens when the premise $c = testop_t(c_1)$ is satisfied.
+
+% control flow
 

### 2024-10-31 paper/2-background.tex
Minor fix

@@
 that target for specific areas including embedded systems and edge computing.
-However, WebAssembly should be portable across these implementations, it risks
+However, as WebAssembly should be portable across these implementations, it risks
 the implementations divergence.
@@
 by step.
-People such as WebAssembly engine developers who are not familiar with
-mathematical rules utilize it.
+Most WebAssembly users and engine developers are not familiar with
+mathematical rules, so they utilize the prose notation.
 
 % structure
-A WebAssembly program is sequence of \textit{modules}.
+A WebAssembly program is comprised of \textit{modules}.
 A module is the unit of deployment, loading, and compilation, which consists of
@@
 Prose notation is written in the upper part, and formal notation is written in
-the grey box.
+the lower grey box.
 In the prose notation, it pops a value from the stack, performs a test

### 2024-10-31 paper/2-background.tex
Add control structure

@@
 definitions such as functions, globals, and memories.
-A module is instantiated to validate the module and allocate each definition to
+A module is instantiated to validate the module, allocate each definition to
 the data structure named \textit{store}.
+This results in module instance, a runtime representation for the module.
 After the instantiation, functions can be invoked for the execution.
@@
 $(i32.const ~ c)$, and $c = testop_t(c_1)$.
-The LHS means that when the value at the top of the stack $(t.const ~ c_1)$
+The LHS means that the value at the top of the stack is $(t.const ~ c_1)$,
 and an instruction to execute is $t.testop$.
@@
 
-% control flow
+% control structure
+The stack stores not only the input and output values of the instructions but also
+structures related to the control flow.
+A \textit{call frame} and a \textit{label} are used for function call and
+branching, respectively.
+In exception handling proposal, there is a \textit{handler} for exception
+handling, and other strucutres could be introduced in the future.
+How these control structures are used for the control flow will be discussed in
+\cref{sec:control flow in official prose}. \\
+\red{TODO: remove dangling section header}
 

### 2024-11-02 paper/2-background.tex
Minor fix
spectec background version 1

@@
 the W3C~\cite{wasm-w3c}.
-Especially, WebAssembly specification is written very rigorously.
+Especially, WebAssembly specification document is written very rigorously.
 It describes the semantics of WebAssembly in two forms: formal notation and
@@
 and it is used for proofs such as type soundness.
-On the other hand, prose notation uses psudocodes to explain the semantics step
-by step.
+On the other hand, prose notation uses psudocodes to explain the semantics
+through a step by step instruction.
 Most WebAssembly users and engine developers are not familiar with
@@
 How these control structures are used for the control flow will be discussed in
-\cref{sec:control flow in official prose}. \\
-\red{TODO: remove dangling section header}
+\cref{sec:control flow in official prose}.
 
 \section{SpecTec}
+
+% SpecTec
+SpecTec is a tool for \textit{mechanizing} WebAssembly specification.
+Specification mechanization is a technique that treats the specification as
+data that can be manipulated by a computer.
+SpecTec provides a domain specific language(DSL) for specification, and it
+automatically generates a specification documents including the semantics
+described in formal notation and prose noatation from the DSL.
+The DSL is a declarative language that resembles textbook style notation, and
+it is type checked to prevent meta-level errors such as notation misuses and
+dimension mismatches.
+
+% prose language
+However, The DSL cannot generate the prose notation just as it is because it is
+declarative.
+Therefore, SpecTec also defines another imperative language for the prose
+notation, and translates the DSL into the prose language to generate the prose
+notation.
+
+% interpreter
+What is especially intriguing here is that the prose language is executable.
+As a result, the prose program generated from the DSL serves as a WebAssembly
+interpter.
+That is to say, if WebAssembly semantics is written in the DSL, SpecTec
+automatically generates the WebAssembly interpreter following the written
+semantics.
+The generated interpreter passes all the execution tests in the official test
+suites except for the tests related to the infinte loop.
+
+% other backends
+Furthermore, SpecTec automatically translates the DSL into Coq\cite{il2coq} for
+mathematical proofs and it is working on generating other artifacts such as
+test suites required to the WebAssembly standization.

### 2024-11-02 paper/2-background.tex
Add explanation for function body

@@
 This results in module instance, a runtime representation for the module.
-After the instantiation, functions can be invoked for the execution.
+After the instantiation, functions can be invoked so that the function body,
+which is a sequence of instructions, is executed.
 
@@
 WebAssembly execution is based on a stack machine.
-An instruction consumes values in the stack as operands, produces the
-resulting values, and pushes them to the stack.
+An instruction consumes values in the stack as operands, performs some
+operation, produces resulting values, and pushes them to the stack.
 \cref{fig:testop} is the WebAssembly specification for the \texttt{testop}

### 2024-11-02 paper/2-background.tex
spectec background version 2

@@
 
+\red{TODO: fix fig name}
+
 \section{WebAssembly}
+\label{sec:webassembly}
 
@@
     \caption[Enter the caption title here]{\texttt{testop} instruction} \label{fig:testop}
+    \centerline{\includegraphics[width=15cm]{fig/spectec-testop}}
+    \caption[Enter the caption title here]{SpecTec \texttt{testop} instruction} \label{fig:spectec-testop}
 \end{figure}
@@
 operation, produces resulting values, and pushes them to the stack.
-\cref{fig:testop} is the WebAssembly specification for the \texttt{testop}
-instruction.
+\cref{fig:testop} is the WebAssembly specification document for the
+\texttt{testop} instruction.
 Prose notation is written in the upper part, and formal notation is written in
@@
 \section{SpecTec}
+\label{sec:spectec}
 
-% SpecTec
-SpecTec is a tool for \textit{mechanizing} WebAssembly specification.
-Specification mechanization is a technique that treats the specification as
-data that can be manipulated by a computer.
-SpecTec provides a domain specific language(DSL) for specification, and it
-automatically generates a specification documents including the semantics
-described in formal notation and prose noatation from the DSL.
-The DSL is a declarative language that resembles textbook style notation, and
-it is type checked to prevent meta-level errors such as notation misuses and
-dimension mismatches.
+% SpecTec: specification mechanization tool
+SpecTec is a tool for \textit{mechanizing} WebAssembly specification
+~\cite{spectec}.
+Specification mechanization is a technique that treats a specification as data
+that can be manipulated by a computer to automatically generate parser and
+interpreter ~\cite{jiset}, to generate tests and perform differential tests
+~\cite{jest}, to find meta-level errors in the specification ~\cite{jstar}, and
+to perform meta-level static analysis for the static analysis of the defined
+language ~\cite{jsaver}.
+This technique is especially crucial in growing languages such as JavaScript
+and WebAssembly.
+
+% DSL & generated document
+SpecTec provides a domain specific language (DSL) for the specification, and it
+is a declarative language that resembles textbook style notation.
+It is type checked to prevent meta-level errors such as notation misuses and
+dimension mismatches. The code below is the specification for \texttt{testop}
+instruction written in the DSL, which is designed to be similar to the formal
+notation, while obtaining better user experience. \\
+\red{TODO: better character for $\sim$ in the code}
+\begin{verbatim}
+  rule Step_pure/testop:
+  (CONST nt c_1) (TESTOP nt testop)  ~>  (CONST I32 c)
+  -- if c = $testop_(nt, testop, c_1)
+\end{verbatim}
 
@@
 notation.
+\cref{fig:spectec-testop} is the specification document automatically generated
+from the DSL which depicts both formal notation and prose noatation.
+The automatically generated document also closely resembles the official
+document except for some minor notation changes and some missing hyperlinks.
 
-% interpreter
+% meta-level interpreter
 What is especially intriguing here is that the prose language is executable.
@@
 
-% other backends
-Furthermore, SpecTec automatically translates the DSL into Coq\cite{il2coq} for
-mathematical proofs and it is working on generating other artifacts such as
-test suites required to the WebAssembly standization.
+% other backends and other language
+SpecTec also translates the DSL into Coq\cite{il2coq} for mathematical proofs
+and it is also working on generating other artifacts such as test suites
+required to the WebAssembly standization.
+Moreover, some researches are ongoing to deploy SpecTec to mechanize other
+languages \cite{p4-cherry-workshop}.

### 2024-11-02 paper/2-background.tex
Minor change

@@
 notation.
-\cref{fig:spectec-testop} is the specification document automatically generated
-from the DSL which depicts both formal notation and prose noatation.
+\cref{fig:spectec-testop} is the specification document automatically
+generated from the DSL which depicts both formal notation and prose noatation.
 The automatically generated document also closely resembles the official
@@
 % other backends and other language
-SpecTec also translates the DSL into Coq\cite{il2coq} for mathematical proofs
+SpecTec also translates the DSL into Coq for mathematical proofs ~\cite{il2coq}
 and it is also working on generating other artifacts such as test suites
 required to the WebAssembly standization.
+% TODO: check any presentation about west, component model, or rust using SpecTec
 Moreover, some researches are ongoing to deploy SpecTec to mechanize other
-languages \cite{p4-cherry-workshop}.
+languages ~\cite{p4-cherry-workshop}.

### 2024-11-02 paper/2-background.tex
Fix cites

@@
 % breif overview
-WebAssembly is a low-level bytecode language that is safe, fast, portable, and
-compact~\cite{wasm}.
+WebAssembly~\cite{wasm} is a low-level bytecode language that is safe, fast,
+portable, and compact.
 It is widely used as a compile target for web applications.
@@
 There are dozens of WebAssembly engines; all the browsers have there own
-implementations of WebAssembly with multiple tiers, and there are many engines
-that target for specific areas including embedded systems and edge computing.
+implementations of WebAssembly with multiple tiers ~\cite{v8}
+\cite{spidermonkey} \cite{webkit}, and there are many engines that target for
+specific areas including embedded systems and edge computing.
 However, as WebAssembly should be portable across these implementations, it risks

### 2024-11-04 paper/2-background.tex
Update background

@@
 and it is used for proofs such as type soundness.
-On the other hand, prose notation uses psudocodes to explain the semantics
-through a step by step instruction.
+On the other hand, prose notation uses psudocodes algorithm to explain the
+semantics through a step by step instruction.
 Most WebAssembly users and engine developers are not familiar with
@@
 generated from the DSL which depicts both formal notation and prose noatation.
-The automatically generated document also closely resembles the official
-document except for some minor notation changes and some missing hyperlinks.
+The generated document closely resembles the official document except for some
+minor notation changes and some missing hyperlinks.
 
@@
 What is especially intriguing here is that the prose language is executable.
+It can perform each step written in the prose program one by one, which
+WebAssembly interpreter should follow.
 As a result, the prose program generated from the DSL serves as a WebAssembly
@@
 % other backends and other language
-SpecTec also translates the DSL into Coq for mathematical proofs ~\cite{il2coq}
-and it is also working on generating other artifacts such as test suites
-required to the WebAssembly standization.
-% TODO: check any presentation about west, component model, or rust using SpecTec
-Moreover, some researches are ongoing to deploy SpecTec to mechanize other
+SpecTec also translates the DSL into Coq for mathematical proofs and is working
+on generating other artifacts, such as test suites required for WebAssembly
+standardization.
+Additionally, research is ongoing to extend SpecTec for mechanizing other
 languages ~\cite{p4-cherry-workshop}.

### 2024-11-04 paper/2-background.tex
spectec background version 3

@@
 definitions such as functions, globals, and memories.
-A module is instantiated to validate the module, allocate each definition to
+A module is instantiated to validate the module and allocate each definition to
 the data structure named \textit{store}.
-This results in module instance, a runtime representation for the module.
+This results in a module instance, a runtime representation for the module.
 After the instantiation, functions can be invoked so that the function body,
@@
 
-% DSL & generated document
-SpecTec provides a domain specific language (DSL) for the specification, and it
-is a declarative language that resembles textbook style notation.
+% DSL high-level explanation
+SpecTec provides a domain specific language (DSL) for the specification.
+The DSL is a declarative language that resembles textbook style notation.
 It is type checked to prevent meta-level errors such as notation misuses and
@@
 instruction written in the DSL, which is designed to be similar to the formal
-notation, while obtaining better user experience. \\
+notation when describing the semantics with compact and user-friendly
+notation.
+\\
 \red{TODO: better character for $\sim$ in the code}
@@
 
-% prose language
+% DSL
+The DSL consists of DSL definitions.
+There are mainly 4 kinds of DSL definitions: type, grammar, relation, and
+functions.
+Type definitions are used to define abstract syntax, while grammar definitions
+are used to define concrete syntax.
+Relation definitions are used to define semantics; the code above is also
+described in relation definition.
+Function definitions can be utilized for other definitions such as
+\texttt{\$testop\_} in the code.
+In addition, they can be also used to describe semantics.
+Especially, SpecTec uses function definitions to describe module instantiation
+and function invocation.
+
+% AL
 However, The DSL cannot generate the prose notation just as it is because it is
 declarative.
-Therefore, SpecTec also defines another imperative language for the prose
-notation, and translates the DSL into the prose language to generate the prose
-notation.
-\cref{fig:spectec-testop} is the specification document automatically
-generated from the DSL which depicts both formal notation and prose noatation.
-The generated document closely resembles the official document except for some
-minor notation changes and some missing hyperlinks.
+Therefore, SpecTec also defines another imperative language \textit{AL}, which
+stands for algorithmic language, for the prose notation.
+SpecTec translates the semantics describing part of the DSL, which is the
+relation defintions and the function definitions, into the AL.
+Both definitions are expressed in the form of algorithms in AL that takes some
+inputs and performs a step by step instructions as in the prose notation.
 
 % meta-level interpreter
-What is especially intriguing here is that the prose language is executable.
+What is especially intriguing here is that AL is executable.
 It can perform each step written in the prose program one by one, which
 WebAssembly interpreter should follow.
-As a result, the prose program generated from the DSL serves as a WebAssembly
-interpter.
+As a result, the prose program serves as a WebAssembly interpter.
 That is to say, if WebAssembly semantics is written in the DSL, SpecTec
@@
 
-% other backends and other language
-SpecTec also translates the DSL into Coq for mathematical proofs and is working
+% artifacts SpecTec generates & SpecTec for other language
+\cref{fig:spectec-testop} is the specification document generated by the
+SpecTec using the DSL and the AL.
+The generated document closely resembles the official document except for some
+minor notation changes and some missing hyperlinks.
+Additionally, SpecTec can generate many other artifacts automatically from the
+DSL.
+SpecTec translates the DSL into Coq for mathematical proofs and is working
 on generating other artifacts, such as test suites required for WebAssembly
 standardization.
-Additionally, research is ongoing to extend SpecTec for mechanizing other
+Furthermore, research is ongoing to extend SpecTec for mechanizing other
 languages ~\cite{p4-cherry-workshop}.

### 2024-11-05 paper/5-eval.tex
WIP: Add Wasm background
Fix rule and notation

@@
 \texttt{br}, and \texttt{returning from a function}.
-By using the semantics defined in \cref{ch:semantics}, the example 5.1 results in the same result with the \officialp{}; $call(\$brAndReturning)$ does nothing.
+By using the semantics defined in \cref{ch:semantics}, the example 5.1 results
+in the same result with the \officialp{}; $call(\$brAndReturning)$ does
+nothing.
 
@@
 \begin{align}
-  &c^* \vdash call(\$brAndReturning); ~ \wasm ~ (n, ~ c)
+  &c^* \vdash call(\$brAndReturning); ~ \wasm ~ (\seq{c}, ~ k)
   \label{eq:resolve-1} \\
-  &\leadsto c^* \vdash \epsilon; ~ \al ~ ([ ~ \pushctxi, ~ \enteri ~ ], ~ \epsilon, ~ \wasm ~ (n, ~ c))
+  &\leadsto c^* \vdash \epsilon; ~ \algo ~ ([ ~ \pushctxi, ~ \enteri ~ ], ~ \epsilon, ~ \wasm ~ (\seq{c}, ~ k))
   \label{eq:resolve-2} \\
-  &\leadsto Frame, ~ c^* \vdash \epsilon; ~ \al ~ ([ ~ \enteri ~ ], ~ [ ~ Frame ~ ], ~ \wasm ~ (n, ~ c))
+  &\leadsto Frame, ~ c^* \vdash \epsilon; ~ \algo ~ ([ ~ \enteri ~ ], ~ [ ~ Frame ~ ], ~ \wasm ~ (\seq{c}, ~ k))
   \label{eq:resolve-3} \\
   &\leadsto Label(\epsilon), ~ Frame, ~ c^* \vdash (br ~ \epsilon), ~ (unreachable), ~ (exiting), ~ (returning); \\
-  &\quad\quad \wasm ~ ([ ~ Label, ~ Frame ~ ], ~ \al ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (n, ~ c)))
+  &\quad\quad \wasm ~ ([ ~ Label(\epsilon), ~ Frame ~ ], ~ \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (\seq{c}, ~ k)))
   \label{eq:resolve-4} \\
   &\leadsto Label(\epsilon), ~ Frame, ~ c^* \vdash (unreachable), ~ (exiting), ~ (returning); \\
-  &\quad\quad \al ~ ([ ~ \red{\symbol{PopCtxI}}, ~ \executeseqi ~ ], ~ \epsilon, ~ \wasm ~ ([ ~ Label, ~ Frame ~ ], ~ \al ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (n, ~ c))))
+  &\quad\quad \algo ~ ([ ~ \popctxi, ~ \executeseqi ~ ], ~ \epsilon, ~ \wasm ~ ([ ~ Label(\epsilon), ~ Frame ~ ], ~ \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (\seq{c}, ~ k))))
   \label{eq:resolve-5} \\
   &\leadsto Frame, ~ c^* \vdash (returning); ~
-    \al ~ ([ ~ \executeseqi ~ ], ~ \epsilon, ~ \wasm ~ ([ ~ Frame ~ ], ~ \al ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (n, ~ c))))
+    \algo ~ ([ ~ \executeseqi ~ ], ~ \epsilon, ~ \wasm ~ ([ ~ Frame ~ ], ~ \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (\seq{c}, ~ k))))
   \label{eq:resolve-6} \\
   &\leadsto Frame, ~ c^* \vdash (returning); ~
-    \al ~ (\epsilon, ~ \epsilon, ~ \wasm ~ ([ ~ Frame ~ ], ~ \al ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (n, ~ c))))
+    \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ ([ ~ Frame ~ ], ~ \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (\seq{c}, ~ k))))
   \label{eq:resolve-7} \\
-  &\leadsto Frame, ~ c^* \vdash (returning); ~ \wasm ~ ([ ~ Frame ~ ], ~ \al ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (n, ~ c)))
+  &\leadsto Frame, ~ c^* \vdash (returning); ~ \wasm ~ ([ ~ Frame ~ ], ~ \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (\seq{c}, ~ k)))
   \label{eq:resolve-8} \\
   &\leadsto Frame, ~ c^* \vdash \epsilon; ~
-    \al ~ ([ ~ \red{\symbol{PopCtxI}} ~ ], ~ \epsilon, ~ \wasm ~ (\epsilon, ~ \al ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (n, ~ c))))
+    \algo ~ ([ ~ \popctxi ~ ], ~ \epsilon, ~ \wasm ~ (\epsilon, ~ \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (\seq{c}, ~ k))))
   \label{eq:resolve-9} \\
   &\leadsto c^* \vdash \epsilon; ~
-    \al ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (\epsilon, ~ \al ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (n, ~ c))))
+    \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (\epsilon, ~ \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (\seq{c}, ~ k))))
   \label{eq:resolve-10} \\
-  &\leadsto c^* \vdash \epsilon; ~ \wasm ~ (\epsilon, ~ \al ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (n, ~ c)))
+  &\leadsto c^* \vdash \epsilon; ~ \wasm ~ (\epsilon, ~ \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (\seq{c}, ~ k)))
   \label{eq:resolve-11} \\
-  &\leadsto c^* \vdash \epsilon; ~ \al ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (n, ~ c))
+  &\leadsto c^* \vdash \epsilon; ~ \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (\seq{c}, ~ k))
   \label{eq:resolve-12} \\
-  &\leadsto c^* \vdash \epsilon; ~ \wasm ~ (n, ~ c)
+  &\leadsto c^* \vdash \epsilon; ~ \wasm ~ (\seq{c}, ~ k)
   \label{eq:resolve-13}

### 2024-11-12 paper/5-eval.tex
Trim

@@
   &\leadsto Label(\epsilon), ~ Frame, ~ c^* \vdash (unreachable), ~ (exiting), ~ (returning); \\
-  &\quad\quad \algo ~ ([ ~ \popctxi, ~ \executeseqi ~ ], ~ \epsilon, ~ \wasm ~ ([ ~ Label(\epsilon), ~ Frame ~ ], ~ \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (\seq{c}, ~ k))))
+  &\quad\quad \algo ~ ([ ~ \popctxi, ~ \executei ~ ], ~ \epsilon, ~ \wasm ~ ([ ~ Label(\epsilon), ~ Frame ~ ], ~ \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (\seq{c}, ~ k))))
   \label{eq:resolve-5} \\
   &\leadsto Frame, ~ c^* \vdash (returning); ~
-    \algo ~ ([ ~ \executeseqi ~ ], ~ \epsilon, ~ \wasm ~ ([ ~ Frame ~ ], ~ \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (\seq{c}, ~ k))))
+    \algo ~ ([ ~ \executei ~ ], ~ \epsilon, ~ \wasm ~ ([ ~ Frame ~ ], ~ \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (\seq{c}, ~ k))))
   \label{eq:resolve-6} \\

### 2024-11-16 paper/1-intro.tex
WIP: write intro

@@
 \end{itemize}
+
+% breif overview
+WebAssembly (Wasm)~\cite{wasm} is a low-level bytecode language that is safe, fast,
+portable, and compact.
+It is widely used as a compilation target for web applications.
+Beyond web development, Wasm's advantages are also deployed in areas such as
+edge computing~\cite{wasm-edge}, IoT~\cite{wasm-iot}, and
+blockchain~\cite{wasm-block}.
+
+% risk of implementation divergence
+There are numerous Wasm engines, with all major browsers implementing their own
+multi-tiered versions~\cite{v8, spidermonkey, webkit}.
+Additionally, specialized engines target specific domains, such as embedded
+systems and edge computing.
+However, ensuring portability across these diverse implementations introduces
+the risk of divergence among them.
+
+% rigorous standardization -> spec is rigorous: formal notation & prose notation
+To address this challenge, Wasm has been rigorously standardized by the
+W3C~\cite{wasm-w3c}.
+The Wasm specification is particularly rigorous, describing its semantics in
+two complementary forms: formal notation and prose notation.
+The formal notation employs mathematical rules to succinctly define the
+semantics, supporting proofs such as type soundness.
+Conversely, the prose notation uses pseudocode-like algorithms to explain the
+semantics through step-by-step instructions.
+Since most Wasm users and engine developers are less familiar with mathematical
+formalism, they primarily rely on the prose notation.
+
+% challenging specification process
+The demanding standardization process places a significant burden on
+specification authors.
+Crafting this specification document is labor-intensive, and as Wasm evolves,
+the manual effort required becomes increasingly challenging to scale.
+Moreover, the dual requirement of maintaining both the formal notation and
+prose notation exacerbates these difficulties.
+The formal notation is written in LaTeX, and the prose notation is authored
+in reStructuredText; neither of which is particularly user-friendly for
+collaborative review.
+This lack of accessibility in the specification's tooling increases the
+likelihood of inconsistensies and errors, further complicating the
+standardization effort.
+
+% SpecTec
+To alleviate this problem, we had been developing SpecTec, a framework for
+mechanizing WebAssembly specification.
+It provides DSL to define Wasm syntax and semantics in a declarative style
+similar to the formal notation.
+It type-checks DSL to prevent meta-level specification errors, and generates
+many artifacts including the specification document, a Wasm interpreter, and
+mechanized definitions for theorem provers.
+
+% AL
+To express
+
+% problem
+However, AL cannot describe Wasm control flow correctly
+
+% contribution
+build AL's computation model for Wasm, formalize AL semantics

### 2024-11-16 paper/1-intro.tex
Add SpecTec and problem in intro

@@
 % SpecTec
-To alleviate this problem, we had been developing SpecTec, a framework for
+To mitigate this problem, we \red{had been} developing SpecTec, a framework for
 mechanizing WebAssembly specification.
-It provides DSL to define Wasm syntax and semantics in a declarative style
-similar to the formal notation.
-It type-checks DSL to prevent meta-level specification errors, and generates
-many artifacts including the specification document, a Wasm interpreter, and
-mechanized definitions for theorem provers.
-
-% AL
-To express
+It provides domain specific language (DSL) that enables the declarative
+definition of Wasm syntax and semantics, akin to the formal notation.
+Additionally, to capture the pseudocode-like structure of the prose notation,
+SpecTec incorporates an imperative language, \textit{AL}.
+SpecTec performs type-checking on the DSL to prevent meta-level specification
+errors, translates the DSL into AL, and generates various artifacts, including
+the specification document, a Wasm interpreter, and mechanized definitions for
+theorem provers, along with the DSL and AL.
+Notably, AL plays a central role in the generation of the Wasm interpreter,
+bridging the gap between formal specification and executable code.
 
 % problem
-However, AL cannot describe Wasm control flow correctly
+However, at that time, AL was unable to capture the Wasm control flow
+correctly.
+The Wasm interpreter generated using AL produced wrong results when running
+Wasm official test suites.
+Additionally, the lack of formal semantics or a formalized definition for AL
+made it difficult to identify and address the fundamental issues in its design.
+Since running tests is the primary method for assessing the correctness of the
+specification, this limitation hindered the development of both the generated
+interpreter and the specification itself.
+This underscored the need for a more robust approach to formally define AL and
+ensure its alignment with Wasm's complex semantics.
 

### 2024-11-16 paper/1-intro.tex
Write intro

@@
 \noindent
-
-\begin{itemize}
-\item Wasm
-  \begin{itemize}
-  \item is low-level bytecode language, platform-indep, ...
-  \item is compile target for Web applications
-  \item is used in wide area such as IoT, ...
-  \end{itemize}
-\item Wasm Spec
-  \begin{itemize}
-  \item descibes syntax and semantics of Wasm
-  \item should be rigorous because ... (brief explanation for two format)
-  \item has two format to describe semantics (formal\&prose notation)
-  \item formal notation is written in math expression: concise, clear, amenable to proofs, ...
-  \item prose notation is step by step instruction written English: vague (in some part), amenable to developing engines
-  \end{itemize}
-\item SpecTec
-  \begin{itemize}
-  \item is proposed to alleviate spec writing process
-  \item is accepted by Wasm committee to be an official Wasm specification tool
-  \item define AL to describe Wasm semantics in prose notation
-  \item AL is designed to be similar to prose notation (describe with fig)
-  \item implement AL interpreter to make the spec executable
-  \item executable spec passes official wasm test $\leftarrow$ correctness
-  \end{itemize}
-\item Problem
-  \begin{itemize}
-  \item no semantics in AL
-  \item prose notation is still vague unless read AL interpreter code
-  \item ???
-  \end{itemize}
-\item Contribution
-  \begin{itemize}
-  \item formalize AL syntax$\&$semantics
-  \item make prose notation clear to understand
-  \item ???
-  \end{itemize}
-\end{itemize}
 
@@
 SpecTec performs type-checking on the DSL to prevent meta-level specification
-errors, translates the DSL into AL, and generates various artifacts, including
-the specification document, a Wasm interpreter, and mechanized definitions for
-theorem provers, along with the DSL and AL.
-Notably, AL plays a central role in the generation of the Wasm interpreter,
-bridging the gap between formal specification and executable code.
+errors, translates the DSL into AL.
+One notable aspect of AL is that it is executable, meaning any specification
+written in AL serves as a Wasm interpreter program.
+This bridges the gap between formal specification and executable code.
+Futhermore, SpecTec generates various artifacts, including the specification
+document, mechanized definitions for theorem provers.
 
@@
 correctly.
-The Wasm interpreter generated using AL produced wrong results when running
-Wasm official test suites.
+The Wasm interpreter defined in AL produced incorrect results when running Wasm
+official test suites.
 Additionally, the lack of formal semantics or a formalized definition for AL
@@
 
+% in this paper
+In this paper, we present a formal model for how AL describes Wasm control flow.
+Using this model, we identify the underlying issues and the necessary changes
+to improve it.
+Based on these insight, we formalize AL's syntax and semantics to accurately
+capture Wasm control flow.
+We then implement an AL interpreter according to this formalization and
+evaluate its correctness.
+The interpreter successfully passes all tests in the Wasm official test suite,
+as well as tests from proposals.
+
 % contribution
-build AL's computation model for Wasm, formalize AL semantics
+The contributions of this paper are as follows:
+\begin{itemize}
+  \item We propose a formal model for Wasm control flow in AL (\S3)
+  \item We formalize syntax and semantics of AL (\S4)
+  \item We implement an AL interpreter based on the formalization (\S5)
+\end{itemize}

### 2024-11-17 paper/3-motivation.tex
Add a buggy example & explanation Step by step reduction of the example
Add control structure
Move prev eval to motivation

@@
 \section{Control flow in official prose}
+\label{sec:control flow in official prose}
 
@@
 
+\red{TODO: Change to array} \\
 \begin{align}
@@
 the problem.
+
+%%%%%%%%
+
+% an example of buggy test case
+\textbf{Example 5.1}
+\begin{verbatim}
+  // function definition of $br-returning
+  (func  $brAndReturning (br 0) (unreachable))
+
+  // function call
+  (call $brAndReturning)
+\end{verbatim}
+
+Example 5.1 contains a function that has only a \texttt{br} instruction, and a
+\texttt{call} insstruction.
+According to the \officialp{}, when the function \texttt{\$brAndReturning} is
+called, a frame is pushed to the stack, and the block of the function
+body is entered with a label.
+When the \texttt{br} is executed, it pops the label from the stack.
+As the end of the function is reached after the execution, returning from a
+function is performed.
+As a result, the frame is popped, so the call actually does nothing.
+
+% explaination of the bug in SpecTec prose
+\begin{align}
+  &c^* \vdash call(\$brAndReturning) \label{eq:bug-1} \\
+&\leadsto
+  Label(\epsilon), ~ Frame, ~ c^* \vdash (br ~ 0) ~ (unreachable), ~ end \label{eq:bug-2} \\
+&\leadsto
+  Frame, ~ c^* \vdash \epsilon \label{eq:bug-3}
+\end{align}
+
+In \spectecp{} the call instruction in \cref{eq:bug-1} pushes a frame and
+enters the function body \cref{eq:bug-2}.
+When the \texttt{br} is executed, it pops the label and removes the input
+instructions until the end in \cref{eq:bug-3}.
+As a result, the frame is remained after the executing the call.
+The problem is that the \texttt{br} should remove the end of the block but
+should not remove the end of the function, but \texttt{end} express the end of
+the block and the function at the same time.
+
+% Fix
+To fix the problem, semantics of \enteri is changed.
+The sequence of contexts that is pushed in the AL algorithm is stored,
+and when a block is entered, exiting label and returning from a function is
+appended according to the sequence.
+
+% simplified version of AL algorithm of call, br, and returning from a function
+\begin{verbatim}
+Call fname:
+  PushCtx Frame
+  Enter get_function_body(fname) with Label(\epsilon)
+
+Br 0:
+  PopCtx Label(instr*)
+  ExecuteSeq instr*
+
+Returning from a function:
+  PopCtx Frame
+\end{verbatim}
+
+The code above is simplified version of \spectecp{} of \texttt{call},
+\texttt{br}, and \texttt{returning from a function}.
+By using the semantics defined in \cref{ch:semantics}, the example 5.1 results
+in the same result with the \officialp{}; $call(\$brAndReturning)$ does
+nothing.
+
+\red{TODO: Change to array} \\
+\red{TODO: Change naming for $c$}
+\begin{align}
+  &c^* \vdash call(\$brAndReturning); ~ \wasm ~ (\seq{c}, ~ k)
+  \label{eq:resolve-1} \\
+  &\leadsto c^* \vdash \epsilon; ~ \algo ~ ([ ~ \pushctxi, ~ \enteri ~ ], ~ \epsilon, ~ \wasm ~ (\seq{c}, ~ k))
+  \label{eq:resolve-2} \\
+  &\leadsto Frame, ~ c^* \vdash \epsilon; ~ \algo ~ ([ ~ \enteri ~ ], ~ [ ~ Frame ~ ], ~ \wasm ~ (\seq{c}, ~ k))
+  \label{eq:resolve-3} \\
+  &\leadsto Label(\epsilon), ~ Frame, ~ c^* \vdash (br ~ \epsilon), ~ (unreachable), ~ (exiting), ~ (returning); \\
+  &\quad\quad \wasm ~ ([ ~ Label(\epsilon), ~ Frame ~ ], ~ \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (\seq{c}, ~ k)))
+  \label{eq:resolve-4} \\
+  &\leadsto Label(\epsilon), ~ Frame, ~ c^* \vdash (unreachable), ~ (exiting), ~ (returning); \\
+  &\quad\quad \algo ~ ([ ~ \popctxi, ~ \executei ~ ], ~ \epsilon, ~ \wasm ~ ([ ~ Label(\epsilon), ~ Frame ~ ], ~ \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (\seq{c}, ~ k))))
+  \label{eq:resolve-5} \\
+  &\leadsto Frame, ~ c^* \vdash (returning); ~
+    \algo ~ ([ ~ \executei ~ ], ~ \epsilon, ~ \wasm ~ ([ ~ Frame ~ ], ~ \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (\seq{c}, ~ k))))
+  \label{eq:resolve-6} \\
+  &\leadsto Frame, ~ c^* \vdash (returning); ~
+    \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ ([ ~ Frame ~ ], ~ \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (\seq{c}, ~ k))))
+  \label{eq:resolve-7} \\
+  &\leadsto Frame, ~ c^* \vdash (returning); ~ \wasm ~ ([ ~ Frame ~ ], ~ \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (\seq{c}, ~ k)))
+  \label{eq:resolve-8} \\
+  &\leadsto Frame, ~ c^* \vdash \epsilon; ~
+    \algo ~ ([ ~ \popctxi ~ ], ~ \epsilon, ~ \wasm ~ (\epsilon, ~ \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (\seq{c}, ~ k))))
+  \label{eq:resolve-9} \\
+  &\leadsto c^* \vdash \epsilon; ~
+    \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (\epsilon, ~ \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (\seq{c}, ~ k))))
+  \label{eq:resolve-10} \\
+  &\leadsto c^* \vdash \epsilon; ~ \wasm ~ (\epsilon, ~ \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (\seq{c}, ~ k)))
+  \label{eq:resolve-11} \\
+  &\leadsto c^* \vdash \epsilon; ~ \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (\seq{c}, ~ k))
+  \label{eq:resolve-12} \\
+  &\leadsto c^* \vdash \epsilon; ~ \wasm ~ (\seq{c}, ~ k)
+  \label{eq:resolve-13}
+\end{align}

### 2024-11-17 paper/3-motivation.tex
Move command definition

@@
 \begin{align}
-  &c^* \vdash call(\$brAndReturning); ~ \wasm ~ (\seq{c}, ~ k)
+  &c^* \vdash call(\$brAndReturning); ~ \wasm ~ (c^*, ~ k)
   \label{eq:resolve-1} \\
-  &\leadsto c^* \vdash \epsilon; ~ \algo ~ ([ ~ \pushctxi, ~ \enteri ~ ], ~ \epsilon, ~ \wasm ~ (\seq{c}, ~ k))
+  &\leadsto c^* \vdash \epsilon; ~ \algo ~ ([ ~ \pushctxi, ~ \enteri ~ ], ~ \epsilon, ~ \wasm ~ (c^*, ~ k))
   \label{eq:resolve-2} \\
-  &\leadsto Frame, ~ c^* \vdash \epsilon; ~ \algo ~ ([ ~ \enteri ~ ], ~ [ ~ Frame ~ ], ~ \wasm ~ (\seq{c}, ~ k))
+  &\leadsto Frame, ~ c^* \vdash \epsilon; ~ \algo ~ ([ ~ \enteri ~ ], ~ [ ~ Frame ~ ], ~ \wasm ~ (c^*, ~ k))
   \label{eq:resolve-3} \\
   &\leadsto Label(\epsilon), ~ Frame, ~ c^* \vdash (br ~ \epsilon), ~ (unreachable), ~ (exiting), ~ (returning); \\
-  &\quad\quad \wasm ~ ([ ~ Label(\epsilon), ~ Frame ~ ], ~ \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (\seq{c}, ~ k)))
+  &\quad\quad \wasm ~ ([ ~ Label(\epsilon), ~ Frame ~ ], ~ \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (c^*, ~ k)))
   \label{eq:resolve-4} \\
   &\leadsto Label(\epsilon), ~ Frame, ~ c^* \vdash (unreachable), ~ (exiting), ~ (returning); \\
-  &\quad\quad \algo ~ ([ ~ \popctxi, ~ \executei ~ ], ~ \epsilon, ~ \wasm ~ ([ ~ Label(\epsilon), ~ Frame ~ ], ~ \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (\seq{c}, ~ k))))
+  &\quad\quad \algo ~ ([ ~ \popctxi, ~ \executei ~ ], ~ \epsilon, ~ \wasm ~ ([ ~ Label(\epsilon), ~ Frame ~ ], ~ \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (c^*, ~ k))))
   \label{eq:resolve-5} \\
   &\leadsto Frame, ~ c^* \vdash (returning); ~
-    \algo ~ ([ ~ \executei ~ ], ~ \epsilon, ~ \wasm ~ ([ ~ Frame ~ ], ~ \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (\seq{c}, ~ k))))
+    \algo ~ ([ ~ \executei ~ ], ~ \epsilon, ~ \wasm ~ ([ ~ Frame ~ ], ~ \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (c^*, ~ k))))
   \label{eq:resolve-6} \\
   &\leadsto Frame, ~ c^* \vdash (returning); ~
-    \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ ([ ~ Frame ~ ], ~ \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (\seq{c}, ~ k))))
+    \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ ([ ~ Frame ~ ], ~ \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (c^*, ~ k))))
   \label{eq:resolve-7} \\
-  &\leadsto Frame, ~ c^* \vdash (returning); ~ \wasm ~ ([ ~ Frame ~ ], ~ \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (\seq{c}, ~ k)))
+  &\leadsto Frame, ~ c^* \vdash (returning); ~ \wasm ~ ([ ~ Frame ~ ], ~ \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (c^*, ~ k)))
   \label{eq:resolve-8} \\
   &\leadsto Frame, ~ c^* \vdash \epsilon; ~
-    \algo ~ ([ ~ \popctxi ~ ], ~ \epsilon, ~ \wasm ~ (\epsilon, ~ \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (\seq{c}, ~ k))))
+    \algo ~ ([ ~ \popctxi ~ ], ~ \epsilon, ~ \wasm ~ (\epsilon, ~ \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (c^*, ~ k))))
   \label{eq:resolve-9} \\
   &\leadsto c^* \vdash \epsilon; ~
-    \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (\epsilon, ~ \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (\seq{c}, ~ k))))
+    \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (\epsilon, ~ \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (c^*, ~ k))))
   \label{eq:resolve-10} \\
-  &\leadsto c^* \vdash \epsilon; ~ \wasm ~ (\epsilon, ~ \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (\seq{c}, ~ k)))
+  &\leadsto c^* \vdash \epsilon; ~ \wasm ~ (\epsilon, ~ \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (c^*, ~ k)))
   \label{eq:resolve-11} \\
-  &\leadsto c^* \vdash \epsilon; ~ \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (\seq{c}, ~ k))
+  &\leadsto c^* \vdash \epsilon; ~ \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (c^*, ~ k))
   \label{eq:resolve-12} \\
-  &\leadsto c^* \vdash \epsilon; ~ \wasm ~ (\seq{c}, ~ k)
+  &\leadsto c^* \vdash \epsilon; ~ \wasm ~ (c^*, ~ k)
   \label{eq:resolve-13}

### 2024-11-17 paper/3-motivation.tex
Write eval

--- 2024-10-29 paper/3-motivation.tex
@@
 \section{Control flow in SpecTec prose}
+\label{sec:spectecp}
 
@@
 the problem.
+
+%%%%%%%%
+
+% an example of buggy test case
+\begin{example}
+\begin{verbatim}
+  // function definition of $br-returning
+  (func  $brAndReturning (br 0) (unreachable))
+
+  // function call
+  (call $brAndReturning)
+\end{verbatim}
+\label{ex:bug}
+\end{example}
+
+\cref{ex:bug} contains a function that has only a \texttt{br} instruction, and a
+\texttt{call} insstruction.
+According to the \officialp{}, when the function \texttt{\$brAndReturning} is
+called, a frame is pushed to the stack, and the block of the function
+body is entered with a label.
+When the \texttt{br} is executed, it pops the label from the stack.
+As the end of the function is reached after the execution, returning from a
+function is performed.
+As a result, the frame is popped, so the call actually does nothing.
+
+% explaination of the bug in SpecTec prose
+\begin{align}
+  &c^* \vdash call(\$brAndReturning) \label{eq:bug-1} \\
+&\leadsto
+  Label(\epsilon), ~ Frame, ~ c^* \vdash (br ~ 0) ~ (unreachable), ~ end \label{eq:bug-2} \\
+&\leadsto
+  Frame, ~ c^* \vdash \epsilon \label{eq:bug-3}
+\end{align}
+
+In \spectecp{} the call instruction in \cref{eq:bug-1} pushes a frame and
+enters the function body \cref{eq:bug-2}.
+When the \texttt{br} is executed, it pops the label and removes the input
+instructions until the end in \cref{eq:bug-3}.
+As a result, the frame is remained after the executing the call.
+The problem is that the \texttt{br} should remove the end of the block but
+should not remove the end of the function, but \texttt{end} express the end of
+the block and the function at the same time.
+
+% Fix
+To fix the problem, semantics of \enteri is changed.
+The sequence of contexts that is pushed in the AL algorithm is stored,
+and when a block is entered, exiting label and returning from a function is
+appended according to the sequence.

### 2024-11-17 paper/5-eval.tex
Write eval

--- 2024-10-30 paper/5-eval.tex
@@
 
-% an example of buggy test case
-\textbf{Example 5.1}
+We evaluate our approach with the following three research questions.
+\begin{itemize}
+  \item RQ1. (Correctness) Does the formalization correctly capture the Wasm
+    control flow? (\S 5.1)
+  \item RQ2. (Generality) Does the formalization capture the fundamental
+    control flow semantics rather than being tailored to the current version of
+    the specification? (\S 5.2)
+  \item RQ3. \red{(Efficiency) Does our approach efficient? (or scalable?) (\S
+    5.3)}
+\end{itemize}
+
+\section{RQ1: Correctness}
+To demonstrate that our formalization correctly solves the problem, we revisit
+\cref{ex:bug}.
+The code below is a simplified version of Wasm interpreter program written in AL.
 \begin{verbatim}
-  // function definition of $br-returning
-  (func  $brAndReturning (br 0) (unreachable))
+(* Top-level function: invoke *)
+invoke funidx:
+  PushCtxI Frame
+  EnterI (Label(\epsilon), get_function_body(funidx))
+  ReturnFunI get_return_values(funidx)
 
-  // function call
-  (call $brAndReturning)
+(* Wasm instructions *)
+Br 0:
+  PopCtxI Label(instr*)
+  ExecuteI instr*
+
+Unreachable:
+  TrapI
+
+(* Admin instructions *)
+Exiting:
+  PopCtxI Label(instr*)
+
+Returning:
+  PopCtxI Frame
+
 \end{verbatim}
+Using the semantics defined in \cref{ch:formal} together with the AL program,
+\cref{ex:bug} produces the same as the \officialp{}; invoking the function has
+no effect.
+We adopt the simplified notation $wc \vdash wi^*$ for a Wasm state, $L$ for a
+label and $F$ for a frame, as in the notation from \cref{sec:spectecp}.
+Additionally, we omit the environment argument in \algo{} and the arguments of AL
+instructions.
 
-Example 5.1 contains a function that has only a \texttt{br} instruction, and a
-\texttt{call} insstruction.
-According to the \officialp{}, when the function \texttt{\$brAndReturning} is
-called, a frame is pushed to the stack, and the block of the function
-body is entered with a label.
-When the \texttt{br} is executed, it pops the label from the stack.
-As the end of the function is reached after the execution, returning from a
-function is performed.
-As a result, the frame is popped, so the call actually does nothing.
+\[
+\begin{array}{l}
+  \epsilon \vdash \epsilon; ~ \toplevelcall(\text{``invoke"}, \numv ~ (\helper{funidx\_of}(\text{``\$brAndReturning"})))
+  \\
+  \leadsto \epsilon \vdash \epsilon; ~ \algo ~ (\pushctxi ~ \enteri ~ \returnfuni, ~ \epsilon, ~ \mt)
+  \\
+  \leadsto F \vdash \epsilon; ~ \algo ~ (\enteri ~ \returnfuni, ~ F, ~ \mt)
+  \\
+  \leadsto L ~ F \vdash (br ~ \epsilon) ~ (unreachable) ~ (exiting) ~ (returning); \wasm ~ (L ~ F, ~ \algo ~ (\returnfuni, ~ \epsilon, ~ \mt))
+  \\
+  \leadsto L ~ F \vdash (unreachable) ~ (exiting) ~ (returnfuning); \\
+  \quad\qquad\qquad\qquad\qquad \algo ~ (\popctxi ~ \executei, ~ \epsilon, ~ \wasm ~ (L ~ F, ~ \algo ~ (\returnfuni, ~ \epsilon, ~ \mt)))
+  \\
+  \leadsto F \vdash (returning); ~
+    \algo ~ (\executei, ~ \epsilon, ~ \wasm ~ (F, ~ \algo ~ (\returnfuni, ~ \epsilon, ~ \mt)))
+  \\
+  \leadsto F \vdash (returning); ~
+    \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (F, ~ \algo ~ (\returnfuni, ~ \epsilon, ~ \mt)))
+  \\
+  \leadsto F \vdash (returning); ~ \wasm ~ (F, ~ \algo ~ (\returnfuni, ~ \epsilon, ~ \mt))
+  \\
+  \leadsto F \vdash \epsilon; ~
+    \algo ~ (\popctxi, ~ \epsilon, ~ \wasm ~ (\epsilon, ~ \algo ~ (\returnfuni, ~ \epsilon, ~ \mt)))
+  \\
+  \leadsto \epsilon \vdash \epsilon; ~
+    \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (\epsilon, ~ \algo ~ (\returnfuni, ~ \epsilon, ~ \mt)))
+  \\
+  \leadsto \epsilon \vdash \epsilon; ~ \wasm ~ (\epsilon, ~ \algo ~ (\returnfuni, ~ \epsilon, ~ \mt))
+  \\
+  \leadsto \epsilon \vdash \epsilon; ~ \algo ~ (\returnfuni, ~ \epsilon, ~ \mt)
+  \\
+  \leadsto \epsilon \vdash \epsilon; ~ \ret ~ (\epsilon, ~ \mt)
+  \\
+\end{array}
+\]
 
-% explaination of the bug in SpecTec prose
-\begin{align}
-  &c^* \vdash call(\$brAndReturning) \label{eq:bug-1} \\
-&\leadsto
-  Label(\epsilon), ~ Frame, ~ c^* \vdash (br ~ 0) ~ (unreachable), ~ end \label{eq:bug-2} \\
-&\leadsto
-  Frame, ~ c^* \vdash \epsilon \label{eq:bug-3}
-\end{align}
+Furthermore, we implement AL interpreter according to the formalization to
+evaluate its correctness.
+We write the Wasm specification using the DSL and generat a Wasm interpreter
+program written in AL.
+By executing the program with the AL interpreter against the official Wasm test
+suite, we indirectly test the AL interpreter.
+We test it using execution tests from official Wasm test suite.
+Tests related to inifinite loop detection are excluded, as the specification
+does not define this behavior.
+The result shows that the interpreter successfully pass all 51,549 tests in 70
+seconds on \red{(lambda 9)}.
 
-In \spectecp{} the call instruction in \cref{eq:bug-1} pushes a frame and
-enters the function body \cref{eq:bug-2}.
-When the \texttt{br} is executed, it pops the label and removes the input
-instructions until the end in \cref{eq:bug-3}.
-As a result, the frame is remained after the executing the call.
-The problem is that the \texttt{br} should remove the end of the block but
-should not remove the end of the function, but \texttt{end} express the end of
-the block and the function at the same time.
+\section{RQ2: Generality}
+The Wasm exception handler proposal introduces a new control structure,
+\textit{handler}, along with new control instructions.
+In our formalization, we initially assumed that a context is either a label or
+a frame.
+By simply extending the formalization to include a handler case, as shown
+below, AL can effectively describe the exception handler semantics:
 
-% Fix
-To fix the problem, semantics of \enteri is changed.
-The sequence of contexts that is pushed in the AL algorithm is stored,
-and when a block is entered, exiting label and returning from a function is
-appended according to the sequence.
+\begin{align*}
+%
+% Wasm Context
+  \text{Wasm Context}\quad ~ WC ~\ni~ wc ~::=~
+    ~ \casev ~ (\text{``Label"}, \seq v)
+    ~ | ~ \casev ~ (\text{``Frame"}, \seq v)
+    ~ | ~ \casev ~ (\text{``Handler"}, \seq v) \\
+\end{align*}
+\begin{align*}
+%
+% get_end_algo
+\newline \\
+  &\getendinstr(wc) =
+  \begin{cases}
+    \casev ~ (\text{``Exiting"}, \epsilon) &\quad\quad\premise{if}\quad wc = \casev ~ (\text{``Label"}, \seq v) \\
+    \casev ~ (\text{``Returning"}, \epsilon) &\quad\quad\premise{if}\quad wc = \casev ~ (\text{``Frame"}, \seq v) \\
+    \casev ~ (\text{``Exiting-handler"}, \epsilon) &\quad\quad\premise{if}\quad wc = \casev ~ (\text{``Handler"}, \seq v) \\
+  \end{cases} \\
+%
+% is_end_instr
+\newline \\
+  &\isendinstr(wi) =
+  \begin{cases}
+    true &\quad\quad\premise{if}\quad wi = \casev ~ (\text{``Exiting"}, \epsilon) \\
+    &\quad\quad\lor\quad
+    wi = \casev ~ (\text{``Returning"}, \epsilon) \\
+    &\quad\quad\lor\quad
+    wi = \casev ~ (\text{``Exiting-handler"}, \epsilon) \\
+    false &\quad\quad\premise{otherwise}
+  \end{cases} \\
+\end{align*}
 
-% simplified version of AL algorithm of call, br, and returning from a function
-\begin{verbatim}
-Call fname:
-  PushCtx Frame
-  Enter get_function_body(fname) with Label(\epsilon)
+To evaluate the extended formalization, we updated the AL interpreter
+accordingly.
+We incorporated the exception handler feature into the specification and
+generated a Wasm interpreter.
+The interpreter successfully passed all 51,614 execution tests in the exception handlers
+proposal, excluding infinite loop tests, in 116 seconds.
 
-Br 0:
-  PopCtx Label(instr*)
-  ExecuteSeq instr*
-
-Returning from a function:
-  PopCtx Frame
-\end{verbatim}
-
-The code above is simplified version of \spectecp{} of \texttt{call},
-\texttt{br}, and \texttt{returning from a function}.
-By using the semantics defined in \cref{ch:semantics}, the example 5.1 results in the same result with the \officialp{}; $call(\$brAndReturning)$ does nothing.
-
-\red{TODO: Change to array} \\
-\red{TODO: Change naming for $c$}
-\begin{align}
-  &c^* \vdash call(\$brAndReturning); ~ \wasm ~ (n, ~ c)
-  \label{eq:resolve-1} \\
-  &\leadsto c^* \vdash \epsilon; ~ \al ~ ([ ~ \pushctxi, ~ \enteri ~ ], ~ \epsilon, ~ \wasm ~ (n, ~ c))
-  \label{eq:resolve-2} \\
-  &\leadsto Frame, ~ c^* \vdash \epsilon; ~ \al ~ ([ ~ \enteri ~ ], ~ [ ~ Frame ~ ], ~ \wasm ~ (n, ~ c))
-  \label{eq:resolve-3} \\
-  &\leadsto Label(\epsilon), ~ Frame, ~ c^* \vdash (br ~ \epsilon), ~ (unreachable), ~ (exiting), ~ (returning); \\
-  &\quad\quad \wasm ~ ([ ~ Label, ~ Frame ~ ], ~ \al ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (n, ~ c)))
-  \label{eq:resolve-4} \\
-  &\leadsto Label(\epsilon), ~ Frame, ~ c^* \vdash (unreachable), ~ (exiting), ~ (returning); \\
-  &\quad\quad \al ~ ([ ~ \red{\symbol{PopCtxI}}, ~ \executeseqi ~ ], ~ \epsilon, ~ \wasm ~ ([ ~ Label, ~ Frame ~ ], ~ \al ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (n, ~ c))))
-  \label{eq:resolve-5} \\
-  &\leadsto Frame, ~ c^* \vdash (returning); ~
-    \al ~ ([ ~ \executeseqi ~ ], ~ \epsilon, ~ \wasm ~ ([ ~ Frame ~ ], ~ \al ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (n, ~ c))))
-  \label{eq:resolve-6} \\
-  &\leadsto Frame, ~ c^* \vdash (returning); ~
-    \al ~ (\epsilon, ~ \epsilon, ~ \wasm ~ ([ ~ Frame ~ ], ~ \al ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (n, ~ c))))
-  \label{eq:resolve-7} \\
-  &\leadsto Frame, ~ c^* \vdash (returning); ~ \wasm ~ ([ ~ Frame ~ ], ~ \al ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (n, ~ c)))
-  \label{eq:resolve-8} \\
-  &\leadsto Frame, ~ c^* \vdash \epsilon; ~
-    \al ~ ([ ~ \red{\symbol{PopCtxI}} ~ ], ~ \epsilon, ~ \wasm ~ (\epsilon, ~ \al ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (n, ~ c))))
-  \label{eq:resolve-9} \\
-  &\leadsto c^* \vdash \epsilon; ~
-    \al ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (\epsilon, ~ \al ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (n, ~ c))))
-  \label{eq:resolve-10} \\
-  &\leadsto c^* \vdash \epsilon; ~ \wasm ~ (\epsilon, ~ \al ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (n, ~ c)))
-  \label{eq:resolve-11} \\
-  &\leadsto c^* \vdash \epsilon; ~ \al ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (n, ~ c))
-  \label{eq:resolve-12} \\
-  &\leadsto c^* \vdash \epsilon; ~ \wasm ~ (n, ~ c)
-  \label{eq:resolve-13}
-\end{align}
+\section{\red{RQ3: Efficiency}}
+The Wasm tail call proposal introduces tail call optimization so that it
+contains tests with heavy recursive calls.
+As a result, running the test with previous AL interpreter results in a call
+stack overflow.
+To evaluate the \red{efficiency} of our approach, we similarly generate a Wasm
+interpreter including tail call features.
+It successfully passes all 51,639 execution tests in the tail call proposal,
+excluding infinite loop tests, in 1,284 seconds.

### 2024-11-18 paper/3-motivation.tex
Use cref for example

@@
 % an example of Wasm control flow in official prose
-\textbf{Example 1}
+\begin{example}
+\label{ex:br}
 \begin{verbatim}
@@
 \end{verbatim}
+\end{example}
 
-Example 1 is a WebAssembly code example that has \texttt{loop} and
+\cref{ex:br} is a WebAssembly code example that has \texttt{loop} and
 \texttt{unreachable}
@@
 % an example of exiting label in official prose
-\textbf{Example 2}
+\begin{example}
+\label{ex:exit}
 \begin{verbatim}
@@
 \end{verbatim}
+\end{example}
 
-Example 2 is a WebAssembly code example that has \texttt{loop} whose block is
+\cref{ex:exit} is a WebAssembly code example that has \texttt{loop} whose block is
 only \texttt{i32.const}, and \texttt{f32.const}
@@
 
-Consider example 1 above.
+Consider \cref{ex:br} above.
 If we assume some control structure $c^*$ is given, the code can be expressed
@@
 % an example of function call in official prose
-\textbf{Example 3}
+\begin{example}
+\label{ex:invoke}
 \begin{verbatim}
@@
 \end{verbatim}
+\end{example}
 
@@
 
-Example 3 contains a function definition whose body is \texttt{i32.const}, a
+\cref{ex:invoke} contains a function definition whose body is \texttt{i32.const}, a
 \texttt{call} instruction followed by a \texttt{f32.const} instruction, which
@@
 \begin{example}
+\label{ex:bug}
 \begin{verbatim}
@@
 \end{verbatim}
-\label{ex:bug}
 \end{example}

### 2024-11-19 paper/1-intro.tex
Move prev eval to motivation
Fix intro

@@
 blockchain~\cite{wasm-block}.
-
 % risk of implementation divergence
@@
 two complementary forms: formal notation and prose notation.
+\red{\cref{fig:testop} illustrates the specification document for $testop$
+instruction, where the upper part contains the prose notation, and the lower
+part contains the formal notation.}
 The formal notation employs mathematical rules to succinctly define the
@@
 formalism, they primarily rely on the prose notation.
+
+\begin{figure}[t]
+    \centerline{\includegraphics[width=15cm]{fig/testop}}
+    \caption[Enter the caption title here]{\texttt{testop} instruction} \label{fig:testop}
+    \centerline{\includegraphics[width=15cm]{fig/spectec-testop}}
+    \caption[Enter the caption title here]{SpecTec \texttt{testop} instruction} \label{fig:spectec-testop}
+\end{figure}
 
@@
 % SpecTec
-To mitigate this problem, we \red{had been} developing SpecTec, a framework for
-mechanizing WebAssembly specification.
+To mitigate this problem, we \red{propose} SpecTec, a framework for mechanizing
+WebAssembly specification.
 It provides domain specific language (DSL) that enables the declarative
 definition of Wasm syntax and semantics, akin to the formal notation.
+\red{
+SpecTec performs type-checking on the DSL to prevent meta-level specification
+errors and generates LaTeX for the formal notation.
 Additionally, to capture the pseudocode-like structure of the prose notation,
-SpecTec incorporates an imperative language, \textit{AL}.
-SpecTec performs type-checking on the DSL to prevent meta-level specification
-errors, translates the DSL into AL.
+SpecTec incorporates an imperative language, \textit{AL}, which stands for
+Algorithmic Language.
+SpecTec translates the DSL into AL and generate reStructuredText for the prose
+notation.
+\cref{fig:spectec-testop} is the specification document generated by the
+SpecTec using the DSL and the AL.
+The generated document closely resembles the official document except for some
+minor notation changes and some missing hyperlinks.
 One notable aspect of AL is that it is executable, meaning any specification
 written in AL serves as a Wasm interpreter program.
+By testing the Wasm interpreter program, we can check the correctness of the
+specification.
 This bridges the gap between formal specification and executable code.
-Futhermore, SpecTec generates various artifacts, including the specification
-document, mechanized definitions for theorem provers.
+Futhermore, SpecTec generates mechanized definitions for theorem provers.}
 
-% problem
-However, at that time, AL was unable to capture the Wasm control flow
-correctly.
-The Wasm interpreter defined in AL produced incorrect results when running Wasm
-official test suites.
-Additionally, the lack of formal semantics or a formalized definition for AL
+% challenge
+\red{
+One challenge of our approach is making AL executable.
+Since the prose notation in the official specification is written in informal
+pseudocode, the interpretation of the prose notation is not straightforward.
+The naive implementation of AL interpreter produced incorrect results when
+running Wasm official test suites, particularly those related to control flow.
+}
+Additionally, the absence of a formalized definition or formal semantics for AL
 made it difficult to identify and address the fundamental issues in its design.
@@
 
-% in this paper
-In this paper, we present a formal model for how AL describes Wasm control flow.
+% solution
+To address this, we present a formal model for how AL describes Wasm control flow.
 Using this model, we identify the underlying issues and the necessary changes
@@
 \begin{itemize}
-  \item We propose a formal model for Wasm control flow in AL (\S3)
-  \item We formalize syntax and semantics of AL (\S4)
-  \item We implement an AL interpreter based on the formalization (\S5)
+  \item We propose a formal model for Wasm control flow in AL (\cref{ch:motivation})
+  \item We formalize syntax and semantics of AL (\cref{ch:formal})
+  \item We implement an AL interpreter based on the formalization (\cref{ch:eval})
 \end{itemize}

### 2024-11-19 paper/related.tex
Rename files

--- 2024-10-07 kw/6-related.tex
@@
 \label{ch:related}
-
-\paragraph{Formal Semantics for Programming Languages}
-
-Researchers have defined formal semantics for various real-world programming
-languages.
-%
-\Cref{ch:intro} already provides the list of studies of formal
-semantics for each language.
-%
-Therefore, this section emphasizes their methodologies and compares them with
-the approach proposed in this work.
-
-While numerous studies have directly defined the formal semantics of a complete
-real-world language, this approach necessitates significant manual effort.
-%
-To circumvent this issue, some studies have chosen to define formal semantics
-by defining a small core language and then desugaring the surface language to
-the core
-language~\cite{guha2010essence,politz2012tested,memarian2016depths,politz2013python}.
-%
-This approach aligns with our work, where we introduce OpenQASMCore and perform
-desugaring from OpenQASM to OpenQASMCore.
-
-Another way to reduce the burden of defining formal semantics is to utilize
-semantic frameworks, such as K~\cite{rosu2010overview}, PLT
-Redex~\cite{felleisen2009semantics}, and Lem~\cite{mulligan2014lem}.
-%
-These semantic frameworks are also beneficial because they allow the semantics
-to be executable without any additional work.
-%
-K is a rewrite-based semantic framework, in which the semantics of many
-languages have been
-formalized~\cite{park2015kjs,ellison2012executable,bogdanas2015kjava,filaretti2014executable,jiao2020semantic}.
-%
-PLT Redex is a domain-specific language for specifying operational semantics
-and has been used to describe the semantics of various
-languages~\cite{guha2010essence,politz2013python}.
-
-Researchers have defined formal semantics using proof assistants as well.
-%
-Coq~\cite{barras1997coq} has been widely employed to mechanize formal
-semantics~\cite{bodin2014trusted,blazy2009mechanized,jung2017rustbelt,watt2021two}.
-%
-Isabelle/HOL~\cite{nipkow2002isabelle} is another popular proof assistant used
-for defining formal semantics~\cite{klein2006machine,watt2018mechanising}.
-%
-Semantics defined with proof assistants are often executable, such as shown by
-\citet{bodin2014trusted}, who extract OCaml code from Coq.
-%
-In this work, we also mechanize the semantics in Coq and extract OCaml code.
-
-One notable approach to obtaining formal semantics with low manual effort is to
-automatically extract mechanized semantics from a structured natural language
-specification, as proposed and adopted in JavaScript by \citet{park2021jiset}.
-%
-However, this approach is not applicable to OpenQASM due to its lack of
-sufficient structure in the specification.
-
-When formal semantics is executable, it is a standard approach to validate its
-conformance using existing test suites.
-%
-Most traditional programming languages provide official conformance test
-suites, such as Test262~\cite{test262} for JavaScript and the Java
-Compatibility Kit~\cite{jck} for Java.
-%
-However, quantum computing is still in its early stages, and OpenQASM 2.0 does
-not have an official test suite.
-%
-Consequently, we assess the conformance of the semantics using example circuits
-from the OpenQASM 2.0 specification~\cite{openqasm2} and benchmark circuits
-from QASMBench~\cite{li2023qasmbench}.
-%
-Several studies have demonstrated the practical utility of executable semantics
-in identifying bugs in language
-implementations~\cite{watt2023wasmref,schumi2021spectest,park2021jest} and
-automatically generating test suites of high coverage~\cite{park2023feature}.
-
-\paragraph{Quantum Programming Languages}
-
-Researchers have proposed high-level quantum programming languages, which are
-complementary to quantum circuit description languages, such as OpenQASM, the
-main focus of this work.
-%
-While high-level languages allow programmers to express complex quantum
-algorithms using abstractions, circuit description languages are tightly
-coupled with real quantum hardware and often serve as the compilation targets
-for high-level languages.
-
-In the early days of quantum computing, quantum programming languages were
-defined by incorporating quantum computations into classical languages.
-%
-For instance, QCL~\cite{omer2003structured} and qGCL~\cite{sanders2000quantum}
-introduce imperative features alongside quantum computations, while quantum
-lambda calculus~\cite{maymin1997extending},
-QML~\cite{altenkirch2005functional}, and Quipper~\cite{green2013quipper} enable
-quantum computations within functional languages.
-
-Recently, researchers have proposed quantum programming languages to address
-specific difficulties in quantum programming.
-%
-Here, we discuss a few notable examples.
-%
-Silq~\cite{bichsel2020silq} enables programmers to safely and concisely perform
-\emph{uncomputation}, which means ignoring temporary qubits during computation
-without any side effects.
-%
-Twist~\cite{yuan2022twist} aids in reasoning about \emph{purity}, i.e., the
-absence of entanglement, through its novel type system.
-%
-Tower~\cite{yuan2022tower} allows the use of random-access memory for
-implementing pointer-based data structures.
-%
-Qunity~\cite{voichick2023qunity} generalizes language constructs in classical
-programming, thereby establishing a unified approach to quantum and classical
-computing.
-%
-\citet{Feng2021Hoare} propose a Hoare logic for quantum programs by defining
-classical-quantum states, which are capable of describing states of dynamic
-circuits.
-%
-\textsc{voqc}~\cite{Hietala2021voqc} is the first verified optimizer for quantum
-circuits, utilizing a language called \textsc{sqir}, compatible with OpenQASM.
-%
-Embedded in Coq, \textsc{sqir} facilitates the verification of \textsc{voqc}'s
-correctness.
-
-A notable exception is the work of \citet{paykin2017qwire}, which introduces
-\textsc{Qwire}, a quantum circuit description language that can be manipulated
-using a classical language.
-%
-Like our semantics, \textsc{Qwire}'s semantics is defined in terms of density
-matrices.
-%
-Its type system guarantees the preservation of the trace of the density matrix,
-ensuring that any well-typed circuit maintains a trace of 1 during execution.
-%
-This aligns with the last condition specified by \cref{pos:density}.
-%
-However, \textsc{Qwire} does not enforce the other properties required by the
-density matrix postulate, whereas in this work, we prove the physical
-consistency of the proposed semantics.
-
-\paragraph{Testing Quantum Programming Platforms}
-
-As mentioned in \Cref{ch:evaluation:rq2}, \textsc{QDiff}~\cite{wang2022qdiff}
-and MorphQ~\cite{paltenghi2023morphq} are differential testing and metamorphic
-testing techniques for quantum programming platforms, respectively.
-%
-Although these techniques are effective in identifying bugs in quantum
-programming platforms, all the identified bugs are due to software crashes, but
-not due to outcomes not following the correct probability distributions.
-%
-These approaches do not utilize a testing oracle to compute the correct
-probability distributions, thereby limiting their bug detection capabilities.
+\noindent

### 2024-11-19 paper/intro.tex
Fix problem

--- 2024-11-19 paper/1-intro.tex
@@
     \centerline{\includegraphics[width=15cm]{fig/testop}}
-    \caption[Enter the caption title here]{\texttt{testop} instruction} \label{fig:testop}
+    \caption[\texttt{testop} instruction in official specification document]
+      {\texttt{testop} instruction in official specification document}
+      \label{fig:testop}
     \centerline{\includegraphics[width=15cm]{fig/spectec-testop}}
-    \caption[Enter the caption title here]{SpecTec \texttt{testop} instruction} \label{fig:spectec-testop}
+    \caption[\texttt{testop} instruction in SpecTec specification document]
+      {\texttt{testop} instruction in SpecTec specification document}
+      \label{fig:spectec-testop}
 \end{figure}
@@
 The demanding standardization process places a significant burden on
-specification authors.
-Crafting this specification document is labor-intensive, and as Wasm evolves,
-the manual effort required becomes increasingly challenging to scale.
-Moreover, the dual requirement of maintaining both the formal notation and
-prose notation exacerbates these difficulties.
-The formal notation is written in LaTeX, and the prose notation is authored
-in reStructuredText; neither of which is particularly user-friendly for
-collaborative review.
-This lack of accessibility in the specification's tooling increases the
-likelihood of inconsistensies and errors, further complicating the
-standardization effort.
+specification authors, \red{particularly in maintaining consistency and correctness.
+Crafting the specification document is labor-intensive, which becomes
+increasingly difficult to scale as Wasm evolves.
+This challenge is compounded by the dual requirement to maintain both a formal
+notation, written in LaTeX, and a prose notation, authored in reStructuredText.
+The separate and user-unfriendly nature of these formats hinders collaborative
+review and increases the risk of discrepancies and errors between the two
+notations, further complicating the standardization process.}
 

### 2024-11-19 paper/intro.tex
Add reference

@@
 Beyond web development, Wasm's advantages are also deployed in areas such as
-edge computing~\cite{wasm-edge}, IoT~\cite{wasm-iot}, and
-blockchain~\cite{wasm-block}.
+cloud and edge computing~\cite{lucet, cloudflare}, IoT~\cite{wasm-iot}, and
+blockchains~\cite{wasm-blockchain}.
 % risk of implementation divergence

### 2024-11-20 paper/intro.tex
Update reference
Fix Korean title

@@
 two complementary forms: formal notation and prose notation.
-\red{\cref{fig:testop} illustrates the specification document for $testop$
+\cref{fig:testop} illustrates the specification document for $testop$
 instruction, where the upper part contains the prose notation, and the lower
-part contains the formal notation.}
+part contains the formal notation.
 The formal notation employs mathematical rules to succinctly define the
@@
 The demanding standardization process places a significant burden on
-specification authors, \red{particularly in maintaining consistency and correctness.
+specification authors, particularly in maintaining consistency and correctness.
 Crafting the specification document is labor-intensive, which becomes
@@
 review and increases the risk of discrepancies and errors between the two
-notations, further complicating the standardization process.}
+notations, further complicating the standardization process.
 
 % SpecTec
-To mitigate this problem, we \red{propose} SpecTec, a framework for mechanizing
-WebAssembly specification.
+To mitigate this problem, we propose SpecTec~\cite{spectec}, a framework for
+mechanizing WebAssembly specification.
 It provides domain specific language (DSL) that enables the declarative
 definition of Wasm syntax and semantics, akin to the formal notation.
-\red{
 SpecTec performs type-checking on the DSL to prevent meta-level specification
@@
 This bridges the gap between formal specification and executable code.
-Futhermore, SpecTec generates mechanized definitions for theorem provers.}
+Futhermore, SpecTec generates mechanized definitions for theorem provers.
 
 % challenge
-\red{
 One challenge of our approach is making AL executable.
@@
 running Wasm official test suites, particularly those related to control flow.
-}
 Additionally, the absence of a formalized definition or formal semantics for AL

### 2024-11-20 paper/spectec.tex
Change file name

% !TEX root = main.tex

\chapter{SpecTec}
\label{ch:spectec}
\noindent

\newcommand{\seq}[1]{#1^*}

%% Syntax
\section{Syntax of AL}
\label{syntax}

{
\renewcommand{\arraystretch}{0.8835}  % Decrease row spacing locally
\begin{align*}
\begin{array}{lcccrlr}
%
% Algorithm
  \text{Algorithm}\quad& A &\ni& a &::=& ~ \rel ~ (s, \seq e, \seq i) ~ | ~ \fun ~ (s, \seq e, \seq i) \\
%
% Instruction
  \text{Instruction}\quad& I &\ni& i &::=& ~ \ifi ~ (e, \seq i, \seq i) &\quad\text{(If)} \\
    &&&& | & ~ \eitheri ~ (\seq i, \seq i) &\quad\text{(Either)} \\
    &&&& | & ~ \enteri ~ (e, e) &\quad\text{(Enter)} \\
    &&&& | & ~ \pushctxi ~ e &\quad\text{(Push Context)} \\
    &&&& | & ~ \pushi ~ e &\quad\text{(Push)} \\
    &&&& | & ~ \popctxi ~ e &\quad\text{(Pop Context)} \\
    &&&& | & ~ \popi ~ e &\quad\text{(Pop)} \\
    &&&& | & ~ \popni ~ (e, e) &\quad\text{(Pop N)} \\
    &&&& | & ~ \popalli ~ e &\quad\text{(Pop All)} \\
    &&&& | & ~ \leti ~ (e, e) &\quad\text{(Let)} \\
    &&&& | & ~ \trapi &\quad\text{(Trap)} \\
    &&&& | & ~ \returnreli &\quad\text{(Return Relation)} \\
    &&&& | & ~ \returnfuni ~ e &\quad\text{(Return Function)} \\
    &&&& | & ~ \executei ~ e &\quad\text{(Execute)} \\
    &&&& | & ~ \calli ~ (s, s, \seq{e}) &\quad\text{(Let Call)} \\
    &&&& | & ~ \replaceframei ~ (p^+, e) &\quad\text{(Replace Frame)} \\
    &&&& | & ~ \replacestorei ~ (p^+, e) &\quad\text{(Replace Store)} \\
%
% Expression
  \text{Expression}\quad& E &\ni& e &::=& ~ \vare ~ s &\quad\text{(Variable)} \\
    &&&& | & ~ \nume ~ n &\quad\text{(Number)} \\
    &&&& | & ~ \boole ~ b &\quad\text{(Boolean)} \\
    &&&& | & ~ \fnamee ~ s &\quad\text{(Function Name)} \\
    &&&& | & ~ \liste ~ \seq e &\quad\text{(List)} \\
    &&&& | & ~ \stre ~ \seq{(s, e)} &\quad\text{(Record)} \\
    &&&& | & ~ \tupe ~ \seq e &\quad\text{(Tuple)} \\
    &&&& | & ~ \casee ~ (s, \seq e) &\quad\text{(Tagged Tuple)} \\
    &&&& | & ~ \une ~ (unop, e) &\quad\text{(Unary Operation)} \\
    &&&& | & ~ \bine ~ (binop, e, e) &\quad\text{(Binary Operation)} \\
    &&&& | & ~ \acce ~ (e, p) &\quad\text{(Access)} \\
    &&&& | & ~ \upde ~ (e, p^+, e) &\quad\text{(Update)} \\
    &&&& | & ~ \cate ~ (e, e) &\quad\text{(Concatenation)} \\
    &&&& | & ~ \compe ~ (e, e) &\quad\text{(Composition)} \\
    &&&& | & ~ \meme ~ (e, e) &\quad\text{(Membership)} \\
    &&&& | & ~ \choosee ~ e &\quad\text{(Choose)} \\
    &&&& | & ~ \lene ~ e &\quad\text{(Length)} \\
    &&&& | & ~ \iscaseofe ~ (e, s) &\quad\text{(Check Tag)} \\
    &&&& | & ~ \getcurctxe  &\quad\text{(Get Current Context)} \\
    &&&& | & ~ \ctxkinde ~ s &\quad\text{(Context Kind)} \\
    &&&& | & ~ \itere ~ (e, iter, \seq{s}) &\quad\text{(Iteration)} \\
    &&&& | & ~ \matche ~ (e, e) &\quad\text{(Match)} \\
    &&&& | & ~ \hastypee ~ (e, s) &\quad\text{(Has Type)} \\
\end{array}
\end{align*}
}

\newpage
\begin{align*}
\begin{array}{lcccrlr}
%
% Path
  \text{Path}\quad& P &\ni& p &::=& ~ \idxp ~ e ~ | ~ \slicep ~ (e, e) ~ | ~ \dotp ~ s \\
%
% Iter
  \text{Iter}\quad& Iter &\ni& iter &::=& ~ \listiter ~ | ~ \listniter ~ e ~|~ \listidxiter ~ (s, e) \\
%
% Operator
  \text{Unary Operator}\quad& Unop &\ni& unop &::=& ~ \notop ~ | ~ \minusop \\
  \text{Binary Operator}\quad& Binop &\ni& binop &::=& ~ \addop ~ | ~ \subop ~ | ~ \mulop \\
    &&&& | & \divop ~ | ~ \modop ~ | ~ \expop \\
    &&&& | & \implop ~ | ~ \equivop ~ | ~ \andop \\
    &&&& | & \orop ~ | ~ \eqop ~ | ~ \neop ~ | ~ \ltop \\
    &&&& | & \gtop ~ | ~ \leop ~ | ~ \geop \\
% Primitives
  \text{String}\quad& \mathbb S &\ni& s \\
  \text{Integer}\quad& \mathbb Z &\ni& n \\
  \text{Boolean}\quad& \mathbb B &\ni& b \\
\end{array}
\end{align*}

% algorithm
An algorithm is either an AL relation or an AL function, corresponding to the
relation and the function definitions in the DSL.
% instruction
An AL instruction is \ifi{} for conditionally choosing a branch, \eitheri{} for
nondeterministically choosing a branch, \enteri{} for entering a block,
\pushctxi{} for pushing a Wasm context, \pushi{} for pushing a Wasm value,
\popctxi{} for popping a Wasm context, \popi{} for popping a Wasm value,
\popni{} for popping $n$ Wasm values, \popalli{} for popping all Wasm values
within a Wasm context, \leti{} for let binding, \trapi{} for trapping,
\returnreli{} for returning from an AL relation, \returnfuni{} for returning
from an AL function, \executei{} for executing given Wasm instructions,
\calli{} for an AL function call with let binding, \replaceframei{} for
replacing a frame, or \replacestorei{} for replacing a store.
% expression
An expression is a variable, number, boolean, AL function name, list, record
(or struct), tuple, tagged tuple, unary or binary operation, data structure
access or update, list or record concatenation, membership check, element
selection, length retrieval, tag check, Wasm context retrieval, Wasm context
kind check, iteration, match relation, or type relation.

\begin{align*}
\begin{array}{lcccrlr}
%
% State
  \text{State}\quad& \Sigma &\ni& \sigma & ::=& ~ \seq a, w, k \\
%
% Wasm state
  \text{Wasm State}\quad& W &\ni& w &::=& ~ \seq{we}, \seq{wi}, sto \\
%
% Wasm Entry
  \text{Wasm Entry}\quad& WE &\ni& we &::=& ~ wv ~ | ~ wc \\
%
% Wasm Value
  \text{Wasm Value}\quad& WV &\ni& wv &::=& ~ \casev ~ (s, \seq v) \\
%
% Wasm Context
  \text{Wasm Context}\quad& WC &\ni& wc &::=& ~ \casev ~ (\text{``Label"}, \seq v) ~ | ~ \casev ~ (\text{``Frame"}, \seq v) \\
%
% Wasm Instruction
  \text{Wasm Instruction}\quad& WI &\ni& wi &::=& ~ \casev ~ (s, \seq v) \\
%
% Store
  \text{Store}\quad& Store &\ni& sto &::=& ~ \seq{(s, v)} \\
%
% Continuation
  \text{Continuation}\quad & K &\ni& k &::=& ~ \mt &\quad\text{(Empty)} \\
    &&&& | & ~ \toplevelcall ~ (s, \seq v) &\quad\text{(Top-level Call)} \\
    &&&& | & ~ \call ~ (s, s, \seq v, k) &\quad\text{(Call)} \\
    &&&& | & ~ \exe ~ (\seq{wi}, k) &\quad\text{(Execute)} \\
    &&&& | & ~ \wasm ~ (\seq{wc}, k) &\quad\text{(Wasm)} \\
    &&&& | & ~ \algo ~ c &\quad\text{(Algorithm)} \\
    &&&& | & ~ \ret ~ (v, k) &\quad\text{(Return)} \\
%
% Context
  \text{Context}\quad& C &\ni& c &::=& ~ (\seq i, \mu, \seq{wc}, k) \\
%
% Environment
  \text{Environment}\quad& M &\ni& \mu &::=& ~ \seq{[s \mapsto v]} \\
\end{array}
\end{align*}

\newpage
\begin{align*}
\begin{array}{lcccrlr}
%
% Value
  \text{Value}\quad& V &\ni& v &::=& ~ \numv ~ n &\quad\text{(Number)} \\
    &&&& | & ~ \boolv ~ b &\quad\text{(Boolean)} \\
    &&&& | & ~ \fnamev ~ s &\quad\text{(Function Name)} \\
    &&&& | & ~ \listv ~ \seq v &\quad\text{(List)} \\
    &&&& | & ~ \strv ~ \seq{(s, v)} &\quad\text{(Record)} \\
    &&&& | & ~ \tupv ~ \seq v &\quad\text{(Tuple)} \\
    &&&& | & ~ \casev ~ (s, \seq v) &\quad\text{(Tagged Tuple)} \\
    &&&& | & ~ \trapv &\quad\text{(Trap)} \\
    &&&& | & ~ \storev &\quad\text{(Store)} \\
\end{array}
\end{align*}

% State
An AL state consists of a sequence of algorithms, a Wasm state, and a
continuation.
% Wasm State
A Wasm state consists of an interleaved stack, a Wasm instruction stack, and a
store.
% Wasm Entry
An interleaved stack contains two types of entries: a Wasm value and a Wasm
context.
% Wasm Value & Wasm Context & Wasm Instruction
Each of a Wasm value, a Wasm context, and a Wasm instruction is expressed as an
AL value $\casev$.
In particular, the tag of a Wasm context is either ``Label" or ``Frame".
% Store
A store is expressed as a sequence of pairs, each consisting of a field name
and an AL value.
% Continuation
A continuation is \mt{} for an empty continuation, \toplevelcall{} for
top-level function call, \call{} for a function call with let binding, \exe{}
for executing given Wasm instructions, \wasm{} for executing a Wasm instruction
in the instruction stack, \algo{} for interpreting AL instructions in an
algorithm, or \ret{} for returning from an AL function.
% Env
An environment is a finite mapping from variable names to AL values.
% Context
An AL context consists of a sequence of Wasm instruction, an environment, a
sequence of Wasm contexts, and a continuation.
% Value
An AL value is a number, boolean, function name, list, record, tuple, tagged tuple,
trap, or store.

% Terminology
In this section, AL-specific terms (state, context, value, function, and
instruction) are referred to simply as state, context, value, function, and
instruction, while Wasm terms remain unchanged.

\newpage
%% Semantics
\section{Semantics of AL}
\label{semantics}

%% Continuation

\begin{gather*}
\boxed{\leadsto \subseteq \Sigma \times \Sigma} \\
%
% TopLevelCall
\newline \\
  \\
  \hline
  (\seq{a}, w, \toplevelcall ~ (s, \seq v)) \leadsto (\seq{a}, w, \createalgo(\seq a, s, \seq v, \mt)) \\
%
% Call
\newline \\
  \\
  \hline
  (\seq{a}, w, \call ~ (s_{var}, s_{name}, \seq v, k)) \leadsto
  (\seq{a}, w, \createalgo(\seq a, s_{name}, \seq v, \call ~ (s_{var}, s_{name}, \seq v, k))) \\
%
% Execute-empty
\newline \\
  \\
  \hline
  (\seq{a}, w, \exe ~ (\epsilon, k)) \leadsto (\seq{a}, w, k) \\
%
% Execute-instr
\newline \\
  \casev ~ (s, \seq v) = wi \\
  \hline
  (\seq{a}, w, \exe ~ (wi ~ \seq{wi}, k)) \leadsto
  (\seq{a}, w, \createalgo(\seq a, s, \seq v, \exe ~ (\seq{wi}, k))) \\
%
% Wasm-empty
\newline \\
  \\
  \hline
  (\seq{a}, w, \wasm ~ (\epsilon, k)) \leadsto (\seq{a}, w, k) \\
%
% Wasm-instr
\newline \\
  (w', \casev ~ (s, \seq v)) = \popwasminstr(w) \\
  \hline
  (\seq{a}, w, \wasm ~ (wc^+, k))
  \leadsto
  (\seq{a}, w', \createalgo(\seq a, s, \seq v, \wasm ~ (wc^+, k))) \\
%
% Al-empty
\newline \\
  \\
  \hline
  (\seq{a}, w, \algo ~ (\epsilon, \mu, \seq{wc}, k)) \leadsto (\seq{a}, w, k) \\
%
% Al-instr
\newline \\
  \seq{a}, w, (\seq{i_1}, \mu, \seq{wc}, k) \vdash i_0 \Rightarrow (w', k') \\
  \hline
  (\seq{a}, w, \algo ~ (i_0 ~ \seq{i_1}, \mu, \seq{wc}, k)) \leadsto (\seq{a}, w', k') \\
%
% Return
\newline \\
  \mu = \getenv(c) \qquad c' = \setenv(c, \mu[s_{var} \mapsto v]) \\
  \hline
  (\seq{a}, w, \ret ~ (v, \call ~ (s_{var}, s_{name}, \seq v, \algo ~ c)) \leadsto
  (\seq a, w, \algo ~ c') \\
\end{gather*}

The semantics of AL is defined by a state transition system.
% TopLevelCall
Given algorithms $\seq a$ generated from the DSL and a Wasm state $w$, a
top-level function $s$ can be invoked with arguments $\seq v$ to perform
either module instantiation or function invocation:
$(\seq a, w, \toplevelcall ~ (s, \seq v))$.
$\toplevelcall ~ (s, \seq v)${} transitions to \algo{} with the function name
$s$, the arguments $\seq v$, and \mt{}.
% Empty
\mt{} is an empty continuation, representing the end of the transition, so no
further transition occurs.
% Call
Similarly to the top-level call, $\call~ (s_{var}, s_{name}, \seq v, k)${}
transitionsto \algo{} with the function name $s_{name}$ and the arguments $\seq
v$, but with the current continuation nested within it.
% Execute
$\exe ~ (\seq{wi}, k)${} executes each Wasm instruction in the sequence
$\seq{wi}$ one by one, each containing a relation name and arguments.
The continuation transitions to \algo{} with the relation name, the arguments,
and the current continuation, with the Wasm instruction being popped.
% Wasm
In $\wasm ~ (\seq{wc}, k)${}, a Wasm instruction is popped from
the Wasm instruction stack and executed until the Wasm context sequence
$\seq{wc}$ is exhausted.
This execution transitions the continuation to \algo{}, with the relation name
and the arguments of the Wasm instruction, and the current continuation nested
within it.
% Algo
In $\algo ~ (\seq i, \mu, \seq{wc}, k)${}, the body instructions $\seq i$
execute sequentially, allowing transitions to \call{}, \exe{}, \wasm{}, \ret{},
or \algo{}.
% Return
The execution of a function call concludes with a return instruction, which
changes the continuation to $\ret ~ v${}.
If the function call originates from \call{}, the return value $v$ is assigned
to the variable $s_{var}$.
If it originates from \toplevelcall{}, the inner continuation $k$ is always
\mt, so the entire execution concludes with the return value $v$:
$
(\seq a, w, \toplevelcall ~ (s, \seq v))
\leadsto^*
(\seq a, w', \ret ~ (v, \mt))
$.

%% Instruction

\begin{gather*}
  \boxed{\seq{a}, ~ w, ~ c \vdash i \Rightarrow w, ~ k} \\
%
% If-true
\newline \\
  w, \getenv(c) \vdash e \Rightarrow v \qquad
  \istrue(v) \\
  \hline
  \seq{a}, w, c \vdash \ifi ~ (e, \seq{i_1}, \seq{i_2}) \Rightarrow
  (w, \algo ~ (\prependinstr(c, \seq{i_1}))) \\
%
% If-false
\newline \\
  w, \getenv(k) \vdash e \Rightarrow v \qquad
  \neg \istrue(v) \\
  \hline
  \seq{a}, w, c \vdash \ifi ~ (e, \seq{i_1}, \seq{i_2}) \Rightarrow
  (w, \algo ~ (\prependinstr(c, \seq{i_2}))) \\
%
% Either-1
\newline \\
  \\
  \hline
  \seq{a}, w, c \vdash \eitheri ~ (\seq{i_1}, \seq{i_2}) \Rightarrow
  (w, \algo ~ (\prependinstr(c, \seq{i_1}))) \\
%
% Either-2
\newline \\
  \\
  \hline
  \seq{a}, w, c \vdash \eitheri ~ (\seq{i_1}, \seq{i_2}) \Rightarrow
  (w, \algo ~ (\prependinstr(c, \seq{i_2}))) \\
%
% Enter
\newline \\
  (\seq{we}, \seq{wi}, sto) = w \qquad
  (i^*, \mu, wc_1 ~ ... ~ wc_n, k) = c \\
  w, \mu \vdash e_1 \Rightarrow wc \qquad
  w, \mu \vdash e_2 \Rightarrow \listv ~ \seq{wi_1} \\
  wi = \getendinstr(wc) \qquad
  wi_1 = \getendinstr(wc_1) \quad ... \quad wi_n = \getendinstr(wc_n) \\
  wi_2^* = wi_1^* ~ wi ~ wi_1 ~ ... ~ wi_n \\
  \hline
  \seq{a}, w, c \vdash \enteri ~ (e_1, e_2)
  \Rightarrow
  (
    (wc ~ \seq{we}, wi_2^* ~ \seq{wi}, sto),
    \wasm ~ (wc ~ wc_1 ~ ... ~ wc_n, \algo ~ (i^*, \mu, \epsilon, k))
  ) \\
%
% PushCtx
\newline \\
  w, \getenv(c) \vdash e \Rightarrow wc \\
  \hline
  \seq{a}, w, c \vdash \pushctxi ~ e
  \Rightarrow
  (\push(w, wc), \algo ~ (\addctx(c, wc))) \\
%
% Push
\newline \\
  w, \getenv(c) \vdash e \Rightarrow wv \\
  \hline
  \seq{a}, w, c \vdash \pushi ~ e \Rightarrow (\push(w, wv), \algo ~ c) \\
%
% PopCtx
\newline \\
  (\seq{we}, \seq{wi}, sto) = w \qquad
  (\seq{wv}, wc, \seq{we'}) = \splitctx(\seq{we}) \\
  \mu = \getenv(c) \cdot \assign(e, wc) \qquad
  \seq{wi'} = \exit(\seq{wi}) \qquad
  c' = \popwasmctx_{C}(c) \\
  \hline
  \seq{a}, w, c \vdash \popctxi ~ e
  \Rightarrow
  ((\seq{we'}, \seq{wi'}, sto), \algo ~ (\setenv(c', \mu))) \\
%
% Pop
\newline \\
  (w', wv) = \pop(w) \qquad
  \mu' = \getenv(c) \cdot \assign(e, wv) \\
  \hline
  \seq{a}, w, c \vdash \popi ~ e \Rightarrow (w', \algo ~ (\setenv(c, \mu'))) \\
%
% PopN
\newline \\
  \mu = \getenv(c) \qquad
  w, \mu \vdash e_2 \Rightarrow \numv ~ n \\
  (w', wv^n) = \popn(w, n) \qquad
  \mu' = \mu \cdot \assign(e_1, \listv ~ wv^n) \\
  \hline
  \seq{a}, w, c \vdash \popni ~ (e_1, e_2) \Rightarrow
  (w', \algo ~ (\setenv(c, \mu'))) \\
%
% PopAll
\newline \\
  (\seq{wv}, wc, \seq{we'}) = \splitctx(\seq{we}) \qquad
  \mu' = \getenv(c) \cdot \assign(e, \liste ~ \seq{wv}) \\
  \hline
  \seq{a}, (\seq{we}, \seq{wi}, sto), c \vdash \popalli ~ e
  \Rightarrow
  ((wc ~ \seq{we'}, \seq{wi}, sto), \algo ~ (\setenv(c, \mu'))) \\
%
% Let
\newline \\
  \mu = \getenv(c) \qquad
  w, \mu \vdash e_2 \Rightarrow v \qquad
  \mu' = \mu \cdot \assign(e_1, v) \\
  \hline
  \seq{a}, w, c \vdash \leti ~ (e_1, e_2)
  \Rightarrow
  (w, \setenv(c, \mu')) \\
%
% Trap
\newline \\
  \\
  \hline
  \seq{a}, w, c \vdash \trapi \Rightarrow (w, \ret ~ (\trapv, \mt)) \\
%
% Return-rule
\newline \\
  \\
  \hline
  \seq{a}, w, (\seq{i}, \mu, wc^*, k) \vdash \returnreli \Rightarrow (w, k) \\
%
% Return-func
\newline \\
  w, \mu \vdash e \Rightarrow v \\
  \hline
  \seq{a}, w, (\seq{i}, \mu, wc^*, k) \vdash \returnfuni ~ e \Rightarrow
  (w, \ret ~ (v, k)) \\
%
% Execute
\newline \\
  w, \getenv(c) \vdash e \Rightarrow \listv ~ wi^* \\
  \hline
  \seq{a}, w, c \vdash \executei ~ e \Rightarrow
  (w, \exe ~ (wi^*, \algo ~ c)) \\
%
% Call-fname
\newline \\
  \mu = \getenv(c) \qquad
  \fnamev ~ s_3 = \mu(s_2) \qquad
  w, \mu \vdash e_1 \Rightarrow v_1 \quad ... \quad
  w, \mu \vdash e_n \Rightarrow v_n \\
  \hline
  \seq{a}, w, c \vdash \calli ~ (s_1, s_2, e_1 ~ ... ~ e_n) \Rightarrow
  (w, \call ~ (s_1, s_3, v_1 ~ ... ~ v_n, \algo ~ c)) \\
%
% Call-fname
\newline \\
  \mu = \getenv(c) \qquad
  s_2 \not\in \domain(\mu) \qquad
  w, \mu \vdash e_1 \Rightarrow v_1 \quad ... \quad
  w, \mu \vdash e_n \Rightarrow v_n \\
  \hline
  \seq{a}, w, c \vdash \calli ~ (s_1, s_2, \seq e) \Rightarrow
  (w, \call ~ (s_1, s_2, v_1 ~ ... ~ v_n, \algo ~ c)) \\
%
% Replace-frame
\newline \\
\text{NOTE: $e_1$ is either store or current frame} \\
  \mu = \getenv(c) \qquad
  \casev ~ (\text{``Frame"}, v_1 ~ v_2) = \getcurframe(w) \qquad
  w, \mu \vdash e \Rightarrow v \\
  wc = \casev ~ (\text{``Frame"}, v_1 ~ \update(w, \mu, v_2, p^+, v)) \\
  \hline
  \seq{a}, w, c \vdash \replaceframei ~ (p^+, e) \Rightarrow
  (\setcurframe(w, wc), \algo ~ c) \\
%
% Replace-store
\newline \\
  \mu = \getenv(c) \qquad
  w, \mu \vdash e \Rightarrow v \\
  \strv ~ sto = \update(w, \mu, \strv ~ (\getstore(w)), p^+, v) \\
  \hline
  \seq{a}, w, c \vdash \replacestorei ~ (p^+, e)
  \Rightarrow (\setstore(w, sto), c) \\
\end{gather*}

Given algorithms $\seq a$, a Wasm state $w$, and a context $c$, executing an
instruction $i$ produces an updated Wasm state and a continuation.
% If
$\ifi ~ (e, i_1^*, i_2^*)${} evaluates $e$ and branches to $\seq i_1$ if the
result is true, or $\seq i_2$ otherwise.
% Either
$\eitheri ~ (i_1^*, i_2^*)${} nondeterministically selects either $i_1^*$ or
$i_2^*$.
% Enter
$\enteri ~ (e_1, e_2)${} evaluates $e_1$, pushing the resulting Wasm context
$wc$ to the interleaved stack.
It then evaluates $e_2$ to obtain the Wasm instruction sequence $wi_1^*$.
Then, the end instructions of the Wasm context $wc$ and the Wasm contexts
$wc_1 ~ ... ~ wc_n$ in the context $c$ are appended to $wi_1^*$ and pushed to
the Wasm instruction stack.
Finally, it changes the continuation to
$\wasm ~ (wc ~ wc_1 ~ ... ~ wc_n, \algo ~ (i^*, \mu, \epsilon, k))${}.
% PushCtx
$\pushctxi ~ e${} evaluates $e$, pushing the resulting Wasm context $wc$ to the
interleaved stack and the context $c$.
% Push
$\pushi ~ e${} evaluates $e$, pushing the resulting value $wv$ to the
interleaved stack.
% PopCtx
$\popctxi ~ e${} pops all Wasm values up to and including a Wasm context $wc$
from the interleaved stack and all Wasm instructions up to and including an end
instruction from the Wasm instruction stack.
It also pops a Wasm context from the context $c$.
Then, it assign the Wasm context $wc$ to $e$.
% Pop
$\popi ~ e${} pops a Wasm value $wv$ from the interleaved stack and assign the
value to $e$.
% PopN
$\popni ~ (e_1, e_2)${} evaluates $e_2$ to obtain a number $n$, pops $n$ Wasm
values from the interleaved stack, and assigns the values to $e_1$.
% PopAll
$\popalli ~ e${} pops all the Wasm values up to, but not including, a
Wasm context from the interleaved stack and assign the values to $e$.
% Let
$\leti ~ (e_1, e_2)${} evaluates $e_2$ and assigns the resulting value to
$e_1$.
% Trap
\trapi{} results in $\ret ~ (\trapv{}, \mt{})$ to terminate the execution.
% ReturnRel
\returnreli{} results in the continuation $k$ in the context $c$, indicating a
return from a relation algorithm.
% ReturnFunc
$\returnfuni ~ e${} evaluates $e$ to $v$, resulting in $\ret ~ (v, k)${} where
$k$ is the continuation in the context $c$.
% Execute
\executei{} evaluates $e$ to obtain a Wasm instruction sequence $wi^*$ and
result in $\exe ~ (wi^*, \algo ~ c)${}.
% Call
$\calli ~ (s_1, s_2, e_1 ~ ... ~ e_n)${} checks if $s_2$ is bound in the
environment.
If it is, it retrieves the function name $s_3$ from the environment $\mu$;
otherwise, it uses $s_2$ as the function name.
It evaluates $\seq e$ to $v^*$, resulting in $\call ~ (s_1, s_{name}, v_1 ~ ...
~ v_n)${} where $s_{name}$ is the function name.
% ReplaceFrame
\replaceframei{} retrieves the current frame $\casev ~ (\text{``Frame"}, v_1
v_2)${} in the Wasm context $w$.
It then evaluates $e$ to $v$, updating the value at location $v_2$ along paths
$p^+$ to $v$.
% ReplaceFrame
\replacestorei{} evaluates $e$ to $v$, updating the value at location $\strv ~
sto$ along paths $p^+$ to v.

%% Expression

\begin{gather*}
  \boxed{w, \mu \vdash e \Rightarrow v} \\
%
% Var
\newline \\
  v = \mu(s) \\
  \hline
  w, \mu \vdash \vare ~ s \Rightarrow v \\
%
% Num
\newline \\
  \hline
  w, \mu \vdash \nume ~ n \Rightarrow \numv ~ n \\
%
% Bool
\newline \\
  \hline
  w, \mu \vdash \boole ~ b \Rightarrow \boolv ~ b \\
%
% Fname
\newline \\
  \hline
  w, \mu \vdash \fnamee ~ s \Rightarrow \fnamev ~ s \\
%
% List
\newline \\
  w, \mu \vdash e_1 \Rightarrow v_1 \quad ... \quad
  w, \mu \vdash e_n \Rightarrow v_n \\
  \hline
  w, \mu \vdash \liste ~ e_1 ~ ... ~ e_n \Rightarrow \listv ~ v_1 ~ ... ~ v_n \\
%
% Str
\newline \\
  w, \mu \vdash e_1 \Rightarrow v_1 \quad ... \quad
  w, \mu \vdash e_n \Rightarrow v_n \\
  \hline
  w, \mu \vdash \stre ~ (s_1, e_1) ~ ... ~ (s_n, e_n) \Rightarrow
  \strv ~ (s_1, v_1) ~ ... ~ (s_n, v_n) \\
%
% Tup
\newline \\
  w, \mu \vdash e_1 \Rightarrow v_1 \quad ... \quad
  w, \mu \vdash e_n \Rightarrow v_n \\
  \hline
  w, \mu \vdash \tupe ~ e_1 ~ ... ~ e_n \Rightarrow \tupv ~ v_1 ~ ... ~ v_n \\
%
% Case
\newline \\
  w, \mu \vdash e_1 \Rightarrow v_1 \quad ... \quad
  w, \mu \vdash e_n \Rightarrow v_n \\
  \hline
  w, \mu \vdash \casee ~ (s, e_1 ~ ... ~ e_n) \Rightarrow \casev ~ (s, v_1 ~ ... ~ v_n) \\
%
% Un
\newline \\
  w, \mu \vdash e \Rightarrow v \\
  \hline
  w, \mu \vdash \une ~ (unop, e) \Rightarrow \unop(unop, v) \\
%
% Bin
\newline \\
  w, \mu \vdash e_1 \Rightarrow v_1 \qquad w, \mu \vdash e_2 \Rightarrow v_2 \\
  \hline
  w, \mu \vdash \bine ~ (binop, e_1, e_2) \Rightarrow \binop(binop, v_1, v_2) \\
%
% Acc
\newline \\
  w, \mu \vdash e \Rightarrow v \\
  \hline
  w, \mu \vdash \acce ~ (e, p) \Rightarrow \access(w, \mu, v, p) \\
%
% Upd
\newline \\
  w, \mu \vdash e_1 \Rightarrow v_1 \qquad w, \mu \vdash e_2 \Rightarrow v_2 \\
  \hline
  w, \mu \vdash \upde ~ (e_1, p^+, e_2) \Rightarrow \update(w, \mu, v_1, p^+, v_2) \\
%
% Cat
\newline \\
   w, \mu \vdash e_1 \Rightarrow \listv ~ \seq{v_1} \qquad
   w, \mu \vdash e_2 \Rightarrow \listv ~ \seq{v_2} \\
  \hline
  w, \mu \vdash \cate ~ (e_1, e_2) \Rightarrow \listv ~ (\seq{v_1} ~ \seq{v_1}) \\
%
% Comp
\newline \\
   w, \mu \vdash e_1 \Rightarrow \strv ~ \seq{(s_1, v_1)} \qquad
   w, \mu \vdash e_2 \Rightarrow \strv ~ \seq{(s_2, v_2)} \\
  \hline
  w, \mu \vdash \compe ~ (e_1, e_2) \Rightarrow \strv ~ (\seq{(s_1, v_1)} ~ \seq{(s_2, v_2)}) \\
%
% Mem
\newline \\
  w, \mu \vdash e_1 \Rightarrow v_1 \qquad
  w, \mu \vdash e_2 \Rightarrow \listv ~ \seq{v_2} \\
  \hline
  w, \mu \vdash \meme ~ (e_1, e_2) \Rightarrow \boolv ~ (v_1 \in \seq{v_2}) \\
%
% Choose
\newline \\
  w, \mu \vdash e \Rightarrow \listv ~ \seq{v} \qquad
  v \in \seq{v} \\
  \hline
  w, \mu \vdash \choosee ~ e \Rightarrow v \\
%
% Len
\newline \\
  w, \mu \vdash e \Rightarrow \listv ~ \seq{v} \\
  \hline
  w, \mu \vdash \lene ~ e \Rightarrow \numv ~ |\seq v| \\
%
% IsCaseOf
\newline \\
  w, \mu \vdash e \Rightarrow \casev ~ (s', \seq{v}) \\
  \hline
  w, \mu \vdash \iscaseofe ~ (e, s) \Rightarrow \boolv ~ (s = s') \\
%
% GetCurCtx
\newline \\
  wc = \getcurctx(w) \\
  \hline
  w, \mu \vdash \getcurctxe \Rightarrow wc \\
%
% CtxKind
\newline \\
  \casev (s', \seq v) = \getcurctx(w) \\
  \hline
  w, \mu \vdash \ctxkinde ~ s \Rightarrow \boolv ~ (s = s') \\
%
% Iter-
\newline \\
  \,~\mu(s_1) = \listv ~ (v_{11} ~ ... ~ v_{1n}) \quad ... \quad
  \mu(s_m) = \listv ~ (v_{m1} ~ ... ~ v_{mn}) \\
  \mu_1 = \mu[s_1 \mapsto v_{11}, ~ ... ~ s_m \mapsto v_{m1}] \quad ... \quad
  \mu_n = \mu[s_1 \mapsto v_{1n}, ~ ... ~ s_m \mapsto v_{mn}] \\
  w, \mu_1 \vdash e \Rightarrow v_1 \quad ... \quad
  w, \mu_n \vdash e \Rightarrow v_n \\
  \hline
  w, \mu \vdash \itere ~ (e, \listiter, s_1 ~ ... ~ s_m) \Rightarrow
  \listv ~ v_1 ~ ... ~ v_n \\
%
% Iter-n
\newline \\
  w, \mu \vdash e_2 \Rightarrow \numv ~ n \\
  \,~\mu(s_1) = \listv ~ (v_{11} ~ ... ~ v_{1n}) \quad ... \quad
  \mu(s_m) = \listv ~ (v_{m1} ~ ... ~ v_{mn}) \\
  \mu_1 = \mu[s_1 \mapsto v_{11}, ~ ... ~ s_m \mapsto v_{m1}] \quad ... \quad
  \mu_n = \mu[s_1 \mapsto v_{1n}, ~ ... ~ s_m \mapsto v_{mn}] \\
  w, \mu_1 \vdash e \Rightarrow v_1 \quad ... \quad
  w, \mu_n \vdash e \Rightarrow v_n \\
  \hline
  w, \mu \vdash \itere ~ (e_1, \listniter ~ e_2, s_1 ~ ... s_m) \Rightarrow
  \listv ~ v_1 ~ ... ~ v_n \\
%
% Iter-idx
\newline \\
  w, \mu \vdash e_2 \Rightarrow \numv ~ n \\
  \mu_1 = \mu[s_1 \mapsto \numv ~ 0, ~ ... ~ s_m \mapsto \numv ~ 0] \quad ... \quad
  \mu_n = \mu[s_1 \mapsto \numv ~ n, ~ ... ~ s_m \mapsto \numv ~ n] \\
  w, \mu_1 \vdash e \Rightarrow v_1 \quad ... \quad
  w, \mu_n \vdash e \Rightarrow v_n \\
  \hline
  w, \mu \vdash \itere ~ (e_1, \listidxiter ~ e_2, s_1 ~ ... ~ s_m) \Rightarrow
  \listv ~ v_1 ~ ... ~ v_n \\
%
% Match
\newline \\
  w, \mu \vdash e_1 \Rightarrow v_1 \qquad
  w, \mu \vdash e_2 \Rightarrow v_2 \\
  \hline
  w, \mu \vdash \matche ~ (e_1, e_2) \Rightarrow \boolv ~ (\match(v_1, v_2)) \\
%
% HasType
\newline \\
  w, \mu \vdash e \Rightarrow v \\
  \hline
  w, \mu \vdash \hastypee ~ (e, s) \Rightarrow \boolv ~ (\hastype(v, s)) \\
\end{gather*}

Given a Wasm state $w$ and an environment $\mu$, an expression $e$ evaluates
to a value.
% Var
$\vare ~ s${} looks up the environment $\mu$ and gets the value.
% Num Bool Fname
$\nume ~ n${}, $\boole ~ b${}, and $\fnamee ~ s${} directly reduce to $\numv ~
n${}, $\boolv ~ b${}, and $\fnamev ~ s${}, respectively.
% Tuple Case List Struct
$\liste ~ e_1 ~ ... ~ e_n${}, $\stre ~ (s_1, e_1) ~ ... ~ (s_n, e_n)${}, $\tupe
~ e_1 ~ ... ~ e_n${}, and $\casee ~ (s, e_1 ~ ... ~ e_n)${} evaluate their
sub-expressions and reduce to $\listv ~ v_1 ~ ... ~ v_n${}, $\strv ~ (s_1, v_1)
~ ... ~ (s_n, v_n)${}, $\tupv ~ v_1 ~ ... ~ v_n${}, and $\casev ~ (s, v_1 ~ ...
~ v_n)${}, respectively.
% Un Bin
$\une ~ (unop, e)${} and $\bine ~ (binop, e_1, e_2)${} evaluate their
sub-expressions and perform \unop{} and \binop{}, respectively.
% Acc
$\acce ~ (e, p)${} evaluates $e$ and accesses the resulting value $v$ using a
path $p$.
% Upd
$\upde ~ (e_1, p^+, e_2)${} evaluates $e_1$ and $e_2$ to $v_1$ and $v_2$,
updating the value at location $v_1$ along paths $p^+$ to the value $v_2$.
% Cat Comp
$\cate ~ (e_1, e_2)${} and $\compe ~ (e_1, e_2)${} each evaluate two
sub-expressions $e_1$ and $e_2$, concatenating the resulting lists and
resulting records, respectively.
% Mem
$\meme ~ (e_1, e_2)${} evaluates $e_1$ and $e_2$ to $v_1$ and $\listv ~ \seq
v_2$, and checks whether $v_1$ is a member of $\seq v_2$.
% Choose Len
$\choosee ~ e$ and $\lene ~ e${} each evaluate $e$ to the list value $\listv ~
\seq v$, choosing an element from the list and calculating the length of the
list, respectively.
% IsCaseOf
$\iscaseofe ~ (e, s)${} evaluates $e$ to the tagged tuple $\casev ~ (s', \seq
v)$ and checks whether the tag $s'$ matches $s$.
% GetCurCtx
\getcurctxe{} retrieves the current Wasm context from the interleaved stack.
% CtxKind
$\ctxkinde ~ s${} retrieves the current Wasm context $\casev ~ (s', \seq v)$
and checks whether the kind $s'$ matches $s$.
% Iter
$\itere ~ (e, iter, s^*)${} behaves differently depending on $iter$.
\listiter{} and $\listniter ~ e_2${} are essentially equivalent.
They assume that all the values mapped to the variables $\seq s$ by the
environment $\mu${} are list values with the same length $n$.
$\listniter ~ e_2${} checks whether the number $n$ matches the result of
evaluating $e_2$, while \listiter{} allows any number.
It generates a sequence of environments $\mu_1 ~ ... ~ \mu_n$ where the k-th
environment maps each variable to the k-th elements of each list.
Each environment is used to evaluate $e_1$, producing a sequence of values $v_1
~ ... ~ v_n$.
Finally, it results in $\listv ~ v_1 ~ ... ~ v_n$.
$\listidxiter ~ e_2${} behaves similarly except that the variables $\seq s$ are
unbound.
It evaluates $e_2$ to obtain $n$ and then behaves like $\listniter ~ e_2${} as
if a list of values [0 ... n-1] were bound to the variables $s^*$.
% IsValid Match HasType
$\matche ~ (e_1, e_2)${} and $\hastypee ~ (e, s)${} evaluate their
sub-expressions and perform \match{} and \hastype{}, respectively.
Note that \match{} and \hastype{} represent the match relation and the type
relation in the Wasm static semantics.
Their formalization is omitted as they are highly specific to Wasm and
excessively verbose.

%% Helper functions

\begin{align*}
%
% get_algo_name
\newline \\
  &\getalgoname(\rel ~ (s, \seq e, \seq i)) = s \\
  &\getalgoname(\fun ~ (s, \seq e, \seq i)) = s \\
%
% lookup
\newline \\
  &\lookup(\epsilon, s) = \epsilon \\
  &\lookup(a ~ \seq{a}, s) =
    \begin{cases}
      a &\quad\quad \premise{if}\quad ~ s = \getalgoname(a) \\
      \lookup(\seq{a}, s) &\quad\quad \premise{otherwise}
    \end{cases}
  \\
%
% assign
\newline \\
  &\assign(\vare ~ s, v) = [s \mapsto v] \\
  &\assign(\itere ~ (e, iter, s^*), v) = \assign(e, v) \\
  &\assign(\tupe ~ e_1 ~ ... ~ e_n, \tupv ~ v_1 ~ ... ~ v_n) =
    \assign(e_1, v_1) \cdot ~ ... ~ \cdot \assign(e_n, v_n) \\
  &\assign(\liste ~ e_1 ~ ... ~ e_n, \listv ~ v_1 ~ ... ~ v_n) =
    \assign(e_1, v_1) \cdot ~ ... ~ \cdot \assign(e_n, v_n) \\
  &\assign(\casee ~ (s, e_1 ~ ... ~ e_n), \casev ~ (s, v_1 ~ ... ~ v_n)) =
    \assign(e_1, v_1) \cdot ~ ... ~ \cdot \assign(e_n, v_n) \\
  &\assign(
    w,
    \mu,
    \stre ~ (s_1, e_1) ~ ... ~ (s_2, e_n),
    \strv ~ (s_1, v_1) ~ ... ~ (s_n, v_n)
  ) =
    \assign(e_1, v_1) \cdot ~ ... ~ \cdot \assign(e_n, v_n) \\
  &\assign(\cate ~ (\liste ~ e_1 ~ ... ~ e_m, e), \listv v_1 ~ ... ~ v_n) =
    \assign(e_1, v_1) \cdot ~ ... ~ \cdot \assign(e_m, v_m) \cdot
    \assign(e, \listv ~ v_{m+1} ~ ... ~ v_n) \\
  &\assign(\cate ~ (e, \liste ~ e_1 ~ ... ~ e_m), \listv v_1 ~ ... ~ v_n) =
    \assign(e, \listv ~ v_1 ~ ... ~ v_{n-m}) \cdot
    \assign(e_1, v_{n-m+1}) \cdot ~ ... ~ \cdot \assign(e_m, v_n) \\
%
% create_algo
\newline \\
  &\createalgo(\seq a, s, \seq v, k) =
  \algo(\seq i, \mu, \epsilon, k) \\
  &\qquad\qquad\qquad\qquad\premise{if}\quad
  (\rel ~ (s, \seq e, \seq i) \quad\lor\quad \fun ~ (s, \seq e, \seq i)) = \lookup(\seq a, s)
  \quad\land\quad
  \mu = \assign(w, [\text{``s"} \mapsto \storev], \seq e, \seq v) \\
%
% split_ctx
\newline \\
  &\splitctx(\epsilon) = (\epsilon, \epsilon, \epsilon) \\
  &\splitctx(we ~ \seq{we}) =
    \begin{cases}
      (\epsilon, wc, \seq{we})
      &\quad\quad \premise{if}\quad
      we = wc \\
      (wv ~ \seq{wv}, wc, \seq{we'})
      &\quad\quad \premise{if}\quad
      we = wv \quad\land\quad
      (\seq{wv}, wc, \seq{we'}) = \splitctx(\seq{we}) \\
    \end{cases}
  \\
%
% pop-wasm-ctx_context
\newline \\
  &\popwasmctx_{C}((\seq i, \mu, epsilon, k)) = (\seq i, \mu, \epsilon, \popwasmctx_K(k)) \\
  &\popwasmctx_{C}((\seq i, \mu, wc ~ \seq{wc}, k)) = (\seq i, \mu, wc^*, k) \\
%
\newline \\
% pop-wasm-ctx_cont-mt
  &\popwasmctx_{K}(\mt) = \epsilon \\
% pop-wasm-ctx_cont-toplevelcall
  &\popwasmctx_{K}(\toplevelcall ~ (s, v^*)) = \epsilon \\
% pop-wasm-ctx_cont-call
  &\popwasmctx_{K}(\call ~ (s_1, s_2, \seq v, k)) =
    \call ~ (s_1, s_2, \seq v, (\popwasmctx_K(k)) \\
% pop-wasm-ctx_cont-execute
  &\popwasmctx_{K}(\exe ~ (v, k)) = \exe ~ (v, (\popwasmctx_K(k)) \\
% pop-wasm-ctx_cont-wasm_0
  &\popwasmctx_{K}(\wasm ~ (\epsilon, k)) = \wasm ~ (\epsilon, \popwasmctx_{K}(k)) \\
% pop-wasm-ctx_cont-wasm_nonzero
  &\popwasmctx_{K}(\wasm ~ (wc ~ \seq{wc}, k)) = \wasm ~ (\seq{wc}, k) \\
% pop-wasm-ctx_cont-al
  &\popwasmctx_{K}(\algo ~ c) = \algo ~ (\popwasmctx_C(c)) \\
% pop-wasm-ctx_cont-ret
  &\popwasmctx_{K}(\ret ~ (v, k)) = \ret ~ (v, \popwasmctx_K(k)) \\
%
% exit
\newline \\
  &\exit(wi ~ \seq{wi}) =
    \begin{cases}
      \seq{wi} &\quad\quad \premise{if}\quad ~ \isendinstr(wi) \\
      \exit(\seq{wi}) &\quad\quad \premise{otherwise} \\
    \end{cases}
  \\
%
% get_env
\newline \\
  &\getenv((\seq i, \mu, wc^*, k)) = \mu \\
%
% set_env
\newline \\
  &\setenv((\seq i, \mu, wc^*, k), \mu') = (\seq i, \mu', wc^*, k) \\
%
% prepend_instr
\newline \\
  &\prependinstr((\seq i, \mu, wc^*, k), \seq{i'}) = (\seq{i'} ~ \seq i, \mu, wc^*, k) \\
%
% add_ctx
\newline \\
  &\addctx((\seq i, \mu, wc^*, k), wc') = (\seq i, \mu, wc' ~ wc^*, k) \\
%
% get_store
\newline \\
  &\getstore((\seq{we}, \seq{wi}, sto)) = sto \\
%
% set_store
\newline \\
  &\setstore((\seq{we}, \seq{wi}, sto), sto') = (\seq{we}, \seq{wi}, sto') \\
%
% pop_wasm_instr
\newline \\
  &\popwasminstr((\seq{we}, wi ~ \seq{wi}, sto)) = ((\seq{we}, \seq{wi}, sto), wi) \\
%
% push
\newline \\
  &\push((\seq{we}, \seq{wi}, sto), we) = (we ~ \seq{we}, \seq{wi}, sto) \\
%
% pop
\newline \\
  &\pop((we ~ \seq{we}, \seq{wi}, sto)) = ((\seq{we}, \seq{wi}, sto), we) \\
%
% popn
\newline \\
  &\popn((we^n ~ \seq{we}, \seq{wi}, sto), n) = ((\seq{we}, \seq{wi}, sto), we^n) \\
%
% unop
\newline \\
  &\unop(\notop, v) =
    \begin{cases}
      \boolv ~ true &\quad\quad\premise{if}\quad ~ \istrue(v) \\
      \boolv ~ false &\quad\quad\premise{otherwise} \\
    \end{cases}
  \\
  &\unop(\minusop, \numv ~ n) = \numv ~ (-n) \\
%
% get_cur_ctx
\newline \\
  &\getcurctx((\seq{we}, \seq{wi}, sto)) =
  wc \quad\quad\premise{if}\quad ~ (\seq{wv}, wc, \seq{we'}) = \splitctx(\seq{we}) \\
%
% get_cur_frame-frame
\newline \\
  &\getcurframe((\seq{we}, \seq{wi}, sto)) =
    \begin{cases}
      wc &\quad\quad\premise{if}\quad
      (\seq{wv}, wc, \seq{we'}) = \splitctx(\seq{we}) \quad\land\quad \isframe(wc)\\
      \getcurframe(\seq{we'}) &\quad\quad\premise{if}\quad
      (\seq{wv}, wc, \seq{we'}) = \splitctx(\seq{we}) \quad\land\quad \neg \isframe(wc) \\
    \end{cases}
  \\
%
% set_cur_frame-frame
\newline \\
  &\setcurframe((\seq{we}, \seq{wi}, sto), wc_{frame})
  =
  \begin{cases}
    (\seq{wv} ~ wc_{frame} ~ \seq{we'}, \seq{wi}, sto) \\
    \qquad\qquad\qquad\quad\premise{if}\quad
    (\seq{wv}, wc, \seq{we'}) = \splitctx(\seq{we}) \quad \isframe(wc) \\
    (\seq{wv} ~ wc ~ \setcurframe(\seq{we'}, wc_{frame}), \seq{wi}, sto) \\
    \qquad\qquad\qquad\quad\premise{if}\quad
    (\seq{wv}, wc, \seq{we'}) = \splitctx(\seq{we}) \quad \neg \isframe(wc) \\
  \end{cases}
  \\
%
% access-idx
\newline \\
  &\access(w, \mu, \listv ~ \seq{v}, \idxp ~ e) = \seq{v}[n]
  \qquad\qquad\qquad~\,\quad\premise{if}\quad w, \mu \vdash e \Rightarrow \numv ~ n \\
% access-slice
  &\access(w, \mu, \listv ~ \seq{v}, \slicep ~ (e_1, e_2)) = \seq{v}[n_1: n_2]
  \quad\quad\premise{if}\quad
  w, \mu \vdash e_1 \Rightarrow \numv ~ n_1 \quad\land\quad
  w, \mu \vdash e_2 \Rightarrow \numv ~ n_2 \\
% access-dot
  &\access(w, \mu, \strv ~ ((s_0, v_0) ~ \seq{(s_1, v_1)}), \dotp ~ s) =
  \begin{cases}
    v_0 &\quad\quad\premise{if}\quad s_0 = s \\
    \access(w, \mu, \strv ~ \seq{(s_1, v_1)}, \dotp ~ s) &\quad\quad\premise{otherwise} \\
  \end{cases}
  \\
% access-store
  &\access(w, \mu, \storev, p) =\access(w, \mu, \strv ~ sto, p)
  \quad\quad\premise{if}\quad w = (\seq{we}, \seq{wi}, sto)\\
%
% update
\newline \\
  % idx
  &\update(w, \mu, \listv ~ \seq{v}, (\idxp ~ e) ~ \seq{p}, v) =
  \listv ~ (\updateidx(\seq v, n, \update(w, \mu, \seq{v}[n], \seq{p}, v))) \\
  &\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\premise{if}\quad
  w, \mu \vdash e \Rightarrow \numv ~ n \\
  % slice
  &\update(w, \mu, \listv ~ \seq{v}, (\slicep ~ e) ~ \seq{p}, v) =
  \listv ~ (\updateslice(\seq v, n_1, n_2, \update(w, \mu, \seq{v_i}[n_1: n_2], \seq{p}, v))) \\
  &\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\premise{if}\quad
  w, \mu \vdash e_1 \Rightarrow \numv ~ n_1 \quad\land\quad
  w, \mu \vdash e_2 \Rightarrow \numv ~ n_2 \\
  % dot
  &\update(w, \mu, \strv ~ \seq{(s, v)}, (\dotp ~ s') ~ \seq{p}, v)
  =
  \strv ~ (\updatedot(\seq{(s, v)}, s', \update(w, \mu, \access(v, \dotp ~ s'), \seq{p}, v))) \\
%
% update_idx
\newline \\
  &\updateidx(v_0 ~ \seq{v_1}, n, v_2) =
  \begin{cases}
    v_2 ~ \seq{v_1}
    &\quad\quad\premise{if}\quad n = 0 \\
    v_0 ~ \updateidx(\seq{v_1}, n-1, v_2)
    &\quad\quad\premise{if}\quad n > 0 \\
  \end{cases} \\
%
% update_slice
\newline \\
  &\updateslice(v_0^n ~ \seq{v_1}, 0, n, v_2^n) = v_2^n ~ \seq{v_1} \\
  &\updateslice(v_0 ~ \seq{v_1}, m, n, v_2^n) =  v_0 ~ \updateslice(\seq{v_1}, m-1, n, v_2)
  \quad\quad\premise{if}\quad m > 0 \\
%
% update_dot
\newline \\
  &\updatedot((s_0, v_0) ~ \seq{(s_1, v_1)}, s_2, v_2) =
  \begin{cases}
    (s_2, v_2) ~ \seq{(s_1, v_1)}
    &\quad\quad\premise{if}\quad s_0 = s_2 \\
    (s_0, v_0) ~ \updatedot(\seq{(s_1, v_1)}, s_2, v_2)
    &\quad\quad\premise{otherwise}
  \end{cases}
  \\
%
% get_end_algo
\newline \\
  &\getendinstr(wc) =
  \begin{cases}
    \casev ~ (\text{``Exiting"}, \epsilon) &\quad\quad\premise{if}\quad wc = \casev ~ (\text{``Label"}, \seq v)\\
    \casev ~ (\text{``Returning"}, \epsilon) &\quad\quad\premise{if}\quad wc = \casev ~ (\text{``Frame"}, \seq v)
  \end{cases} \\
%
% is_end_instr
\newline \\
  &\isendinstr(wi) =
  \begin{cases}
    true &\quad\quad\premise{if}\quad wi = \casev ~ (\text{``Exiting"}, \epsilon) \quad\lor\quad wi = \casev ~ (\text{``Returning"}, \epsilon)\\
    false &\quad\quad\premise{otherwise}
  \end{cases} \\
%
% is_true
\newline \\
  &\istrue(v) =
  \begin{cases}
  b &\quad\quad\premise{if}\quad v = \boolv ~ b \\
    \istrue(v_1) ~ \land ~ ... ~ \land ~ \istrue(v_n) &\quad\quad\premise{if}\quad
    v = \listv ~ v_1 ~ ... ~ v_n \\
  \end{cases} \\
%
% is_frame
\newline \\
  &\isframe(v) =
  \begin{cases}
    true
    &\quad\quad\premise{if}\quad v = \casev ~ (\text{"FRAME"}, \seq v) \\
    false &\quad\quad\premise{otherwise}\\
  \end{cases}
  \\
\end{align*}

### 2024-11-20 paper/background.tex
Rename files
WIP: write overview

--- 2024-11-04 paper/2-background.tex
@@
 definitions such as functions, globals, and memories.
-A module is instantiated to validate the module, allocate each definition to
+A module is instantiated to validate the module and allocate each definition to
 the data structure named \textit{store}.
-This results in module instance, a runtime representation for the module.
+This results in a module instance, a runtime representation for the module.
 After the instantiation, functions can be invoked so that the function body,
 which is a sequence of instructions, is executed.
-
-\begin{figure}[t]
-    \centerline{\includegraphics[width=15cm]{fig/testop}}
-    \caption[Enter the caption title here]{\texttt{testop} instruction} \label{fig:testop}
-    \centerline{\includegraphics[width=15cm]{fig/spectec-testop}}
-    \caption[Enter the caption title here]{SpecTec \texttt{testop} instruction} \label{fig:spectec-testop}
-\end{figure}
 
@@
 \cref{sec:control flow in official prose}.
-
-\section{SpecTec}
-\label{sec:spectec}
-
-% SpecTec: specification mechanization tool
-SpecTec is a tool for \textit{mechanizing} WebAssembly specification
-~\cite{spectec}.
-Specification mechanization is a technique that treats a specification as data
-that can be manipulated by a computer to automatically generate parser and
-interpreter ~\cite{jiset}, to generate tests and perform differential tests
-~\cite{jest}, to find meta-level errors in the specification ~\cite{jstar}, and
-to perform meta-level static analysis for the static analysis of the defined
-language ~\cite{jsaver}.
-This technique is especially crucial in growing languages such as JavaScript
-and WebAssembly.
-
-% DSL & generated document
-SpecTec provides a domain specific language (DSL) for the specification, and it
-is a declarative language that resembles textbook style notation.
-It is type checked to prevent meta-level errors such as notation misuses and
-dimension mismatches. The code below is the specification for \texttt{testop}
-instruction written in the DSL, which is designed to be similar to the formal
-notation, while obtaining better user experience. \\
-\red{TODO: better character for $\sim$ in the code}
-\begin{verbatim}
-  rule Step_pure/testop:
-  (CONST nt c_1) (TESTOP nt testop)  ~>  (CONST I32 c)
-  -- if c = $testop_(nt, testop, c_1)
-\end{verbatim}
-
-% prose language
-However, The DSL cannot generate the prose notation just as it is because it is
-declarative.
-Therefore, SpecTec also defines another imperative language for the prose
-notation, and translates the DSL into the prose language to generate the prose
-notation.
-\cref{fig:spectec-testop} is the specification document automatically
-generated from the DSL which depicts both formal notation and prose noatation.
-The generated document closely resembles the official document except for some
-minor notation changes and some missing hyperlinks.
-
-% meta-level interpreter
-What is especially intriguing here is that the prose language is executable.
-It can perform each step written in the prose program one by one, which
-WebAssembly interpreter should follow.
-As a result, the prose program generated from the DSL serves as a WebAssembly
-interpter.
-That is to say, if WebAssembly semantics is written in the DSL, SpecTec
-automatically generates the WebAssembly interpreter following the written
-semantics.
-The generated interpreter passes all the execution tests in the official test
-suites except for the tests related to the infinte loop.
-
-% other backends and other language
-SpecTec also translates the DSL into Coq for mathematical proofs and is working
-on generating other artifacts, such as test suites required for WebAssembly
-standardization.
-Additionally, research is ongoing to extend SpecTec for mechanizing other
-languages ~\cite{p4-cherry-workshop}.

### 2024-11-20 paper/overview.tex
WIP: write overview

% !TEX root = main.tex

\chapter{Overview}
\label{ch:overview}
\noindent

\begin{figure}[h]
  \centerline{\includegraphics[width=15cm]{fig/overview}}
  \caption[An overview of the SpecTec architecture]
    {An overview of the SpecTec architecture}
    \label{fig:overview}
\end{figure}

We propose SpecTec ...

% Overview explanation
In this chapter, we explain an overall approach to generate multiple artifacts
from the specification and check correctness of the specification.
\cref{fig:overview} illustrates the overview of our approach.
SpecTec provides a domain specific language (DSL) for the specification.
It is designed to be similar to the formal notation when describing the
semantics with compact and user-friendly notation.
It is type checked to prevent meta-level errors such as notation misuses and
dimension mismatches.
The code below is the specification for \texttt{testop} instruction written in
the DSL.
\begin{lstlisting}
  rule Step_pure/testop:
  (CONST nt c_1) (TESTOP nt testop)  ~>  (CONST I32 c)
  -- if c = $testop_(nt, testop, c_1)
\end{lstlisting}

### 2024-11-20 paper/spectec.tex
WIP: write overview

@@
 \newcommand{\seq}[1]{#1^*}
+
+\section{SpecTec}
+\label{sec:spectec}
+
+% SpecTec: specification mechanization tool
+SpecTec is a tool for \textit{mechanizing} WebAssembly specification
+~\cite{spectec}.
+Specification mechanization is a technique that treats a specification as data
+that can be manipulated by a computer to automatically generate parser and
+interpreter ~\cite{jiset}, to generate tests and perform differential tests
+~\cite{jest}, to find meta-level errors in the specification ~\cite{jstar}, and
+to perform meta-level static analysis for the static analysis of the defined
+language ~\cite{jsaver}.
+This technique is especially crucial in growing languages such as JavaScript
+and WebAssembly.
+
+% DSL high-level explanation
+SpecTec provides a domain specific language (DSL) for the specification.
+The DSL is a declarative language that resembles textbook style notation.
+It is type checked to prevent meta-level errors such as notation misuses and
+dimension mismatches. The code below is the specification for \texttt{testop}
+instruction written in the DSL, which is designed to be similar to the formal
+notation when describing the semantics with compact and user-friendly
+notation.
+\\
+\red{TODO: better character for $\sim$ in the code}
+\begin{verbatim}
+  rule Step_pure/testop:
+  (CONST nt c_1) (TESTOP nt testop)  ~>  (CONST I32 c)
+  -- if c = $testop_(nt, testop, c_1)
+\end{verbatim}
+
+% DSL
+The DSL consists of DSL definitions.
+There are mainly 4 kinds of DSL definitions: type, grammar, relation, and
+functions.
+Type definitions are used to define abstract syntax, while grammar definitions
+are used to define concrete syntax.
+Relation definitions are used to define semantics; the code above is also
+described in relation definition.
+Function definitions can be utilized for other definitions such as
+\texttt{\$testop\_} in the code.
+In addition, they can be also used to describe semantics.
+Especially, SpecTec uses function definitions to describe module instantiation
+and function invocation.
+
+% AL
+However, The DSL cannot generate the prose notation just as it is because it is
+declarative.
+Therefore, SpecTec also defines another imperative language \textit{AL}, which
+stands for algorithmic language, for the prose notation.
+SpecTec translates the semantics describing part of the DSL, which is the
+relation defintions and the function definitions, into the AL.
+Both definitions are expressed in the form of algorithms in AL that takes some
+inputs and performs a step by step instructions as in the prose notation.
+
+% meta-level interpreter
+What is especially intriguing here is that AL is executable.
+It can perform each step written in the prose program one by one, which
+WebAssembly interpreter should follow.
+As a result, the prose program serves as a WebAssembly interpter.
+That is to say, if WebAssembly semantics is written in the DSL, SpecTec
+automatically generates the WebAssembly interpreter following the written
+semantics.
+The generated interpreter passes all the execution tests in the official test
+suites except for the tests related to the infinte loop.
+
+% artifacts SpecTec generates & SpecTec for other language
+\cref{fig:spectec-testop} is the specification document generated by the
+SpecTec using the DSL and the AL.
+The generated document closely resembles the official document except for some
+minor notation changes and some missing hyperlinks.
+Additionally, SpecTec can generate many other artifacts automatically from the
+DSL.
+SpecTec translates the DSL into Coq for mathematical proofs and is working
+on generating other artifacts, such as test suites required for WebAssembly
+standardization.
+Furthermore, research is ongoing to extend SpecTec for mechanizing other
+languages ~\cite{p4-cherry-workshop}.
 

### 2024-11-20 paper/overview.tex
Add code font for tilde and dollar
Write overview

@@
 
-\begin{figure}[h]
+\begin{figure}[t]
   \centerline{\includegraphics[width=15cm]{fig/overview}}
@@
 
-We propose SpecTec ...
+% Overview explanation
+In this chapter, we explain the overall approach to generating multiple
+artifacts from the specification and ensuring the correctness of the
+specification. \Cref{fig:overview} illustrates an overview of our approach.
+% DSL
+SpecTec provides a domain-specific language (DSL) for writing specifications.
+It is a declarative language designed to resemble the formal notation while
+offering a compact and user-friendly syntax.
+The code below shows the specification for the \texttt{testop} instruction
+written in the DSL:
+\begin{lstlisting}[escapechar=\%]
+  rule Step_pure/testop:
+  (CONST nt c_1) (TESTOP nt testop)  %$\tildecode$%>  (CONST I32 c)
+  -- if c = %$\dollarcode$%testop_(nt, testop, c_1)
+\end{lstlisting}
+The DSL is type checked to prevent meta-level errors such as notation misuses and
+dimension mismatches.
+Using the DSL, SpecTec generates LaTeX code and mechanized definitions for theorem provers.
 
-% Overview explanation
-In this chapter, we explain an overall approach to generate multiple artifacts
-from the specification and check correctness of the specification.
-\cref{fig:overview} illustrates the overview of our approach.
-SpecTec provides a domain specific language (DSL) for the specification.
-It is designed to be similar to the formal notation when describing the
-semantics with compact and user-friendly notation.
-It is type checked to prevent meta-level errors such as notation misuses and
-dimension mismatches.
-The code below is the specification for \texttt{testop} instruction written in
-the DSL.
-\begin{lstlisting}
-  rule Step_pure/testop:
-  (CONST nt c_1) (TESTOP nt testop)  ~>  (CONST I32 c)
-  -- if c = $testop_(nt, testop, c_1)
+% AL
+Additionally, SpecTec introduces another imperative language, \textit{AL}, for
+the prose notation.
+In AL, semantics are expressed algorithmically, in the form of step-by-step
+instructions that take inputs and perform operations, mirroring the prose notation.
+SpecTec translates declarative definitions in the DSL into algorithmic
+definitions in AL.
+The AL code below is the specification for the \texttt{testop} instruction
+generated from the DSL:
+\begin{lstlisting}[escapechar=\%]
+Step_pure/testop nt testop {
+ Assert (top_value(nt))
+ Pop (numtype_0.CONST c_1)
+ Let c = %$\dollarcode$%testop_(nt, testop, c_1)
+ Push (I32.CONST c)
+}
 \end{lstlisting}
+Using AL, SpecTec generates reStructuredText code.
+\cref{fig:spectec-testop} shows the generated document for \texttt{testop}
+instruction.
+The generated document closely resembles the official document except for some
+minor notation changes and some missing hyperlinks.
+
+% AL interpreter
+A notable feature of AL is that it is interpretable.
+SpecTec includes an AL interpreter that executes an AL program with given inputs
+and produces results.
+Since an AL program specify the behavior of a Wasm program, the input to an AL
+program is a Wasm program, and the result is the execution outcome of the Wasm
+program.
+In other words, the generated AL program functions as a WebAssembly interpreter.
+What is particularly intriguing is that if specification authors write a Wasm
+specification using the DSL, a WebAssembly interpreter that adheres to the
+defined semantics is automatically generated.
+By testing this generated interpreter, we can assess the specification,
+thereby ensuring its correctness.

### 2024-11-20 paper/background.tex
Split motivation into motiv&model

@@
 \cref{sec:control flow in official prose}.
+
+\newcommand{\officialp}{official prose}
+\newcommand{\spectecp}{SpecTec prose}
+
+\red{TODO: explanation of the terms \officialp{} and \spectecp{}} \\
+\red{TODO: remove redundant part in the fig} \\
+\red{TODO: explanation of context(frame/label): in the background? or here?} \\
+\red{TODO: the notion of meta-level interpreter in background} \\
+\red{TODO: try finding better terminology rather than "model"} \\
+
+\section{Control flow in official prose}
+\label{sec:control flow in official prose}
+
+% control flow structure in official prose
+It might be easy to understand how the \officialp{} explains the control flow
+of WebAssembly, if we assume that a WebAssembly code is loaded on a memory and
+a pc points to the instruction to execute.
+To describe control flow, it uses a structure named \textit{block} which
+consitutes of a instruction sequence.
+When executing the instructions in the block, pc can be changed to the starting
+point of the block or the end of the block.
+
+% an example of Wasm control flow in official prose
+\begin{example}
+\label{ex:br}
+\begin{verbatim}
+  // infinite loop
+  (loop (result i32) (i32.const 42) (br 0) (unreachable) end) (unreachable)
+\end{verbatim}
+\end{example}
+
+\cref{ex:br} is a WebAssembly code example that has \texttt{loop} and
+\texttt{unreachable}
+Here, result type of the \texttt{loop} is \texttt{i32} and the \texttt{loop}
+has a block of three instructions: \texttt{i32.const}, \texttt{br}, and
+\texttt{unreachable}.
+\cref{fig:loop} is the \officialp{} of the \texttt{loop} instruction.
+It says that the continuation is the start of the loop, the information is
+stored in a label, and it \textit{enters} the block with the label.
+The term \textit{enter} means that it pushes the label in the stack and makes
+pc points to the first instruction in the block.
+The \texttt{i32.const} instruction is executed first, which just pushes the
+\texttt{i32} value \texttt{42} in the stack.
+Then, The \texttt{br} instruction is executed.
+\cref{fig:br} is the \officialp{} of the \texttt{br} instruction.
+It pops the label from the stack, and makes pc points to the continuation of
+the label, which is the start of the loop instruction.
+As a result, the code example above is a infinite loop pushing \texttt{42}
+forever without executing \texttt{unreachable} instructions.
+
+\begin{figure}[h!]
+    \centerline{\includegraphics[width=15cm]{fig/loop}}
+    \caption[Enter the caption title here]{\texttt{loop} instruction} \label{fig:loop}
+    \centerline{\includegraphics[width=15cm]{fig/br}}
+    \caption[Enter the caption title here]{\texttt{br} instruction} \label{fig:br}
+\end{figure}
+
+% exiting label in official prose
+There is also a special form of a behavior related to the block: an
+\textit{exiting label}.
+\cref{fig:exiting-label} is the \officialp{} of the \textit{exiting label}.
+It is special because this behavior is performed without an explicit
+WebAssembly instruction.
+Rather, the behavior is performed when \textbf{the end of a block is reached}
+without control instructions or runtime error.
+The behavior is that the label is popped from the stack, and the pc changes to
+the point after the end of the block.
+
+\begin{figure}[h!]
+    \centerline{\includegraphics[width=15cm]{fig/exiting}}
+    \caption[Enter the caption title here]{exiting label} \label{fig:exiting-label}
+\end{figure}
+
+% an example of exiting label in official prose
+\begin{example}
+\label{ex:exit}
+\begin{verbatim}
+  // exiting label
+  (loop (result i32) (i32.const 42) end) (f64.const 3.14)
+\end{verbatim}
+\end{example}
+
+\cref{ex:exit} is a WebAssembly code example that has \texttt{loop} whose block is
+only \texttt{i32.const}, and \texttt{f32.const}
+When the \texttt{loop} instruction is executed, it enters the block with a label.
+After \texttt{i32.const} is executed, the end of the block is reached.
+As a result, exiting label occurs so that the label is popped from the stack,
+pc changes to the point after the end of block: \texttt{f64.const}.
+As a result, after executing the code, the value 42 and 3.14 pushed to the
+stack.
+
+% control flow structure with function call in official prose
+Similar to the control flow using the block and the label, there is a function
+call with a frame.
+\cref{fig:invoke} is the \officialp{} of the function invocation which takes
+place by a \texttt{call} instruction.
+A frame is pushed to the stack and enters the block of the function body with a
+label.
+There is also a special behavior named returning from a function.
+\cref{fig:returning} is the \officialp{} of the returning from a function.
+When end of function is reached without without control instructions or runtime
+error, The frame is popped from the stack, and the pc points to the next
+instruction of the caller instruction.
+
+\begin{figure}[h!]
+    \centerline{\includegraphics[width=15cm]{fig/invoke}}
+    \caption[Enter the caption title here]{function invocation} \label{fig:invoke}
+    \centerline{\includegraphics[width=15cm]{fig/returning}}
+    \caption[Enter the caption title here]{returning from a function} \label{fig:returning}
+\end{figure}

### 2024-11-20 paper/model.tex
Split motivation into motiv&model

--- 2024-11-18 paper/3-motivation.tex
@@
-% !TEX root = main.tex
-
-\chapter{Motivation}
-\label{ch:motivation}
+\chapter{Model}
+\label{ch:Model}
 \noindent
-
-\newcommand{\officialp}{official prose}
-\newcommand{\spectecp}{SpecTec prose}
-
-\red{TODO: explanation of the terms \officialp{} and \spectecp{}} \\
-\red{TODO: remove redundant part in the fig} \\
-\red{TODO: explanation of context(frame/label): in the background? or here?} \\
-\red{TODO: the notion of meta-level interpreter in background} \\
-\red{TODO: try finding better terminology rather than "model"} \\
-
-\section{Control flow in official prose}
-\label{sec:control flow in official prose}
-
-% control flow structure in official prose
-It might be easy to understand how the \officialp{} explains the control flow
-of WebAssembly, if we assume that a WebAssembly code is loaded on a memory and
-a pc points to the instruction to execute.
-To describe control flow, it uses a structure named \textit{block} which
-consitutes of a instruction sequence.
-When executing the instructions in the block, pc can be changed to the starting
-point of the block or the end of the block.
-
-% an example of Wasm control flow in official prose
-\begin{example}
-\label{ex:br}
-\begin{verbatim}
-  // infinite loop
-  (loop (result i32) (i32.const 42) (br 0) (unreachable) end) (unreachable)
-\end{verbatim}
-\end{example}
-
-\cref{ex:br} is a WebAssembly code example that has \texttt{loop} and
-\texttt{unreachable}
-Here, result type of the \texttt{loop} is \texttt{i32} and the \texttt{loop}
-has a block of three instructions: \texttt{i32.const}, \texttt{br}, and
-\texttt{unreachable}.
-\cref{fig:loop} is the \officialp{} of the \texttt{loop} instruction.
-It says that the continuation is the start of the loop, the information is
-stored in a label, and it \textit{enters} the block with the label.
-The term \textit{enter} means that it pushes the label in the stack and makes
-pc points to the first instruction in the block.
-The \texttt{i32.const} instruction is executed first, which just pushes the
-\texttt{i32} value \texttt{42} in the stack.
-Then, The \texttt{br} instruction is executed.
-\cref{fig:br} is the \officialp{} of the \texttt{br} instruction.
-It pops the label from the stack, and makes pc points to the continuation of
-the label, which is the start of the loop instruction.
-As a result, the code example above is a infinite loop pushing \texttt{42}
-forever without executing \texttt{unreachable} instructions.
-
-\begin{figure}[h!]
-    \centerline{\includegraphics[width=15cm]{fig/loop}}
-    \caption[Enter the caption title here]{\texttt{loop} instruction} \label{fig:loop}
-    \centerline{\includegraphics[width=15cm]{fig/br}}
-    \caption[Enter the caption title here]{\texttt{br} instruction} \label{fig:br}
-\end{figure}
-
-% exiting label in official prose
-There is also a special form of a behavior related to the block: an
-\textit{exiting label}.
-\cref{fig:exiting-label} is the \officialp{} of the \textit{exiting label}.
-It is special because this behavior is performed without an explicit
-WebAssembly instruction.
-Rather, the behavior is performed when \textbf{the end of a block is reached}
-without control instructions or runtime error.
-The behavior is that the label is popped from the stack, and the pc changes to
-the point after the end of the block.
-
-\begin{figure}[h!]
-    \centerline{\includegraphics[width=15cm]{fig/exiting}}
-    \caption[Enter the caption title here]{exiting label} \label{fig:exiting-label}
-\end{figure}
-
-% an example of exiting label in official prose
-\begin{example}
-\label{ex:exit}
-\begin{verbatim}
-  // exiting label
-  (loop (result i32) (i32.const 42) end) (f64.const 3.14)
-\end{verbatim}
-\end{example}
-
-\cref{ex:exit} is a WebAssembly code example that has \texttt{loop} whose block is
-only \texttt{i32.const}, and \texttt{f32.const}
-When the \texttt{loop} instruction is executed, it enters the block with a label.
-After \texttt{i32.const} is executed, the end of the block is reached.
-As a result, exiting label occurs so that the label is popped from the stack,
-pc changes to the point after the end of block: \texttt{f64.const}.
-As a result, after executing the code, the value 42 and 3.14 pushed to the
-stack.
-
-% control flow structure with function call in official prose
-Similar to the control flow using the block and the label, there is a function
-call with a frame.
-\cref{fig:invoke} is the \officialp{} of the function invocation which takes
-place by a \texttt{call} instruction.
-A frame is pushed to the stack and enters the block of the function body with a
-label.
-There is also a special behavior named returning from a function.
-\cref{fig:returning} is the \officialp{} of the returning from a function.
-When end of function is reached without without control instructions or runtime
-error, The frame is popped from the stack, and the pc points to the next
-instruction of the caller instruction.
-
-\begin{figure}[h!]
-    \centerline{\includegraphics[width=15cm]{fig/invoke}}
-    \caption[Enter the caption title here]{function invocation} \label{fig:invoke}
-    \centerline{\includegraphics[width=15cm]{fig/returning}}
-    \caption[Enter the caption title here]{returning from a function} \label{fig:returning}
-\end{figure}
 

### 2024-11-20 paper/spectec.tex
Split motivation into motiv&model

@@
 languages ~\cite{p4-cherry-workshop}.
-
-%% Syntax
-\section{Syntax of AL}
-\label{syntax}
-
-{
-\renewcommand{\arraystretch}{0.8835}  % Decrease row spacing locally
-\begin{align*}
-\begin{array}{lcccrlr}
-%
-% Algorithm
-  \text{Algorithm}\quad& A &\ni& a &::=& ~ \rel ~ (s, \seq e, \seq i) ~ | ~ \fun ~ (s, \seq e, \seq i) \\
-%
-% Instruction
-  \text{Instruction}\quad& I &\ni& i &::=& ~ \ifi ~ (e, \seq i, \seq i) &\quad\text{(If)} \\
-    &&&& | & ~ \eitheri ~ (\seq i, \seq i) &\quad\text{(Either)} \\
-    &&&& | & ~ \enteri ~ (e, e) &\quad\text{(Enter)} \\
-    &&&& | & ~ \pushctxi ~ e &\quad\text{(Push Context)} \\
-    &&&& | & ~ \pushi ~ e &\quad\text{(Push)} \\
-    &&&& | & ~ \popctxi ~ e &\quad\text{(Pop Context)} \\
-    &&&& | & ~ \popi ~ e &\quad\text{(Pop)} \\
-    &&&& | & ~ \popni ~ (e, e) &\quad\text{(Pop N)} \\
-    &&&& | & ~ \popalli ~ e &\quad\text{(Pop All)} \\
-    &&&& | & ~ \leti ~ (e, e) &\quad\text{(Let)} \\
-    &&&& | & ~ \trapi &\quad\text{(Trap)} \\
-    &&&& | & ~ \returnreli &\quad\text{(Return Relation)} \\
-    &&&& | & ~ \returnfuni ~ e &\quad\text{(Return Function)} \\
-    &&&& | & ~ \executei ~ e &\quad\text{(Execute)} \\
-    &&&& | & ~ \calli ~ (s, s, \seq{e}) &\quad\text{(Let Call)} \\
-    &&&& | & ~ \replaceframei ~ (p^+, e) &\quad\text{(Replace Frame)} \\
-    &&&& | & ~ \replacestorei ~ (p^+, e) &\quad\text{(Replace Store)} \\
-%
-% Expression
-  \text{Expression}\quad& E &\ni& e &::=& ~ \vare ~ s &\quad\text{(Variable)} \\
-    &&&& | & ~ \nume ~ n &\quad\text{(Number)} \\
-    &&&& | & ~ \boole ~ b &\quad\text{(Boolean)} \\
-    &&&& | & ~ \fnamee ~ s &\quad\text{(Function Name)} \\
-    &&&& | & ~ \liste ~ \seq e &\quad\text{(List)} \\
-    &&&& | & ~ \stre ~ \seq{(s, e)} &\quad\text{(Record)} \\
-    &&&& | & ~ \tupe ~ \seq e &\quad\text{(Tuple)} \\
-    &&&& | & ~ \casee ~ (s, \seq e) &\quad\text{(Tagged Tuple)} \\
-    &&&& | & ~ \une ~ (unop, e) &\quad\text{(Unary Operation)} \\
-    &&&& | & ~ \bine ~ (binop, e, e) &\quad\text{(Binary Operation)} \\
-    &&&& | & ~ \acce ~ (e, p) &\quad\text{(Access)} \\
-    &&&& | & ~ \upde ~ (e, p^+, e) &\quad\text{(Update)} \\
-    &&&& | & ~ \cate ~ (e, e) &\quad\text{(Concatenation)} \\
-    &&&& | & ~ \compe ~ (e, e) &\quad\text{(Composition)} \\
-    &&&& | & ~ \meme ~ (e, e) &\quad\text{(Membership)} \\
-    &&&& | & ~ \choosee ~ e &\quad\text{(Choose)} \\
-    &&&& | & ~ \lene ~ e &\quad\text{(Length)} \\
-    &&&& | & ~ \iscaseofe ~ (e, s) &\quad\text{(Check Tag)} \\
-    &&&& | & ~ \getcurctxe  &\quad\text{(Get Current Context)} \\
-    &&&& | & ~ \ctxkinde ~ s &\quad\text{(Context Kind)} \\
-    &&&& | & ~ \itere ~ (e, iter, \seq{s}) &\quad\text{(Iteration)} \\
-    &&&& | & ~ \matche ~ (e, e) &\quad\text{(Match)} \\
-    &&&& | & ~ \hastypee ~ (e, s) &\quad\text{(Has Type)} \\
-\end{array}
-\end{align*}
-}
-
-\newpage
-\begin{align*}
-\begin{array}{lcccrlr}
-%
-% Path
-  \text{Path}\quad& P &\ni& p &::=& ~ \idxp ~ e ~ | ~ \slicep ~ (e, e) ~ | ~ \dotp ~ s \\
-%
-% Iter
-  \text{Iter}\quad& Iter &\ni& iter &::=& ~ \listiter ~ | ~ \listniter ~ e ~|~ \listidxiter ~ (s, e) \\
-%
-% Operator
-  \text{Unary Operator}\quad& Unop &\ni& unop &::=& ~ \notop ~ | ~ \minusop \\
-  \text{Binary Operator}\quad& Binop &\ni& binop &::=& ~ \addop ~ | ~ \subop ~ | ~ \mulop \\
-    &&&& | & \divop ~ | ~ \modop ~ | ~ \expop \\
-    &&&& | & \implop ~ | ~ \equivop ~ | ~ \andop \\
-    &&&& | & \orop ~ | ~ \eqop ~ | ~ \neop ~ | ~ \ltop \\
-    &&&& | & \gtop ~ | ~ \leop ~ | ~ \geop \\
-% Primitives
-  \text{String}\quad& \mathbb S &\ni& s \\
-  \text{Integer}\quad& \mathbb Z &\ni& n \\
-  \text{Boolean}\quad& \mathbb B &\ni& b \\
-\end{array}
-\end{align*}
-
-% algorithm
-An algorithm is either an AL relation or an AL function, corresponding to the
-relation and the function definitions in the DSL.
-% instruction
-An AL instruction is \ifi{} for conditionally choosing a branch, \eitheri{} for
-nondeterministically choosing a branch, \enteri{} for entering a block,
-\pushctxi{} for pushing a Wasm context, \pushi{} for pushing a Wasm value,
-\popctxi{} for popping a Wasm context, \popi{} for popping a Wasm value,
-\popni{} for popping $n$ Wasm values, \popalli{} for popping all Wasm values
-within a Wasm context, \leti{} for let binding, \trapi{} for trapping,
-\returnreli{} for returning from an AL relation, \returnfuni{} for returning
-from an AL function, \executei{} for executing given Wasm instructions,
-\calli{} for an AL function call with let binding, \replaceframei{} for
-replacing a frame, or \replacestorei{} for replacing a store.
-% expression
-An expression is a variable, number, boolean, AL function name, list, record
-(or struct), tuple, tagged tuple, unary or binary operation, data structure
-access or update, list or record concatenation, membership check, element
-selection, length retrieval, tag check, Wasm context retrieval, Wasm context
-kind check, iteration, match relation, or type relation.
-
-\begin{align*}
-\begin{array}{lcccrlr}
-%
-% State
-  \text{State}\quad& \Sigma &\ni& \sigma & ::=& ~ \seq a, w, k \\
-%
-% Wasm state
-  \text{Wasm State}\quad& W &\ni& w &::=& ~ \seq{we}, \seq{wi}, sto \\
-%
-% Wasm Entry
-  \text{Wasm Entry}\quad& WE &\ni& we &::=& ~ wv ~ | ~ wc \\
-%
-% Wasm Value
-  \text{Wasm Value}\quad& WV &\ni& wv &::=& ~ \casev ~ (s, \seq v) \\
-%
-% Wasm Context
-  \text{Wasm Context}\quad& WC &\ni& wc &::=& ~ \casev ~ (\text{``Label"}, \seq v) ~ | ~ \casev ~ (\text{``Frame"}, \seq v) \\
-%
-% Wasm Instruction
-  \text{Wasm Instruction}\quad& WI &\ni& wi &::=& ~ \casev ~ (s, \seq v) \\
-%
-% Store
-  \text{Store}\quad& Store &\ni& sto &::=& ~ \seq{(s, v)} \\
-%
-% Continuation
-  \text{Continuation}\quad & K &\ni& k &::=& ~ \mt &\quad\text{(Empty)} \\
-    &&&& | & ~ \toplevelcall ~ (s, \seq v) &\quad\text{(Top-level Call)} \\
-    &&&& | & ~ \call ~ (s, s, \seq v, k) &\quad\text{(Call)} \\
-    &&&& | & ~ \exe ~ (\seq{wi}, k) &\quad\text{(Execute)} \\
-    &&&& | & ~ \wasm ~ (\seq{wc}, k) &\quad\text{(Wasm)} \\
-    &&&& | & ~ \algo ~ c &\quad\text{(Algorithm)} \\
-    &&&& | & ~ \ret ~ (v, k) &\quad\text{(Return)} \\
-%
-% Context
-  \text{Context}\quad& C &\ni& c &::=& ~ (\seq i, \mu, \seq{wc}, k) \\
-%
-% Environment
-  \text{Environment}\quad& M &\ni& \mu &::=& ~ \seq{[s \mapsto v]} \\
-\end{array}
-\end{align*}
-
-\newpage
-\begin{align*}
-\begin{array}{lcccrlr}
-%
-% Value
-  \text{Value}\quad& V &\ni& v &::=& ~ \numv ~ n &\quad\text{(Number)} \\
-    &&&& | & ~ \boolv ~ b &\quad\text{(Boolean)} \\
-    &&&& | & ~ \fnamev ~ s &\quad\text{(Function Name)} \\
-    &&&& | & ~ \listv ~ \seq v &\quad\text{(List)} \\
-    &&&& | & ~ \strv ~ \seq{(s, v)} &\quad\text{(Record)} \\
-    &&&& | & ~ \tupv ~ \seq v &\quad\text{(Tuple)} \\
-    &&&& | & ~ \casev ~ (s, \seq v) &\quad\text{(Tagged Tuple)} \\
-    &&&& | & ~ \trapv &\quad\text{(Trap)} \\
-    &&&& | & ~ \storev &\quad\text{(Store)} \\
-\end{array}
-\end{align*}
-
-% State
-An AL state consists of a sequence of algorithms, a Wasm state, and a
-continuation.
-% Wasm State
-A Wasm state consists of an interleaved stack, a Wasm instruction stack, and a
-store.
-% Wasm Entry
-An interleaved stack contains two types of entries: a Wasm value and a Wasm
-context.
-% Wasm Value & Wasm Context & Wasm Instruction
-Each of a Wasm value, a Wasm context, and a Wasm instruction is expressed as an
-AL value $\casev$.
-In particular, the tag of a Wasm context is either ``Label" or ``Frame".
-% Store
-A store is expressed as a sequence of pairs, each consisting of a field name
-and an AL value.
-% Continuation
-A continuation is \mt{} for an empty continuation, \toplevelcall{} for
-top-level function call, \call{} for a function call with let binding, \exe{}
-for executing given Wasm instructions, \wasm{} for executing a Wasm instruction
-in the instruction stack, \algo{} for interpreting AL instructions in an
-algorithm, or \ret{} for returning from an AL function.
-% Env
-An environment is a finite mapping from variable names to AL values.
-% Context
-An AL context consists of a sequence of Wasm instruction, an environment, a
-sequence of Wasm contexts, and a continuation.
-% Value
-An AL value is a number, boolean, function name, list, record, tuple, tagged tuple,
-trap, or store.
-
-% Terminology
-In this section, AL-specific terms (state, context, value, function, and
-instruction) are referred to simply as state, context, value, function, and
-instruction, while Wasm terms remain unchanged.
-
-\newpage
-%% Semantics
-\section{Semantics of AL}
-\label{semantics}
-
-%% Continuation
-
-\begin{gather*}
-\boxed{\leadsto \subseteq \Sigma \times \Sigma} \\
-%
-% TopLevelCall
-\newline \\
-  \\
-  \hline
-  (\seq{a}, w, \toplevelcall ~ (s, \seq v)) \leadsto (\seq{a}, w, \createalgo(\seq a, s, \seq v, \mt)) \\
-%
-% Call
-\newline \\
-  \\
-  \hline
-  (\seq{a}, w, \call ~ (s_{var}, s_{name}, \seq v, k)) \leadsto
-  (\seq{a}, w, \createalgo(\seq a, s_{name}, \seq v, \call ~ (s_{var}, s_{name}, \seq v, k))) \\
-%
-% Execute-empty
-\newline \\
-  \\
-  \hline
-  (\seq{a}, w, \exe ~ (\epsilon, k)) \leadsto (\seq{a}, w, k) \\
-%
-% Execute-instr
-\newline \\
-  \casev ~ (s, \seq v) = wi \\
-  \hline
-  (\seq{a}, w, \exe ~ (wi ~ \seq{wi}, k)) \leadsto
-  (\seq{a}, w, \createalgo(\seq a, s, \seq v, \exe ~ (\seq{wi}, k))) \\
-%
-% Wasm-empty
-\newline \\
-  \\
-  \hline
-  (\seq{a}, w, \wasm ~ (\epsilon, k)) \leadsto (\seq{a}, w, k) \\
-%
-% Wasm-instr
-\newline \\
-  (w', \casev ~ (s, \seq v)) = \popwasminstr(w) \\
-  \hline
-  (\seq{a}, w, \wasm ~ (wc^+, k))
-  \leadsto
-  (\seq{a}, w', \createalgo(\seq a, s, \seq v, \wasm ~ (wc^+, k))) \\
-%
-% Al-empty
-\newline \\
-  \\
-  \hline
-  (\seq{a}, w, \algo ~ (\epsilon, \mu, \seq{wc}, k)) \leadsto (\seq{a}, w, k) \\
-%
-% Al-instr
-\newline \\
-  \seq{a}, w, (\seq{i_1}, \mu, \seq{wc}, k) \vdash i_0 \Rightarrow (w', k') \\
-  \hline
-  (\seq{a}, w, \algo ~ (i_0 ~ \seq{i_1}, \mu, \seq{wc}, k)) \leadsto (\seq{a}, w', k') \\
-%
-% Return
-\newline \\
-  \mu = \getenv(c) \qquad c' = \setenv(c, \mu[s_{var} \mapsto v]) \\
-  \hline
-  (\seq{a}, w, \ret ~ (v, \call ~ (s_{var}, s_{name}, \seq v, \algo ~ c)) \leadsto
-  (\seq a, w, \algo ~ c') \\
-\end{gather*}
-
-The semantics of AL is defined by a state transition system.
-% TopLevelCall
-Given algorithms $\seq a$ generated from the DSL and a Wasm state $w$, a
-top-level function $s$ can be invoked with arguments $\seq v$ to perform
-either module instantiation or function invocation:
-$(\seq a, w, \toplevelcall ~ (s, \seq v))$.
-$\toplevelcall ~ (s, \seq v)${} transitions to \algo{} with the function name
-$s$, the arguments $\seq v$, and \mt{}.
-% Empty
-\mt{} is an empty continuation, representing the end of the transition, so no
-further transition occurs.
-% Call
-Similarly to the top-level call, $\call~ (s_{var}, s_{name}, \seq v, k)${}
-transitionsto \algo{} with the function name $s_{name}$ and the arguments $\seq
-v$, but with the current continuation nested within it.
-% Execute
-$\exe ~ (\seq{wi}, k)${} executes each Wasm instruction in the sequence
-$\seq{wi}$ one by one, each containing a relation name and arguments.
-The continuation transitions to \algo{} with the relation name, the arguments,
-and the current continuation, with the Wasm instruction being popped.
-% Wasm
-In $\wasm ~ (\seq{wc}, k)${}, a Wasm instruction is popped from
-the Wasm instruction stack and executed until the Wasm context sequence
-$\seq{wc}$ is exhausted.
-This execution transitions the continuation to \algo{}, with the relation name
-and the arguments of the Wasm instruction, and the current continuation nested
-within it.
-% Algo
-In $\algo ~ (\seq i, \mu, \seq{wc}, k)${}, the body instructions $\seq i$
-execute sequentially, allowing transitions to \call{}, \exe{}, \wasm{}, \ret{},
-or \algo{}.
-% Return
-The execution of a function call concludes with a return instruction, which
-changes the continuation to $\ret ~ v${}.
-If the function call originates from \call{}, the return value $v$ is assigned
-to the variable $s_{var}$.
-If it originates from \toplevelcall{}, the inner continuation $k$ is always
-\mt, so the entire execution concludes with the return value $v$:
-$
-(\seq a, w, \toplevelcall ~ (s, \seq v))
-\leadsto^*
-(\seq a, w', \ret ~ (v, \mt))
-$.
-
-%% Instruction
-
-\begin{gather*}
-  \boxed{\seq{a}, ~ w, ~ c \vdash i \Rightarrow w, ~ k} \\
-%
-% If-true
-\newline \\
-  w, \getenv(c) \vdash e \Rightarrow v \qquad
-  \istrue(v) \\
-  \hline
-  \seq{a}, w, c \vdash \ifi ~ (e, \seq{i_1}, \seq{i_2}) \Rightarrow
-  (w, \algo ~ (\prependinstr(c, \seq{i_1}))) \\
-%
-% If-false
-\newline \\
-  w, \getenv(k) \vdash e \Rightarrow v \qquad
-  \neg \istrue(v) \\
-  \hline
-  \seq{a}, w, c \vdash \ifi ~ (e, \seq{i_1}, \seq{i_2}) \Rightarrow
-  (w, \algo ~ (\prependinstr(c, \seq{i_2}))) \\
-%
-% Either-1
-\newline \\
-  \\
-  \hline
-  \seq{a}, w, c \vdash \eitheri ~ (\seq{i_1}, \seq{i_2}) \Rightarrow
-  (w, \algo ~ (\prependinstr(c, \seq{i_1}))) \\
-%
-% Either-2
-\newline \\
-  \\
-  \hline
-  \seq{a}, w, c \vdash \eitheri ~ (\seq{i_1}, \seq{i_2}) \Rightarrow
-  (w, \algo ~ (\prependinstr(c, \seq{i_2}))) \\
-%
-% Enter
-\newline \\
-  (\seq{we}, \seq{wi}, sto) = w \qquad
-  (i^*, \mu, wc_1 ~ ... ~ wc_n, k) = c \\
-  w, \mu \vdash e_1 \Rightarrow wc \qquad
-  w, \mu \vdash e_2 \Rightarrow \listv ~ \seq{wi_1} \\
-  wi = \getendinstr(wc) \qquad
-  wi_1 = \getendinstr(wc_1) \quad ... \quad wi_n = \getendinstr(wc_n) \\
-  wi_2^* = wi_1^* ~ wi ~ wi_1 ~ ... ~ wi_n \\
-  \hline
-  \seq{a}, w, c \vdash \enteri ~ (e_1, e_2)
-  \Rightarrow
-  (
-    (wc ~ \seq{we}, wi_2^* ~ \seq{wi}, sto),
-    \wasm ~ (wc ~ wc_1 ~ ... ~ wc_n, \algo ~ (i^*, \mu, \epsilon, k))
-  ) \\
-%
-% PushCtx
-\newline \\
-  w, \getenv(c) \vdash e \Rightarrow wc \\
-  \hline
-  \seq{a}, w, c \vdash \pushctxi ~ e
-  \Rightarrow
-  (\push(w, wc), \algo ~ (\addctx(c, wc))) \\
-%
-% Push
-\newline \\
-  w, \getenv(c) \vdash e \Rightarrow wv \\
-  \hline
-  \seq{a}, w, c \vdash \pushi ~ e \Rightarrow (\push(w, wv), \algo ~ c) \\
-%
-% PopCtx
-\newline \\
-  (\seq{we}, \seq{wi}, sto) = w \qquad
-  (\seq{wv}, wc, \seq{we'}) = \splitctx(\seq{we}) \\
-  \mu = \getenv(c) \cdot \assign(e, wc) \qquad
-  \seq{wi'} = \exit(\seq{wi}) \qquad
-  c' = \popwasmctx_{C}(c) \\
-  \hline
-  \seq{a}, w, c \vdash \popctxi ~ e
-  \Rightarrow
-  ((\seq{we'}, \seq{wi'}, sto), \algo ~ (\setenv(c', \mu))) \\
-%
-% Pop
-\newline \\
-  (w', wv) = \pop(w) \qquad
-  \mu' = \getenv(c) \cdot \assign(e, wv) \\
-  \hline
-  \seq{a}, w, c \vdash \popi ~ e \Rightarrow (w', \algo ~ (\setenv(c, \mu'))) \\
-%
-% PopN
-\newline \\
-  \mu = \getenv(c) \qquad
-  w, \mu \vdash e_2 \Rightarrow \numv ~ n \\
-  (w', wv^n) = \popn(w, n) \qquad
-  \mu' = \mu \cdot \assign(e_1, \listv ~ wv^n) \\
-  \hline
-  \seq{a}, w, c \vdash \popni ~ (e_1, e_2) \Rightarrow
-  (w', \algo ~ (\setenv(c, \mu'))) \\
-%
-% PopAll
-\newline \\
-  (\seq{wv}, wc, \seq{we'}) = \splitctx(\seq{we}) \qquad
-  \mu' = \getenv(c) \cdot \assign(e, \liste ~ \seq{wv}) \\
-  \hline
-  \seq{a}, (\seq{we}, \seq{wi}, sto), c \vdash \popalli ~ e
-  \Rightarrow
-  ((wc ~ \seq{we'}, \seq{wi}, sto), \algo ~ (\setenv(c, \mu'))) \\
-%
-% Let
-\newline \\
-  \mu = \getenv(c) \qquad
-  w, \mu \vdash e_2 \Rightarrow v \qquad
-  \mu' = \mu \cdot \assign(e_1, v) \\
-  \hline
-  \seq{a}, w, c \vdash \leti ~ (e_1, e_2)
-  \Rightarrow
-  (w, \setenv(c, \mu')) \\
-%
-% Trap
-\newline \\
-  \\
-  \hline
-  \seq{a}, w, c \vdash \trapi \Rightarrow (w, \ret ~ (\trapv, \mt)) \\
-%
-% Return-rule
-\newline \\
-  \\
-  \hline
-  \seq{a}, w, (\seq{i}, \mu, wc^*, k) \vdash \returnreli \Rightarrow (w, k) \\
-%
-% Return-func
-\newline \\
-  w, \mu \vdash e \Rightarrow v \\
-  \hline
-  \seq{a}, w, (\seq{i}, \mu, wc^*, k) \vdash \returnfuni ~ e \Rightarrow
-  (w, \ret ~ (v, k)) \\
-%
-% Execute
-\newline \\
-  w, \getenv(c) \vdash e \Rightarrow \listv ~ wi^* \\
-  \hline
-  \seq{a}, w, c \vdash \executei ~ e \Rightarrow
-  (w, \exe ~ (wi^*, \algo ~ c)) \\
-%
-% Call-fname
-\newline \\
-  \mu = \getenv(c) \qquad
-  \fnamev ~ s_3 = \mu(s_2) \qquad
-  w, \mu \vdash e_1 \Rightarrow v_1 \quad ... \quad
-  w, \mu \vdash e_n \Rightarrow v_n \\
-  \hline
-  \seq{a}, w, c \vdash \calli ~ (s_1, s_2, e_1 ~ ... ~ e_n) \Rightarrow
-  (w, \call ~ (s_1, s_3, v_1 ~ ... ~ v_n, \algo ~ c)) \\
-%
-% Call-fname
-\newline \\
-  \mu = \getenv(c) \qquad
-  s_2 \not\in \domain(\mu) \qquad
-  w, \mu \vdash e_1 \Rightarrow v_1 \quad ... \quad
-  w, \mu \vdash e_n \Rightarrow v_n \\
-  \hline
-  \seq{a}, w, c \vdash \calli ~ (s_1, s_2, \seq e) \Rightarrow
-  (w, \call ~ (s_1, s_2, v_1 ~ ... ~ v_n, \algo ~ c)) \\
-%
-% Replace-frame
-\newline \\
-\text{NOTE: $e_1$ is either store or current frame} \\
-  \mu = \getenv(c) \qquad
-  \casev ~ (\text{``Frame"}, v_1 ~ v_2) = \getcurframe(w) \qquad
-  w, \mu \vdash e \Rightarrow v \\
-  wc = \casev ~ (\text{``Frame"}, v_1 ~ \update(w, \mu, v_2, p^+, v)) \\
-  \hline
-  \seq{a}, w, c \vdash \replaceframei ~ (p^+, e) \Rightarrow
-  (\setcurframe(w, wc), \algo ~ c) \\
-%
-% Replace-store
-\newline \\
-  \mu = \getenv(c) \qquad
-  w, \mu \vdash e \Rightarrow v \\
-  \strv ~ sto = \update(w, \mu, \strv ~ (\getstore(w)), p^+, v) \\
-  \hline
-  \seq{a}, w, c \vdash \replacestorei ~ (p^+, e)
-  \Rightarrow (\setstore(w, sto), c) \\
-\end{gather*}
-
-Given algorithms $\seq a$, a Wasm state $w$, and a context $c$, executing an
-instruction $i$ produces an updated Wasm state and a continuation.
-% If
-$\ifi ~ (e, i_1^*, i_2^*)${} evaluates $e$ and branches to $\seq i_1$ if the
-result is true, or $\seq i_2$ otherwise.
-% Either
-$\eitheri ~ (i_1^*, i_2^*)${} nondeterministically selects either $i_1^*$ or
-$i_2^*$.
-% Enter
-$\enteri ~ (e_1, e_2)${} evaluates $e_1$, pushing the resulting Wasm context
-$wc$ to the interleaved stack.
-It then evaluates $e_2$ to obtain the Wasm instruction sequence $wi_1^*$.
-Then, the end instructions of the Wasm context $wc$ and the Wasm contexts
-$wc_1 ~ ... ~ wc_n$ in the context $c$ are appended to $wi_1^*$ and pushed to
-the Wasm instruction stack.
-Finally, it changes the continuation to
-$\wasm ~ (wc ~ wc_1 ~ ... ~ wc_n, \algo ~ (i^*, \mu, \epsilon, k))${}.
-% PushCtx
-$\pushctxi ~ e${} evaluates $e$, pushing the resulting Wasm context $wc$ to the
-interleaved stack and the context $c$.
-% Push
-$\pushi ~ e${} evaluates $e$, pushing the resulting value $wv$ to the
-interleaved stack.
-% PopCtx
-$\popctxi ~ e${} pops all Wasm values up to and including a Wasm context $wc$
-from the interleaved stack and all Wasm instructions up to and including an end
-instruction from the Wasm instruction stack.
-It also pops a Wasm context from the context $c$.
-Then, it assign the Wasm context $wc$ to $e$.
-% Pop
-$\popi ~ e${} pops a Wasm value $wv$ from the interleaved stack and assign the
-value to $e$.
-% PopN
-$\popni ~ (e_1, e_2)${} evaluates $e_2$ to obtain a number $n$, pops $n$ Wasm
-values from the interleaved stack, and assigns the values to $e_1$.
-% PopAll
-$\popalli ~ e${} pops all the Wasm values up to, but not including, a
-Wasm context from the interleaved stack and assign the values to $e$.
-% Let
-$\leti ~ (e_1, e_2)${} evaluates $e_2$ and assigns the resulting value to
-$e_1$.
-% Trap
-\trapi{} results in $\ret ~ (\trapv{}, \mt{})$ to terminate the execution.
-% ReturnRel
-\returnreli{} results in the continuation $k$ in the context $c$, indicating a
-return from a relation algorithm.
-% ReturnFunc
-$\returnfuni ~ e${} evaluates $e$ to $v$, resulting in $\ret ~ (v, k)${} where
-$k$ is the continuation in the context $c$.
-% Execute
-\executei{} evaluates $e$ to obtain a Wasm instruction sequence $wi^*$ and
-result in $\exe ~ (wi^*, \algo ~ c)${}.
-% Call
-$\calli ~ (s_1, s_2, e_1 ~ ... ~ e_n)${} checks if $s_2$ is bound in the
-environment.
-If it is, it retrieves the function name $s_3$ from the environment $\mu$;
-otherwise, it uses $s_2$ as the function name.
-It evaluates $\seq e$ to $v^*$, resulting in $\call ~ (s_1, s_{name}, v_1 ~ ...
-~ v_n)${} where $s_{name}$ is the function name.
-% ReplaceFrame
-\replaceframei{} retrieves the current frame $\casev ~ (\text{``Frame"}, v_1
-v_2)${} in the Wasm context $w$.
-It then evaluates $e$ to $v$, updating the value at location $v_2$ along paths
-$p^+$ to $v$.
-% ReplaceFrame
-\replacestorei{} evaluates $e$ to $v$, updating the value at location $\strv ~
-sto$ along paths $p^+$ to v.
-
-%% Expression
-
-\begin{gather*}
-  \boxed{w, \mu \vdash e \Rightarrow v} \\
-%
-% Var
-\newline \\
-  v = \mu(s) \\
-  \hline
-  w, \mu \vdash \vare ~ s \Rightarrow v \\
-%
-% Num
-\newline \\
-  \hline
-  w, \mu \vdash \nume ~ n \Rightarrow \numv ~ n \\
-%
-% Bool
-\newline \\
-  \hline
-  w, \mu \vdash \boole ~ b \Rightarrow \boolv ~ b \\
-%
-% Fname
-\newline \\
-  \hline
-  w, \mu \vdash \fnamee ~ s \Rightarrow \fnamev ~ s \\
-%
-% List
-\newline \\
-  w, \mu \vdash e_1 \Rightarrow v_1 \quad ... \quad
-  w, \mu \vdash e_n \Rightarrow v_n \\
-  \hline
-  w, \mu \vdash \liste ~ e_1 ~ ... ~ e_n \Rightarrow \listv ~ v_1 ~ ... ~ v_n \\
-%
-% Str
-\newline \\
-  w, \mu \vdash e_1 \Rightarrow v_1 \quad ... \quad
-  w, \mu \vdash e_n \Rightarrow v_n \\
-  \hline
-  w, \mu \vdash \stre ~ (s_1, e_1) ~ ... ~ (s_n, e_n) \Rightarrow
-  \strv ~ (s_1, v_1) ~ ... ~ (s_n, v_n) \\
-%
-% Tup
-\newline \\
-  w, \mu \vdash e_1 \Rightarrow v_1 \quad ... \quad
-  w, \mu \vdash e_n \Rightarrow v_n \\
-  \hline
-  w, \mu \vdash \tupe ~ e_1 ~ ... ~ e_n \Rightarrow \tupv ~ v_1 ~ ... ~ v_n \\
-%
-% Case
-\newline \\
-  w, \mu \vdash e_1 \Rightarrow v_1 \quad ... \quad
-  w, \mu \vdash e_n \Rightarrow v_n \\
-  \hline
-  w, \mu \vdash \casee ~ (s, e_1 ~ ... ~ e_n) \Rightarrow \casev ~ (s, v_1 ~ ... ~ v_n) \\
-%
-% Un
-\newline \\
-  w, \mu \vdash e \Rightarrow v \\
-  \hline
-  w, \mu \vdash \une ~ (unop, e) \Rightarrow \unop(unop, v) \\
-%
-% Bin
-\newline \\
-  w, \mu \vdash e_1 \Rightarrow v_1 \qquad w, \mu \vdash e_2 \Rightarrow v_2 \\
-  \hline
-  w, \mu \vdash \bine ~ (binop, e_1, e_2) \Rightarrow \binop(binop, v_1, v_2) \\
-%
-% Acc
-\newline \\
-  w, \mu \vdash e \Rightarrow v \\
-  \hline
-  w, \mu \vdash \acce ~ (e, p) \Rightarrow \access(w, \mu, v, p) \\
-%
-% Upd
-\newline \\
-  w, \mu \vdash e_1 \Rightarrow v_1 \qquad w, \mu \vdash e_2 \Rightarrow v_2 \\
-  \hline
-  w, \mu \vdash \upde ~ (e_1, p^+, e_2) \Rightarrow \update(w, \mu, v_1, p^+, v_2) \\
-%
-% Cat
-\newline \\
-   w, \mu \vdash e_1 \Rightarrow \listv ~ \seq{v_1} \qquad
-   w, \mu \vdash e_2 \Rightarrow \listv ~ \seq{v_2} \\
-  \hline
-  w, \mu \vdash \cate ~ (e_1, e_2) \Rightarrow \listv ~ (\seq{v_1} ~ \seq{v_1}) \\
-%
-% Comp
-\newline \\
-   w, \mu \vdash e_1 \Rightarrow \strv ~ \seq{(s_1, v_1)} \qquad
-   w, \mu \vdash e_2 \Rightarrow \strv ~ \seq{(s_2, v_2)} \\
-  \hline
-  w, \mu \vdash \compe ~ (e_1, e_2) \Rightarrow \strv ~ (\seq{(s_1, v_1)} ~ \seq{(s_2, v_2)}) \\
-%
-% Mem
-\newline \\
-  w, \mu \vdash e_1 \Rightarrow v_1 \qquad
-  w, \mu \vdash e_2 \Rightarrow \listv ~ \seq{v_2} \\
-  \hline
-  w, \mu \vdash \meme ~ (e_1, e_2) \Rightarrow \boolv ~ (v_1 \in \seq{v_2}) \\
-%
-% Choose
-\newline \\
-  w, \mu \vdash e \Rightarrow \listv ~ \seq{v} \qquad
-  v \in \seq{v} \\
-  \hline
-  w, \mu \vdash \choosee ~ e \Rightarrow v \\
-%
-% Len
-\newline \\
-  w, \mu \vdash e \Rightarrow \listv ~ \seq{v} \\
-  \hline
-  w, \mu \vdash \lene ~ e \Rightarrow \numv ~ |\seq v| \\
-%
-% IsCaseOf
-\newline \\
-  w, \mu \vdash e \Rightarrow \casev ~ (s', \seq{v}) \\
-  \hline
-  w, \mu \vdash \iscaseofe ~ (e, s) \Rightarrow \boolv ~ (s = s') \\
-%
-% GetCurCtx
-\newline \\
-  wc = \getcurctx(w) \\
-  \hline
-  w, \mu \vdash \getcurctxe \Rightarrow wc \\
-%
-% CtxKind
-\newline \\
-  \casev (s', \seq v) = \getcurctx(w) \\
-  \hline
-  w, \mu \vdash \ctxkinde ~ s \Rightarrow \boolv ~ (s = s') \\
-%
-% Iter-
-\newline \\
-  \,~\mu(s_1) = \listv ~ (v_{11} ~ ... ~ v_{1n}) \quad ... \quad
-  \mu(s_m) = \listv ~ (v_{m1} ~ ... ~ v_{mn}) \\
-  \mu_1 = \mu[s_1 \mapsto v_{11}, ~ ... ~ s_m \mapsto v_{m1}] \quad ... \quad
-  \mu_n = \mu[s_1 \mapsto v_{1n}, ~ ... ~ s_m \mapsto v_{mn}] \\
-  w, \mu_1 \vdash e \Rightarrow v_1 \quad ... \quad
-  w, \mu_n \vdash e \Rightarrow v_n \\
-  \hline
-  w, \mu \vdash \itere ~ (e, \listiter, s_1 ~ ... ~ s_m) \Rightarrow
-  \listv ~ v_1 ~ ... ~ v_n \\
-%
-% Iter-n
-\newline \\
-  w, \mu \vdash e_2 \Rightarrow \numv ~ n \\
-  \,~\mu(s_1) = \listv ~ (v_{11} ~ ... ~ v_{1n}) \quad ... \quad
-  \mu(s_m) = \listv ~ (v_{m1} ~ ... ~ v_{mn}) \\
-  \mu_1 = \mu[s_1 \mapsto v_{11}, ~ ... ~ s_m \mapsto v_{m1}] \quad ... \quad
-  \mu_n = \mu[s_1 \mapsto v_{1n}, ~ ... ~ s_m \mapsto v_{mn}] \\
-  w, \mu_1 \vdash e \Rightarrow v_1 \quad ... \quad
-  w, \mu_n \vdash e \Rightarrow v_n \\
-  \hline
-  w, \mu \vdash \itere ~ (e_1, \listniter ~ e_2, s_1 ~ ... s_m) \Rightarrow
-  \listv ~ v_1 ~ ... ~ v_n \\
-%
-% Iter-idx
-\newline \\
-  w, \mu \vdash e_2 \Rightarrow \numv ~ n \\
-  \mu_1 = \mu[s_1 \mapsto \numv ~ 0, ~ ... ~ s_m \mapsto \numv ~ 0] \quad ... \quad
-  \mu_n = \mu[s_1 \mapsto \numv ~ n, ~ ... ~ s_m \mapsto \numv ~ n] \\
-  w, \mu_1 \vdash e \Rightarrow v_1 \quad ... \quad
-  w, \mu_n \vdash e \Rightarrow v_n \\
-  \hline
-  w, \mu \vdash \itere ~ (e_1, \listidxiter ~ e_2, s_1 ~ ... ~ s_m) \Rightarrow
-  \listv ~ v_1 ~ ... ~ v_n \\
-%
-% Match
-\newline \\
-  w, \mu \vdash e_1 \Rightarrow v_1 \qquad
-  w, \mu \vdash e_2 \Rightarrow v_2 \\
-  \hline
-  w, \mu \vdash \matche ~ (e_1, e_2) \Rightarrow \boolv ~ (\match(v_1, v_2)) \\
-%
-% HasType
-\newline \\
-  w, \mu \vdash e \Rightarrow v \\
-  \hline
-  w, \mu \vdash \hastypee ~ (e, s) \Rightarrow \boolv ~ (\hastype(v, s)) \\
-\end{gather*}
-
-Given a Wasm state $w$ and an environment $\mu$, an expression $e$ evaluates
-to a value.
-% Var
-$\vare ~ s${} looks up the environment $\mu$ and gets the value.
-% Num Bool Fname
-$\nume ~ n${}, $\boole ~ b${}, and $\fnamee ~ s${} directly reduce to $\numv ~
-n${}, $\boolv ~ b${}, and $\fnamev ~ s${}, respectively.
-% Tuple Case List Struct
-$\liste ~ e_1 ~ ... ~ e_n${}, $\stre ~ (s_1, e_1) ~ ... ~ (s_n, e_n)${}, $\tupe
-~ e_1 ~ ... ~ e_n${}, and $\casee ~ (s, e_1 ~ ... ~ e_n)${} evaluate their
-sub-expressions and reduce to $\listv ~ v_1 ~ ... ~ v_n${}, $\strv ~ (s_1, v_1)
-~ ... ~ (s_n, v_n)${}, $\tupv ~ v_1 ~ ... ~ v_n${}, and $\casev ~ (s, v_1 ~ ...
-~ v_n)${}, respectively.
-% Un Bin
-$\une ~ (unop, e)${} and $\bine ~ (binop, e_1, e_2)${} evaluate their
-sub-expressions and perform \unop{} and \binop{}, respectively.
-% Acc
-$\acce ~ (e, p)${} evaluates $e$ and accesses the resulting value $v$ using a
-path $p$.
-% Upd
-$\upde ~ (e_1, p^+, e_2)${} evaluates $e_1$ and $e_2$ to $v_1$ and $v_2$,
-updating the value at location $v_1$ along paths $p^+$ to the value $v_2$.
-% Cat Comp
-$\cate ~ (e_1, e_2)${} and $\compe ~ (e_1, e_2)${} each evaluate two
-sub-expressions $e_1$ and $e_2$, concatenating the resulting lists and
-resulting records, respectively.
-% Mem
-$\meme ~ (e_1, e_2)${} evaluates $e_1$ and $e_2$ to $v_1$ and $\listv ~ \seq
-v_2$, and checks whether $v_1$ is a member of $\seq v_2$.
-% Choose Len
-$\choosee ~ e$ and $\lene ~ e${} each evaluate $e$ to the list value $\listv ~
-\seq v$, choosing an element from the list and calculating the length of the
-list, respectively.
-% IsCaseOf
-$\iscaseofe ~ (e, s)${} evaluates $e$ to the tagged tuple $\casev ~ (s', \seq
-v)$ and checks whether the tag $s'$ matches $s$.
-% GetCurCtx
-\getcurctxe{} retrieves the current Wasm context from the interleaved stack.
-% CtxKind
-$\ctxkinde ~ s${} retrieves the current Wasm context $\casev ~ (s', \seq v)$
-and checks whether the kind $s'$ matches $s$.
-% Iter
-$\itere ~ (e, iter, s^*)${} behaves differently depending on $iter$.
-\listiter{} and $\listniter ~ e_2${} are essentially equivalent.
-They assume that all the values mapped to the variables $\seq s$ by the
-environment $\mu${} are list values with the same length $n$.
-$\listniter ~ e_2${} checks whether the number $n$ matches the result of
-evaluating $e_2$, while \listiter{} allows any number.
-It generates a sequence of environments $\mu_1 ~ ... ~ \mu_n$ where the k-th
-environment maps each variable to the k-th elements of each list.
-Each environment is used to evaluate $e_1$, producing a sequence of values $v_1
-~ ... ~ v_n$.
-Finally, it results in $\listv ~ v_1 ~ ... ~ v_n$.
-$\listidxiter ~ e_2${} behaves similarly except that the variables $\seq s$ are
-unbound.
-It evaluates $e_2$ to obtain $n$ and then behaves like $\listniter ~ e_2${} as
-if a list of values [0 ... n-1] were bound to the variables $s^*$.
-% IsValid Match HasType
-$\matche ~ (e_1, e_2)${} and $\hastypee ~ (e, s)${} evaluate their
-sub-expressions and perform \match{} and \hastype{}, respectively.
-Note that \match{} and \hastype{} represent the match relation and the type
-relation in the Wasm static semantics.
-Their formalization is omitted as they are highly specific to Wasm and
-excessively verbose.
-
-%% Helper functions
-
-\begin{align*}
-%
-% get_algo_name
-\newline \\
-  &\getalgoname(\rel ~ (s, \seq e, \seq i)) = s \\
-  &\getalgoname(\fun ~ (s, \seq e, \seq i)) = s \\
-%
-% lookup
-\newline \\
-  &\lookup(\epsilon, s) = \epsilon \\
-  &\lookup(a ~ \seq{a}, s) =
-    \begin{cases}
-      a &\quad\quad \premise{if}\quad ~ s = \getalgoname(a) \\
-      \lookup(\seq{a}, s) &\quad\quad \premise{otherwise}
-    \end{cases}
-  \\
-%
-% assign
-\newline \\
-  &\assign(\vare ~ s, v) = [s \mapsto v] \\
-  &\assign(\itere ~ (e, iter, s^*), v) = \assign(e, v) \\
-  &\assign(\tupe ~ e_1 ~ ... ~ e_n, \tupv ~ v_1 ~ ... ~ v_n) =
-    \assign(e_1, v_1) \cdot ~ ... ~ \cdot \assign(e_n, v_n) \\
-  &\assign(\liste ~ e_1 ~ ... ~ e_n, \listv ~ v_1 ~ ... ~ v_n) =
-    \assign(e_1, v_1) \cdot ~ ... ~ \cdot \assign(e_n, v_n) \\
-  &\assign(\casee ~ (s, e_1 ~ ... ~ e_n), \casev ~ (s, v_1 ~ ... ~ v_n)) =
-    \assign(e_1, v_1) \cdot ~ ... ~ \cdot \assign(e_n, v_n) \\
-  &\assign(
-    w,
-    \mu,
-    \stre ~ (s_1, e_1) ~ ... ~ (s_2, e_n),
-    \strv ~ (s_1, v_1) ~ ... ~ (s_n, v_n)
-  ) =
-    \assign(e_1, v_1) \cdot ~ ... ~ \cdot \assign(e_n, v_n) \\
-  &\assign(\cate ~ (\liste ~ e_1 ~ ... ~ e_m, e), \listv v_1 ~ ... ~ v_n) =
-    \assign(e_1, v_1) \cdot ~ ... ~ \cdot \assign(e_m, v_m) \cdot
-    \assign(e, \listv ~ v_{m+1} ~ ... ~ v_n) \\
-  &\assign(\cate ~ (e, \liste ~ e_1 ~ ... ~ e_m), \listv v_1 ~ ... ~ v_n) =
-    \assign(e, \listv ~ v_1 ~ ... ~ v_{n-m}) \cdot
-    \assign(e_1, v_{n-m+1}) \cdot ~ ... ~ \cdot \assign(e_m, v_n) \\
-%
-% create_algo
-\newline \\
-  &\createalgo(\seq a, s, \seq v, k) =
-  \algo(\seq i, \mu, \epsilon, k) \\
-  &\qquad\qquad\qquad\qquad\premise{if}\quad
-  (\rel ~ (s, \seq e, \seq i) \quad\lor\quad \fun ~ (s, \seq e, \seq i)) = \lookup(\seq a, s)
-  \quad\land\quad
-  \mu = \assign(w, [\text{``s"} \mapsto \storev], \seq e, \seq v) \\
-%
-% split_ctx
-\newline \\
-  &\splitctx(\epsilon) = (\epsilon, \epsilon, \epsilon) \\
-  &\splitctx(we ~ \seq{we}) =
-    \begin{cases}
-      (\epsilon, wc, \seq{we})
-      &\quad\quad \premise{if}\quad
-      we = wc \\
-      (wv ~ \seq{wv}, wc, \seq{we'})
-      &\quad\quad \premise{if}\quad
-      we = wv \quad\land\quad
-      (\seq{wv}, wc, \seq{we'}) = \splitctx(\seq{we}) \\
-    \end{cases}
-  \\
-%
-% pop-wasm-ctx_context
-\newline \\
-  &\popwasmctx_{C}((\seq i, \mu, epsilon, k)) = (\seq i, \mu, \epsilon, \popwasmctx_K(k)) \\
-  &\popwasmctx_{C}((\seq i, \mu, wc ~ \seq{wc}, k)) = (\seq i, \mu, wc^*, k) \\
-%
-\newline \\
-% pop-wasm-ctx_cont-mt
-  &\popwasmctx_{K}(\mt) = \epsilon \\
-% pop-wasm-ctx_cont-toplevelcall
-  &\popwasmctx_{K}(\toplevelcall ~ (s, v^*)) = \epsilon \\
-% pop-wasm-ctx_cont-call
-  &\popwasmctx_{K}(\call ~ (s_1, s_2, \seq v, k)) =
-    \call ~ (s_1, s_2, \seq v, (\popwasmctx_K(k)) \\
-% pop-wasm-ctx_cont-execute
-  &\popwasmctx_{K}(\exe ~ (v, k)) = \exe ~ (v, (\popwasmctx_K(k)) \\
-% pop-wasm-ctx_cont-wasm_0
-  &\popwasmctx_{K}(\wasm ~ (\epsilon, k)) = \wasm ~ (\epsilon, \popwasmctx_{K}(k)) \\
-% pop-wasm-ctx_cont-wasm_nonzero
-  &\popwasmctx_{K}(\wasm ~ (wc ~ \seq{wc}, k)) = \wasm ~ (\seq{wc}, k) \\
-% pop-wasm-ctx_cont-al
-  &\popwasmctx_{K}(\algo ~ c) = \algo ~ (\popwasmctx_C(c)) \\
-% pop-wasm-ctx_cont-ret
-  &\popwasmctx_{K}(\ret ~ (v, k)) = \ret ~ (v, \popwasmctx_K(k)) \\
-%
-% exit
-\newline \\
-  &\exit(wi ~ \seq{wi}) =
-    \begin{cases}
-      \seq{wi} &\quad\quad \premise{if}\quad ~ \isendinstr(wi) \\
-      \exit(\seq{wi}) &\quad\quad \premise{otherwise} \\
-    \end{cases}
-  \\
-%
-% get_env
-\newline \\
-  &\getenv((\seq i, \mu, wc^*, k)) = \mu \\
-%
-% set_env
-\newline \\
-  &\setenv((\seq i, \mu, wc^*, k), \mu') = (\seq i, \mu', wc^*, k) \\
-%
-% prepend_instr
-\newline \\
-  &\prependinstr((\seq i, \mu, wc^*, k), \seq{i'}) = (\seq{i'} ~ \seq i, \mu, wc^*, k) \\
-%
-% add_ctx
-\newline \\
-  &\addctx((\seq i, \mu, wc^*, k), wc') = (\seq i, \mu, wc' ~ wc^*, k) \\
-%
-% get_store
-\newline \\
-  &\getstore((\seq{we}, \seq{wi}, sto)) = sto \\
-%
-% set_store
-\newline \\
-  &\setstore((\seq{we}, \seq{wi}, sto), sto') = (\seq{we}, \seq{wi}, sto') \\
-%
-% pop_wasm_instr
-\newline \\
-  &\popwasminstr((\seq{we}, wi ~ \seq{wi}, sto)) = ((\seq{we}, \seq{wi}, sto), wi) \\
-%
-% push
-\newline \\
-  &\push((\seq{we}, \seq{wi}, sto), we) = (we ~ \seq{we}, \seq{wi}, sto) \\
-%
-% pop
-\newline \\
-  &\pop((we ~ \seq{we}, \seq{wi}, sto)) = ((\seq{we}, \seq{wi}, sto), we) \\
-%
-% popn
-\newline \\
-  &\popn((we^n ~ \seq{we}, \seq{wi}, sto), n) = ((\seq{we}, \seq{wi}, sto), we^n) \\
-%
-% unop
-\newline \\
-  &\unop(\notop, v) =
-    \begin{cases}
-      \boolv ~ true &\quad\quad\premise{if}\quad ~ \istrue(v) \\
-      \boolv ~ false &\quad\quad\premise{otherwise} \\
-    \end{cases}
-  \\
-  &\unop(\minusop, \numv ~ n) = \numv ~ (-n) \\
-%
-% get_cur_ctx
-\newline \\
-  &\getcurctx((\seq{we}, \seq{wi}, sto)) =
-  wc \quad\quad\premise{if}\quad ~ (\seq{wv}, wc, \seq{we'}) = \splitctx(\seq{we}) \\
-%
-% get_cur_frame-frame
-\newline \\
-  &\getcurframe((\seq{we}, \seq{wi}, sto)) =
-    \begin{cases}
-      wc &\quad\quad\premise{if}\quad
-      (\seq{wv}, wc, \seq{we'}) = \splitctx(\seq{we}) \quad\land\quad \isframe(wc)\\
-      \getcurframe(\seq{we'}) &\quad\quad\premise{if}\quad
-      (\seq{wv}, wc, \seq{we'}) = \splitctx(\seq{we}) \quad\land\quad \neg \isframe(wc) \\
-    \end{cases}
-  \\
-%
-% set_cur_frame-frame
-\newline \\
-  &\setcurframe((\seq{we}, \seq{wi}, sto), wc_{frame})
-  =
-  \begin{cases}
-    (\seq{wv} ~ wc_{frame} ~ \seq{we'}, \seq{wi}, sto) \\
-    \qquad\qquad\qquad\quad\premise{if}\quad
-    (\seq{wv}, wc, \seq{we'}) = \splitctx(\seq{we}) \quad \isframe(wc) \\
-    (\seq{wv} ~ wc ~ \setcurframe(\seq{we'}, wc_{frame}), \seq{wi}, sto) \\
-    \qquad\qquad\qquad\quad\premise{if}\quad
-    (\seq{wv}, wc, \seq{we'}) = \splitctx(\seq{we}) \quad \neg \isframe(wc) \\
-  \end{cases}
-  \\
-%
-% access-idx
-\newline \\
-  &\access(w, \mu, \listv ~ \seq{v}, \idxp ~ e) = \seq{v}[n]
-  \qquad\qquad\qquad~\,\quad\premise{if}\quad w, \mu \vdash e \Rightarrow \numv ~ n \\
-% access-slice
-  &\access(w, \mu, \listv ~ \seq{v}, \slicep ~ (e_1, e_2)) = \seq{v}[n_1: n_2]
-  \quad\quad\premise{if}\quad
-  w, \mu \vdash e_1 \Rightarrow \numv ~ n_1 \quad\land\quad
-  w, \mu \vdash e_2 \Rightarrow \numv ~ n_2 \\
-% access-dot
-  &\access(w, \mu, \strv ~ ((s_0, v_0) ~ \seq{(s_1, v_1)}), \dotp ~ s) =
-  \begin{cases}
-    v_0 &\quad\quad\premise{if}\quad s_0 = s \\
-    \access(w, \mu, \strv ~ \seq{(s_1, v_1)}, \dotp ~ s) &\quad\quad\premise{otherwise} \\
-  \end{cases}
-  \\
-% access-store
-  &\access(w, \mu, \storev, p) =\access(w, \mu, \strv ~ sto, p)
-  \quad\quad\premise{if}\quad w = (\seq{we}, \seq{wi}, sto)\\
-%
-% update
-\newline \\
-  % idx
-  &\update(w, \mu, \listv ~ \seq{v}, (\idxp ~ e) ~ \seq{p}, v) =
-  \listv ~ (\updateidx(\seq v, n, \update(w, \mu, \seq{v}[n], \seq{p}, v))) \\
-  &\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\premise{if}\quad
-  w, \mu \vdash e \Rightarrow \numv ~ n \\
-  % slice
-  &\update(w, \mu, \listv ~ \seq{v}, (\slicep ~ e) ~ \seq{p}, v) =
-  \listv ~ (\updateslice(\seq v, n_1, n_2, \update(w, \mu, \seq{v_i}[n_1: n_2], \seq{p}, v))) \\
-  &\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\premise{if}\quad
-  w, \mu \vdash e_1 \Rightarrow \numv ~ n_1 \quad\land\quad
-  w, \mu \vdash e_2 \Rightarrow \numv ~ n_2 \\
-  % dot
-  &\update(w, \mu, \strv ~ \seq{(s, v)}, (\dotp ~ s') ~ \seq{p}, v)
-  =
-  \strv ~ (\updatedot(\seq{(s, v)}, s', \update(w, \mu, \access(v, \dotp ~ s'), \seq{p}, v))) \\
-%
-% update_idx
-\newline \\
-  &\updateidx(v_0 ~ \seq{v_1}, n, v_2) =
-  \begin{cases}
-    v_2 ~ \seq{v_1}
-    &\quad\quad\premise{if}\quad n = 0 \\
-    v_0 ~ \updateidx(\seq{v_1}, n-1, v_2)
-    &\quad\quad\premise{if}\quad n > 0 \\
-  \end{cases} \\
-%
-% update_slice
-\newline \\
-  &\updateslice(v_0^n ~ \seq{v_1}, 0, n, v_2^n) = v_2^n ~ \seq{v_1} \\
-  &\updateslice(v_0 ~ \seq{v_1}, m, n, v_2^n) =  v_0 ~ \updateslice(\seq{v_1}, m-1, n, v_2)
-  \quad\quad\premise{if}\quad m > 0 \\
-%
-% update_dot
-\newline \\
-  &\updatedot((s_0, v_0) ~ \seq{(s_1, v_1)}, s_2, v_2) =
-  \begin{cases}
-    (s_2, v_2) ~ \seq{(s_1, v_1)}
-    &\quad\quad\premise{if}\quad s_0 = s_2 \\
-    (s_0, v_0) ~ \updatedot(\seq{(s_1, v_1)}, s_2, v_2)
-    &\quad\quad\premise{otherwise}
-  \end{cases}
-  \\
-%
-% get_end_algo
-\newline \\
-  &\getendinstr(wc) =
-  \begin{cases}
-    \casev ~ (\text{``Exiting"}, \epsilon) &\quad\quad\premise{if}\quad wc = \casev ~ (\text{``Label"}, \seq v)\\
-    \casev ~ (\text{``Returning"}, \epsilon) &\quad\quad\premise{if}\quad wc = \casev ~ (\text{``Frame"}, \seq v)
-  \end{cases} \\
-%
-% is_end_instr
-\newline \\
-  &\isendinstr(wi) =
-  \begin{cases}
-    true &\quad\quad\premise{if}\quad wi = \casev ~ (\text{``Exiting"}, \epsilon) \quad\lor\quad wi = \casev ~ (\text{``Returning"}, \epsilon)\\
-    false &\quad\quad\premise{otherwise}
-  \end{cases} \\
-%
-% is_true
-\newline \\
-  &\istrue(v) =
-  \begin{cases}
-  b &\quad\quad\premise{if}\quad v = \boolv ~ b \\
-    \istrue(v_1) ~ \land ~ ... ~ \land ~ \istrue(v_n) &\quad\quad\premise{if}\quad
-    v = \listv ~ v_1 ~ ... ~ v_n \\
-  \end{cases} \\
-%
-% is_frame
-\newline \\
-  &\isframe(v) =
-  \begin{cases}
-    true
-    &\quad\quad\premise{if}\quad v = \casev ~ (\text{"FRAME"}, \seq v) \\
-    false &\quad\quad\premise{otherwise}\\
-  \end{cases}
-  \\
-\end{align*}

### 2024-11-21 paper/overview.tex
Add mechanized specification explanation

@@
 artifacts from the specification and ensuring the correctness of the
-specification. \Cref{fig:overview} illustrates an overview of our approach.
+specification.
+\Cref{fig:overview} illustrates an overview of our approach.
+% SpecTec: specification mechanization tool
+SpecTec is a tool for \textit{mechanizing} WebAssembly specification
+~\cite{spectec}.
+Specification mechanization is a technique that treats a specification as data
+that can be manipulated by a machine to automatically generate parser and
+interpreter ~\cite{jiset}, to generate tests and perform differential tests
+~\cite{jest}, to find meta-level errors in the specification ~\cite{jstar}, and
+to perform meta-level static analysis for the static analysis of the defined
+language ~\cite{jsaver}.
+This technique is especially crucial in growing languages such as JavaScript
+and WebAssembly.
+
 % DSL
@@
 thereby ensuring its correctness.
+
+\red{(Explanation for each chapter for SpecTec)}

### 2024-11-21 paper/background.tex
Write background

--- 2024-11-20 paper/background.tex
@@
 
-\chapter{Background}
-\label{ch:background}
-\noindent
+\chapter{Background} \label{ch:background} \noindent
 
-\red{TODO: fix fig name}
+\section{WebAssembly} \label{sec:webassembly}
 
-\section{WebAssembly}
-\label{sec:webassembly}
+WebAssembly (Wasm) is a portable, binary instruction format designed for safe
+and efficient execution across diverse environments.
+Initially developed to enhance performance in web browsers, diverse ecosystems
+leverages the advantages of Wasm, including cloud and edge
+computing~\cite{lucet, cloudflare}, the Internet of Things (IoT)\cite{wasm-iot}, and
+blockchain technologies\cite{wasm-blockchain}.
+Wasm programs are compiled from high-level languages (e.g., C, C++, Rust) into
+a compact, platform-independent binary format, enabling consistent execution
+across different hardware and operating systems.
 
-% breif overview
-WebAssembly~\cite{wasm} is a low-level bytecode language that is safe, fast,
-portable, and compact.
-It is widely used as a compile target for web applications.
-Not only that, other areas such as edge computing~\cite{wasm-edge},
-IoT~\cite{wasm-iot}, and block chains~\cite{wasm-block} deploy the advantages
-of WebAssembly.
+% Structure
+A Wasm program is comprised of \textit{modules}.
+A module is the unit of deployment, loading, and compilation, which contains
+definitions such as functions, tables, memories, and globals.
+Functions represent units of executable code.
+Each function has a type specifying its parameter and result types, along with
+a body composed of Wasm instructions.
+Tables represent indexed collections of references, such as function
+references.
+Tables are commonly used for indirect function calls, enabling features like
+dynamic dispatch.
+Memories represent linear memory, a contiguous, resizable array of bytes that
+serves as the module’s heap.
+Memory is used for storing and accessing data such as arrays, strings, or other
+data structures.
+Globals represent mutable or immutable variables that hold single values.
+Globals can store values such as integers or floating-point numbers and are
+accessible throughout the module.
 
-% risk of implementation divergence
-There are dozens of WebAssembly engines; all the browsers have there own
-implementations of WebAssembly with multiple tiers ~\cite{v8}
-\cite{spidermonkey} \cite{webkit}, and there are many engines that target for
-specific areas including embedded systems and edge computing.
-However, as WebAssembly should be portable across these implementations, it risks
-the implementations divergence.
-
-% rigorous standardization -> spec is rigorous: formal notation & prose notation
-To mitigate this problem, WebAssembly has been standardized very rigourously by
-the W3C~\cite{wasm-w3c}.
-Especially, WebAssembly specification document is written very rigorously.
-It describes the semantics of WebAssembly in two forms: formal notation and
-prose notation.
-Formal notation uses mathematical rules to compactly describe the semantics,
-and it is used for proofs such as type soundness.
-On the other hand, prose notation uses psudocodes algorithm to explain the
-semantics through a step by step instruction.
-Most WebAssembly users and engine developers are not familiar with
-mathematical rules, so they utilize the prose notation.
-
-% structure
-A WebAssembly program is comprised of \textit{modules}.
-A module is the unit of deployment, loading, and compilation, which consists of
-definitions such as functions, globals, and memories.
-A module is instantiated to validate the module and allocate each definition to
-the data structure named \textit{store}.
-This results in a module instance, a runtime representation for the module.
-After the instantiation, functions can be invoked so that the function body,
-which is a sequence of instructions, is executed.
+% Instantiation & store
+Before executing the code in a module, the module must be instantiated to
+validate its content and allocate its definitions into a runtime data structure
+called the \textit{store}.
+The store acts as the memory space where executable code for functions are
+stored, where function references are placed in tables, where the linear memory
+is allocated for data, and where global variables are maintained.
+Once instantiated, functions in the module can be invoked.
+Invoking a function triggers the execution of its body, a sequence of
+Wasm instructions.
+These instructions operate on the module’s internal state, including the values
+stored in the store.
 
 % execution
-WebAssembly execution is based on a stack machine.
-An instruction consumes values in the stack as operands, performs some
-operation, produces resulting values, and pushes them to the stack.
-\cref{fig:testop} is the WebAssembly specification document for the
+Wasm execution is based on a stack machine.
+An instruction consumes values from the stack as operands, performs an
+operation, and pushes the resulting values onto the stack.
+For example, the \texttt{testop} instruction exemplifies how this execution
+model operates in practice.
+\cref{fig:testop} illustrates the Wasm specification document for the
 \texttt{testop} instruction.
-Prose notation is written in the upper part, and formal notation is written in
-the lower grey box.
-In the prose notation, it pops a value from the stack, performs a test
-operation, and pushes the result to the stack.
-It is written compactly in the form of reduction rule in the formal notation
-which consists of LHS, RHS, and premises: $(t.const ~ c_1) ~ t.testop$,
-$(i32.const ~ c)$, and $c = testop_t(c_1)$.
-The LHS means that the value at the top of the stack is $(t.const ~ c_1)$,
-and an instruction to execute is $t.testop$.
-The RHS means that the value at the top of the stack is changed to $(i32.const
-~ c)$, and this happens when the premise $c = testop_t(c_1)$ is satisfied.
-
-% control structure
-The stack stores not only the input and output values of the instructions but also
-structures related to the control flow.
-A \textit{call frame} and a \textit{label} are used for function call and
-branching, respectively.
-In exception handling proposal, there is a \textit{handler} for exception
-handling, and other strucutres could be introduced in the future.
-How these control structures are used for the control flow will be discussed in
-\cref{sec:control flow in official prose}.
+It describes its semantics using two complementary forms: prose notation and
+formal notation.
+The prose notation is written in the upper part, and the formal notation is
+written in the lower grey box.
+In the prose notation, the instruction pops a value from the stack, performs a
+test operation, and pushes the result back onto the stack.
+In the formal notation, this is compactly expressed as a reduction rule
+consisting of the left-hand side (LHS), right-hand side (RHS), and premises:
+$(t.const ~ c_1) ~ t.testop$, $(i32.const ~ c)$, and $c = testop_t(c_1)$,
+respectively.
+The LHS indicates that the value at the top of the stack is $(t.const ~ c_1)$,
+and the instruction to execute is $t.testop$.
+The RHS shows that the value at the top of the stack is updated to $(i32.const
+~ c)$, which occurs if the premise $c = testop_t(c_1)$ holds true.
+% Additionally, the stack also stores structures related to the control flow
+% along with the input and output values of the instructions: A \textit{frame}
+% for function calls and a \textit{label} for branching.

### 2024-11-21 paper/dsl.tex
Write background

--- 2024-11-20 paper/overview.tex
@@
 
-\chapter{Overview}
-\label{ch:overview}
+\chapter{DSL}
+\label{ch:dsl}
 \noindent
-
-\begin{figure}[h]
-  \centerline{\includegraphics[width=15cm]{fig/overview}}
-  \caption[An overview of the SpecTec architecture]
-    {An overview of the SpecTec architecture}
-    \label{fig:overview}
-\end{figure}
-
-We propose SpecTec ...
-
-% Overview explanation
-In this chapter, we explain an overall approach to generate multiple artifacts
-from the specification and check correctness of the specification.
-\cref{fig:overview} illustrates the overview of our approach.
-SpecTec provides a domain specific language (DSL) for the specification.
-It is designed to be similar to the formal notation when describing the
-semantics with compact and user-friendly notation.
-It is type checked to prevent meta-level errors such as notation misuses and
-dimension mismatches.
-The code below is the specification for \texttt{testop} instruction written in
-the DSL.
-\begin{lstlisting}
-  rule Step_pure/testop:
-  (CONST nt c_1) (TESTOP nt testop)  ~>  (CONST I32 c)
-  -- if c = $testop_(nt, testop, c_1)
-\end{lstlisting}

### 2024-11-21 paper/model.tex
Write background

--- 2024-11-18 paper/3-motivation.tex
@@
-% !TEX root = main.tex
-
-\chapter{Motivation}
-\label{ch:motivation}
+\chapter{Model}
+\label{ch:Model}
 \noindent
 
+\section{Control flow in SpecTec prose}
+\label{sec:spectecp}
+
+% TODO: Remove these commands
 \newcommand{\officialp}{official prose}
@@
 
-\red{TODO: explanation of the terms \officialp{} and \spectecp{}} \\
-\red{TODO: remove redundant part in the fig} \\
-\red{TODO: explanation of context(frame/label): in the background? or here?} \\
-\red{TODO: the notion of meta-level interpreter in background} \\
-\red{TODO: try finding better terminology rather than "model"} \\
-
-\section{Control flow in official prose}
-\label{sec:control flow in official prose}
-
-% control flow structure in official prose
-It might be easy to understand how the \officialp{} explains the control flow
-of WebAssembly, if we assume that a WebAssembly code is loaded on a memory and
-a pc points to the instruction to execute.
-To describe control flow, it uses a structure named \textit{block} which
+% control structure
+It might be easy to understand how the control flow of Wasm works, if we assume
+that a Wasm code is loaded on a memory space and a pc points to the instruction
+to execute.
+To describe control flow, it introduces a structure named \textit{block} which
 consitutes of a instruction sequence.
-When executing the instructions in the block, pc can be changed to the starting
-point of the block or the end of the block.
+When branching instruction in the block is performed, pc can be changed to the
+starting point of the block or the end of the block.
 
@@
 
-\cref{ex:br} is a WebAssembly code example that has \texttt{loop} and
-\texttt{unreachable}
+\cref{ex:br} is a Wasm code example that has \texttt{loop} and
+\texttt{unreachable}.
 Here, result type of the \texttt{loop} is \texttt{i32} and the \texttt{loop}
@@
 \texttt{unreachable}.
-\cref{fig:loop} is the \officialp{} of the \texttt{loop} instruction.
-It says that the continuation is the start of the loop, the information is
-stored in a label, and it \textit{enters} the block with the label.
-The term \textit{enter} means that it pushes the label in the stack and makes
+\cref{fig:loop} is the specification for the \texttt{loop} instruction in the
+prose notation.
+In forth step, it says that the continuation is the start of the loop, the
+information is stored in a label.
+In the last step, it \textit{enters} the block with the label.
+The term \textit{enter} means that it pushes the label onto the stack and makes
 pc points to the first instruction in the block.
@@
 It is special because this behavior is performed without an explicit
-WebAssembly instruction.
+Wasm instruction.
 Rather, the behavior is performed when \textbf{the end of a block is reached}
@@
 
-\cref{ex:exit} is a WebAssembly code example that has \texttt{loop} whose block is
+\cref{ex:exit} is a Wasm code example that has \texttt{loop} whose block is
 only \texttt{i32.const}, and \texttt{f32.const}
@@
 \end{figure}
-
-\section{Control flow in SpecTec prose}
-\label{sec:spectecp}
 

### 2024-11-21 paper/dsl.tex
Add al and dsl style WIP: write dsl

@@
 \noindent
+
+% DSL high-level explanation
+SpecTec offers a domain-specific language (DSL) tailored for specification
+purposes.
+It is a declarative language, providing a compact and user-friendly syntax.
+It consists of DSL definitions: type, grammar, relation, and functions.
+Type definitions are used to define abstract syntax, while grammar definitions
+are used to define concrete syntax.
+The below code represents the abstract syntax and concrete syntax for
+\texttt{numtype}.
+\begin{lstlisting}[style=dsl]
+syntax numtype = I32 | I64 | F32 | F64 | V128
+grammar Bnumtype : numtype =
+  | 0x7F => I32
+  | 0x7E => I64
+  | 0x7D => F32
+  | 0x7C => F64
+\end{lstlisting}
+Relation definitions are used to define semantics, while function definitions
+are used auxiliary definitions.
+
+It is type checked to prevent meta-level errors such as notation misuses and
+dimension mismatches. The code below is the specification for \texttt{testop}
+instruction written in the DSL, which is designed to be similar to the formal
+notation when describing the semantics with compact and user-friendly
+notation.
+\begin{lstlisting}[style=dsl]
+  rule Step_pure/testop:
+  (CONST nt c_1) (TESTOP nt testop)  @$\tildecode$@>  (CONST I32 c)
+  -- if c = @$\dollarcode$@testop_(nt, testop, c_1)
+\end{lstlisting}
+
+% DSL
+The DSL consists of DSL definitions.
+There are mainly 4 kinds of DSL definitions: type, grammar, relation, and
+functions.
+Type definitions are used to define abstract syntax, while grammar definitions
+are used to define concrete syntax.
+Relation definitions are used to define semantics; the code above is also
+described in relation definition.
+Function definitions can be utilized for other definitions such as
+\texttt{\$testop\_} in the code.
+In addition, they can be also used to describe semantics.
+Especially, SpecTec uses function definitions to describe module instantiation
+and function invocation.

### 2024-11-21 paper/overview.tex
Write background
Add al and dsl style WIP: write dsl

@@
 
-% Overview explanation
-In this chapter, we explain the overall approach to generating multiple
-artifacts from the specification and ensuring the correctness of the
-specification.
-\Cref{fig:overview} illustrates an overview of our approach.
 % SpecTec: specification mechanization tool
@@
 language ~\cite{jsaver}.
-This technique is especially crucial in growing languages such as JavaScript
-and WebAssembly.
+% Overview explanation
+In this chapter, we explain the overall approach to generating multiple
+artifacts from the specification and ensuring the correctness of the
+specification.
+\Cref{fig:overview} illustrates an overview of our approach.
 
 % DSL
-SpecTec provides a domain-specific language (DSL) for writing specifications.
-It is a declarative language designed to resemble the formal notation while
-offering a compact and user-friendly syntax.
+SpecTec offers a domain specific language (DSL) designed for writing
+Wasm specifications.
+It is a declarative language, providing a compact and user-friendly syntax.
+This DSL can represent Wasm semantics in a manner closely resembling the formal
+notation as shown in the following code for specification of testop
+instruction.
 The code below shows the specification for the \texttt{testop} instruction
 written in the DSL:
-\begin{lstlisting}[escapechar=\%]
+\begin{lstlisting}[style=dsl]
   rule Step_pure/testop:
-  (CONST nt c_1) (TESTOP nt testop)  %$\tildecode$%>  (CONST I32 c)
-  -- if c = %$\dollarcode$%testop_(nt, testop, c_1)
+  (CONST nt c_1) (TESTOP nt testop)  @$\tildecode$@>  (CONST I32 c)
+  -- if c = @$\dollarcode$@testop_(nt, testop, c_1)
 \end{lstlisting}
@@
 generated from the DSL:
-\begin{lstlisting}[escapechar=\%]
-Step_pure/testop nt testop {
+\begin{lstlisting}[style=al]
+@\textbf{Step\_pure/testop}@ nt testop {
     Assert (top_value(nt))
     Pop (numtype_0.CONST c_1)
- Let c = %$\dollarcode$%testop_(nt, testop, c_1)
+    Let c = @$\dollarcode$@testop_(nt, testop, c_1)
     Push (I32.CONST c)
@@
 thereby ensuring its correctness.
-
-\red{(Explanation for each chapter for SpecTec)}

### 2024-11-21 paper/dsl.tex
Write type and relation in DSL

--- 2024-11-21 paper/dsl.tex
@@
 \noindent
+
+% DSL high-level explanation
+SpecTec offers a domain-specific language (DSL) tailored for specification
+purposes.
+The DSL is a declarative language, providing a compact and user-friendly syntax.
+It includes definitions for type, relation, and functions.
+Type definitions define the abstract syntax in a mannar analogous to the BNF
+format, where terminals are denoted by capital letters, while non-terminals are
+represented by lowercase letters.
+The below code represents the abstract syntax for \texttt{testop},
+\texttt{numtype} and \texttt{instr}.
+\begin{lstlisting}[style=dsl]
+syntax testop = EQZ
+syntax numtype = I32 | I64 | F32 | F64 | V128
+syntax instr =
+  | CONST numtype number
+  | TESTOP numtype testop
+  | ...
+\end{lstlisting}
+
+With relation, user can define the formal semantics of Wasm.
+The below code illustrates the execution semantics for \texttt{testop}
+instruction described in a relation definition.
+\begin{lstlisting}[style=dsl]
+  relation Step_pure: instr* ~> instr*
+  rule Step_pure/testop:
+  (CONST numtype c_1) (TESTOP numtype testop)  ~>  (CONST I32 c)
+  -- if c = @$\dollarcode$@testop_(numtype, testop, c_1)
+\end{lstlisting}
+The signature of the step relation $\tildecode$\texttt{>} is defined first.
+It then specifies a step rule that describes the semantics of the
+\texttt{testop} instruction with premises under which the relation holds.
+SpecTec performs type checking to ensure that any incorrect symbols or
+ill-typed relations like below code are rejected.
+\begin{lstlisting}[style=dsl]
+  (* Wrong type *)
+  rule Step_pure/testop:
+  (CONST numtype c_1) (TESTOP numtype testop)  ~>  c
+  -- if c = @$\dollarcode$@testop_(numtype, testop, c_1)
+
+  (* Notation misuse *)
+  rule Step_pure/testop:
+  (CONST numtype c_1) (TESTOP numtype testop)  =>  (CONST I32 c)
+  -- if c = @$\dollarcode$@testop_(numtype, testop, c_1)
+\end{lstlisting}
+
+TODO: funciton definition
+Function definitions can be utilized for other definitions such as
+\texttt{\$testop\_} in the code.
+In addition, they can be also used to describe semantics.
+Especially, SpecTec uses function definitions to describe module instantiation
+and function invocation.
+
+TODO: latex and mechanized definitions

### 2024-11-21 paper/overview.tex
Write type and relation in DSL

@@
   rule Step_pure/testop:
-  (CONST nt c_1) (TESTOP nt testop)  @$\tildecode$@>  (CONST I32 c)
+  (CONST nt c_1) (TESTOP nt testop)  ~>  (CONST I32 c)
   -- if c = @$\dollarcode$@testop_(nt, testop, c_1)

### 2024-11-22 paper/model.tex
Fix chapter title

@@
-\chapter{Model}
+\chapter{Algorithmic Specification Language}
 \label{ch:Model}
@@
 appended according to the sequence.
+
+\newcommand{\seq}[1]{#1^*}
+
+%% Syntax
+\section{Syntax of AL}
+\label{syntax}
+
+{
+\renewcommand{\arraystretch}{0.8835}  % Decrease row spacing locally
+\begin{align*}
+\begin{array}{lcccrlr}
+%
+% Algorithm
+  \text{Algorithm}\quad& A &\ni& a &::=& ~ \rel ~ (s, \seq e, \seq i) ~ | ~ \fun ~ (s, \seq e, \seq i) \\
+%
+% Instruction
+  \text{Instruction}\quad& I &\ni& i &::=& ~ \ifi ~ (e, \seq i, \seq i) &\quad\text{(If)} \\
+    &&&& | & ~ \eitheri ~ (\seq i, \seq i) &\quad\text{(Either)} \\
+    &&&& | & ~ \enteri ~ (e, e) &\quad\text{(Enter)} \\
+    &&&& | & ~ \pushctxi ~ e &\quad\text{(Push Context)} \\
+    &&&& | & ~ \pushi ~ e &\quad\text{(Push)} \\
+    &&&& | & ~ \popctxi ~ e &\quad\text{(Pop Context)} \\
+    &&&& | & ~ \popi ~ e &\quad\text{(Pop)} \\
+    &&&& | & ~ \popni ~ (e, e) &\quad\text{(Pop N)} \\
+    &&&& | & ~ \popalli ~ e &\quad\text{(Pop All)} \\
+    &&&& | & ~ \leti ~ (e, e) &\quad\text{(Let)} \\
+    &&&& | & ~ \trapi &\quad\text{(Trap)} \\
+    &&&& | & ~ \returnreli &\quad\text{(Return Relation)} \\
+    &&&& | & ~ \returnfuni ~ e &\quad\text{(Return Function)} \\
+    &&&& | & ~ \executei ~ e &\quad\text{(Execute)} \\
+    &&&& | & ~ \calli ~ (s, s, \seq{e}) &\quad\text{(Let Call)} \\
+    &&&& | & ~ \replaceframei ~ (p^+, e) &\quad\text{(Replace Frame)} \\
+    &&&& | & ~ \replacestorei ~ (p^+, e) &\quad\text{(Replace Store)} \\
+%
+% Expression
+  \text{Expression}\quad& E &\ni& e &::=& ~ \vare ~ s &\quad\text{(Variable)} \\
+    &&&& | & ~ \nume ~ n &\quad\text{(Number)} \\
+    &&&& | & ~ \boole ~ b &\quad\text{(Boolean)} \\
+    &&&& | & ~ \fnamee ~ s &\quad\text{(Function Name)} \\
+    &&&& | & ~ \liste ~ \seq e &\quad\text{(List)} \\
+    &&&& | & ~ \stre ~ \seq{(s, e)} &\quad\text{(Record)} \\
+    &&&& | & ~ \tupe ~ \seq e &\quad\text{(Tuple)} \\
+    &&&& | & ~ \casee ~ (s, \seq e) &\quad\text{(Tagged Tuple)} \\
+    &&&& | & ~ \une ~ (unop, e) &\quad\text{(Unary Operation)} \\
+    &&&& | & ~ \bine ~ (binop, e, e) &\quad\text{(Binary Operation)} \\
+    &&&& | & ~ \acce ~ (e, p) &\quad\text{(Access)} \\
+    &&&& | & ~ \upde ~ (e, p^+, e) &\quad\text{(Update)} \\
+    &&&& | & ~ \cate ~ (e, e) &\quad\text{(Concatenation)} \\
+    &&&& | & ~ \compe ~ (e, e) &\quad\text{(Composition)} \\
+    &&&& | & ~ \meme ~ (e, e) &\quad\text{(Membership)} \\
+    &&&& | & ~ \choosee ~ e &\quad\text{(Choose)} \\
+    &&&& | & ~ \lene ~ e &\quad\text{(Length)} \\
+    &&&& | & ~ \iscaseofe ~ (e, s) &\quad\text{(Check Tag)} \\
+    &&&& | & ~ \getcurctxe  &\quad\text{(Get Current Context)} \\
+    &&&& | & ~ \ctxkinde ~ s &\quad\text{(Context Kind)} \\
+    &&&& | & ~ \itere ~ (e, iter, \seq{s}) &\quad\text{(Iteration)} \\
+    &&&& | & ~ \matche ~ (e, e) &\quad\text{(Match)} \\
+    &&&& | & ~ \hastypee ~ (e, s) &\quad\text{(Has Type)} \\
+\end{array}
+\end{align*}
+}
+
+\newpage
+\begin{align*}
+\begin{array}{lcccrlr}
+%
+% Path
+  \text{Path}\quad& P &\ni& p &::=& ~ \idxp ~ e ~ | ~ \slicep ~ (e, e) ~ | ~ \dotp ~ s \\
+%
+% Iter
+  \text{Iter}\quad& Iter &\ni& iter &::=& ~ \listiter ~ | ~ \listniter ~ e ~|~ \listidxiter ~ (s, e) \\
+%
+% Operator
+  \text{Unary Operator}\quad& Unop &\ni& unop &::=& ~ \notop ~ | ~ \minusop \\
+  \text{Binary Operator}\quad& Binop &\ni& binop &::=& ~ \addop ~ | ~ \subop ~ | ~ \mulop \\
+    &&&& | & \divop ~ | ~ \modop ~ | ~ \expop \\
+    &&&& | & \implop ~ | ~ \equivop ~ | ~ \andop \\
+    &&&& | & \orop ~ | ~ \eqop ~ | ~ \neop ~ | ~ \ltop \\
+    &&&& | & \gtop ~ | ~ \leop ~ | ~ \geop \\
+% Primitives
+  \text{String}\quad& \mathbb S &\ni& s \\
+  \text{Integer}\quad& \mathbb Z &\ni& n \\
+  \text{Boolean}\quad& \mathbb B &\ni& b \\
+\end{array}
+\end{align*}
+
+% algorithm
+An algorithm is either an AL relation or an AL function, corresponding to the
+relation and the function definitions in the DSL.
+% instruction
+An AL instruction is \ifi{} for conditionally choosing a branch, \eitheri{} for
+nondeterministically choosing a branch, \enteri{} for entering a block,
+\pushctxi{} for pushing a Wasm context, \pushi{} for pushing a Wasm value,
+\popctxi{} for popping a Wasm context, \popi{} for popping a Wasm value,
+\popni{} for popping $n$ Wasm values, \popalli{} for popping all Wasm values
+within a Wasm context, \leti{} for let binding, \trapi{} for trapping,
+\returnreli{} for returning from an AL relation, \returnfuni{} for returning
+from an AL function, \executei{} for executing given Wasm instructions,
+\calli{} for an AL function call with let binding, \replaceframei{} for
+replacing a frame, or \replacestorei{} for replacing a store.
+% expression
+An expression is a variable, number, boolean, AL function name, list, record
+(or struct), tuple, tagged tuple, unary or binary operation, data structure
+access or update, list or record concatenation, membership check, element
+selection, length retrieval, tag check, Wasm context retrieval, Wasm context
+kind check, iteration, match relation, or type relation.
+
+\begin{align*}
+\begin{array}{lcccrlr}
+%
+% State
+  \text{State}\quad& \Sigma &\ni& \sigma & ::=& ~ \seq a, w, k \\
+%
+% Wasm state
+  \text{Wasm State}\quad& W &\ni& w &::=& ~ \seq{we}, \seq{wi}, sto \\
+%
+% Wasm Entry
+  \text{Wasm Entry}\quad& WE &\ni& we &::=& ~ wv ~ | ~ wc \\
+%
+% Wasm Value
+  \text{Wasm Value}\quad& WV &\ni& wv &::=& ~ \casev ~ (s, \seq v) \\
+%
+% Wasm Context
+  \text{Wasm Context}\quad& WC &\ni& wc &::=& ~ \casev ~ (\text{``Label"}, \seq v) ~ | ~ \casev ~ (\text{``Frame"}, \seq v) \\
+%
+% Wasm Instruction
+  \text{Wasm Instruction}\quad& WI &\ni& wi &::=& ~ \casev ~ (s, \seq v) \\
+%
+% Store
+  \text{Store}\quad& Store &\ni& sto &::=& ~ \seq{(s, v)} \\
+%
+% Continuation
+  \text{Continuation}\quad & K &\ni& k &::=& ~ \mt &\quad\text{(Empty)} \\
+    &&&& | & ~ \toplevelcall ~ (s, \seq v) &\quad\text{(Top-level Call)} \\
+    &&&& | & ~ \call ~ (s, s, \seq v, k) &\quad\text{(Call)} \\
+    &&&& | & ~ \exe ~ (\seq{wi}, k) &\quad\text{(Execute)} \\
+    &&&& | & ~ \wasm ~ (\seq{wc}, k) &\quad\text{(Wasm)} \\
+    &&&& | & ~ \algo ~ c &\quad\text{(Algorithm)} \\
+    &&&& | & ~ \ret ~ (v, k) &\quad\text{(Return)} \\
+%
+% Context
+  \text{Context}\quad& C &\ni& c &::=& ~ (\seq i, \mu, \seq{wc}, k) \\
+%
+% Environment
+  \text{Environment}\quad& M &\ni& \mu &::=& ~ \seq{[s \mapsto v]} \\
+\end{array}
+\end{align*}
+
+\newpage
+\begin{align*}
+\begin{array}{lcccrlr}
+%
+% Value
+  \text{Value}\quad& V &\ni& v &::=& ~ \numv ~ n &\quad\text{(Number)} \\
+    &&&& | & ~ \boolv ~ b &\quad\text{(Boolean)} \\
+    &&&& | & ~ \fnamev ~ s &\quad\text{(Function Name)} \\
+    &&&& | & ~ \listv ~ \seq v &\quad\text{(List)} \\
+    &&&& | & ~ \strv ~ \seq{(s, v)} &\quad\text{(Record)} \\
+    &&&& | & ~ \tupv ~ \seq v &\quad\text{(Tuple)} \\
+    &&&& | & ~ \casev ~ (s, \seq v) &\quad\text{(Tagged Tuple)} \\
+    &&&& | & ~ \trapv &\quad\text{(Trap)} \\
+    &&&& | & ~ \storev &\quad\text{(Store)} \\
+\end{array}
+\end{align*}
+
+% State
+An AL state consists of a sequence of algorithms, a Wasm state, and a
+continuation.
+% Wasm State
+A Wasm state consists of an interleaved stack, a Wasm instruction stack, and a
+store.
+% Wasm Entry
+An interleaved stack contains two types of entries: a Wasm value and a Wasm
+context.
+% Wasm Value & Wasm Context & Wasm Instruction
+Each of a Wasm value, a Wasm context, and a Wasm instruction is expressed as an
+AL value $\casev$.
+In particular, the tag of a Wasm context is either ``Label" or ``Frame".
+% Store
+A store is expressed as a sequence of pairs, each consisting of a field name
+and an AL value.
+% Continuation
+A continuation is \mt{} for an empty continuation, \toplevelcall{} for
+top-level function call, \call{} for a function call with let binding, \exe{}
+for executing given Wasm instructions, \wasm{} for executing a Wasm instruction
+in the instruction stack, \algo{} for interpreting AL instructions in an
+algorithm, or \ret{} for returning from an AL function.
+% Env
+An environment is a finite mapping from variable names to AL values.
+% Context
+An AL context consists of a sequence of Wasm instruction, an environment, a
+sequence of Wasm contexts, and a continuation.
+% Value
+An AL value is a number, boolean, function name, list, record, tuple, tagged tuple,
+trap, or store.
+
+% Terminology
+In this section, AL-specific terms (state, context, value, function, and
+instruction) are referred to simply as state, context, value, function, and
+instruction, while Wasm terms remain unchanged.
+
+\newpage
+%% Semantics
+\section{Semantics of AL}
+\label{semantics}
+
+%% Continuation
+
+\begin{gather*}
+\boxed{\leadsto \subseteq \Sigma \times \Sigma} \\
+%
+% TopLevelCall
+\newline \\
+  \\
+  \hline
+  (\seq{a}, w, \toplevelcall ~ (s, \seq v)) \leadsto (\seq{a}, w, \createalgo(\seq a, s, \seq v, \mt)) \\
+%
+% Call
+\newline \\
+  \\
+  \hline
+  (\seq{a}, w, \call ~ (s_{var}, s_{name}, \seq v, k)) \leadsto
+  (\seq{a}, w, \createalgo(\seq a, s_{name}, \seq v, \call ~ (s_{var}, s_{name}, \seq v, k))) \\
+%
+% Execute-empty
+\newline \\
+  \\
+  \hline
+  (\seq{a}, w, \exe ~ (\epsilon, k)) \leadsto (\seq{a}, w, k) \\
+%
+% Execute-instr
+\newline \\
+  \casev ~ (s, \seq v) = wi \\
+  \hline
+  (\seq{a}, w, \exe ~ (wi ~ \seq{wi}, k)) \leadsto
+  (\seq{a}, w, \createalgo(\seq a, s, \seq v, \exe ~ (\seq{wi}, k))) \\
+%
+% Wasm-empty
+\newline \\
+  \\
+  \hline
+  (\seq{a}, w, \wasm ~ (\epsilon, k)) \leadsto (\seq{a}, w, k) \\
+%
+% Wasm-instr
+\newline \\
+  (w', \casev ~ (s, \seq v)) = \popwasminstr(w) \\
+  \hline
+  (\seq{a}, w, \wasm ~ (wc^+, k))
+  \leadsto
+  (\seq{a}, w', \createalgo(\seq a, s, \seq v, \wasm ~ (wc^+, k))) \\
+%
+% Al-empty
+\newline \\
+  \\
+  \hline
+  (\seq{a}, w, \algo ~ (\epsilon, \mu, \seq{wc}, k)) \leadsto (\seq{a}, w, k) \\
+%
+% Al-instr
+\newline \\
+  \seq{a}, w, (\seq{i_1}, \mu, \seq{wc}, k) \vdash i_0 \Rightarrow (w', k') \\
+  \hline
+  (\seq{a}, w, \algo ~ (i_0 ~ \seq{i_1}, \mu, \seq{wc}, k)) \leadsto (\seq{a}, w', k') \\
+%
+% Return
+\newline \\
+  \mu = \getenv(c) \qquad c' = \setenv(c, \mu[s_{var} \mapsto v]) \\
+  \hline
+  (\seq{a}, w, \ret ~ (v, \call ~ (s_{var}, s_{name}, \seq v, \algo ~ c)) \leadsto
+  (\seq a, w, \algo ~ c') \\
+\end{gather*}
+
+The semantics of AL is defined by a state transition system.
+% TopLevelCall
+Given algorithms $\seq a$ generated from the DSL and a Wasm state $w$, a
+top-level function $s$ can be invoked with arguments $\seq v$ to perform
+either module instantiation or function invocation:
+$(\seq a, w, \toplevelcall ~ (s, \seq v))$.
+$\toplevelcall ~ (s, \seq v)${} transitions to \algo{} with the function name
+$s$, the arguments $\seq v$, and \mt{}.
+% Empty
+\mt{} is an empty continuation, representing the end of the transition, so no
+further transition occurs.
+% Call
+Similarly to the top-level call, $\call~ (s_{var}, s_{name}, \seq v, k)${}
+transitionsto \algo{} with the function name $s_{name}$ and the arguments $\seq
+v$, but with the current continuation nested within it.
+% Execute
+$\exe ~ (\seq{wi}, k)${} executes each Wasm instruction in the sequence
+$\seq{wi}$ one by one, each containing a relation name and arguments.
+The continuation transitions to \algo{} with the relation name, the arguments,
+and the current continuation, with the Wasm instruction being popped.
+% Wasm
+In $\wasm ~ (\seq{wc}, k)${}, a Wasm instruction is popped from
+the Wasm instruction stack and executed until the Wasm context sequence
+$\seq{wc}$ is exhausted.
+This execution transitions the continuation to \algo{}, with the relation name
+and the arguments of the Wasm instruction, and the current continuation nested
+within it.
+% Algo
+In $\algo ~ (\seq i, \mu, \seq{wc}, k)${}, the body instructions $\seq i$
+execute sequentially, allowing transitions to \call{}, \exe{}, \wasm{}, \ret{},
+or \algo{}.
+% Return
+The execution of a function call concludes with a return instruction, which
+changes the continuation to $\ret ~ v${}.
+If the function call originates from \call{}, the return value $v$ is assigned
+to the variable $s_{var}$.
+If it originates from \toplevelcall{}, the inner continuation $k$ is always
+\mt, so the entire execution concludes with the return value $v$:
+$
+(\seq a, w, \toplevelcall ~ (s, \seq v))
+\leadsto^*
+(\seq a, w', \ret ~ (v, \mt))
+$.
+
+%% Instruction
+
+\begin{gather*}
+  \boxed{\seq{a}, ~ w, ~ c \vdash i \Rightarrow w, ~ k} \\
+%
+% If-true
+\newline \\
+  w, \getenv(c) \vdash e \Rightarrow v \qquad
+  \istrue(v) \\
+  \hline
+  \seq{a}, w, c \vdash \ifi ~ (e, \seq{i_1}, \seq{i_2}) \Rightarrow
+  (w, \algo ~ (\prependinstr(c, \seq{i_1}))) \\
+%
+% If-false
+\newline \\
+  w, \getenv(k) \vdash e \Rightarrow v \qquad
+  \neg \istrue(v) \\
+  \hline
+  \seq{a}, w, c \vdash \ifi ~ (e, \seq{i_1}, \seq{i_2}) \Rightarrow
+  (w, \algo ~ (\prependinstr(c, \seq{i_2}))) \\
+%
+% Either-1
+\newline \\
+  \\
+  \hline
+  \seq{a}, w, c \vdash \eitheri ~ (\seq{i_1}, \seq{i_2}) \Rightarrow
+  (w, \algo ~ (\prependinstr(c, \seq{i_1}))) \\
+%
+% Either-2
+\newline \\
+  \\
+  \hline
+  \seq{a}, w, c \vdash \eitheri ~ (\seq{i_1}, \seq{i_2}) \Rightarrow
+  (w, \algo ~ (\prependinstr(c, \seq{i_2}))) \\
+%
+% Enter
+\newline \\
+  (\seq{we}, \seq{wi}, sto) = w \qquad
+  (i^*, \mu, wc_1 ~ ... ~ wc_n, k) = c \\
+  w, \mu \vdash e_1 \Rightarrow wc \qquad
+  w, \mu \vdash e_2 \Rightarrow \listv ~ \seq{wi_1} \\
+  wi = \getendinstr(wc) \qquad
+  wi_1 = \getendinstr(wc_1) \quad ... \quad wi_n = \getendinstr(wc_n) \\
+  wi_2^* = wi_1^* ~ wi ~ wi_1 ~ ... ~ wi_n \\
+  \hline
+  \seq{a}, w, c \vdash \enteri ~ (e_1, e_2)
+  \Rightarrow
+  (
+    (wc ~ \seq{we}, wi_2^* ~ \seq{wi}, sto),
+    \wasm ~ (wc ~ wc_1 ~ ... ~ wc_n, \algo ~ (i^*, \mu, \epsilon, k))
+  ) \\
+%
+% PushCtx
+\newline \\
+  w, \getenv(c) \vdash e \Rightarrow wc \\
+  \hline
+  \seq{a}, w, c \vdash \pushctxi ~ e
+  \Rightarrow
+  (\push(w, wc), \algo ~ (\addctx(c, wc))) \\
+%
+% Push
+\newline \\
+  w, \getenv(c) \vdash e \Rightarrow wv \\
+  \hline
+  \seq{a}, w, c \vdash \pushi ~ e \Rightarrow (\push(w, wv), \algo ~ c) \\
+%
+% PopCtx
+\newline \\
+  (\seq{we}, \seq{wi}, sto) = w \qquad
+  (\seq{wv}, wc, \seq{we'}) = \splitctx(\seq{we}) \\
+  \mu = \getenv(c) \cdot \assign(e, wc) \qquad
+  \seq{wi'} = \exit(\seq{wi}) \qquad
+  c' = \popwasmctx_{C}(c) \\
+  \hline
+  \seq{a}, w, c \vdash \popctxi ~ e
+  \Rightarrow
+  ((\seq{we'}, \seq{wi'}, sto), \algo ~ (\setenv(c', \mu))) \\
+%
+% Pop
+\newline \\
+  (w', wv) = \pop(w) \qquad
+  \mu' = \getenv(c) \cdot \assign(e, wv) \\
+  \hline
+  \seq{a}, w, c \vdash \popi ~ e \Rightarrow (w', \algo ~ (\setenv(c, \mu'))) \\
+%
+% PopN
+\newline \\
+  \mu = \getenv(c) \qquad
+  w, \mu \vdash e_2 \Rightarrow \numv ~ n \\
+  (w', wv^n) = \popn(w, n) \qquad
+  \mu' = \mu \cdot \assign(e_1, \listv ~ wv^n) \\
+  \hline
+  \seq{a}, w, c \vdash \popni ~ (e_1, e_2) \Rightarrow
+  (w', \algo ~ (\setenv(c, \mu'))) \\
+%
+% PopAll
+\newline \\
+  (\seq{wv}, wc, \seq{we'}) = \splitctx(\seq{we}) \qquad
+  \mu' = \getenv(c) \cdot \assign(e, \liste ~ \seq{wv}) \\
+  \hline
+  \seq{a}, (\seq{we}, \seq{wi}, sto), c \vdash \popalli ~ e
+  \Rightarrow
+  ((wc ~ \seq{we'}, \seq{wi}, sto), \algo ~ (\setenv(c, \mu'))) \\
+%
+% Let
+\newline \\
+  \mu = \getenv(c) \qquad
+  w, \mu \vdash e_2 \Rightarrow v \qquad
+  \mu' = \mu \cdot \assign(e_1, v) \\
+  \hline
+  \seq{a}, w, c \vdash \leti ~ (e_1, e_2)
+  \Rightarrow
+  (w, \setenv(c, \mu')) \\
+%
+% Trap
+\newline \\
+  \\
+  \hline
+  \seq{a}, w, c \vdash \trapi \Rightarrow (w, \ret ~ (\trapv, \mt)) \\
+%
+% Return-rule
+\newline \\
+  \\
+  \hline
+  \seq{a}, w, (\seq{i}, \mu, wc^*, k) \vdash \returnreli \Rightarrow (w, k) \\
+%
+% Return-func
+\newline \\
+  w, \mu \vdash e \Rightarrow v \\
+  \hline
+  \seq{a}, w, (\seq{i}, \mu, wc^*, k) \vdash \returnfuni ~ e \Rightarrow
+  (w, \ret ~ (v, k)) \\
+%
+% Execute
+\newline \\
+  w, \getenv(c) \vdash e \Rightarrow \listv ~ wi^* \\
+  \hline
+  \seq{a}, w, c \vdash \executei ~ e \Rightarrow
+  (w, \exe ~ (wi^*, \algo ~ c)) \\
+%
+% Call-fname
+\newline \\
+  \mu = \getenv(c) \qquad
+  \fnamev ~ s_3 = \mu(s_2) \qquad
+  w, \mu \vdash e_1 \Rightarrow v_1 \quad ... \quad
+  w, \mu \vdash e_n \Rightarrow v_n \\
+  \hline
+  \seq{a}, w, c \vdash \calli ~ (s_1, s_2, e_1 ~ ... ~ e_n) \Rightarrow
+  (w, \call ~ (s_1, s_3, v_1 ~ ... ~ v_n, \algo ~ c)) \\
+%
+% Call-fname
+\newline \\
+  \mu = \getenv(c) \qquad
+  s_2 \not\in \domain(\mu) \qquad
+  w, \mu \vdash e_1 \Rightarrow v_1 \quad ... \quad
+  w, \mu \vdash e_n \Rightarrow v_n \\
+  \hline
+  \seq{a}, w, c \vdash \calli ~ (s_1, s_2, \seq e) \Rightarrow
+  (w, \call ~ (s_1, s_2, v_1 ~ ... ~ v_n, \algo ~ c)) \\
+%
+% Replace-frame
+\newline \\
+\text{NOTE: $e_1$ is either store or current frame} \\
+  \mu = \getenv(c) \qquad
+  \casev ~ (\text{``Frame"}, v_1 ~ v_2) = \getcurframe(w) \qquad
+  w, \mu \vdash e \Rightarrow v \\
+  wc = \casev ~ (\text{``Frame"}, v_1 ~ \update(w, \mu, v_2, p^+, v)) \\
+  \hline
+  \seq{a}, w, c \vdash \replaceframei ~ (p^+, e) \Rightarrow
+  (\setcurframe(w, wc), \algo ~ c) \\
+%
+% Replace-store
+\newline \\
+  \mu = \getenv(c) \qquad
+  w, \mu \vdash e \Rightarrow v \\
+  \strv ~ sto = \update(w, \mu, \strv ~ (\getstore(w)), p^+, v) \\
+  \hline
+  \seq{a}, w, c \vdash \replacestorei ~ (p^+, e)
+  \Rightarrow (\setstore(w, sto), c) \\
+\end{gather*}
+
+Given algorithms $\seq a$, a Wasm state $w$, and a context $c$, executing an
+instruction $i$ produces an updated Wasm state and a continuation.
+% If
+$\ifi ~ (e, i_1^*, i_2^*)${} evaluates $e$ and branches to $\seq i_1$ if the
+result is true, or $\seq i_2$ otherwise.
+% Either
+$\eitheri ~ (i_1^*, i_2^*)${} nondeterministically selects either $i_1^*$ or
+$i_2^*$.
+% Enter
+$\enteri ~ (e_1, e_2)${} evaluates $e_1$, pushing the resulting Wasm context
+$wc$ to the interleaved stack.
+It then evaluates $e_2$ to obtain the Wasm instruction sequence $wi_1^*$.
+Then, the end instructions of the Wasm context $wc$ and the Wasm contexts
+$wc_1 ~ ... ~ wc_n$ in the context $c$ are appended to $wi_1^*$ and pushed to
+the Wasm instruction stack.
+Finally, it changes the continuation to
+$\wasm ~ (wc ~ wc_1 ~ ... ~ wc_n, \algo ~ (i^*, \mu, \epsilon, k))${}.
+% PushCtx
+$\pushctxi ~ e${} evaluates $e$, pushing the resulting Wasm context $wc$ to the
+interleaved stack and the context $c$.
+% Push
+$\pushi ~ e${} evaluates $e$, pushing the resulting value $wv$ to the
+interleaved stack.
+% PopCtx
+$\popctxi ~ e${} pops all Wasm values up to and including a Wasm context $wc$
+from the interleaved stack and all Wasm instructions up to and including an end
+instruction from the Wasm instruction stack.
+It also pops a Wasm context from the context $c$.
+Then, it assign the Wasm context $wc$ to $e$.
+% Pop
+$\popi ~ e${} pops a Wasm value $wv$ from the interleaved stack and assign the
+value to $e$.
+% PopN
+$\popni ~ (e_1, e_2)${} evaluates $e_2$ to obtain a number $n$, pops $n$ Wasm
+values from the interleaved stack, and assigns the values to $e_1$.
+% PopAll
+$\popalli ~ e${} pops all the Wasm values up to, but not including, a
+Wasm context from the interleaved stack and assign the values to $e$.
+% Let
+$\leti ~ (e_1, e_2)${} evaluates $e_2$ and assigns the resulting value to
+$e_1$.
+% Trap
+\trapi{} results in $\ret ~ (\trapv{}, \mt{})$ to terminate the execution.
+% ReturnRel
+\returnreli{} results in the continuation $k$ in the context $c$, indicating a
+return from a relation algorithm.
+% ReturnFunc
+$\returnfuni ~ e${} evaluates $e$ to $v$, resulting in $\ret ~ (v, k)${} where
+$k$ is the continuation in the context $c$.
+% Execute
+\executei{} evaluates $e$ to obtain a Wasm instruction sequence $wi^*$ and
+result in $\exe ~ (wi^*, \algo ~ c)${}.
+% Call
+$\calli ~ (s_1, s_2, e_1 ~ ... ~ e_n)${} checks if $s_2$ is bound in the
+environment.
+If it is, it retrieves the function name $s_3$ from the environment $\mu$;
+otherwise, it uses $s_2$ as the function name.
+It evaluates $\seq e$ to $v^*$, resulting in $\call ~ (s_1, s_{name}, v_1 ~ ...
+~ v_n)${} where $s_{name}$ is the function name.
+% ReplaceFrame
+\replaceframei{} retrieves the current frame $\casev ~ (\text{``Frame"}, v_1
+v_2)${} in the Wasm context $w$.
+It then evaluates $e$ to $v$, updating the value at location $v_2$ along paths
+$p^+$ to $v$.
+% ReplaceFrame
+\replacestorei{} evaluates $e$ to $v$, updating the value at location $\strv ~
+sto$ along paths $p^+$ to v.
+
+%% Expression
+
+\begin{gather*}
+  \boxed{w, \mu \vdash e \Rightarrow v} \\
+%
+% Var
+\newline \\
+  v = \mu(s) \\
+  \hline
+  w, \mu \vdash \vare ~ s \Rightarrow v \\
+%
+% Num
+\newline \\
+  \hline
+  w, \mu \vdash \nume ~ n \Rightarrow \numv ~ n \\
+%
+% Bool
+\newline \\
+  \hline
+  w, \mu \vdash \boole ~ b \Rightarrow \boolv ~ b \\
+%
+% Fname
+\newline \\
+  \hline
+  w, \mu \vdash \fnamee ~ s \Rightarrow \fnamev ~ s \\
+%
+% List
+\newline \\
+  w, \mu \vdash e_1 \Rightarrow v_1 \quad ... \quad
+  w, \mu \vdash e_n \Rightarrow v_n \\
+  \hline
+  w, \mu \vdash \liste ~ e_1 ~ ... ~ e_n \Rightarrow \listv ~ v_1 ~ ... ~ v_n \\
+%
+% Str
+\newline \\
+  w, \mu \vdash e_1 \Rightarrow v_1 \quad ... \quad
+  w, \mu \vdash e_n \Rightarrow v_n \\
+  \hline
+  w, \mu \vdash \stre ~ (s_1, e_1) ~ ... ~ (s_n, e_n) \Rightarrow
+  \strv ~ (s_1, v_1) ~ ... ~ (s_n, v_n) \\
+%
+% Tup
+\newline \\
+  w, \mu \vdash e_1 \Rightarrow v_1 \quad ... \quad
+  w, \mu \vdash e_n \Rightarrow v_n \\
+  \hline
+  w, \mu \vdash \tupe ~ e_1 ~ ... ~ e_n \Rightarrow \tupv ~ v_1 ~ ... ~ v_n \\
+%
+% Case
+\newline \\
+  w, \mu \vdash e_1 \Rightarrow v_1 \quad ... \quad
+  w, \mu \vdash e_n \Rightarrow v_n \\
+  \hline
+  w, \mu \vdash \casee ~ (s, e_1 ~ ... ~ e_n) \Rightarrow \casev ~ (s, v_1 ~ ... ~ v_n) \\
+%
+% Un
+\newline \\
+  w, \mu \vdash e \Rightarrow v \\
+  \hline
+  w, \mu \vdash \une ~ (unop, e) \Rightarrow \unop(unop, v) \\
+%
+% Bin
+\newline \\
+  w, \mu \vdash e_1 \Rightarrow v_1 \qquad w, \mu \vdash e_2 \Rightarrow v_2 \\
+  \hline
+  w, \mu \vdash \bine ~ (binop, e_1, e_2) \Rightarrow \binop(binop, v_1, v_2) \\
+%
+% Acc
+\newline \\
+  w, \mu \vdash e \Rightarrow v \\
+  \hline
+  w, \mu \vdash \acce ~ (e, p) \Rightarrow \access(w, \mu, v, p) \\
+%
+% Upd
+\newline \\
+  w, \mu \vdash e_1 \Rightarrow v_1 \qquad w, \mu \vdash e_2 \Rightarrow v_2 \\
+  \hline
+  w, \mu \vdash \upde ~ (e_1, p^+, e_2) \Rightarrow \update(w, \mu, v_1, p^+, v_2) \\
+%
+% Cat
+\newline \\
+   w, \mu \vdash e_1 \Rightarrow \listv ~ \seq{v_1} \qquad
+   w, \mu \vdash e_2 \Rightarrow \listv ~ \seq{v_2} \\
+  \hline
+  w, \mu \vdash \cate ~ (e_1, e_2) \Rightarrow \listv ~ (\seq{v_1} ~ \seq{v_1}) \\
+%
+% Comp
+\newline \\
+   w, \mu \vdash e_1 \Rightarrow \strv ~ \seq{(s_1, v_1)} \qquad
+   w, \mu \vdash e_2 \Rightarrow \strv ~ \seq{(s_2, v_2)} \\
+  \hline
+  w, \mu \vdash \compe ~ (e_1, e_2) \Rightarrow \strv ~ (\seq{(s_1, v_1)} ~ \seq{(s_2, v_2)}) \\
+%
+% Mem
+\newline \\
+  w, \mu \vdash e_1 \Rightarrow v_1 \qquad
+  w, \mu \vdash e_2 \Rightarrow \listv ~ \seq{v_2} \\
+  \hline
+  w, \mu \vdash \meme ~ (e_1, e_2) \Rightarrow \boolv ~ (v_1 \in \seq{v_2}) \\
+%
+% Choose
+\newline \\
+  w, \mu \vdash e \Rightarrow \listv ~ \seq{v} \qquad
+  v \in \seq{v} \\
+  \hline
+  w, \mu \vdash \choosee ~ e \Rightarrow v \\
+%
+% Len
+\newline \\
+  w, \mu \vdash e \Rightarrow \listv ~ \seq{v} \\
+  \hline
+  w, \mu \vdash \lene ~ e \Rightarrow \numv ~ |\seq v| \\
+%
+% IsCaseOf
+\newline \\
+  w, \mu \vdash e \Rightarrow \casev ~ (s', \seq{v}) \\
+  \hline
+  w, \mu \vdash \iscaseofe ~ (e, s) \Rightarrow \boolv ~ (s = s') \\
+%
+% GetCurCtx
+\newline \\
+  wc = \getcurctx(w) \\
+  \hline
+  w, \mu \vdash \getcurctxe \Rightarrow wc \\
+%
+% CtxKind
+\newline \\
+  \casev (s', \seq v) = \getcurctx(w) \\
+  \hline
+  w, \mu \vdash \ctxkinde ~ s \Rightarrow \boolv ~ (s = s') \\
+%
+% Iter-
+\newline \\
+  \,~\mu(s_1) = \listv ~ (v_{11} ~ ... ~ v_{1n}) \quad ... \quad
+  \mu(s_m) = \listv ~ (v_{m1} ~ ... ~ v_{mn}) \\
+  \mu_1 = \mu[s_1 \mapsto v_{11}, ~ ... ~ s_m \mapsto v_{m1}] \quad ... \quad
+  \mu_n = \mu[s_1 \mapsto v_{1n}, ~ ... ~ s_m \mapsto v_{mn}] \\
+  w, \mu_1 \vdash e \Rightarrow v_1 \quad ... \quad
+  w, \mu_n \vdash e \Rightarrow v_n \\
+  \hline
+  w, \mu \vdash \itere ~ (e, \listiter, s_1 ~ ... ~ s_m) \Rightarrow
+  \listv ~ v_1 ~ ... ~ v_n \\
+%
+% Iter-n
+\newline \\
+  w, \mu \vdash e_2 \Rightarrow \numv ~ n \\
+  \,~\mu(s_1) = \listv ~ (v_{11} ~ ... ~ v_{1n}) \quad ... \quad
+  \mu(s_m) = \listv ~ (v_{m1} ~ ... ~ v_{mn}) \\
+  \mu_1 = \mu[s_1 \mapsto v_{11}, ~ ... ~ s_m \mapsto v_{m1}] \quad ... \quad
+  \mu_n = \mu[s_1 \mapsto v_{1n}, ~ ... ~ s_m \mapsto v_{mn}] \\
+  w, \mu_1 \vdash e \Rightarrow v_1 \quad ... \quad
+  w, \mu_n \vdash e \Rightarrow v_n \\
+  \hline
+  w, \mu \vdash \itere ~ (e_1, \listniter ~ e_2, s_1 ~ ... s_m) \Rightarrow
+  \listv ~ v_1 ~ ... ~ v_n \\
+%
+% Iter-idx
+\newline \\
+  w, \mu \vdash e_2 \Rightarrow \numv ~ n \\
+  \mu_1 = \mu[s_1 \mapsto \numv ~ 0, ~ ... ~ s_m \mapsto \numv ~ 0] \quad ... \quad
+  \mu_n = \mu[s_1 \mapsto \numv ~ n, ~ ... ~ s_m \mapsto \numv ~ n] \\
+  w, \mu_1 \vdash e \Rightarrow v_1 \quad ... \quad
+  w, \mu_n \vdash e \Rightarrow v_n \\
+  \hline
+  w, \mu \vdash \itere ~ (e_1, \listidxiter ~ e_2, s_1 ~ ... ~ s_m) \Rightarrow
+  \listv ~ v_1 ~ ... ~ v_n \\
+%
+% Match
+\newline \\
+  w, \mu \vdash e_1 \Rightarrow v_1 \qquad
+  w, \mu \vdash e_2 \Rightarrow v_2 \\
+  \hline
+  w, \mu \vdash \matche ~ (e_1, e_2) \Rightarrow \boolv ~ (\match(v_1, v_2)) \\
+%
+% HasType
+\newline \\
+  w, \mu \vdash e \Rightarrow v \\
+  \hline
+  w, \mu \vdash \hastypee ~ (e, s) \Rightarrow \boolv ~ (\hastype(v, s)) \\
+\end{gather*}
+
+Given a Wasm state $w$ and an environment $\mu$, an expression $e$ evaluates
+to a value.
+% Var
+$\vare ~ s${} looks up the environment $\mu$ and gets the value.
+% Num Bool Fname
+$\nume ~ n${}, $\boole ~ b${}, and $\fnamee ~ s${} directly reduce to $\numv ~
+n${}, $\boolv ~ b${}, and $\fnamev ~ s${}, respectively.
+% Tuple Case List Struct
+$\liste ~ e_1 ~ ... ~ e_n${}, $\stre ~ (s_1, e_1) ~ ... ~ (s_n, e_n)${}, $\tupe
+~ e_1 ~ ... ~ e_n${}, and $\casee ~ (s, e_1 ~ ... ~ e_n)${} evaluate their
+sub-expressions and reduce to $\listv ~ v_1 ~ ... ~ v_n${}, $\strv ~ (s_1, v_1)
+~ ... ~ (s_n, v_n)${}, $\tupv ~ v_1 ~ ... ~ v_n${}, and $\casev ~ (s, v_1 ~ ...
+~ v_n)${}, respectively.
+% Un Bin
+$\une ~ (unop, e)${} and $\bine ~ (binop, e_1, e_2)${} evaluate their
+sub-expressions and perform \unop{} and \binop{}, respectively.
+% Acc
+$\acce ~ (e, p)${} evaluates $e$ and accesses the resulting value $v$ using a
+path $p$.
+% Upd
+$\upde ~ (e_1, p^+, e_2)${} evaluates $e_1$ and $e_2$ to $v_1$ and $v_2$,
+updating the value at location $v_1$ along paths $p^+$ to the value $v_2$.
+% Cat Comp
+$\cate ~ (e_1, e_2)${} and $\compe ~ (e_1, e_2)${} each evaluate two
+sub-expressions $e_1$ and $e_2$, concatenating the resulting lists and
+resulting records, respectively.
+% Mem
+$\meme ~ (e_1, e_2)${} evaluates $e_1$ and $e_2$ to $v_1$ and $\listv ~ \seq
+v_2$, and checks whether $v_1$ is a member of $\seq v_2$.
+% Choose Len
+$\choosee ~ e$ and $\lene ~ e${} each evaluate $e$ to the list value $\listv ~
+\seq v$, choosing an element from the list and calculating the length of the
+list, respectively.
+% IsCaseOf
+$\iscaseofe ~ (e, s)${} evaluates $e$ to the tagged tuple $\casev ~ (s', \seq
+v)$ and checks whether the tag $s'$ matches $s$.
+% GetCurCtx
+\getcurctxe{} retrieves the current Wasm context from the interleaved stack.
+% CtxKind
+$\ctxkinde ~ s${} retrieves the current Wasm context $\casev ~ (s', \seq v)$
+and checks whether the kind $s'$ matches $s$.
+% Iter
+$\itere ~ (e, iter, s^*)${} behaves differently depending on $iter$.
+\listiter{} and $\listniter ~ e_2${} are essentially equivalent.
+They assume that all the values mapped to the variables $\seq s$ by the
+environment $\mu${} are list values with the same length $n$.
+$\listniter ~ e_2${} checks whether the number $n$ matches the result of
+evaluating $e_2$, while \listiter{} allows any number.
+It generates a sequence of environments $\mu_1 ~ ... ~ \mu_n$ where the k-th
+environment maps each variable to the k-th elements of each list.
+Each environment is used to evaluate $e_1$, producing a sequence of values $v_1
+~ ... ~ v_n$.
+Finally, it results in $\listv ~ v_1 ~ ... ~ v_n$.
+$\listidxiter ~ e_2${} behaves similarly except that the variables $\seq s$ are
+unbound.
+It evaluates $e_2$ to obtain $n$ and then behaves like $\listniter ~ e_2${} as
+if a list of values [0 ... n-1] were bound to the variables $s^*$.
+% IsValid Match HasType
+$\matche ~ (e_1, e_2)${} and $\hastypee ~ (e, s)${} evaluate their
+sub-expressions and perform \match{} and \hastype{}, respectively.
+Note that \match{} and \hastype{} represent the match relation and the type
+relation in the Wasm static semantics.
+Their formalization is omitted as they are highly specific to Wasm and
+excessively verbose.
+
+%% Helper functions
+
+\begin{align*}
+%
+% get_algo_name
+\newline \\
+  &\getalgoname(\rel ~ (s, \seq e, \seq i)) = s \\
+  &\getalgoname(\fun ~ (s, \seq e, \seq i)) = s \\
+%
+% lookup
+\newline \\
+  &\lookup(\epsilon, s) = \epsilon \\
+  &\lookup(a ~ \seq{a}, s) =
+    \begin{cases}
+      a &\quad\quad \premise{if}\quad ~ s = \getalgoname(a) \\
+      \lookup(\seq{a}, s) &\quad\quad \premise{otherwise}
+    \end{cases}
+  \\
+%
+% assign
+\newline \\
+  &\assign(\vare ~ s, v) = [s \mapsto v] \\
+  &\assign(\itere ~ (e, iter, s^*), v) = \assign(e, v) \\
+  &\assign(\tupe ~ e_1 ~ ... ~ e_n, \tupv ~ v_1 ~ ... ~ v_n) =
+    \assign(e_1, v_1) \cdot ~ ... ~ \cdot \assign(e_n, v_n) \\
+  &\assign(\liste ~ e_1 ~ ... ~ e_n, \listv ~ v_1 ~ ... ~ v_n) =
+    \assign(e_1, v_1) \cdot ~ ... ~ \cdot \assign(e_n, v_n) \\
+  &\assign(\casee ~ (s, e_1 ~ ... ~ e_n), \casev ~ (s, v_1 ~ ... ~ v_n)) =
+    \assign(e_1, v_1) \cdot ~ ... ~ \cdot \assign(e_n, v_n) \\
+  &\assign(
+    w,
+    \mu,
+    \stre ~ (s_1, e_1) ~ ... ~ (s_2, e_n),
+    \strv ~ (s_1, v_1) ~ ... ~ (s_n, v_n)
+  ) =
+    \assign(e_1, v_1) \cdot ~ ... ~ \cdot \assign(e_n, v_n) \\
+  &\assign(\cate ~ (\liste ~ e_1 ~ ... ~ e_m, e), \listv v_1 ~ ... ~ v_n) =
+    \assign(e_1, v_1) \cdot ~ ... ~ \cdot \assign(e_m, v_m) \cdot
+    \assign(e, \listv ~ v_{m+1} ~ ... ~ v_n) \\
+  &\assign(\cate ~ (e, \liste ~ e_1 ~ ... ~ e_m), \listv v_1 ~ ... ~ v_n) =
+    \assign(e, \listv ~ v_1 ~ ... ~ v_{n-m}) \cdot
+    \assign(e_1, v_{n-m+1}) \cdot ~ ... ~ \cdot \assign(e_m, v_n) \\
+%
+% create_algo
+\newline \\
+  &\createalgo(\seq a, s, \seq v, k) =
+  \algo(\seq i, \mu, \epsilon, k) \\
+  &\qquad\qquad\qquad\qquad\premise{if}\quad
+  (\rel ~ (s, \seq e, \seq i) \quad\lor\quad \fun ~ (s, \seq e, \seq i)) = \lookup(\seq a, s)
+  \quad\land\quad
+  \mu = \assign(w, [\text{``s"} \mapsto \storev], \seq e, \seq v) \\
+%
+% split_ctx
+\newline \\
+  &\splitctx(\epsilon) = (\epsilon, \epsilon, \epsilon) \\
+  &\splitctx(we ~ \seq{we}) =
+    \begin{cases}
+      (\epsilon, wc, \seq{we})
+      &\quad\quad \premise{if}\quad
+      we = wc \\
+      (wv ~ \seq{wv}, wc, \seq{we'})
+      &\quad\quad \premise{if}\quad
+      we = wv \quad\land\quad
+      (\seq{wv}, wc, \seq{we'}) = \splitctx(\seq{we}) \\
+    \end{cases}
+  \\
+%
+% pop-wasm-ctx_context
+\newline \\
+  &\popwasmctx_{C}((\seq i, \mu, epsilon, k)) = (\seq i, \mu, \epsilon, \popwasmctx_K(k)) \\
+  &\popwasmctx_{C}((\seq i, \mu, wc ~ \seq{wc}, k)) = (\seq i, \mu, wc^*, k) \\
+%
+\newline \\
+% pop-wasm-ctx_cont-mt
+  &\popwasmctx_{K}(\mt) = \epsilon \\
+% pop-wasm-ctx_cont-toplevelcall
+  &\popwasmctx_{K}(\toplevelcall ~ (s, v^*)) = \epsilon \\
+% pop-wasm-ctx_cont-call
+  &\popwasmctx_{K}(\call ~ (s_1, s_2, \seq v, k)) =
+    \call ~ (s_1, s_2, \seq v, (\popwasmctx_K(k)) \\
+% pop-wasm-ctx_cont-execute
+  &\popwasmctx_{K}(\exe ~ (v, k)) = \exe ~ (v, (\popwasmctx_K(k)) \\
+% pop-wasm-ctx_cont-wasm_0
+  &\popwasmctx_{K}(\wasm ~ (\epsilon, k)) = \wasm ~ (\epsilon, \popwasmctx_{K}(k)) \\
+% pop-wasm-ctx_cont-wasm_nonzero
+  &\popwasmctx_{K}(\wasm ~ (wc ~ \seq{wc}, k)) = \wasm ~ (\seq{wc}, k) \\
+% pop-wasm-ctx_cont-al
+  &\popwasmctx_{K}(\algo ~ c) = \algo ~ (\popwasmctx_C(c)) \\
+% pop-wasm-ctx_cont-ret
+  &\popwasmctx_{K}(\ret ~ (v, k)) = \ret ~ (v, \popwasmctx_K(k)) \\
+%
+% exit
+\newline \\
+  &\exit(wi ~ \seq{wi}) =
+    \begin{cases}
+      \seq{wi} &\quad\quad \premise{if}\quad ~ \isendinstr(wi) \\
+      \exit(\seq{wi}) &\quad\quad \premise{otherwise} \\
+    \end{cases}
+  \\
+%
+% get_env
+\newline \\
+  &\getenv((\seq i, \mu, wc^*, k)) = \mu \\
+%
+% set_env
+\newline \\
+  &\setenv((\seq i, \mu, wc^*, k), \mu') = (\seq i, \mu', wc^*, k) \\
+%
+% prepend_instr
+\newline \\
+  &\prependinstr((\seq i, \mu, wc^*, k), \seq{i'}) = (\seq{i'} ~ \seq i, \mu, wc^*, k) \\
+%
+% add_ctx
+\newline \\
+  &\addctx((\seq i, \mu, wc^*, k), wc') = (\seq i, \mu, wc' ~ wc^*, k) \\
+%
+% get_store
+\newline \\
+  &\getstore((\seq{we}, \seq{wi}, sto)) = sto \\
+%
+% set_store
+\newline \\
+  &\setstore((\seq{we}, \seq{wi}, sto), sto') = (\seq{we}, \seq{wi}, sto') \\
+%
+% pop_wasm_instr
+\newline \\
+  &\popwasminstr((\seq{we}, wi ~ \seq{wi}, sto)) = ((\seq{we}, \seq{wi}, sto), wi) \\
+%
+% push
+\newline \\
+  &\push((\seq{we}, \seq{wi}, sto), we) = (we ~ \seq{we}, \seq{wi}, sto) \\
+%
+% pop
+\newline \\
+  &\pop((we ~ \seq{we}, \seq{wi}, sto)) = ((\seq{we}, \seq{wi}, sto), we) \\
+%
+% popn
+\newline \\
+  &\popn((we^n ~ \seq{we}, \seq{wi}, sto), n) = ((\seq{we}, \seq{wi}, sto), we^n) \\
+%
+% unop
+\newline \\
+  &\unop(\notop, v) =
+    \begin{cases}
+      \boolv ~ true &\quad\quad\premise{if}\quad ~ \istrue(v) \\
+      \boolv ~ false &\quad\quad\premise{otherwise} \\
+    \end{cases}
+  \\
+  &\unop(\minusop, \numv ~ n) = \numv ~ (-n) \\
+%
+% get_cur_ctx
+\newline \\
+  &\getcurctx((\seq{we}, \seq{wi}, sto)) =
+  wc \quad\quad\premise{if}\quad ~ (\seq{wv}, wc, \seq{we'}) = \splitctx(\seq{we}) \\
+%
+% get_cur_frame-frame
+\newline \\
+  &\getcurframe((\seq{we}, \seq{wi}, sto)) =
+    \begin{cases}
+      wc &\quad\quad\premise{if}\quad
+      (\seq{wv}, wc, \seq{we'}) = \splitctx(\seq{we}) \quad\land\quad \isframe(wc)\\
+      \getcurframe(\seq{we'}) &\quad\quad\premise{if}\quad
+      (\seq{wv}, wc, \seq{we'}) = \splitctx(\seq{we}) \quad\land\quad \neg \isframe(wc) \\
+    \end{cases}
+  \\
+%
+% set_cur_frame-frame
+\newline \\
+  &\setcurframe((\seq{we}, \seq{wi}, sto), wc_{frame})
+  =
+  \begin{cases}
+    (\seq{wv} ~ wc_{frame} ~ \seq{we'}, \seq{wi}, sto) \\
+    \qquad\qquad\qquad\quad\premise{if}\quad
+    (\seq{wv}, wc, \seq{we'}) = \splitctx(\seq{we}) \quad \isframe(wc) \\
+    (\seq{wv} ~ wc ~ \setcurframe(\seq{we'}, wc_{frame}), \seq{wi}, sto) \\
+    \qquad\qquad\qquad\quad\premise{if}\quad
+    (\seq{wv}, wc, \seq{we'}) = \splitctx(\seq{we}) \quad \neg \isframe(wc) \\
+  \end{cases}
+  \\
+%
+% access-idx
+\newline \\
+  &\access(w, \mu, \listv ~ \seq{v}, \idxp ~ e) = \seq{v}[n]
+  \qquad\qquad\qquad~\,\quad\premise{if}\quad w, \mu \vdash e \Rightarrow \numv ~ n \\
+% access-slice
+  &\access(w, \mu, \listv ~ \seq{v}, \slicep ~ (e_1, e_2)) = \seq{v}[n_1: n_2]
+  \quad\quad\premise{if}\quad
+  w, \mu \vdash e_1 \Rightarrow \numv ~ n_1 \quad\land\quad
+  w, \mu \vdash e_2 \Rightarrow \numv ~ n_2 \\
+% access-dot
+  &\access(w, \mu, \strv ~ ((s_0, v_0) ~ \seq{(s_1, v_1)}), \dotp ~ s) =
+  \begin{cases}
+    v_0 &\quad\quad\premise{if}\quad s_0 = s \\
+    \access(w, \mu, \strv ~ \seq{(s_1, v_1)}, \dotp ~ s) &\quad\quad\premise{otherwise} \\
+  \end{cases}
+  \\
+% access-store
+  &\access(w, \mu, \storev, p) =\access(w, \mu, \strv ~ sto, p)
+  \quad\quad\premise{if}\quad w = (\seq{we}, \seq{wi}, sto)\\
+%
+% update
+\newline \\
+  % idx
+  &\update(w, \mu, \listv ~ \seq{v}, (\idxp ~ e) ~ \seq{p}, v) =
+  \listv ~ (\updateidx(\seq v, n, \update(w, \mu, \seq{v}[n], \seq{p}, v))) \\
+  &\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\premise{if}\quad
+  w, \mu \vdash e \Rightarrow \numv ~ n \\
+  % slice
+  &\update(w, \mu, \listv ~ \seq{v}, (\slicep ~ e) ~ \seq{p}, v) =
+  \listv ~ (\updateslice(\seq v, n_1, n_2, \update(w, \mu, \seq{v_i}[n_1: n_2], \seq{p}, v))) \\
+  &\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\premise{if}\quad
+  w, \mu \vdash e_1 \Rightarrow \numv ~ n_1 \quad\land\quad
+  w, \mu \vdash e_2 \Rightarrow \numv ~ n_2 \\
+  % dot
+  &\update(w, \mu, \strv ~ \seq{(s, v)}, (\dotp ~ s') ~ \seq{p}, v)
+  =
+  \strv ~ (\updatedot(\seq{(s, v)}, s', \update(w, \mu, \access(v, \dotp ~ s'), \seq{p}, v))) \\
+%
+% update_idx
+\newline \\
+  &\updateidx(v_0 ~ \seq{v_1}, n, v_2) =
+  \begin{cases}
+    v_2 ~ \seq{v_1}
+    &\quad\quad\premise{if}\quad n = 0 \\
+    v_0 ~ \updateidx(\seq{v_1}, n-1, v_2)
+    &\quad\quad\premise{if}\quad n > 0 \\
+  \end{cases} \\
+%
+% update_slice
+\newline \\
+  &\updateslice(v_0^n ~ \seq{v_1}, 0, n, v_2^n) = v_2^n ~ \seq{v_1} \\
+  &\updateslice(v_0 ~ \seq{v_1}, m, n, v_2^n) =  v_0 ~ \updateslice(\seq{v_1}, m-1, n, v_2)
+  \quad\quad\premise{if}\quad m > 0 \\
+%
+% update_dot
+\newline \\
+  &\updatedot((s_0, v_0) ~ \seq{(s_1, v_1)}, s_2, v_2) =
+  \begin{cases}
+    (s_2, v_2) ~ \seq{(s_1, v_1)}
+    &\quad\quad\premise{if}\quad s_0 = s_2 \\
+    (s_0, v_0) ~ \updatedot(\seq{(s_1, v_1)}, s_2, v_2)
+    &\quad\quad\premise{otherwise}
+  \end{cases}
+  \\
+%
+% get_end_algo
+\newline \\
+  &\getendinstr(wc) =
+  \begin{cases}
+    \casev ~ (\text{``Exiting"}, \epsilon) &\quad\quad\premise{if}\quad wc = \casev ~ (\text{``Label"}, \seq v)\\
+    \casev ~ (\text{``Returning"}, \epsilon) &\quad\quad\premise{if}\quad wc = \casev ~ (\text{``Frame"}, \seq v)
+  \end{cases} \\
+%
+% is_end_instr
+\newline \\
+  &\isendinstr(wi) =
+  \begin{cases}
+    true &\quad\quad\premise{if}\quad wi = \casev ~ (\text{``Exiting"}, \epsilon) \quad\lor\quad wi = \casev ~ (\text{``Returning"}, \epsilon)\\
+    false &\quad\quad\premise{otherwise}
+  \end{cases} \\
+%
+% is_true
+\newline \\
+  &\istrue(v) =
+  \begin{cases}
+  b &\quad\quad\premise{if}\quad v = \boolv ~ b \\
+    \istrue(v_1) ~ \land ~ ... ~ \land ~ \istrue(v_n) &\quad\quad\premise{if}\quad
+    v = \listv ~ v_1 ~ ... ~ v_n \\
+  \end{cases} \\
+%
+% is_frame
+\newline \\
+  &\isframe(v) =
+  \begin{cases}
+    true
+    &\quad\quad\premise{if}\quad v = \casev ~ (\text{"FRAME"}, \seq v) \\
+    false &\quad\quad\premise{otherwise}\\
+  \end{cases}
+  \\
+\end{align*}

### 2024-11-22 paper/dsl.tex
Fix chapter title
Write function def and backends

--- 2024-11-21 paper/dsl.tex
@@
 
-\chapter{DSL}
+\chapter{Declarative specification Language}
 \label{ch:dsl}
 \noindent
+
+% DSL high-level explanation
+SpecTec offers a domain-specific language (DSL) tailored for specification
+purposes.
+The DSL is a declarative language, providing a compact and user-friendly syntax.
+It includes definitions for type, relation, and functions.
+% Type definition
+Types define the abstract syntax in a mannar analogous to the BNF format, where
+terminals are represented by capital letters, while non-terminals are denoted
+by lowercase letters.
+The below code represents the abstract syntax for \texttt{testop},
+\texttt{numtype} and \texttt{instr}.
+\begin{lstlisting}[style=dsl]
+syntax testop = EQZ
+syntax numtype = I32 | I64 | F32 | F64 | V128
+syntax instr =
+  | CONST numtype num
+  | TESTOP numtype testop
+  | ...
+\end{lstlisting}
+
+% Relation definition
+With relation, user can define the formal semantics of Wasm.
+The below code illustrates the execution semantics for \texttt{testop}
+instruction described in a relation definition.
+\begin{lstlisting}[style=dsl]
+  relation Step_pure: instr* ~> instr*
+  rule Step_pure/testop:
+  (CONST nt c_1) (TESTOP nt testop)  ~>  (CONST I32 c)
+  -- if c = @\dollarcode@testop_(nt, testop, c_1)
+\end{lstlisting}
+The signature of the \texttt{Step\_pure} relation $\tildecode$\texttt{>} is
+defined first.
+A \texttt{Step\_pure} rule is then specified to describe the semantics of the
+\texttt{testop} instruction with premises under which the relation holds.
+SpecTec performs type checking to reject any incorrect symbols or ill-typed
+relations like below codes.
+\begin{lstlisting}[style=dsl]
+  (* Wrong type *)
+  rule Step_pure/testop:
+  (CONST nt c_1) (TESTOP nt testop)  ~>  c
+  -- if c = @\dollarcode@testop_(nt, testop, c_1)
+
+  (* Wrong symbol *)
+  rule Step_pure/testop:
+  (CONST nt c_1) (TESTOP nt testop)  =>  (CONST I32 c)
+  -- if c = @\dollarcode@testop_(nt, testop, c_1)
+\end{lstlisting}
+
+% Function definition
+Functions enable auxiliary definitions that are used for other definitions such
+as \texttt{\$testop\_} in the relation for \texttt{testop} in the code example
+above.
+As with relation definitions, defining a function begins by specifying the
+signature of the function.
+The function's actual definition is then specified, including output value and
+the premises under which the function is applicable.
+The below code represents the function definition for \texttt{\$testop\_}.
+% XXX: Hack
+\newpage
+\begin{lstlisting}[style=dsl]
+  def @\dollarcode@testop_(nt, testop, num) : num
+  def @\dollarcode@testop_(nt, testop, c) = 1
+  -- if c = 0
+  def @\dollarcode@testop_(nt, testop, c) = 1
+  -- otherwise
+\end{lstlisting}
+In addition, functions themselves can be used to describe semantics, rather
+than merely serving as auxiliary definitions.
+Specifically, SpecTec employs function definitions to describe semantics of module
+instantiation and function invocation.
+
+% LaTeX and mechanized definitions
+SpecTec generates LaTeX code from these DSL definitions.
+The DSL allows users to include rendering hints, enabling them to customize how
+the definitions are rendered.
+The below grey box in \cref{fig:spectec-testop} is rendered result of
+LaTeX code generated from the relation definition.
+By using rendering hints, \texttt{(CONST nt c)} appears as \texttt{(nt.const
+c)}, and \texttt{(TESTOP nt testop)} is displayed as \texttt{(nt.testop)}.
+In addition, SpecTec generates mechanized definitions for theorem provers
+enabling the formal verification of properties such as type soundness.

### 2024-11-22 paper/overview.tex
Write function def and backends

@@
   (CONST nt c_1) (TESTOP nt testop)  ~>  (CONST I32 c)
-  -- if c = @$\dollarcode$@testop_(nt, testop, c_1)
+  -- if c = @\dollarcode@testop_(nt, testop, c_1)
 \end{lstlisting}
@@
 @\textbf{Step\_pure/testop}@ nt testop {
-    Assert (top_value(nt))
-    Pop (numtype_0.CONST c_1)
+    Assert (check_type_of_stack_top(nt))
+    Pop (CONST nt c_1)
     Let c = @$\dollarcode$@testop_(nt, testop, c_1)
-    Push (I32.CONST c)
+    Push (CONST I32 c)
 }

### 2024-11-22 paper/intro.tex
WIP: write overview
Write background
Add al and dsl style WIP: write dsl
Minor organize

@@
 
-% breif overview
+% Brief overview
 WebAssembly (Wasm)~\cite{wasm} is a low-level bytecode language that is safe, fast,
@@
 blockchains~\cite{wasm-blockchain}.
-% risk of implementation divergence
+% Risk of implementation divergence
 There are numerous Wasm engines, with all major browsers implementing their own
@@
 
-% rigorous standardization -> spec is rigorous: formal notation & prose notation
+% Rigorous standardization -> spec is rigorous: formal notation & prose notation
 To address this challenge, Wasm has been rigorously standardized by the
@@
 The Wasm specification is particularly rigorous, describing its semantics in
-two complementary forms: formal notation and prose notation.
+two complementary forms: \textit{formal notation} and \textit{prose notation}.
 \cref{fig:testop} illustrates the specification document for $testop$
@@
 
-% challenging specification process
+% Challenging specification process
 The demanding standardization process places a significant burden on
@@
 
-% challenge
+% Challenge
 One challenge of our approach is making AL executable.
@@
 
-% solution
+% Solution
 To address this, we present a formal model for how AL describes Wasm control flow.
@@
 
-% contribution
+% TODO: cref
+% Contribution
 The contributions of this paper are as follows:
 \begin{itemize}
-  \item We propose a formal model for Wasm control flow in AL (\cref{ch:motivation})
-  \item We formalize syntax and semantics of AL (\cref{ch:formal})
-  \item We implement an AL interpreter based on the formalization (\cref{ch:eval})
+  \item We propose a formal model for Wasm control flow in AL (\cref{sec:model})
+  \item We formalize syntax and semantics of AL (\cref{sec:formal})
+  \item We implement an AL interpreter based on the formalization (\cref{ch:interpreter})
 \end{itemize}

### 2024-11-22 paper/dsl.tex
Minor changes
Add ex env

@@
 
-\chapter{Declarative specification Language}
+\chapter{Declarative Specification Language}
 \label{ch:dsl}
@@
 by lowercase letters.
-The below code represents the abstract syntax for \texttt{testop},
+\cref{ex:dsl-syntax} represents the abstract syntax for \texttt{testop},
 \texttt{numtype} and \texttt{instr}.
+\\
+\begin{example}
 \begin{lstlisting}[style=dsl]
@@
 \end{lstlisting}
+\label{ex:dsl-syntax}
+\end{example}
 
@@
 With relation, user can define the formal semantics of Wasm.
-The below code illustrates the execution semantics for \texttt{testop}
+\cref{ex:dsl-testop} illustrates the execution semantics for \texttt{testop}
 instruction described in a relation definition.
+\\
+\begin{example}
 \begin{lstlisting}[style=dsl]
@@
 \end{lstlisting}
+  \label{ex:dsl-testop}
+\end{example}
 The signature of the \texttt{Step\_pure} relation $\tildecode$\texttt{>} is
@@
 relations like below codes.
+\\
+\begin{example}
 \begin{lstlisting}[style=dsl]
@@
 \end{lstlisting}
+  \label{ex:dsl-error}
+\end{example}
 
@@
 the premises under which the function is applicable.
-The below code represents the function definition for \texttt{\$testop\_}.
+\cref{ex:dsl-fun} represents the function definition for \texttt{\$testop\_}.
 % XXX: Hack
 \newpage
+\begin{example}
 \begin{lstlisting}[style=dsl]
@@
 \end{lstlisting}
+  \label{ex:dsl-fun}
+\end{example}
 In addition, functions themselves can be used to describe semantics, rather

### 2024-11-22 paper/overview.tex
Add ex env
Minor update
Minor fix

@@
 \end{lstlisting}
-The DSL is type checked to prevent meta-level errors such as notation misuses and
-dimension mismatches.
-Using the DSL, SpecTec generates LaTeX code and mechanized definitions for theorem provers.
+The DSL is type checked to prevent meta-level errors such as notation misuses
+and dimension mismatches.
+Using the DSL, SpecTec generates LaTeX code and mechanized definitions for
+theorem provers.
 
@@
 Additionally, SpecTec introduces another imperative language, \textit{AL}, for
-the prose notation.
+representing the prose notation.
 In AL, semantics are expressed algorithmically, in the form of step-by-step
-instructions that take inputs and perform operations, mirroring the prose notation.
+instructions that take inputs and perform operations.
 SpecTec translates declarative definitions in the DSL into algorithmic
 definitions in AL.
-The AL code below is the specification for the \texttt{testop} instruction
-generated from the DSL:
+The following code is generated from the DSL, illustrating the algorithmic
+specification for the \texttt{testop} instruction using AL:
 \begin{lstlisting}[style=al]

### 2024-11-22 paper/al.tex
Change file name
Minor organize
Minor changes
WIP: write al

--- 2024-11-22 paper/model.tex
@@
 \chapter{Algorithmic Specification Language}
-\label{ch:Model}
+\label{ch:al}
 \noindent
 
-\section{Control flow in SpecTec prose}
-\label{sec:spectecp}
+SpecTec introduces AL, an algorithmic specification language, for specifying
+the prose notation.
+Given that the prose notation in the official specification is written
+informally, it is essential to define AL with rigour.
+We begin by how the official specification describes Wasm control flow in
+\cref{sec:???}
+Next, we then define a formal model for Wasm control flow in AL in
+\cref{sec:model}.
+We then formalize the syntax and semantics of AL in \cref{sec:formal}.
+\red{Finally, we ??? in \cref{sec:spectec-prose}.}
+
+\section{Formal Model for WebAssembly Control Flow}
+\label{sec:model}
+
+The way the official specification describe Wasm control flow differs b/w the
+prose notation and the formal notation.
+In the prose notation, it implicitly assumes a program counter, so it
+\textit{jumps} to certain point of the code.
+\begin{lstlisting}[style=wasm]
+(block (br 0) (unreachable) end)
+@\hatcode@
+(block (br 0) (unreachable) end)
+        @\hatcode@
+(block (br 0) (unreachable) end)
+                                  @\hatcode@
+\end{lstlisting}
+
+\begin{lstlisting}[style=wasm]
+(i32.const 1) (if (br 0) (unreachable) else (unreachable) end)
+@\hatcode@
+(i32.const 1) (if (br 0) (unreachable) else (unreachable) end)
+                @\hatcode@
+(block (br 0) (unreachable) end)
+@\hatcode@
+(block (br 0) (unreachable) end)
+        @\hatcode@
+(block (br 0) (unreachable) end)
+                                  @\hatcode@
+\end{lstlisting}
+Official specification explains Wasm control flow in a different manner in the prose notation and the formal notation.
+The keyword is "jump" and "execute"
 
@@
 % control structure
+A program counter is imp
 It might be easy to understand how the control flow of Wasm works, if we assume
@@
 to execute.
+
 To describe control flow, it introduces a structure named \textit{block} which
@@
 
+\section{Formalization for Algorithmic Specification Language}
+\label{sec:formal}
+
+% TODO: remove seq command
 \newcommand{\seq}[1]{#1^*}
@@
 %% Syntax
-\section{Syntax of AL}
+\subsection{Syntax}
 \label{syntax}
@@
 %% Semantics
-\section{Semantics of AL}
+\subsection{Semantics}
 \label{semantics}
@@
 \end{align*}
+
+\section{Prose Notation Specification for WebAssembly}
+\label{sec:spectec-prose}
+
+SpecTec translates declarative definitions in the DSL into algorithmic
+definitions in AL.
+\cref{ex:al-testop} is generated from \cref{ex:dsl-testop}, representing the
+algorithmic specification for the \texttt{testop} instruction using AL:
+\\
+\begin{example}
+\begin{lstlisting}[style=al]
+@\textbf{Step\_pure/testop}@ nt testop {
+    Assert (check_type_of_stack_top(nt))
+    Pop (CONST nt c_1)
+    Let c = @$\dollarcode$@testop_(nt, testop, c_1)
+    Push (CONST I32 c)
+}
+\end{lstlisting}
+  \label{ex:al-testop}
+\end{example}
+Using AL, SpecTec generates reStructuredText code.
+\cref{fig:spectec-testop} shows the generated document for \texttt{testop}
+instruction.
+The generated document closely resembles the official document except for some
+minor notation changes and some missing hyperlinks.

### 2024-11-23 paper/background.tex
Add explanation for structured control flow

@@
 
-% execution
+% Execution
 Wasm execution is based on a stack machine.
@@
 % for function calls and a \textit{label} for branching.
+
+% Structured Control Flow
+Wasm supports control instructions for structured control flow, which
+eliminates the need for arbitrary jumps, ensuring structured and predictable
+execution.
+Control instructions include \texttt{blocks}, \texttt{call}, \texttt{br}, and
+\texttt{return}, which are explicitly delimited in the code.
+Each control instruction is associated with \textit{contexts} that facilitate
+structured execution and aid in handling nested or complex control flows.
+For instance, the \texttt{block} instruction groups a sequence of instructions
+and provides a scoped execution context with a \textit{label}, while the
+\texttt{call} instruction executes body of a function with a \textit{frame}.
+The \texttt{br} and \texttt{return} instructions jumps to the beginning or
+the end of the block and the function, specified within the contexts, providing
+targeted control transfers.

### 2024-11-23 paper/background.tex
Add red mark for change

@@
 % Structured Control Flow
-Wasm supports control instructions for structured control flow, which
+\red{Wasm supports control instructions for structured control flow, which
 eliminates the need for arbitrary jumps, ensuring structured and predictable
@@
 \texttt{call} instruction executes body of a function with a \textit{frame}.
-The \texttt{br} and \texttt{return} instructions jumps to the beginning or
+The \texttt{br} and \texttt{return} instructions jump to the beginning or
 the end of the block and the function, specified within the contexts, providing
 targeted control transfers.
+}

### 2024-11-23 paper/background.tex
Minor fix

@@
 Each control instruction is associated with \textit{contexts} that facilitate
-structured execution and aid in handling nested or complex control flows.
+structured execution and support handling nested or complex control flows.
 For instance, the \texttt{block} instruction groups a sequence of instructions
@@
 The \texttt{br} and \texttt{return} instructions jump to the beginning or
-the end of the block and the function, specified within the contexts, providing
-targeted control transfers.
+the end of the block and the function, specified within the contexts.
 }

### 2024-11-23 paper/dsl.tex
Add missing cref

@@
 SpecTec performs type checking to reject any incorrect symbols or ill-typed
-relations like below codes.
+relations like \cref{ex:dsl-error}.
 \\
@@
 Functions enable auxiliary definitions that are used for other definitions such
-as \texttt{\$testop\_} in the relation for \texttt{testop} in the code example
-above.
+as \texttt{\$testop\_} in the premise of \cref{ex:dsl-testop}.
 As with relation definitions, defining a function begins by specifying the

### 2024-11-23 paper/background.tex
Minor change

@@
 execution.
-Control instructions include \texttt{blocks}, \texttt{call}, \texttt{br}, and
+Control instructions include \texttt{block}, \texttt{call}, \texttt{br}, and
 \texttt{return}, which are explicitly delimited in the code.
-Each control instruction is associated with \textit{contexts} that facilitate
-structured execution and support handling nested or complex control flows.
+Each control instruction is associated with \textit{contexts} in the stack that
+facilitate structured execution and support handling nested or complex control
+flows.
 For instance, the \texttt{block} instruction groups a sequence of instructions
-and provides a scoped execution context with a \textit{label}, while the
-\texttt{call} instruction executes body of a function with a \textit{frame}.
+and provides a scoped execution with pusing a \textit{label} onto the stack,
+while the \texttt{call} instruction executes the body of a function with
+pushing a \textit{frame} onto the stack.
 The \texttt{br} and \texttt{return} instructions jump to the beginning or
-the end of the block and the function, specified within the contexts.
+the end of the block and the function, specified in the contexts.
 }

### 2024-11-23 paper/intro.tex
Minor changes
WIP: write al
Minor change

@@
 % Brief overview
-WebAssembly (Wasm)~\cite{wasm} is a low-level bytecode language that is safe, fast,
-portable, and compact.
+WebAssembly (Wasm)~\cite{wasm} is a low-level bytecode language that is safe,
+fast, portable, and compact.
 It is widely used as a compilation target for web applications.
@@
 
-% Rigorous standardization -> spec is rigorous: formal notation & prose notation
+% Rigorous standardization -> formal notation & prose notation
 To address this challenge, Wasm has been rigorously standardized by the
@@
 This challenge is compounded by the dual requirement to maintain both a formal
-notation, written in LaTeX, and a prose notation, authored in reStructuredText.
+specification, written in LaTeX, and a prose specification, authored in
+reStructuredText.
 The separate and user-unfriendly nature of these formats hinders collaborative
@@
 One challenge of our approach is making AL executable.
-Since the prose notation in the official specification is written in informal
-pseudocode, the interpretation of the prose notation is not straightforward.
+Since the prose specification is written in informal pseudocode, its
+interpretation is not straightforward.
 The naive implementation of AL interpreter produced incorrect results when
@@
 % Solution
-To address this, we present a formal model for how AL describes Wasm control flow.
+To address this, we present a formal model for how AL describes Wasm control
+flow.
 Using this model, we identify the underlying issues and the necessary changes
@@
 
-% TODO: cref
 % Contribution
@@
 \begin{itemize}
-  \item We propose a formal model for Wasm control flow in AL (\cref{sec:model})
+  \item We propose a formal model for Wasm control flow in AL
+    (\cref{sec:model})
   \item We formalize syntax and semantics of AL (\cref{sec:formal})
-  \item We implement an AL interpreter based on the formalization (\cref{ch:interpreter})
+  \item We implement an AL interpreter based on the formalization
+    (\cref{ch:interpreter})
 \end{itemize}

### 2024-11-23 paper/al.tex
Add ambiguity section

@@
 SpecTec introduces AL, an algorithmic specification language, for specifying
-the prose notation.
-Given that the prose notation in the official specification is written
-informally, it is essential to define AL with rigour.
-We begin by how the official specification describes Wasm control flow in
-\cref{sec:???}
-Next, we then define a formal model for Wasm control flow in AL in
+the prose specification.
+Since the prose specification is written informally, it is essential to define
+AL with rigour.
+We begin by outlining the ambiguity of the prose specification in
+\cref{sec:prose-informal}.
+Next, we define a formal model for Wasm control flow in AL in
 \cref{sec:model}.
@@
 
+\section{Ambiguity in the prose specification}
+\label{sec:prose-informal}
+
+The prose specification describes Wasm semantics in an informal manner,
+particularly when addressing control flow.
+% Jump
+Wasm control instructions \textit{jump} to certain code point for control
+transfer.
+While it does not explicitly explain the mechanics of jumping, the simplest
+interpretation involves assuming a program counter.
+\cref{fig:return} illustrates the prose specification for the \texttt{return}
+instruction.
+In the final step, it jumps to the instruction after the original call that
+pushed the frame, which likely corresponds to updating the program counter to
+the return address stored in the frame.
+
+% Execute
+On the other hand, some control instructions \textit{execute} a Wasm
+instruction.
+\cref{fig:if} represents the prose specification for the \texttt{if}
+instruction, which executes a block instruction with instructions in the then
+branch or the else branch based on the value in the stack.
+However, the block instruction itself does not appeared in the actual code,
+making its execution unclear in this context.
+The situation become even more complicated when the block contains control
+instructions, such as function call.
+While the intended meaning can still be inferred, this informality introduces
+ambiguity, leaving room for multiple interpretations.
+
+\begin{figure}[h]
+  \centerline{\includegraphics[width=15cm]{fig/return}}
+  \caption[\texttt{return} instruction]{\texttt{return} instruction}
+    \label{fig:return}
+  \centerline{\includegraphics[width=15cm]{fig/if}}
+  \caption[\texttt{if} instruction]{\texttt{if} instruction}
+    \label{fig:if}
+\end{figure}
+
 \section{Formal Model for WebAssembly Control Flow}
 \label{sec:model}
-
-The way the official specification describe Wasm control flow differs b/w the
-prose notation and the formal notation.
-In the prose notation, it implicitly assumes a program counter, so it
-\textit{jumps} to certain point of the code.
-\begin{lstlisting}[style=wasm]
-(block (br 0) (unreachable) end)
-@\hatcode@
-(block (br 0) (unreachable) end)
-        @\hatcode@
-(block (br 0) (unreachable) end)
-                                  @\hatcode@
-\end{lstlisting}
-
-\begin{lstlisting}[style=wasm]
-(i32.const 1) (if (br 0) (unreachable) else (unreachable) end)
-@\hatcode@
-(i32.const 1) (if (br 0) (unreachable) else (unreachable) end)
-                @\hatcode@
-(block (br 0) (unreachable) end)
-@\hatcode@
-(block (br 0) (unreachable) end)
-        @\hatcode@
-(block (br 0) (unreachable) end)
-                                  @\hatcode@
-\end{lstlisting}
-Official specification explains Wasm control flow in a different manner in the prose notation and the formal notation.
-The keyword is "jump" and "execute"
 

### 2024-11-24 paper/al.tex
Add 5.2

@@
 the prose specification.
-Since the prose specification is written informally, it is essential to define
-AL with rigour.
+Since the prose notation of the official specification is informal, it is
+essential to rigorously define AL to ensure an unambiguous and correct
+specification, rather than merely imitating the original.
 We begin by outlining the ambiguity of the prose specification in
 \cref{sec:prose-informal}.
-Next, we define a formal model for Wasm control flow in AL in
+Next, we propose a formal model for Wasm control flow in AL in
 \cref{sec:model}.
-We then formalize the syntax and semantics of AL in \cref{sec:formal}.
-\red{Finally, we ??? in \cref{sec:spectec-prose}.}
+We then define the syntax and the semantics of AL in \cref{sec:definition}.
+\red{Finally, we ??? in \cref{???}.}
 
@@
 
-\section{Formal Model for WebAssembly Control Flow}
+\section{A Formal Model for WebAssembly Control Flow}
 \label{sec:model}
 
-% TODO: Remove these commands
-\newcommand{\officialp}{official prose}
-\newcommand{\spectecp}{SpecTec prose}
-
-% control structure
-A program counter is imp
-It might be easy to understand how the control flow of Wasm works, if we assume
-that a Wasm code is loaded on a memory space and a pc points to the instruction
-to execute.
-
-To describe control flow, it introduces a structure named \textit{block} which
-consitutes of a instruction sequence.
-When branching instruction in the block is performed, pc can be changed to the
-starting point of the block or the end of the block.
-
-% an example of Wasm control flow in official prose
-\begin{example}
-\label{ex:br}
-\begin{verbatim}
-  // infinite loop
-  (loop (result i32) (i32.const 42) (br 0) (unreachable) end) (unreachable)
-\end{verbatim}
-\end{example}
-
-\cref{ex:br} is a Wasm code example that has \texttt{loop} and
-\texttt{unreachable}.
-Here, result type of the \texttt{loop} is \texttt{i32} and the \texttt{loop}
-has a block of three instructions: \texttt{i32.const}, \texttt{br}, and
-\texttt{unreachable}.
-\cref{fig:loop} is the specification for the \texttt{loop} instruction in the
-prose notation.
-In forth step, it says that the continuation is the start of the loop, the
-information is stored in a label.
-In the last step, it \textit{enters} the block with the label.
-The term \textit{enter} means that it pushes the label onto the stack and makes
-pc points to the first instruction in the block.
-The \texttt{i32.const} instruction is executed first, which just pushes the
-\texttt{i32} value \texttt{42} in the stack.
-Then, The \texttt{br} instruction is executed.
-\cref{fig:br} is the \officialp{} of the \texttt{br} instruction.
-It pops the label from the stack, and makes pc points to the continuation of
-the label, which is the start of the loop instruction.
-As a result, the code example above is a infinite loop pushing \texttt{42}
-forever without executing \texttt{unreachable} instructions.
-
-\begin{figure}[h!]
-    \centerline{\includegraphics[width=15cm]{fig/loop}}
-    \caption[Enter the caption title here]{\texttt{loop} instruction} \label{fig:loop}
-    \centerline{\includegraphics[width=15cm]{fig/br}}
-    \caption[Enter the caption title here]{\texttt{br} instruction} \label{fig:br}
+% TODO: Clever way to distinguish jargon rather then \textit
+% Formal model
+To formally specify the prose specification, we construct a formal model to
+describe Wasm control flow.
+% Execute
+Rather than relying on a program counter, we employ an instruction stack to
+represent execution progress.
+Execution proceeds by popping an instruction from the stack and performing its
+operation.
+To avoid confusion with the stack that contains values interleaved with
+contexts, we refer to the latter as the \textit{interleaved stack}.
+This approach allows us to define the \textit{execute} operation without
+introducing a program counter.
+
+% Jump
+We then split two distinct usages of the \textit{jump} operation:
+\textit{enter} and \textit{exit}.
+Since Wasm supports structured control flow, the \textit{jump} operation is
+used either for entering or exiting control structures such as a function body
+and a block of instructions.
+% Enter
+We define the \textit{enter} as the operation of pushing the context associated
+with the jump onto the interleaved stack and the instructions within
+the structure onto the instruction stack.
+On the contrary, we define \textit{exit} as the operation of popping the
+corresponding context and instructions.
+
+% State transition
+To express the \textit{enter} and \textit{exit} operation formally, we model
+them as state transitions.
+We define a Wasm state as a tuple of the interleaved stack, the instruction
+stack, and the store.
+For brevity, we omit the store in this discussion.
+Let the interleaved stack initially contains $n$ entries $we_1 ~ ... ~ we_n$,
+and the instruction stack contains $m$ entries $wi_1 ~ ... ~ wi_m$.
+Suppose $wi_1$ is a control instruction that enters a control structure
+containing a sequence of k instrucitons $wi_{11} ~ ... ~ wi_{1k}$ with an
+associated context $wc$.
+The \textit{enter} operation can be expressed as the following state transition:
+\[
+(we_1 ~ ... ~ we_n, ~ wi_1 ~ ... ~ wi_m) \quad\leadsto\quad
+  (
+    wc ~ we_1 ~ ...  ~ we_n, ~
+    wi_{11} ~ ... ~ wi_{1k} ~ \textbf{end}_{wc} ~ wi_2 ~ ... ~ wi_m
+  )
+\]
+
+% TODO: Fix end instruction in the 5.3
+
+% End
+Note that we introduce an administrative instruction, \textbf{end}, to mark the
+end of the structure in the instruction stack.
+The prose specification describes a behavior that must take place when the end
+of a structure is reached without control transfer.
+\cref{fig:exiting-label} illustrates this behavior for a function.
+While not a Wasm instruction itself, it is presented in the same format as Wasm
+instructions.
+Therefore, we formalize this behavior as an administrative instruction,
+\textbf{end}.
+
+\begin{figure}[h]
+  \centerline{\includegraphics[width=15cm]{fig/returning}}
+  \caption[Behavior for the end a block]{Behavior for the end of a block}
+    \label{fig:exiting-label}
 \end{figure}
 
-% exiting label in official prose
-There is also a special form of a behavior related to the block: an
-\textit{exiting label}.
-\cref{fig:exiting-label} is the \officialp{} of the \textit{exiting label}.
-It is special because this behavior is performed without an explicit
-Wasm instruction.
-Rather, the behavior is performed when \textbf{the end of a block is reached}
-without control instructions or runtime error.
-The behavior is that the label is popped from the stack, and the pc changes to
-the point after the end of the block.
-
-\begin{figure}[h!]
-    \centerline{\includegraphics[width=15cm]{fig/exiting}}
-    \caption[Enter the caption title here]{exiting label} \label{fig:exiting-label}
-\end{figure}
-
-% an example of exiting label in official prose
-\begin{example}
-\label{ex:exit}
-\begin{verbatim}
-  // exiting label
-  (loop (result i32) (i32.const 42) end) (f64.const 3.14)
-\end{verbatim}
-\end{example}
-
-\cref{ex:exit} is a Wasm code example that has \texttt{loop} whose block is
-only \texttt{i32.const}, and \texttt{f32.const}
-When the \texttt{loop} instruction is executed, it enters the block with a label.
-After \texttt{i32.const} is executed, the end of the block is reached.
-As a result, exiting label occurs so that the label is popped from the stack,
-pc changes to the point after the end of block: \texttt{f64.const}.
-As a result, after executing the code, the value 42 and 3.14 pushed to the
-stack.
-
-% control flow structure with function call in official prose
-Similar to the control flow using the block and the label, there is a function
-call with a frame.
-\cref{fig:invoke} is the \officialp{} of the function invocation which takes
-place by a \texttt{call} instruction.
-A frame is pushed to the stack and enters the block of the function body with a
-label.
-There is also a special behavior named returning from a function.
-\cref{fig:returning} is the \officialp{} of the returning from a function.
-When end of function is reached without without control instructions or runtime
-error, The frame is popped from the stack, and the pc points to the next
-instruction of the caller instruction.
-
-\begin{figure}[h!]
-    \centerline{\includegraphics[width=15cm]{fig/invoke}}
-    \caption[Enter the caption title here]{function invocation} \label{fig:invoke}
-    \centerline{\includegraphics[width=15cm]{fig/returning}}
-    \caption[Enter the caption title here]{returning from a function} \label{fig:returning}
-\end{figure}
-
-% control flow structure in SpecTec prose
-However, the \spectecp{} assumes a different model to explain the control flow
-of WebAssembly.
-Rather than using the notion of a pc, \spectecp{} assumes that WebAssembly
-instructions are just given one by one.
-This is because the \spectecp{} is generated automatically from the
-\red{SpecTec DSL}.
-\red{SpecTec DSL} uses rewrite rule to describe WebAssembly semantics, and it
-is interpreted in \spectecp{} as consuming a WebAssembly instruction, pushing
-or poping a control structures, and inputing new WebAssembly instructions.
-It can be expressed in the following form:
-$c^* \vdash i_0, i_1, ..., i_m \leadsto c'^* \vdash i'_0, ...i'_{m'}, i_1, ..., i_m$.
-It means that given a sequence of contexts $c^*$, executing $i_0$ results in a
-new sequence of contexts $c'^*$ and instructions $i'_0, ...i'_{m'}$.
-To simply model the control flow of WebAssembly, other things not closely
-related to the control flow are omitted in this notation.
-
-% challenges in SpecTec prose
-However, this viewpoint is challenged by the exiting label.
-As the block structure doesn't remain intact in this model, it is hard to
-describe \textbf{the end of the block}.
-Furthermore, the exiting label is not performed by a specific WebAssembly
-instruction, which makes it hard to model the behavior.
-To handle this problem, an administrative instruction \texttt{end} is
-introduced in \spectecp.
-When entering a block, an \texttt{end} is appended to the end of the block.
-
-% an example of Wasm control flow in official prose
-\begin{figure}[h!]
-    \centerline{\includegraphics[width=15cm]{fig/spectec-loop}}
-    \caption[Enter the caption title here]{SpecTec \texttt{loop}} \label{fig:spectec-loop}
-\end{figure}
-\begin{figure}[h!]
-    \centerline{\includegraphics[width=15cm]{fig/spectec-br}}
-    \caption[Enter the caption title here]{SpecTec \texttt{br}} \label{fig:spectec-br}
-\end{figure}
-
-\red{TODO: Change to array} \\
-\begin{align}
-  &c^*
-  \vdash
-  loop([ ~ i32.const ~ 42, ~ br ~ 0, ~ unreachable ~ ]), ~ unreachable
-  \label{eq:loop-1} \\
-&\leadsto
-  label(loop([ ~ i32.const ~ 42; ~ br ~ 0; ~ unreachable ~ ])), c^*
-  \vdash
-  i32.const ~ 42, ~ br ~ 0, ~ unreachable, ~ end, ~ unreachable
-  \label{eq:loop-2} \\
-  &\leadsto
-  label(loop([ ~ i32.const ~ 42; ~ br ~ 0; ~ unreachable ~ ])), c^*
-  \vdash
-  ~ br ~ 0, ~ unreachable, ~ end, ~ unreachable
-  \label{eq:loop-3} \\
-&\leadsto
-  c^*
-  \vdash
-  loop([ ~ i32.const ~ 42; ~ br ~ 0; ~ unreachable ~ ]), unreachable
-  \label{eq:loop-4}
-\end{align}
-
-Consider \cref{ex:br} above.
-If we assume some control structure $c^*$ is given, the code can be expressed
-like \cref{eq:loop-1}.
-\cref{fig:spectec-loop} is the \spectecp{} of the \texttt{loop} instruction.
-Rather than storing the point to jump as a continuation in the label, it stores
-the \texttt{loop} instruction itself in the label, and \textit{enters} the
-block.
-Here, \textit{enter} means that it pushes the label in the stack and executes
-the instructions in the block, which means that \texttt{i32.const},
-\texttt{br}, \texttt{unreachable}, \texttt{end} and \texttt{unreachable} become
-the new inputs in \cref{eq:loop-2}.
-When \texttt{i32.const} is excuted, it pushes 42 as \officialp{} does
-in \cref{eq:loop-3}.
-The value is omitted for brevity.
-However, \texttt{br} behaves a bit differently.
-\cref{fig:spectec-br} is the \spectecp{} of the \texttt{br} instruction.
-It pops the label from the stack, removes the input instructions until the end
-of the block including the \texttt{end}, considers the loop instruction in the
-label as a new input instruction.
-Consequently, remaining inputs are \texttt{loop}, and \texttt{unreachable}
-again in \cref{eq:loop-4}.
-Therefore, it explains the same behavior in a different point of view.
-
-\red{TODO: Also use example 3 to explain function call in official prose}
-
-% an example of function call in official prose
-\begin{example}
-\label{ex:invoke}
-\begin{verbatim}
-  // function definition of $push42
-  (func $push42 (result i32) (i32.const 42))
-
-  // function call
-  (call $push42) (f32.const 3.14)
-\end{verbatim}
-\end{example}
-
-\red{TODO: Change to array} \\
-\begin{align}
-  &c^* \vdash call(\$push42), ~ f32.const ~ 3.14 \label{eq:call-1} \\
-&\leadsto
-  Label(\epsilon), ~ Frame, ~ c^* \vdash i32.const ~ 42, ~ end, ~ f32.const ~ 3.14 \label{eq:call-2} \\
-&\leadsto
-  Label(\epsilon), ~ Frame, ~ c^* \vdash end, ~ f32.const ~ 3.14 \label{eq:call-3} \\
-&\leadsto
-  c^* \vdash f32.const ~ 3.14 \label{eq:call-4} \\
-&\leadsto
-  c^* \vdash \epsilon \label{eq:call-5}
-\end{align}
-
-\cref{ex:invoke} contains a function definition whose body is \texttt{i32.const}, a
-\texttt{call} instruction followed by a \texttt{f32.const} instruction, which
-can be expressed in \cref{eq:call-1}.
-By invoking a function, it pushes a frame and enters the block with a label in
-\cref{eq:call-2}.
-After executing the body in \cref{eq:call-3}, an \texttt{end} is executed.
-In this point, this \texttt{end} represent both the end of the block and the
-function.
-As a consequence, this \text{end} perform both exiting label and returning from
-a function in \cref{eq:call-4}, and the instruction after the \texttt{call} is
-executed in \cref{eq:call-5}
-It results in the values 42 and 3.14 pushed to the stack.
-
-% summary & test
-In short, the behavior of an \texttt{end} instruction 1) performs exiting
-label, 2) looks up the top of the context, and 3) performs returing from a
-function, if it is a frame.
-The \spectecp{} can model the WebAssembly control flow using this \texttt{end}
-instruction.
-To show the correctness of the \spectecp{}, we run the WebAssembly tests
-including official WebAssembly test suite by executing the \spectecp{} with
-\red{definitional interpreter}.
-
-% Problem
-However, tests related to the control instruction are failed.
-At that time, there was no explicit model to describe the control flow in
-\spectecp{}, it was hard to define the problem, making it complicated to fix
-the problem.
-
-%%%%%%%%
-
-% an example of buggy test case
-\begin{example}
-\label{ex:bug}
-\begin{verbatim}
-  // function definition of $br-returning
-  (func  $brAndReturning (br 0) (unreachable))
-
-  // function call
-  (call $brAndReturning)
-\end{verbatim}
-\end{example}
-
-\cref{ex:bug} contains a function that has only a \texttt{br} instruction, and a
-\texttt{call} insstruction.
-According to the \officialp{}, when the function \texttt{\$brAndReturning} is
-called, a frame is pushed to the stack, and the block of the function
-body is entered with a label.
-When the \texttt{br} is executed, it pops the label from the stack.
-As the end of the function is reached after the execution, returning from a
-function is performed.
-As a result, the frame is popped, so the call actually does nothing.
-
-% explaination of the bug in SpecTec prose
-\begin{align}
-  &c^* \vdash call(\$brAndReturning) \label{eq:bug-1} \\
-&\leadsto
-  Label(\epsilon), ~ Frame, ~ c^* \vdash (br ~ 0) ~ (unreachable), ~ end \label{eq:bug-2} \\
-&\leadsto
-  Frame, ~ c^* \vdash \epsilon \label{eq:bug-3}
-\end{align}
-
-In \spectecp{} the call instruction in \cref{eq:bug-1} pushes a frame and
-enters the function body \cref{eq:bug-2}.
-When the \texttt{br} is executed, it pops the label and removes the input
-instructions until the end in \cref{eq:bug-3}.
-As a result, the frame is remained after the executing the call.
-The problem is that the \texttt{br} should remove the end of the block but
-should not remove the end of the function, but \texttt{end} express the end of
-the block and the function at the same time.
-
-% Fix
-To fix the problem, semantics of \enteri is changed.
-The sequence of contexts that is pushed in the AL algorithm is stored,
-and when a block is entered, exiting label and returning from a function is
-appended according to the sequence.
-
-\section{Formalization for Algorithmic Specification Language}
-\label{sec:formal}
+Similarly, the \textit{exit} operation can be described as a transition of Wasm
+state.
+Suppose $wi_1$ is a control instruction that exits a control structure, and $i$
+values $wv_1 ~ ... wv_i$ are at the top of the interleaved stack, followed by a
+context $wc$, and $j$ instructions $wi_1 ~ ... ~ wi_j$ are at the top of the
+instruction stack followed by an \textbf{end} instruction.
+The \textit{exit} operation can be represented as the following state transition:
+\[
+(wv_1 ~ ... ~ wv_i ~ wc ~ we_{i+2} ~ ...  ~ we_n,
+  wi_1 ~ ... ~ wi_j ~ \textbf{end}_{wc} ~ wi_{j+2} ~ ..  ~ wi_m)
+  \quad\leadsto\quad
+  (we_{i+1} ~ ... ~ we_n, wi_{j+1} ~ ... ~ wi_m)
+\]
+This transition model provides the foundation for the formal definition of AL.
+
+\section{Definition of Algorithmic Specification Language}
+\label{sec:definition}
 
@@
 
-\section{Prose Notation Specification for WebAssembly}
-\label{sec:spectec-prose}
+\section{???}
+\label{sec:???}
 

### 2024-11-24 paper/intro.tex
Add 5.2

@@
 to improve it.
-Based on these insight, we formalize AL's syntax and semantics to accurately
-capture Wasm control flow.
+Based on these insight, we formally define AL's syntax and semantics to
+accurately capture Wasm control flow.
 We then implement an AL interpreter according to this formalization and
@@
 \begin{itemize}
-  \item We propose a formal model for Wasm control flow in AL
+  \item We propose a formal model for Wasm control flow.
     (\cref{sec:model})
-  \item We formalize syntax and semantics of AL (\cref{sec:formal})
+  \item We define syntax and semantics of AL (\cref{sec:definition})
   \item We implement an AL interpreter based on the formalization

### 2024-11-25 paper/al.tex
Fix 5.2 and 5.3

@@
 
-\begin{figure}[h]
-  \centerline{\includegraphics[width=15cm]{fig/return}}
-  \caption[\texttt{return} instruction]{\texttt{return} instruction}
-    \label{fig:return}
-  \centerline{\includegraphics[width=15cm]{fig/if}}
-  \caption[\texttt{if} instruction]{\texttt{if} instruction}
-    \label{fig:if}
-\end{figure}
-
 \section{A Formal Model for WebAssembly Control Flow}
@@
 contexts, we refer to the latter as the \textit{interleaved stack}.
-This approach allows us to define the \textit{execute} operation without
-introducing a program counter.
+This approach allows us to define the execute operation without introducing a
+program counter.
 
 % Jump
-We then split two distinct usages of the \textit{jump} operation:
-\textit{enter} and \textit{exit}.
-Since Wasm supports structured control flow, the \textit{jump} operation is
-used either for entering or exiting control structures such as a function body
-and a block of instructions.
+We then split two distinct usages of the jump operation: \textit{enter} and
+\textit{exit}.
+Since Wasm supports structured control flow, the jump operation is used either
+for entering or exiting control structures such as a function body and a block
+of instructions.
 % Enter
-We define the \textit{enter} as the operation of pushing the context associated
-with the jump onto the interleaved stack and the instructions within
-the structure onto the instruction stack.
-On the contrary, we define \textit{exit} as the operation of popping the
-corresponding context and instructions.
+We define the enter as the operation of pushing the context associated with the
+jump onto the interleaved stack and the instructions within the structure onto
+the instruction stack.
+Similarly, we define exit as the operation of popping the corresponding context
+and instructions.
 
 % State transition
-To express the \textit{enter} and \textit{exit} operation formally, we model
-them as state transitions.
+To express the enter and exit operation formally, we model them as state
+transitions.
 We define a Wasm state as a tuple of the interleaved stack, the instruction
@@
 associated context $wc$.
-The \textit{enter} operation can be expressed as the following state transition:
+The enter operation can be expressed as the following state transition:
 \[
@@
 
-% TODO: Fix end instruction in the 5.3
-
 % End
@@
 end of the structure in the instruction stack.
-The prose specification describes a behavior that must take place when the end
-of a structure is reached without control transfer.
-\cref{fig:exiting-label} illustrates this behavior for a function.
-While not a Wasm instruction itself, it is presented in the same format as Wasm
+It models unique behavior described in the prose specification that must take
+place when the end of a structure is reached without control transfer.
+For example, if a function body executes without concluding with a
+\texttt{return} instruction, a behavior that corresponds to the \texttt{return}
+is performed.
+\cref{fig:exiting-label} illustrates this behavior.
+While not a Wasm instruction itself, it is represented in the same way as Wasm
 instructions.
@@
 
-\begin{figure}[h]
-  \centerline{\includegraphics[width=15cm]{fig/returning}}
-  \caption[Behavior for the end a block]{Behavior for the end of a block}
-    \label{fig:exiting-label}
-\end{figure}
-
-Similarly, the \textit{exit} operation can be described as a transition of Wasm
+% Exit
+In the same way, the exit operation can be described as a transition of Wasm
 state.
@@
 instruction stack followed by an \textbf{end} instruction.
-The \textit{exit} operation can be represented as the following state transition:
+The exit operation can be represented as the following state transition:
 \[
@@
 
+% XXX: HACK
+\newpage
+
+\begin{figure}[h]
+  \centerline{\includegraphics[width=15cm]{fig/return}}
+  \caption[\texttt{return} instruction]{\texttt{return} instruction}
+    \label{fig:return}
+  \centerline{\includegraphics[width=15cm]{fig/if}}
+  \caption[\texttt{if} instruction]{\texttt{if} instruction}
+    \label{fig:if}
+  \centerline{\includegraphics[width=15cm]{fig/returning}}
+  \caption[Behavior for the end a block]{Behavior for the end of a block}
+    \label{fig:exiting-label}
+\end{figure}
+
 \section{Definition of Algorithmic Specification Language}
@@
 {
-\renewcommand{\arraystretch}{0.8835}  % Decrease row spacing locally
+\renewcommand{\arraystretch}{0.97}  % Decrease row spacing locally
 \begin{align*}
@@
     &&&& | & ~ \pushi ~ e &\quad\text{(Push)} \\
+    &&&& | & ~ \exiti ~ &\quad\text{(Exit)} \\
     &&&& | & ~ \popctxi ~ e &\quad\text{(Pop Context)} \\
@@
 An AL instruction is \ifi{} for conditionally choosing a branch, \eitheri{} for
-nondeterministically choosing a branch, \enteri{} for entering a block,
-\pushctxi{} for pushing a Wasm context, \pushi{} for pushing a Wasm value,
-\popctxi{} for popping a Wasm context, \popi{} for popping a Wasm value,
-\popni{} for popping $n$ Wasm values, \popalli{} for popping all Wasm values
-within a Wasm context, \leti{} for let binding, \trapi{} for trapping,
-\returnreli{} for returning from an AL relation, \returnfuni{} for returning
-from an AL function, \executei{} for executing given Wasm instructions,
-\calli{} for an AL function call with let binding, \replaceframei{} for
-replacing a frame, or \replacestorei{} for replacing a store.
+nondeterministically choosing a branch, \enteri{} for entering a control
+structure, \pushctxi{} for pushing a Wasm context, \pushi{} for pushing a Wasm
+value, \exiti{} for exiting a control structure, \popctxi{} for popping a Wasm
+context, \popi{} for popping a Wasm value, \popni{} for popping $n$ Wasm
+values, \popalli{} for popping all Wasm values within a Wasm context, \leti{}
+for let binding, \trapi{} for trapping, \returnreli{} for returning from an AL
+relation, \returnfuni{} for returning from an AL function, \executei{} for
+executing given Wasm instructions, \calli{} for an AL function call with let
+binding, \replaceframei{} for replacing a frame, or \replacestorei{} for
+replacing a store.
 % expression
@@
   w, \mu \vdash e_2 \Rightarrow \listv ~ \seq{wi_1} \\
-  wi = \getendinstr(wc) \qquad
-  wi_1 = \getendinstr(wc_1) \quad ... \quad wi_n = \getendinstr(wc_n) \\
+  wi = \casev ~ (\text{"End"}, wc) \qquad
+  wi_1 = \casev ~ (\text{"End"}, wc_1) \quad ... \quad wi_n = \casev ~ (\text{"End"}, wc_n) \\
   wi_2^* = wi_1^* ~ wi ~ wi_1 ~ ... ~ wi_n \\
@@
 %
+% Exit
+\newline \\
+  (\seq{we}, \seq{wi}, sto) = w \qquad
+  (\seq{wv}, wc, \seq{we'}) = \splitctx(\seq{we}) \\
+  \seq{wi'} = \exit(\seq{wi}) \qquad
+  c' = \popwasmctx_{C}(c) \\
+  \hline
+  \seq{a}, w, c \vdash \exiti
+  \Rightarrow
+  ((\seq{we'}, \seq{wi'}, sto), \algo ~ c') \\
+%
 % PopCtx
@@
   \mu = \getenv(c) \cdot \assign(e, wc) \qquad
-  \seq{wi'} = \exit(\seq{wi}) \qquad
   c' = \popwasmctx_{C}(c) \\
@@
   \Rightarrow
-  ((\seq{we'}, \seq{wi'}, sto), \algo ~ (\setenv(c', \mu))) \\
+  ((\seq{wv} ~ \seq{we'}, \seq{wi}, sto), \algo ~ (\setenv(c', \mu))) \\
 %
@@
 interleaved stack.
+% Exit
+$\exiti${} pops all Wasm values up to and including a Wasm context $wc$ from
+the interleaved stack and all Wasm instructions up to and including an end
+instruction from the Wasm instruction stack.
+It also pops a Wasm context from the context $c$.
 % PopCtx
-$\popctxi ~ e${} pops all Wasm values up to and including a Wasm context $wc$
-from the interleaved stack and all Wasm instructions up to and including an end
-instruction from the Wasm instruction stack.
+$\popctxi ~ e${} pops a Wasm context $wc$ from the interleaved stack.
 It also pops a Wasm context from the context $c$.
@@
     \begin{cases}
-      \seq{wi} &\quad\quad \premise{if}\quad ~ \isendinstr(wi) \\
+      \seq{wi} &\quad\quad \premise{if}\quad ~ wi = \casev ~ (\text{"End"}, wc) \\
       \exit(\seq{wi}) &\quad\quad \premise{otherwise} \\
@@
   \\
-%
-% get_end_algo
-\newline \\
-  &\getendinstr(wc) =
-  \begin{cases}
-    \casev ~ (\text{``Exiting"}, \epsilon) &\quad\quad\premise{if}\quad wc = \casev ~ (\text{``Label"}, \seq v)\\
-    \casev ~ (\text{``Returning"}, \epsilon) &\quad\quad\premise{if}\quad wc = \casev ~ (\text{``Frame"}, \seq v)
-  \end{cases} \\
-%
-% is_end_instr
-\newline \\
-  &\isendinstr(wi) =
-  \begin{cases}
-    true &\quad\quad\premise{if}\quad wi = \casev ~ (\text{``Exiting"}, \epsilon) \quad\lor\quad wi = \casev ~ (\text{``Returning"}, \epsilon)\\
-    false &\quad\quad\premise{otherwise}
-  \end{cases} \\
 %

### 2024-11-25 paper/eval.tex
Rename files
Add 5.2
Fix 5.2 and 5.3

--- 2024-11-17 paper/5-eval.tex
@@
 Using the semantics defined in \cref{ch:formal} together with the AL program,
-\cref{ex:bug} produces the same as the \officialp{}; invoking the function has
+\cref{ex:bug} produces the same as the officialp{}; invoking the function has
 no effect.
@@
   \\
-  \leadsto L ~ F \vdash (unreachable) ~ (exiting) ~ (returnfuning); \\
+  \leadsto L ~ F \vdash (unreachable) ~ (exiting) ~ (returning); \\
   \quad\qquad\qquad\qquad\qquad \algo ~ (\popctxi ~ \executei, ~ \epsilon, ~ \wasm ~ (L ~ F, ~ \algo ~ (\returnfuni, ~ \epsilon, ~ \mt)))
@@
 \end{align*}
-\begin{align*}
-%
-% get_end_algo
-\newline \\
-  &\getendinstr(wc) =
-  \begin{cases}
-    \casev ~ (\text{``Exiting"}, \epsilon) &\quad\quad\premise{if}\quad wc = \casev ~ (\text{``Label"}, \seq v) \\
-    \casev ~ (\text{``Returning"}, \epsilon) &\quad\quad\premise{if}\quad wc = \casev ~ (\text{``Frame"}, \seq v) \\
-    \casev ~ (\text{``Exiting-handler"}, \epsilon) &\quad\quad\premise{if}\quad wc = \casev ~ (\text{``Handler"}, \seq v) \\
-  \end{cases} \\
-%
-% is_end_instr
-\newline \\
-  &\isendinstr(wi) =
-  \begin{cases}
-    true &\quad\quad\premise{if}\quad wi = \casev ~ (\text{``Exiting"}, \epsilon) \\
-    &\quad\quad\lor\quad
-    wi = \casev ~ (\text{``Returning"}, \epsilon) \\
-    &\quad\quad\lor\quad
-    wi = \casev ~ (\text{``Exiting-handler"}, \epsilon) \\
-    false &\quad\quad\premise{otherwise}
-  \end{cases} \\
-\end{align*}
+% \begin{align*}
+% %
+% % get_end_algo
+% \newline \\
+%   &\getendinstr(wc) =
+%   \begin{cases}
+%     \casev ~ (\text{``Exiting"}, \epsilon) &\quad\quad\premise{if}\quad wc = \casev ~ (\text{``Label"}, \seq v) \\
+%     \casev ~ (\text{``Returning"}, \epsilon) &\quad\quad\premise{if}\quad wc = \casev ~ (\text{``Frame"}, \seq v) \\
+%     \casev ~ (\text{``Exiting-handler"}, \epsilon) &\quad\quad\premise{if}\quad wc = \casev ~ (\text{``Handler"}, \seq v) \\
+%   \end{cases} \\
+% %
+% % is_end_instr
+% \newline \\
+%   &\isendinstr(wi) =
+%   \begin{cases}
+%     true &\quad\quad\premise{if}\quad wi = \casev ~ (\text{``Exiting"}, \epsilon) \\
+%     &\quad\quad\lor\quad
+%     wi = \casev ~ (\text{``Returning"}, \epsilon) \\
+%     &\quad\quad\lor\quad
+%     wi = \casev ~ (\text{``Exiting-handler"}, \epsilon) \\
+%     false &\quad\quad\premise{otherwise}
+%   \end{cases} \\
+% \end{align*}
 

### 2024-11-25 paper/al.tex
Add 5.4

@@
 We then define the syntax and the semantics of AL in \cref{sec:definition}.
-\red{Finally, we ??? in \cref{???}.}
+Finally, we produce an executable Wasm specification using AL interpreter that
+ensures the correctness of the specification in \cref{sec:exec-spec}.
 
@@
 \newline \\
-\text{NOTE: $e_1$ is either store or current frame} \\
   \mu = \getenv(c) \qquad
@@
 
-\section{???}
-\label{sec:???}
-
-SpecTec translates declarative definitions in the DSL into algorithmic
-definitions in AL.
-\cref{ex:al-testop} is generated from \cref{ex:dsl-testop}, representing the
-algorithmic specification for the \texttt{testop} instruction using AL:
+\section{An Executable WebAssembly Specification}
+\label{sec:exec-spec}
+
+% AL spec
+SpecTec translates declarative definitions in the DSL into algorithms in AL.
+\cref{ex:al-testop} is the generated algorithm from \cref{ex:dsl-testop},
+representing the algorithmic specification for the \texttt{testop} instruction:
 \\
@@
 \end{example}
-Using AL, SpecTec generates reStructuredText code.
-\cref{fig:spectec-testop} shows the generated document for \texttt{testop}
-instruction.
-The generated document closely resembles the official document except for some
-minor notation changes and some missing hyperlinks.
+SpecTec generates reStructuredText code from the AL algorithms.
+The upper part of \cref{fig:spectec-testop} shows the rendered result of
+reStructuredText code for the \texttt{testop} instruction.
+
+% Executable
+A notable feature of AL is executability, supported by SpecTec's AL interpreter
+following the syntax and the semantics of AL.
+This executability allows the specification itself to be executed, enabling the
+execution of the specified language, Wasm.
+The execution of a Wasm program begins with two entry points: module
+instantiation and function invocation, both defined within the specification as
+algorithms.
+Executing these algorithms performs operations for executing the Wasm program,
+ultimately producing its execution result.
+In this way, the execution of a Wasm program is realized by the execution of
+the specification itself.
+
+% Correctness
+That is to say, the prose specification written in AL, together with the AL
+interpreter, functions as a Wasm interpreter which faithfully follows the
+written specification.
+By testing this Wasm interpreter, we can test the specification, thereby
+ensuring its correctness.

### 2024-11-25 paper/intro.tex
Add 5.4

@@
 accurately capture Wasm control flow.
-We then implement an AL interpreter according to this formalization and
+We then implement an AL interpreter according to the formal definition and
 evaluate its correctness.
@@
 \begin{itemize}
-  \item We propose a formal model for Wasm control flow.
-    (\cref{sec:model})
+  \item We propose a formal model for Wasm control flow. (\cref{sec:model})
   \item We define syntax and semantics of AL (\cref{sec:definition})
-  \item We implement an AL interpreter based on the formalization
-    (\cref{ch:interpreter})
+  \item We implement an AL interpreter for executable specification
+    (\cref{sec:exec-spec})
 \end{itemize}

### 2024-11-25 paper/eval.tex
Add eval

--- 2024-11-12 paper/5-eval.tex
@@
 
-% an example of buggy test case
-\textbf{Example 5.1}
-\begin{verbatim}
-  // function definition of $br-returning
-  (func  $brAndReturning (br 0) (unreachable))
+We evaluate our approach with the following two research questions.
+\begin{itemize}
+  \item \textbf{RQ1. (Correctness)} Does AL ensures the correct specification?
+    (\S 5.1)
+  \item \textbf{RQ2. (Generality)} Does AL capture the fundamental operations of
+    Wasm to describe Wasm semantics rather than being tailored to the current
+    version of the specification? (\S 5.2)
+\end{itemize}
 
-  // function call
-  (call $brAndReturning)
-\end{verbatim}
+\section{RQ1: Correctness}
 
-Example 5.1 contains a function that has only a \texttt{br} instruction, and a
-\texttt{call} insstruction.
-According to the \officialp{}, when the function \texttt{\$brAndReturning} is
-called, a frame is pushed to the stack, and the block of the function
-body is entered with a label.
-When the \texttt{br} is executed, it pops the label from the stack.
-As the end of the function is reached after the execution, returning from a
-function is performed.
-As a result, the frame is popped, so the call actually does nothing.
+We define the syntax and semantics of Wasm manually using the DSL.
+The DSL is translated into AL, which serves as a Wasm interpreter that
+faithfully follows the written specification.
+To ensure correctness of the specification, we run this Wasm interpreter
+against the execution tests from the official Wasm test suite.
+The execution tests check whether the execution of a Wasm program produces
+expected values or errors.
+Tests related to inifinite loop detection are excluded, as the specification
+does not define this behavior.
+The results demonstrate that the interpreter successfully pass all 51,549 tests in 70
+seconds on \red{(lambda 9)}.
+Passing all the tests ensures the correctness of the specification.
 
-% explaination of the bug in SpecTec prose
-\begin{align}
-  &c^* \vdash call(\$brAndReturning) \label{eq:bug-1} \\
-&\leadsto
-  Label(\epsilon), ~ Frame, ~ c^* \vdash (br ~ 0) ~ (unreachable), ~ end \label{eq:bug-2} \\
-&\leadsto
-  Frame, ~ c^* \vdash \epsilon \label{eq:bug-3}
-\end{align}
+% 7 + 3 + ??? + 44
+\section{RQ2: Generality}
+We evaluate AL's generality by applying SpecTec to eight proposals prepared for
+inclusion in the next version of Wasm: typed function references
+~\cite{proposal-fr}, garbage collection~\cite{propsal-gc}, tail calls
+~\cite{proposal-tc}, multiple memories~\cite{proposal-mm}, extended constant
+expressions~\cite{proposal-ec}, memory64~\cite{proposal-m64}, relaxed SIMD
+~\cite{proposal-rs}, exception handling~\cite{proposal-eh}.
+This proposals add or modify ??? Wasm instructions.
+In particular, the garbage collection proposal adds substantial complications
+to the Wasm type system, and the exception handling proposal introduces a new
+context, \textit{handler}.
+Notably, specifying all eight proposals in AL requires only trivial
+adjustments, such as adding a handler case to the context.
 
-In \spectecp{} the call instruction in \cref{eq:bug-1} pushes a frame and
-enters the function body \cref{eq:bug-2}.
-When the \texttt{br} is executed, it pops the label and removes the input
-instructions until the end in \cref{eq:bug-3}.
-As a result, the frame is remained after the executing the call.
-The problem is that the \texttt{br} should remove the end of the block but
-should not remove the end of the function, but \texttt{end} express the end of
-the block and the function at the same time.
-
-% Fix
-To fix the problem, semantics of \enteri is changed.
-The sequence of contexts that is pushed in the AL algorithm is stored,
-and when a block is entered, exiting label and returning from a function is
-appended according to the sequence.
-
-% simplified version of AL algorithm of call, br, and returning from a function
-\begin{verbatim}
-Call fname:
-  PushCtx Frame
-  Enter get_function_body(fname) with Label(\epsilon)
-
-Br 0:
-  PopCtx Label(instr*)
-  ExecuteSeq instr*
-
-Returning from a function:
-  PopCtx Frame
-\end{verbatim}
-
-The code above is simplified version of \spectecp{} of \texttt{call},
-\texttt{br}, and \texttt{returning from a function}.
-By using the semantics defined in \cref{ch:semantics}, the example 5.1 results
-in the same result with the \officialp{}; $call(\$brAndReturning)$ does
-nothing.
-
-\red{TODO: Change to array} \\
-\red{TODO: Change naming for $c$}
-\begin{align}
-  &c^* \vdash call(\$brAndReturning); ~ \wasm ~ (\seq{c}, ~ k)
-  \label{eq:resolve-1} \\
-  &\leadsto c^* \vdash \epsilon; ~ \algo ~ ([ ~ \pushctxi, ~ \enteri ~ ], ~ \epsilon, ~ \wasm ~ (\seq{c}, ~ k))
-  \label{eq:resolve-2} \\
-  &\leadsto Frame, ~ c^* \vdash \epsilon; ~ \algo ~ ([ ~ \enteri ~ ], ~ [ ~ Frame ~ ], ~ \wasm ~ (\seq{c}, ~ k))
-  \label{eq:resolve-3} \\
-  &\leadsto Label(\epsilon), ~ Frame, ~ c^* \vdash (br ~ \epsilon), ~ (unreachable), ~ (exiting), ~ (returning); \\
-  &\quad\quad \wasm ~ ([ ~ Label(\epsilon), ~ Frame ~ ], ~ \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (\seq{c}, ~ k)))
-  \label{eq:resolve-4} \\
-  &\leadsto Label(\epsilon), ~ Frame, ~ c^* \vdash (unreachable), ~ (exiting), ~ (returning); \\
-  &\quad\quad \algo ~ ([ ~ \popctxi, ~ \executei ~ ], ~ \epsilon, ~ \wasm ~ ([ ~ Label(\epsilon), ~ Frame ~ ], ~ \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (\seq{c}, ~ k))))
-  \label{eq:resolve-5} \\
-  &\leadsto Frame, ~ c^* \vdash (returning); ~
-    \algo ~ ([ ~ \executei ~ ], ~ \epsilon, ~ \wasm ~ ([ ~ Frame ~ ], ~ \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (\seq{c}, ~ k))))
-  \label{eq:resolve-6} \\
-  &\leadsto Frame, ~ c^* \vdash (returning); ~
-    \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ ([ ~ Frame ~ ], ~ \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (\seq{c}, ~ k))))
-  \label{eq:resolve-7} \\
-  &\leadsto Frame, ~ c^* \vdash (returning); ~ \wasm ~ ([ ~ Frame ~ ], ~ \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (\seq{c}, ~ k)))
-  \label{eq:resolve-8} \\
-  &\leadsto Frame, ~ c^* \vdash \epsilon; ~
-    \algo ~ ([ ~ \popctxi ~ ], ~ \epsilon, ~ \wasm ~ (\epsilon, ~ \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (\seq{c}, ~ k))))
-  \label{eq:resolve-9} \\
-  &\leadsto c^* \vdash \epsilon; ~
-    \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (\epsilon, ~ \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (\seq{c}, ~ k))))
-  \label{eq:resolve-10} \\
-  &\leadsto c^* \vdash \epsilon; ~ \wasm ~ (\epsilon, ~ \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (\seq{c}, ~ k)))
-  \label{eq:resolve-11} \\
-  &\leadsto c^* \vdash \epsilon; ~ \algo ~ (\epsilon, ~ \epsilon, ~ \wasm ~ (\seq{c}, ~ k))
-  \label{eq:resolve-12} \\
-  &\leadsto c^* \vdash \epsilon; ~ \wasm ~ (\seq{c}, ~ k)
-  \label{eq:resolve-13}
-\end{align}
+We manually craft the specification including the proposal features using the
+DSL, and similarly generate a Wasm interpreter using AL.
+This interpreter is tested against the execution tests in the official test
+suites and the proposals, excluding test related to the infinite loop
+detection.
+The interpreter successfully passed all ??? execution tests, in ??? seconds.
+Passing all these tests demonstrates the generality of AL.

### 2024-11-25 paper/related.tex
Add related work

@@
 \noindent
+
+% TODO: Fact check
+
+\textbf{General purpose language framework.}
+% Lem
+Lem~\cite{lem} is a language and tool for developing large-scale, rigorous
+semantics; it supports multiple output formats (e.g., LaTeX, Coq, and OCaml) to
+integrate formal specifications with proof tools and implementations.
+While Lem itself does not directly generate interpreters, its OCaml backend can
+be used to implement one.
+% Ott
+Ott~\cite{ott} is a lightweight tool for writing and maintaining formal
+semantics of programming languages, enabling generation of LaTeX for
+documentation and definitions for theorem provers such as Coq, Isabelle, or
+Twelf.
+It does not directly generate interpreters but serves as a frontend for
+creating formal specifications that can be used to derive interpreters manually
+or via theorem provers.
+% PLT
+PLT Redex~\cite{plt} is a Racket-based domain-specific language for specifying,
+experimenting with, and testing operational semantics, making it particularly
+suited for exploring small-step semantics and reduction systems.
+It provides an interactive environment for evaluating terms, which is akin to
+generating an interpreter.
+% K
+K Framework~\cite{k} is a formal language for defining programming languages
+and formal analysis tools; it enables the generation of interpreters,
+compilers, and formal verification frameworks directly from the semantics,
+making it a powerful tool for executable formal semantics.
+% Sail
+Sail~\cite{sail} is a domain-specific language designed for the formal
+specification of hardware architectures, particularly processors, using an
+abstract, executable semantics.
+While it is primarily used for hardware modeling, its general framework can
+also be applied to other systems for language.
+However, none of these tools are capable of specifying an algorithmic
+specification.
+
+\textbf{Specialized language framework.}
+% ESMeta
+ESMeta~\cite{esmeta} is a language framework specialized for JavaScript.
+It parses the JavaScript specification, ECMAScript, to automatically generate
+parser and interpreter ~\cite{jiset}, to generate tests and perform
+differential tests ~\cite{jest}, to find meta-level errors in the specification
+~\cite{jstar}, and to perform meta-level static analysis for the static
+analysis of the defined language ~\cite{jsaver}.
+However, it is not capable of specifying declarative specification.
+% ASL
+ASL~\cite{asl} is a language framework specialized for specifying the behavior
+and semantics of ARM instruction sets.
+% Only test?
+It also parses the algorithmic specification of ARM and test the specification.
+% ASL to Sail
+ASL to Sail project~\cite{asl2sail} translates ASL into Sail to enable formal
+verification, analysis, and multi-platform compatibility of ARM architecture
+semantics.
+Hence, ASL can specify both declarative and algorithmic specifications;
+however, it takes an opposite approach compared to SpecTec, translating the
+algorithmic specification into a declarative form.

### 2024-11-25 paper/al.tex
Minor fix
Fix figure title

@@
   \centerline{\includegraphics[width=15cm]{fig/returning}}
-  \caption[Behavior for the end a block]{Behavior for the end of a block}
+  \caption[A behavior for returning from a function without \texttt{return} instruction]
+    {A behavior for returning from a function without \texttt{return} instruction}
     \label{fig:exiting-label}
@@
   \seq{a}, w, c \vdash \pushi ~ e \Rightarrow (\push(w, wv), \algo ~ c) \\
+\end{gather*}
+\newpage
+\begin{gather*}
 %
@@
 % ReplaceFrame
-\replaceframei{} retrieves the current frame $\casev ~ (\text{``Frame"}, v_1
-v_2)${} in the Wasm context $w$.
+\replaceframei{} retrieves the current frame $\casev ~ (\text{``Frame"}, v_1 ~
+_2)${} in the Wasm context $w$.
 It then evaluates $e$ to $v$, updating the value at location $v_2$ along paths
@@
 sto$ along paths $p^+$ to v.
+
+\newpage
 
@@
   \listv ~ v_1 ~ ... ~ v_n \\
+\end{gather*}
+\newpage
+\begin{gather*}
 %

### 2024-11-26 paper/eval.tex
Add machine spec

@@
 does not define this behavior.
-The results demonstrate that the interpreter successfully pass all 51,549 tests in 70
-seconds on \red{(lambda 9)}.
+The results demonstrate that the interpreter successfully pass all 51,549 tests
+in 70 seconds on a machine running Ubuntu 18.04.3 LTS, equipped with an Intel
+Core i7-6700K processor (6th generation) and 8GB of Samsung DDR4 PC4-17000
+memory.
 Passing all the tests ensures the correctness of the specification.
@@
 % 7 + 3 + ??? + 44
+% TODO: Fill number
+% TODO: Add cite
 \section{RQ2: Generality}

### 2024-11-26 paper/related.tex
Add reference

@@
 be used to implement one.
-% Ott
-Ott~\cite{ott} is a lightweight tool for writing and maintaining formal
-semantics of programming languages, enabling generation of LaTeX for
-documentation and definitions for theorem provers such as Coq, Isabelle, or
-Twelf.
-It does not directly generate interpreters but serves as a frontend for
-creating formal specifications that can be used to derive interpreters manually
-or via theorem provers.
 % PLT
@@
 % ASL
-ASL~\cite{asl} is a language framework specialized for specifying the behavior
-and semantics of ARM instruction sets.
-% Only test?
-It also parses the algorithmic specification of ARM and test the specification.
+ASL~\cite{asl} is a specialized language framework designed for specifying the
+behavior and formal semantics of ARM instruction sets.
+The tool also parses the algorithmic specification of the ARM architecture,
+enabling verification of ARM processors through bounded model checking.
+Additionally, it refines the specification by leveraging methods such as
+testing.
 % ASL to Sail

### 2024-11-26 paper/conclusion.tex
Write abstract and conclusion

% !TEX root = main.tex

\chapter{Conclusion}
\label{ch:conclusion}
\noindent

This thesis introduces SpecTec, a framework designed to mechanize the
specification of Wasm.
It offers a domain specific language that facilitates the declarative
definition of Wasm's syntax and semantics, closely resembling formal notation.
SpecTec incorporates type-checking within the DSL to prevent errors in
meta-level specifications and translates them into prose specifications.

To formally define the semantics of Wasm in the prose specification, we propose
a rigorous model of Wasm’s control flow, define the syntax and semantics of the
AL language for specifying the prose, and implement an AL interpreter.
By leveraging this interpreter, specifications written in AL are executable,
functioning as a Wasm interpreter.

Executing these specifications enables the direct evaluation of the Wasm
specification.
This framework bridges the gap between formal specification and executable
code, ensuring their correctness and fidelity.

### 2024-11-26 paper/eval.tex
Minor change

@@
 Notably, specifying all eight proposals in AL requires only trivial
-adjustments, such as adding a handler case to the context.
+adjustments: adding a handler case to the context and generalizing runtime
+errors to describe an exception.
 

### 2024-11-26 paper/eval.tex
Minor change

--- 2024-11-26 paper/eval.tex
@@
 Notably, specifying all eight proposals in AL requires only trivial
-adjustments, such as adding a handler case to the context.
+adjustments: adding a handler case to the context and generalizing AL
+instruction for describing Wasm runtime errors.
 

### 2024-11-26 paper/al.tex
Remove ambiguity

@@
 Since the prose notation of the official specification is informal, it is
-essential to rigorously define AL to ensure an unambiguous and correct
-specification, rather than merely imitating the original.
-We begin by outlining the ambiguity of the prose specification in
+essential to rigorously define AL to ensure an correct specification, rather
+than merely imitating the original.
+We begin by outlining the informality of the prose specification in
 \cref{sec:prose-informal}.
@@
 
-\section{Ambiguity in the prose specification}
+\section{Informality in the prose specification}
 \label{sec:prose-informal}
@@
 instructions, such as function call.
-While the intended meaning can still be inferred, this informality introduces
-ambiguity, leaving room for multiple interpretations.
+While the intended meaning can still be inferred, this informality leaves room
+for wrong interpretations.
 

### 2024-11-26 paper/al.tex
Fix model explanation

@@
 \cref{sec:prose-informal}.
-Next, we propose a formal model for Wasm control flow in AL in
-\cref{sec:model}.
+Next, we propose a formal model for the prose specification \cref{sec:model}.
 We then define the syntax and the semantics of AL in \cref{sec:definition}.
@@
 % Formal model
-To formally specify the prose specification, we construct a formal model to
-describe Wasm control flow.
+To formally specify the prose specification, we construct a formal model the
+prose specification.
 % Execute

### 2024-11-26 paper/intro.tex
Fix model explanation

@@
 % Solution
-To address this, we present a formal model for how AL describes Wasm control
-flow.
+To address this, we present a formal model for how the prose specification
+describes Wasm semantics.
 Using this model, we identify the underlying issues and the necessary changes
@@
 Based on these insight, we formally define AL's syntax and semantics to
-accurately capture Wasm control flow.
+accurately capture Wasm semantics.
 We then implement an AL interpreter according to the formal definition and
@@
 \begin{itemize}
-  \item We propose a formal model for Wasm control flow. (\cref{sec:model})
+  \item We propose a formal model for the prose specification.
+    (\cref{sec:model})
   \item We define syntax and semantics of AL (\cref{sec:definition})

### 2024-11-26 paper/al.tex
Minor fix
More consistent wording

@@
 % Execute
-On the other hand, some control instructions \textit{execute} a Wasm
+On the other hand, some control instructions \textit{execute} another Wasm
 instruction.
@@
 
-\section{A Formal Model for WebAssembly Control Flow}
+\section{A Formal Model for the prose specification}
 \label{sec:model}
@@
 % Formal model
-To formally specify the prose specification, we construct a formal model the
-prose specification.
+To formally specify the prose specification, we construct a formal model for
+the prose specification.
 % Execute
@@
 Execution proceeds by popping an instruction from the stack and performing its
-operation.
+operations.
 To avoid confusion with the stack that contains values interleaved with
@@
 program counter.
+We define a Wasm state as a tuple of the interleaved stack, the instruction
+stack, and the store.
+An operation, as described in a step of the prose specification, is then
+formalized as a state transition.
 
 % Jump
-We then split two distinct usages of the jump operation: \textit{enter} and
-\textit{exit}.
+To define the jump formally, we split two distinct usages of the jump
+operation: \textit{enter} and \textit{exit}.
 Since Wasm supports structured control flow, the jump operation is used either
@@
 % State transition
-To express the enter and exit operation formally, we model them as state
-transitions.
-We define a Wasm state as a tuple of the interleaved stack, the instruction
-stack, and the store.
+We then formalize the enter operation as a state transition.
 For brevity, we omit the store in this discussion.

### 2024-11-29 paper/intro.tex
Fix intro

@@
 
+% 느낌 대신 "논리"적인 흐름에 집중 (일관성과 일맥상통)
 % Challenge
-One challenge of our approach is making AL executable.
+When desiging the SpecTec, one of the challenges in our approach is making AL
+executable.
+In this master thesis, we focus on this specific challenge.
 Since the prose specification is written in informal pseudocode, its
 interpretation is not straightforward.
-The naive implementation of AL interpreter produced incorrect results when
-running Wasm official test suites, particularly those related to control flow.
-Additionally, the absence of a formalized definition or formal semantics for AL
-made it difficult to identify and address the fundamental issues in its design.
-Since running tests is the primary method for assessing the correctness of the
-specification, this limitation hindered the development of both the generated
-interpreter and the specification itself.
-This underscored the need for a more robust approach to formally define AL and
-ensure its alignment with Wasm's complex semantics.
+The absence of a formalized definition or formal semantics for the prose
+specification posed significant difficulty in realizing executable AL.
+Since executing the specification against tests is the primary method for
+assessing its correctness, this limitation hindered the development of both the
+generated interpreter and the specification itself.
 
@@
 describes Wasm semantics.
-Using this model, we identify the underlying issues and the necessary changes
-to improve it.
 Based on these insight, we formally define AL's syntax and semantics to

### 2024-11-29 paper/eval.tex
Add refs and # instructions in generality

@@
 inclusion in the next version of Wasm: typed function references
-~\cite{proposal-fr}, garbage collection~\cite{propsal-gc}, tail calls
+~\cite{proposal-fr}, garbage collection~\cite{proposal-gc}, tail calls
 ~\cite{proposal-tc}, multiple memories~\cite{proposal-mm}, extended constant
@@
 ~\cite{proposal-rs}, exception handling~\cite{proposal-eh}.
-This proposals add or modify ??? Wasm instructions.
+These proposals add or modify 54 Wasm instructions out of 108 instructions.
 In particular, the garbage collection proposal adds substantial complications

### 2024-11-29 paper/eval.tex
Add number

@@
 DSL, and similarly generate a Wasm interpreter using AL.
-This interpreter is tested against the execution tests in the official test
-suites and the proposals, excluding test related to the infinite loop
-detection.
-The interpreter successfully passed all ??? execution tests, in ??? seconds.
+This interpreter is tested against the execution tests in the proposals,
+excluding test related to the infinite loop detection.
+The interpreter successfully passed all 59,975 execution tests, in 2,375
+seconds.
 Passing all these tests demonstrates the generality of AL.

### 2024-12-05 paper/background.tex
Minor fix

@@
 % Structured Control Flow
-\red{Wasm supports control instructions for structured control flow, which
+Wasm supports control instructions for structured control flow, which
 eliminates the need for arbitrary jumps, ensuring structured and predictable
@@
 the end of the block and the function, specified in the contexts.
-}
